### 2006-11-13 inst/doc/HowTo-flowCore.Rnw
renamed flowcore flowCore
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21108 bc3139a8-67e5-0310-9ffc-ced21a209358

% NOTE -- ONLY EDIT THE .Rnw FILE!!!  The .tex file is
% likely to be overwritten.
%
%\VignetteIndexEntry{Basic Functions for Flow Cytometry Data}
%\VignetteDepends{flowCore}
%\VignetteKeywords{}
%\VignettePackage{flowCore}
\documentclass[11pt]{article}

\usepackage{times}
\usepackage{hyperref}
\usepackage[authoryear,round]{natbib}
\usepackage{times}
\usepackage{comment}
\usepackage{subfigure}

\textwidth=6.2in
\textheight=8.5in
\oddsidemargin=.1in
\evensidemargin=.1in
\headheight=-.3in

\newlength{\smallfigwidth}
\setlength{\smallfigwidth}{6cm}

\newcommand{\scscst}{\scriptscriptstyle}
\newcommand{\scst}{\scriptstyle}
\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textsf{#1}}}
\newcommand{\Rclass}[1]{{\textit{#1}}}

\bibliographystyle{plainnat}

\title{How to use flowCore package}
\author{N. Le Meur F. Hahne P. Haaland}

\begin{document}
\maketitle

\begin{abstract}
\noindent \textbf{Background} The recent application of modern automation technologies to staining and collecting flow cytometry (FCM) samples has led to many new challenges in data management and analysis. We limit our attention here to the associated problems in the analysis of the massive amounts of FCM data now being collected. From our viewpoint, see two related but substantially different problems arising. On the one hand, there is the problem of adapting existing software to apply standard methods to the increased volume of data. The second problem, which we intend to address here, is the absence of any research platform which bioinformaticians, computer scientists, and statisticians can use to develop novel methods that address both the volume and multidimensionality of the mounting tide of data. In our opinion, such a platform should be Open Source, be focused on visualization,  support rapid prototyping, have a large existing base of users, and have demonstrated suitability for development of new methods. We believe that the Open Source statistical software R in conjunction with the Bioconductor Project fills all of these requirements. Consequently we have developed a Bioconductor package that we call flowcore. The flowcore package is not intended to be a complete analysis package for FCM data, rather, we see it as providing a clear object model and a collection of standard tools that enable R as an informatics research platform for flow cytometry. One of the important issues that we have addressed in the flowcore package is that of using a standardized representation that will insure compatibility with existing technologies for data analysis and will support collaboration and interoperability of new methods as they are developed. In order to do this, we have followed the current standardized descriptions of FCM data analysis as being developed under NIH Grant xxxx [n]. We believe that researchers will find flowcore to be a solid foundation for future development of new methods to attack the many interesting open research questions in FCM data analysis.\\
\textbf{Methods} We propose a variety different data structures. We have implemented the classes and methods in the Bioconductor package flowcore. We illustrate their use with X case studies.\\
\textbf{Results} We hope that those proposed data structures will the base for the development of many tools for the analysis of high throughput flowcytometry.\\
\textbf{keywords} Flow cytometry, high throughtput, software, standard
\end{abstract}

\section{Introduction}
Traditionnally, flow cytometry has been a tube-based technique limited to small-scale laboratory and clinical studies. High throughput methods for flow cytometry have recently been developed for drug discovery and advanced research methods \cite{Gasparetto2004, Young2004}. As an example, the flow cytometry high content screening (FC-HCS) can process up to a thousand samples daily at a single workstation, and the results have been equivalent or superior to traditional manual multiparameter staining and analysis techniques. \\
The amount of information generated by high throughput technologies such as FC-HCS need to be transformed into executive summaries (which are brief enough) for creative studies by a human researcher \cite{Brazma2001}. Standardization is critical when developing new high throughput technologies and their associated information services \cite{Brazma2001, Chicurel2002, Boguski2003}. Standardization efforts have been made in clinical cell analysis by flow cytometry \cite{Keeney2004}, however data interpretation has not been standardized for even low throughput FCM. It is one of the most difficult and time consuming aspects of the entire analytical process as well as a primary source of variation in clinical tests, and investigators have traditionally relied on intuition rather than standardized statistical inference \cite{Bagwell2004,Braylan2004,Parks1997,Suni2003}. In the development of standards in high throughput FCM, few progress has been done in term of Open Source software. In this article we propose data structures to handle flow cytometry data through the main steps of prprocessing: compensation, transformation, filtering.\\

\section{Methodology}
\subsection{Reading data}
We first load the libraries and read the data into R as FCS object 
(rflowcyt data format). Note than this small example works for cytoFrame object. 
<<load library>>=
library("flowCore")
@ 
<<load data,echo=FALSE,results=hide>>=
fcs.loc <- system.file("data", package="flowcore")
file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
b08 <- read.FCS(file.location[1])
e07 <- read.FCS(file.location[2])
f06 <- read.FCS(file.location[3])
@
\begin{verbatim}
> fcs.loc <- system.file("data", package="flowcore")
> file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
> b08 = read.FCS(file.location[1])
> e07 = read.FCS(file.location[2])
> f06 = read.FCS(file.location[3])
\end{verbatim}

<<showB08,echo=TRUE>>=
show(b08)
@

We can plot the raw data for the FSC-H and SSC-H parameters.
<<plotRaw1,eval=TRUE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
plot(b08,plotParameters=c(1,2))
@
<<plotRaw2,eval=FALSE,echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
plot(b08,plotParameters=c(3,4))
@ 
\begin{verbatim}
> basicplot(b08,plotParameters=c(1,2))
> basicplot(b08,plotParameters=c(3,4))
> basicplot(e07,plotParameters=c(1,2))
> basicplot(f06,plotParameters=c(1,2))
\end{verbatim}

\begin{figure}
\centering
\subfigure{\includegraphics[width=40mm]{plotRaw1} A}
\subfigure{\includegraphics[width=40mm]{plotRaw2} B}
\caption{
  \label{fig:Fig01}{Example of smooth scatter plot.(A) FSC-H
    \textit{vs.} SSC-H (B) FL2-H \textit{vs.} FL3-H}} 
\end{figure}

\subsection{Transformation data}
While analyzing flow cytometry data, various parameter transformations are performed to provide user-friendly visualization and/or to perform statistical analyses and interpret the data. Following the standard 8 transformations are pre-defined: linear, quadratic, natural logarithm, logarithm (with a specified base), hyperlog, bi-exponential, logicle, and split-scale.

As example we performed two different tramsformations on "FSC-H" channel of sample b08: a linear tranformation and a log base 2 transformation.

A linear transformation is like parameter= a*parameter +b
<<linearTransform,echo=FALSE,results=hide>>=
linear <- new("linearTransformation",transformationId="Linear-transformation",
                parameters="FSC-H", a=2, b=0)
data08Transform <- applyTransformation(linear,b08)
@ 
\begin{verbatim}
>linear <- new("linearTransformation",transformationId="Linear-transformation",
+                parameters="FSC-H", a=2, b=0)
>data08Transform <- applyTransformation(linear,b08)
\end{verbatim}

<<plotTRansfo,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
plot(data08Transform,plotParameters=c(1,2))
@

A log transformation is of type: parameter=log(parameter,logbase) * r/d
<<logTransform,echo=FALSE,results=hide>>=
logTransform <- new("logTransformation",transformationId="Log-Transformation",
                parameters="FSC-H", r=1, d=1,logbase=2)
data08LogTransform <- applyTransformation(logTransform,b08)
@ 
\begin{verbatim}
>logTransform <- new("logTransformation",transformationId="Log-Transformation",
+                parameters="FSC-H", r=1, d=1,logbase=2)
>data08LogTransform <- applyTransformation(logTransform,b08)
\end{verbatim}

<<ResultTransfoLinear>>=
b08@exprs[1:5]
#Linear transformation
data08Transform@exprs[1:5]
##Log base 2 transformation
data08LogTransform@exprs[1:5]
@

<<plotTRansfo,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
##we should find a way to specified if an parameters has been transformed 
## either in the title or via the label of the axis 
plot(data08LogTransform,plotParameters=c("FSC-H","SSC-H"))
@

\subsection{Filtering data (gating)}
\subsubsection{Rectangular gate}
We will define rectangular gates and apply them to filter the bO8 data.
Figure \ref{fig:Fig02} presents the results of such filtering. 
The plot in Figure \ref{fig:Fig02}A shows the result of 
filtering out the data points that are below 200 in the FSC-H channel.
Figure \ref{fig:Fig02}B shows the result of 
filtering out the data points are between 200 
and 500  in the FSC-H channel.

<<define and apply rectangleGate1,eval=TRUE, echo=FALSE,results=hide>>=
min1 <- 200
max1 <- Inf
fil1 <- new("rectangleGate",filterId="nonDebris",
            parameters="FSC-H",min=min1,max=max1)
b08fil1 <- applyFilter(fil1,b08)
@ 
<<plotfiltered, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(b08,y=b08fil1,pch="*")
@ 
<<define and apply polygonGate2,eval=TRUE, echo=FALSE,results=hide>>=
min2 <- 200
max2 <- 500
centerFilter <- new("rectangleGate",filterId="center",
            parameters=c("FSC-H"), min=min2, max=max2)
b08centerFilter <- applyFilter(centerFilter,b08)
@ 
<<plotCenterfiltered, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(b08,y=b08centerFilter,pch="*")
@

%%\begin{figure}
%%\centering
%%\subfigure{\includegraphics[width=40mm]{plotfiltered} A}
%%\subfigure{\includegraphics[width=40mm]{plotCenterfiltered} B} 
%%\caption{
%%  \label{fig:Fig02}{Example of the effect of filtering with a
    rectangular gate.}} 
%%\end{figure}

\subsubsection{polygon gate}
%%We will define a rectangular gate and apply the filter it to the bO8 data.
<<define and apply rectangleGate,eval=TRUE,echo=FALSE>>=
sqrcut <- matrix(c(300,300,600,600,50,300,300,50),ncol=2,nrow=4)
sqrfilter <- new("polygonGate",filterId="nonDebris",
            parameters=c("FSC-H","SSC-H"),boundaries= sqrcut)
b08sqrfilter <- applyFilter(sqrfilter,b08)
@

<<plotRectanglefilter, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(b08,y=b08sqrfilter,pch="*")
@

<<define and apply polygonGate,eval=TRUE,echo=FALSE>>=
triangleCut <- matrix(c(300,400,600,300,400,300),ncol=2,nrow=3)
trianglefilter <- new("polygonGate",filterId="nonDebris",
            parameters=c("FSC-H","SSC-H"),boundaries= triangleCut)
b08Trgfilter <- applyFilter(trianglefilter,b08)
@

<<plotTrianglefilter, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(x=b08,y=b08Trgfilter,filter=trianglefilter,colSubSet="red",pch="*")
@

%%\begin{figure}
%%\centering
%%\subfigure{\includegraphics[width=40mm]{plotRectanglefilter} A}
%%\subfigure{\includegraphics[width=40mm]{plotTrianglefilter} B} 
%%\caption{
%%  \label{fig:Fig03}{Example of the effect of filtering with a
%%    polygon gate.}} 
%%\end{figure}

\clearpage
\section{Case studies}
\subsection{BD}

<<rawdata1, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(b08,plotParameters=c("FSC-H","SSC-H"),main="B08")
@
<<rawdata2, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>= 
plot(e07,plotParameters=c("FSC-H","SSC-H"),main="E07")
@
<<rawdata3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(f06,plotParameters=c("FSC-H","SSC-H"),main="F06")
@ 
%%\begin{figure}
%%\centering
%%\subfigure{\includegraphics[width=40mm]{rawdata1}}
%%\subfigure{\includegraphics[width=40mm]{rawdata2}} 
%%\subfigure{\includegraphics[width=40mm]{rawdata3}} 
%%\caption{
%%  \label{fig:Fig04}{Example of the effect of filtering with a
%%    polygon gate.}} 
%%\end{figure}

The first gate is a range gate to filter out debris

<<SelectNonDebris,eval=TRUE,echo=FALSE,results=hide>>=
min1 <- c(200,0)
max1 <- c(800,800)
filter1 <- new("rectangleGate",filterId="Nondebris",parameters=c("FSC-H","SSC-H"),
  min=min1,max=max1)

b08.result1 <-  applyFilter(filter1,b08)
e07.result1 <- applyFilter(filter1,e07)
f06.result1 <- applyFilter(filter1,f06)

tab <- c("B08"=sum(b08.result1@subSet),"E07"=sum(e07.result1@subSet),"F06"=sum(f06.result1@subSet))
@

<<plotResFilter1, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(b08,y=b08.result1,main="B08",pch="*")
@ 
<<plotResFilter2, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(e07,y=e07.result1,main="E07",pch="*")
@ 
<<plotResFilter3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(f06,y=f06.result1,main="F06",pch="*")
@ 
%%\begin{figure}
%%\centering
%%\subfigure{\includegraphics[width=40mm]{plotResFilter1}}
%%\subfigure{\includegraphics[width=40mm]{plotResFilter2}} 
%%\subfigure{\includegraphics[width=40mm]{plotResFilter3}} 
%%\caption{
%%  \label{fig:Fig05}{Example of the effect of filtering with a
%%    polygon gate.}} 
%%\end{figure}

The second gate gets the live cells

<<FilterLiveCell,eval=TRUE,echo=FALSE,results=hide>>=
filter2 = new("norm2Filter",filterId="Live Cells",scale.factor=2,method="covMcd",parameters=c("FSC-H","SSC-H"))

b08.result2 = applyFilter(filter2,b08,parent=b08.result1)
e07.result2 = applyFilter(filter2,e07,parent=e07.result1)
f06.result2 = applyFilter(filter2,f06,parent=f06.result1)

tab <- rbind(tab,c(sum(b08.result2@subSet),sum(e07.result2@subSet),sum(f06.result2@subSet)))

@

<<plotLivingCell1, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(b08,y=b08.result2,parent=b08.result1,xlim=c(0,1024),ylim=c(0,1024),main="B08",pch="*")
@ 
<<plotLivingCell2, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(e07,y=e07.result2,parent=e07.result1,xlim=c(0,1024),ylim=c(0,1024),main="E07",pch="*")
@ 
<<plotLivingCell3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(f06,y=f06.result2,parent=f06.result1, xlim=c(0,1024),ylim=c(0,1024),main="F06",pch="*")
@ 
%%\begin{figure}
%%\centering
%%\subfigure{\includegraphics[width=40mm]{plotLivingCell1}}
%%\subfigure{\includegraphics[width=40mm]{plotLivingCell2}} 
%%\subfigure{\includegraphics[width=40mm]{plotLivingCell3}} 
%%\caption{
%%  \label{fig:Fig06}{Example of the effect of filtering with a
%%    polygon gate.}} 
%%\end{figure}

The third-fifth gates get the positive cells for the marker in FL1-H. 
This is a really interesting example because it illustrates that there
are two subpopulations. Naturally we would like to automatically find them
In this case we want to now what percent the positive population in FL1-H 
is of the total population.

<<FL1-H, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(b08,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),ylim=c(0,1024),xlim=c(0,1024))
@ 
%%\begin{figure}
%%\centering
%%\includegraphics{FL1-H}
%%\end{figure}

<<FL1Filter,eval=TRUE,echo=FALSE,results=hide>>=
filter3 <- new("rectangleGate",filterId="FL1-H+",parameters="FL1-H",min=400,max=Inf)
b08.result3 <- applyFilter(filter3,b08,b08.result2)

filter4 <-new("norm2Filter",filterId="FL1-H+",scale.factor=2,method="covMcd",parameters=c("FSC-H","FL1-H"))
b08.result4 <- applyFilter(filter4,b08,b08.result2)

##b08.result5 <- applyFilter(filter4,b08,b08.result2@subSet-b08.result4@subSet)

##ans <- c(sum(b08.result3@subSet)/sum(b08.result2@subSet),
##sum(b08.result4@subSet)/(sum(b08.result4@subSet)+sum(b08.result5@subSet)))
@

<<FL1Filtered1, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(b08,y=b08.result3,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
          xlim=c(0,1024),ylim=c(0,1024),pch="*")
@ 
<<FL1Filtered2, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
plot(b08,y=b08.result4,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
          xlim=c(0,1024),ylim=c(0,1024),pch="*")
@ 
<<FL1Filtered3, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
basicplot(b08,y=b08.result5,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
          xlim=c(0,1024),ylim=c(0,1024),pch="*")
@

%%\begin{figure}
%%\centering
%%\subfigure{\includegraphics[width=40mm]{FL1Filtered1}}
%%\subfigure{\includegraphics[width=40mm]{FL1Filtered2}} 
%%\subfigure{\includegraphics[width=40mm]{FL1Filtered3}} 
%%\caption{
%%  \label{fig:Fig07}{Example of the effect of filtering with a
%%    polygon gate.}} 
%%\end{figure}

\clearpage
\small{
<<sessionInfo>>=
sessionInfo()
@ 
}
\end{document}

### 2006-11-13 R/perry.r

@@ -1,0 +1,24 @@
+## last changed November 2, 2006
+##   I abstracted out the code for checking whether or not the parent exists
+
+##getParentSet = function(parent,ncells) {
+## 	if(missing(parent)) {
+   		## if the parent is missing assume that it is all cells
+##		parentSet = rep(1,ncells)
+## 	}
+##	else {  	
+##		if(class(parent) != "filterResult") {
+##			stop("parent must be of class filterResult.")
+##		}
+##		else {
+##			if(is.null(parent@subSet)) {
+##				stop("The filterResult parent must have a subSet slot.")
+##			}
+##			else {
+##				parentSet = parent@subSet
+##			}  	
+##		}
+##	}
+##	parentSet
+##}
+ 

### 2006-11-26 inst/doc/HowTo-flowCore.Rnw
rename fcsFrame/Set to flowFrame/Set
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21270 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \usepackage{comment}
+\usepackage{graphicx}
 \usepackage{subfigure}
@@
 \headheight=-.3in
-
-\newlength{\smallfigwidth}
-\setlength{\smallfigwidth}{6cm}
 
@@
 
-<<showB08,echo=TRUE>>=
+<<showB08,eval=FALSE,echo=FALSE>>=
 show(b08)
@@
 @
-<<plotRaw2,eval=FALSE,echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
+<<plotRaw2,eval=TRUE,echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
 plot(b08,plotParameters=c(3,4))
@@
 \begin{verbatim}
-> basicplot(b08,plotParameters=c(1,2))
-> basicplot(b08,plotParameters=c(3,4))
-> basicplot(e07,plotParameters=c(1,2))
-> basicplot(f06,plotParameters=c(1,2))
+> plot(b08,plotParameters=c(1,2))
+> plot(b08,plotParameters=c(3,4))
+> plot(e07,plotParameters=c(1,2))
+> plot(f06,plotParameters=c(1,2))
 \end{verbatim}
@@
 \end{verbatim}
-
 <<ResultTransfoLinear>>=
@@
 @ 
-<<plotfiltered, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+<<plotfiltered, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
 plot(b08,y=b08fil1,pch="*")
@@
 @ 
-<<plotCenterfiltered, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+<<plotCenterfiltered, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
 plot(b08,y=b08centerFilter,pch="*")
@@
 
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{plotfiltered} A}
-%%\subfigure{\includegraphics[width=40mm]{plotCenterfiltered} B} 
-%%\caption{
-%%  \label{fig:Fig02}{Example of the effect of filtering with a
-    rectangular gate.}} 
-%%\end{figure}
+\begin{figure}
+\centering
+\subfigure{\includegraphics[width=40mm]{plotfiltered} A}
+\subfigure{\includegraphics[width=40mm]{plotCenterfiltered} B} 
+\caption{
+  \label{fig:Fig02}{Example of the effect of filtering with a rectangular gate.}} 
+\end{figure}
 
@@
 
-<<plotRectanglefilter, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+<<plotRectanglefilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
 plot(b08,y=b08sqrfilter,pch="*")
@@
 
-<<plotTrianglefilter, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+<<plotTrianglefilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
 plot(x=b08,y=b08Trgfilter,filter=trianglefilter,colSubSet="red",pch="*")
@@
 
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{plotRectanglefilter} A}
-%%\subfigure{\includegraphics[width=40mm]{plotTrianglefilter} B} 
-%%\caption{
-%%  \label{fig:Fig03}{Example of the effect of filtering with a
-%%    polygon gate.}} 
-%%\end{figure}
+\begin{figure}
+\centering
+\subfigure{\includegraphics[width=40mm]{plotRectanglefilter} A}
+\subfigure{\includegraphics[width=40mm]{plotTrianglefilter} B} 
+\caption{
+  \label{fig:Fig03}{Example of the effect of filtering with a
+    polygon gate.}} 
+\end{figure}
 
@@
 max1 <- c(800,800)
-filter1 <- new("rectangleGate",filterId="Nondebris",parameters=c("FSC-H","SSC-H"),
+ilter1 <- new("rectangleGate",filterId="Nondebris",parameters=c("FSC-H","SSC-H"),
   min=min1,max=max1)
@@
 @ 
-
 <<plotResFilter1, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
@@
 <<FL1Filtered3, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
-basicplot(b08,y=b08.result5,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
+plot(b08,y=b08.result5,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
           xlim=c(0,1024),ylim=c(0,1024),pch="*")

### 2006-11-26 R/AllGeneric.R

@@ -39,14 +39,14 @@ if(!isGeneric("ncol"))
 ## ===========================================================================
 ## Generic for applyFilter
 ## ---------------------------------------------------------------------------
-setGeneric("applyTransformation", function(transformation,flowSet) standardGeneric("applyTransformation"))
+setGeneric("applyTransformation", function(transformation,flowObject) standardGeneric("applyTransformation"))
 ## ---------------------------------------------------------------------------
 
 
 ## ===========================================================================
 ## Generic for applyFilter
 ## ---------------------------------------------------------------------------
-setGeneric("applyFilter", function(filter,flowSet,parent) standardGeneric("applyFilter"))
+setGeneric("applyFilter", function(filter,flowObject,parent) standardGeneric("applyFilter"))
 ## ---------------------------------------------------------------------------
 
 ## ===========================================================================
@@ -64,6 +64,6 @@ setGeneric("spillover",function(x,...) standardGeneric("spillover"))
 ## ===========================================================================
 ## Generic for applyCompensation
 ## ---------------------------------------------------------------------------
-setGeneric("applyCompensation", function(compensation,flowSet) standardGeneric("applyCompensation"))
+setGeneric("applyCompensation", function(compensation,flowObject) standardGeneric("applyCompensation"))
 ## ---------------------------------------------------------------------------
 

### 2006-11-26 R/filter-functions.R

@@ -5,18 +5,18 @@
 ## ==========================================================================
 ## filtering function for polygonal gates
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-rectangleFiltering <- function(filter,flowSet,parent){
+rectangleFiltering <- function(filter,flowObject,parent){
     parameters=filter@parameters
     min=filter@min
     max=filter@max
     ## test for validity of objects
     
-    if(class(flowSet)!="fcsFrame")
-      stop("flowset must be of class fcsFrame.")
-    if (is.null(flowSet@exprs)) 
+    if(class(flowObject)!="flowFrame")
+      stop("flowObject must be of class flowFrame.")
+    if (is.null(flowObject@exprs)) 
         stop("There is no data to createFilter.")
     
-    data=flowSet@exprs
+    data=flowObject@exprs
     
     
     ## ncells is the size of the original data
@@ -24,7 +24,7 @@ rectangleFiltering <- function(filter,flowSet,parent){
     ## a 1 means the cell (row) was in the filter, a 0 means it wasn't
     selectNew = rep(0,ncells)
     ## check for the parent. If it is a filterResult get the parentSet
-	parentSet = getParentSet(parent,ncells)
+    parentSet = getParentSet(parent,ncells)
     ## a 1 means the cell (row) was in the filter, a 0 means it wasn't
     selectCurrent = (parentSet == 1)
     data = data[selectCurrent,parameters,drop=FALSE]
@@ -48,18 +48,18 @@ rectangleFiltering <- function(filter,flowSet,parent){
 ## ==========================================================================
 ## filtering function for polygonal gates
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-polygonFiltering <- function(filter,flowSet,parent){
+polygonFiltering <- function(filter,flowObject,parent){
     parameters=filter@parameters
     boundaries=filter@boundaries
     ## test for validity of objects
 
 
-    if(class(flowSet)!="fcsFrame")
-      stop("flowset must be of class fcsFrame.")
-    if (is.null(flowSet@exprs)) 
+    if(class(flowObject)!="flowFrame")
+      stop("flowObject must be of class flowFrame.")
+    if (is.null(flowObject@exprs)) 
       stop("There is no data to createFilter.")
     
-    data=flowSet@exprs
+    data=flowObject@exprs
 
   
[…]

### 2006-11-26 R/transformation-functions.R

@@ -1,104 +1,104 @@
 ## ==========================================================================
 ## Linear transformation function
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-linearTransformation <- function(transformation,flowSet){
+linearTransformation <- function(transformation,flowObject){
     parameters=transformation@parameters
     a=transformation@a
     b=transformation@b
     ## test for validity of objects
-    if(class(flowSet)!="fcsFrame"){
-        stop("flowSet must of class fcsFrame.")
+    if(class(flowObject)!="flowFrame"){
+        stop("flowObject must of class flowFrame.")
     }
-    if (is.null(flowSet@exprs)) {
+    if (is.null(flowObject@exprs)) {
         stop("There is no data to createFilter.")
     }
-    data <- flowSet@exprs
-    flowSet@exprs[,parameters] <- a * data[,parameters] + b
-    flowSet@description[["transformation"]] <- paste("A linear transformation was applied on the",
+    data <- flowObject@exprs
+    flowObject@exprs[,parameters] <- a * data[,parameters] + b
+    flowObject@description[["transformation"]] <- paste("A linear transformation was applied on the",
                                                      parameters, "channel",sep="",collapse=" ")
     
-    return(flowSet)
+    return(flowObject)
 }
 
 ## ==========================================================================
 ## Quadratic transformation function
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-quadraticTransformation <- function(transformation,flowSet){
+quadraticTransformation <- function(transformation,flowObject){
     parameters=transformation@parameters
     a=transformation@a
     b=transformation@b
     c=transformation@c
     ## test for validity of objects
     
-    if(class(flowSet)!="fcsFrame"){
-        stop("flowSet must of class fcsFrame.")
+    if(class(flowObject)!="flowFrame"){
+        stop("flowObject must of class flowFrame.")
     }
-    if (is.null(flowSet@exprs)) {
+    if (is.null(flowObject@exprs)) {
         stop("There is no data to createFilter.")
     }
-    data <- flowSet@exprs
-    flowSet@exprs[,parameters] <- a * data[,parameters]^2 + b * data[,parameters] + c
-    flowSet@description[["transformation"]] <- paste("A quadratic transformation was applied on the",
+    data <- flowObject@exprs
[…]

### 2006-11-26 R/flowFrame-methods.R
rename fcsFrame/Set to flowFrame/Set
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21270 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -2,7 +2,7 @@
 ## accessor method for slot exprs
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("exprs",
-  signature="fcsFrame", definition=function(object) object@exprs,
+  signature="flowFrame", definition=function(object) object@exprs,
   valueClass="matrix")
 ## ==========================================================================
 
@@ -11,7 +11,7 @@ setMethod("exprs",
 ## replace method for slot exprs
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setReplaceMethod("exprs",
-  signature=c("fcsFrame", "matrix"), definition=function(object, value) {
+  signature=c("flowFrame", "matrix"), definition=function(object, value) {
     object@exprs <- value
     return(object)})
 ## ==========================================================================
@@ -21,7 +21,7 @@ setReplaceMethod("exprs",
 ## accessor method for slot description
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("description",
-  signature="fcsFrame", definition=function(object) object@description,
+  signature="flowFrame", definition=function(object) object@description,
   valueClass="character")
 ## ==========================================================================
 
@@ -30,7 +30,7 @@ setMethod("description",
 ## replace method for slot description
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setReplaceMethod("description",
-  signature=c("fcsFrame", "character"), definition=function(object, value) {
+  signature=c("flowFrame", "character"), definition=function(object, value) {
     object@description <- value
     return(object)})
 ## ==========================================================================
@@ -40,7 +40,7 @@ setReplaceMethod("description",
 ## accessor method for slot colnames
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("colnames",
-  signature="fcsFrame", definition=function(x, do.NULL="missing",
+  signature="flowFrame", definition=function(x, do.NULL="missing",
   prefix="missing") colnames(exprs(x)), valueClass="character")
 ## ==========================================================================
 
@@ -49,7 +49,7 @@ setMethod("colnames",
[…]

### 2006-11-29 inst/doc/HowTo-flowCore.Rnw
move fcs files from data to extdada modify the vignette and man pages
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21315 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 <<load data,echo=FALSE,results=hide>>=
-fcs.loc <- system.file("data", package="flowcore")
+fcs.loc <- system.file("extdata", package="flowCore")
 file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
@@
 \begin{verbatim}
-> fcs.loc <- system.file("data", package="flowcore")
+> fcs.loc <- system.file("extdata", package="flowCore")
 > file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
@@
 max1 <- c(800,800)
-ilter1 <- new("rectangleGate",filterId="Nondebris",parameters=c("FSC-H","SSC-H"),
+filter1 <- new("rectangleGate",filterId="Nondebris",parameters=c("FSC-H","SSC-H"),
   min=min1,max=max1)

### 2006-12-01 R/AllClasses.R
modify applyTransformation function to use the R function transform remains issues withe the names of the diffrent channel: the transform function does not like names such as "FSC-H"
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21327 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -15,15 +15,15 @@ require("Biobase")
 ## ---------------------------------------------------------------------------
 setClass("flowFrame",                
   representation(exprs="matrix",
-                 description="character"),
+                 description="vector"),
   prototype=list(exprs=matrix(numeric(0), nrow=0, ncol=0),
                  description=c(note="empty")),
  validity=function(object){
    msg <- TRUE
    if(!is.matrix(object@exprs))
       msg <- "\nslot 'exprs' must be matrix"
-   if(!is.character(object@description))
-      msg <- "\nslot 'description' must be character vector"
+   if(!is.vector(object@description))
+      msg <- "\nslot 'description' must be a vector"
 })
 
 ## ===========================================================================

### 2006-12-01 R/readFCS-functions.R

@@ -19,7 +19,7 @@ read.FCS <- function(filename,transformation=FALSE,debug=FALSE)
   if(transformation==TRUE){
       txt[["tranformation"]] <-"applied" 
   }
-  return(new("flowFrame", exprs=mat, description=txt))
+  return(new("flowFrame", exprs=mat, description=c(txt)))
 }
 
 readFCSgetPar <- function(x, pnam) {

### 2006-12-01 R/transformation-methods.R

@@ -2,93 +2,110 @@
 ## ==========================================================================
 ## Apply linear transformation on fcsframe Object
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("applyTransformation",
-          signature=signature(transformation="linearTransformation",flowObject="flowFrame"),
-          definition=function(transformation,flowObject) {
-            transformNew <-linearTransformation(transformation,flowObject)
-             msg <- paste("Linear transformation applied on ",
-                           deparse(substitute(flowset)),
-                           " (file:",deparse(substitute(flowObject)),
-                           ") a ", class(flowObject)," object", sep="")
-            print(msg)
-            return(transformNew)
-          }
-          
- )
+##setMethod("applyTransformation",
+##          signature=signature(transformation="linearTransformation",flowObject="flowFrame"),
+##          definition=function(transformation,flowObject) {
+##            transformNew <-linearTransformation(transformation,flowObject)
+##             msg <- paste("Linear transformation applied on ",
+##                           deparse(substitute(flowset)),
+###                           " (file:",deparse(substitute(flowObject)),
+##                           ") a ", class(flowObject)," object", sep="")
+##            print(msg)
+##            return(transformNew)
+##         }
+##          
+## )
 
 ## ==========================================================================
 ## Apply quadratic transformation on flowFrame Object
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("applyTransformation",
-          signature=signature(transformation="quadraticTransformation",flowObject="flowFrame"),
-          definition=function(transformation,flowObject) {
-            transformNew <-quadraticTransformation(transformation,flowObject)
-             msg <- paste("Quadratic transformation applied on ",
-                           deparse(substitute(flowObject)),
-                           " (file:",deparse(substitute(flowObject)),
-                           ") a ", class(flowObject)," object", sep="")
-            print(msg)
-            return(transformNew)
-          }
-          
- )
+##setMethod("applyTransformation",
+##          signature=signature(transformation="quadraticTransformation",flowObject="flowFrame"),
+##          definition=function(transformation,flowObject) {
+##            transformNew <-quadraticTransformation(transformation,flowObject)
+##             msg <- paste("Quadratic transformation applied on ",
+##                           deparse(substitute(flowObject)),
+##                           " (file:",deparse(substitute(flowObject)),
[…]

### 2006-12-01 inst/doc/HowTo-flowCore.Rnw
modify applyTransformation function to use the R function transform remains issues withe the names of the diffrent channel: the transform function does not like names such as "FSC-H"
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21327 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 <<linearTransform,echo=FALSE,results=hide>>=
-linear <- new("linearTransformation",transformationId="Linear-transformation",
-                parameters="FSC-H", a=2, b=0)
-data08Transform <- applyTransformation(linear,b08)
+linear <- linearTransform(transformationId="linear-transformation",a=2, b=0)
+Bb <- b08
+colnames(Bb@exprs) <- paste("c",1:8,sep="")
+data08Transform <- transform(Bb,c1=linear(c1))
 @ 
 \begin{verbatim}
->linear <- new("linearTransformation",transformationId="Linear-transformation",
-+                parameters="FSC-H", a=2, b=0)
->data08Transform <- applyTransformation(linear,b08)
+>linear <- linearTransform(transformationId="linear-transformation",a=2, b=0)
+>data08Transform <- transform(b08,FCS-H=linear(FCS-H))
 \end{verbatim}
@@
 A log transformation is of type: parameter=log(parameter,logbase) * r/d
-<<logTransform,echo=FALSE,results=hide>>=
-logTransform <- new("logTransformation",transformationId="Log-Transformation",
-                parameters="FSC-H", r=1, d=1,logbase=2)
-data08LogTransform <- applyTransformation(logTransform,b08)
+<<logTransform,eval=FALSE,echo=FALSE,results=hide>>=
+logTransform <- logTransformation(transformationId="Log-Transformation",
+                r=1, d=1,logbase=2)
+data08LogTransform <- transform(Bb,c1=logTransform(c1))
 @ 
 \begin{verbatim}
->logTransform <- new("logTransformation",transformationId="Log-Transformation",
-+                parameters="FSC-H", r=1, d=1,logbase=2)
->data08LogTransform <- applyTransformation(logTransform,b08)
+>logTransform <- logTransformation(transformationId="Log-Transformation", r=1, d=1,logbase=2)
+>data08LogTransform <- transform(b08,FCS-H=logTransform(FCS-H))
 \end{verbatim}
@@
 ##Log base 2 transformation
-data08LogTransform@exprs[1:5]
+##data08LogTransform@exprs[1:5]
 @

### 2006-12-01 R/AllGeneric.R

@@ -39,10 +39,11 @@ if(!isGeneric("ncol"))
   setGeneric("ncol", function(x) standardGeneric("ncol"))
 ## ---------------------------------------------------------------------------
 
+
 ## ===========================================================================
-## Generic for applyFilter
+## Generic for applyTransformation
 ## ---------------------------------------------------------------------------
-setGeneric("applyTransformation", function(transformation,flowObject) standardGeneric("applyTransformation"))
+  setGeneric("transform")
 ## ---------------------------------------------------------------------------
 
 
@@ -52,11 +53,6 @@ setGeneric("applyTransformation", function(transformation,flowObject) standardGe
 setGeneric("applyFilter", function(filter,flowObject,parent) standardGeneric("applyFilter"))
 ## ---------------------------------------------------------------------------
 
-## ===========================================================================
-## Generic for applyFilter
-## ---------------------------------------------------------------------------
-setGeneric("basicplot", function(x,y,...) standardGeneric("basicplot"))
-## ---------------------------------------------------------------------------
 
 ## ===========================================================================
 ## Generic for spillover
@@ -67,5 +63,5 @@ setGeneric("spillover",function(x,...) standardGeneric("spillover"))
 ## ===========================================================================
 ## Generic for applyCompensation
 ## ---------------------------------------------------------------------------
-setGeneric("applyCompensation", function(compensation,flowObject) standardGeneric("applyCompensation"))
+setGeneric("applyCompensation", function(flowObject,compensation) standardGeneric("applyCompensation"))
 ## ---------------------------------------------------------------------------

### 2006-12-01 inst/doc/HowTo-flowCore.Rnw
modify the other transformation functions
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21330 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \title{How to use flowCore package}
-\author{N. Le Meur F. Hahne P. Haaland}
+\author{N. Le Meur F. Hahne B. Ellis P. Haaland}
 
@@
 linear <- linearTransform(transformationId="linear-transformation",a=2, b=0)
-Bb <- b08
-colnames(Bb@exprs) <- paste("c",1:8,sep="")
-data08Transform <- transform(Bb,c1=linear(c1))
+data08Transform <- transform(b08,`FSC-H`=linear(`FSC-H`))
 @ 
@@
 >linear <- linearTransform(transformationId="linear-transformation",a=2, b=0)
->data08Transform <- transform(b08,FCS-H=linear(FCS-H))
+>data08Transform <- transform(b08,`FSC-H`=linear(`FSC-H`))
 \end{verbatim}
@@
 <<logTransform,eval=FALSE,echo=FALSE,results=hide>>=
-logTransform <- logTransformation(transformationId="Log-Transformation",
-                r=1, d=1,logbase=2)
-data08LogTransform <- transform(Bb,c1=logTransform(c1))
+logTransform <- logTransform(transformationId="Log-Transformation",logbase=2,r=1,d=1)
+data08LogTransform <- transform(b08,`FSC-H`=logTransform(`FSC-H`))
 @ 
@@
 >logTransform <- logTransformation(transformationId="Log-Transformation", r=1, d=1,logbase=2)
->data08LogTransform <- transform(b08,FCS-H=logTransform(FCS-H))
+>data08LogTransform <- transformb08,`FSC-H`=logTransform(`FSC-H`))
 \end{verbatim}
+
 <<ResultTransfoLinear>>=

### 2006-12-09 inst/doc/filtering-flowCore.Rnw
BIG set of changes! This changes the implementation of the "filter" method by moving most of the work out of the functions and into the %in% method, which is exposed to the user. This is designed to be easier for programmers and future filter implementors.
filter() itself is now a single function that dispatches the appropriate %in%. Also, we're now carrying around some extra filterResult data to help us make filterResult a type of filter class that can be used in conjunction with other unresolved filters or with sets of filter results using the usual sorts of logical operators. 

We also have

--- 2006-12-01 inst/doc/HowTo-flowCore.Rnw
@@
 %
-%\VignetteIndexEntry{Basic Functions for Flow Cytometry Data}
+%\VignetteIndexEntry{Filtering Flow Cytometry Data}
 %\VignetteDepends{flowCore}
@@
 
-\title{How to use flowCore package}
+\title{Filtering in the flowCore package}
 \author{N. Le Meur F. Hahne B. Ellis P. Haaland}
@@
 \begin{abstract}
-\noindent \textbf{Background} The recent application of modern automation technologies to staining and collecting flow cytometry (FCM) samples has led to many new challenges in data management and analysis.
-We limit our attention here to the associated problems in the analysis of the massive amounts of FCM data now being collected.
-From our viewpoint, see two related but substantially different problems arising.
-On the one hand, there is the problem of adapting existing software to apply standard methods to the increased volume of data.
-The second problem, which we intend to address here, is the absence of any research platform which bioinformaticians, computer scientists, and statisticians can use to develop novel methods that address both the volume and multidimensionality of the mounting tide of data.
-In our opinion, such a platform should be Open Source, be focused on visualization,  support rapid prototyping, have a large existing base of users, and have demonstrated suitability for development of new methods.
-We believe that the Open Source statistical software R in conjunction with the Bioconductor Project fills all of these requirements.
-Consequently we have developed a Bioconductor package that we call flowcore.
-The flowcore package is not intended to be a complete analysis package for FCM data, rather, we see it as providing a clear object model and a collection of standard tools that enable R as an informatics research platform for flow cytometry.
-One of the important issues that we have addressed in the flowcore package is that of using a standardized representation that will insure compatibility with existing technologies for data analysis and will support collaboration and interoperability of new methods as they are developed.
-In order to do this, we have followed the current standardized descriptions of FCM data analysis as being developed under NIH Grant xxxx [n].
-We believe that researchers will find flowcore to be a solid foundation for future development of new methods to attack the many interesting open research questions in FCM data analysis.\\
-\textbf{Methods} We propose a variety different data structures. We have implemented the classes and methods in the Bioconductor package flowcore. We illustrate their use with X case studies.\\
-\textbf{Results} We hope that those proposed data structures will the base for the development of many tools for the analysis of high throughput flowcytometry.\\
+\noindent One of the more common operations in flow cytometry data analysis is a process of filtering, also known as gating.
+This is often either the target of analysis in the case of surface staining or a necessary pre-processing step for further analysis on the cell population of interest.
+For example, a researcher may be interested in CD4+ T cells, which typically requires a lymphocyte gate and a CD4 gate to obtain a relatively clean population and other cell types might be even more complex.
+In the case where the number of cells within a gate is of interest, it is not uncommon to require a fairly complex composition of gates for analysis.
+This document describes the rich filter construction and composition facilities in the flowCore package as well as what is needed for developing your own filter objects.
+\\ \\
 \textbf{keywords} Flow cytometry, high throughtput, software, standard
@@
 
-\section{Introduction}
-Traditionnally, flow cytometry has been a tube-based technique limited to small-scale laboratory and clinical studies.
-High throughput methods for flow cytometry have recently been developed for drug discovery and advanced research methods \cite{Gasparetto2004, Young2004}.
-As an example, the flow cytometry high content screening (FC-HCS) can process up to a thousand samples daily at a single workstation, and the results have been equivalent or superior to traditional manual multiparameter staining and analysis techniques.
-\\
-The amount of information generated by high throughput technologies such as FC-HCS need to be transformed into executive summaries (which are brief enough) for creative studies by a human researcher \cite{Brazma2001}.
-Standardization is critical when developing new high throughput technologies and their associated information services \cite{Brazma2001, Chicurel2002, Boguski2003}.
-Standardization efforts have been made in clinical cell analysis by flow cytometry \cite{Keeney2004}, however data interpretation has not been standardized for even low throughput FCM.
-It is one of the most difficult and time consuming aspects of the entire analytical process as well as a primary source of variation in clinical tests, and investigators have traditionally relied on intuition rather than standardized statistical inference \cite{Bagwell2004,Braylan2004,Parks1997,Suni2003}.
-In the development of standards in high throughput FCM, few progress has been done in term of Open Source software.
-In this article we propose data structures to handle flow cytometry data through the main steps of prprocessing: compensation, transformation, filtering.\\
-
-\section{Methodology}
-\subsection{Reading data}
-We first load the libraries and read the data into R as FCS object 
-(rflowcyt data format). Note than this small example works for cytoFrame object. 
-<<load library>>=
-library("flowCore")
-@ 
-<<load data,echo=FALSE,results=hide>>=
-fcs.loc <- system.file("extdata", package="flowCore")
-file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
-b08 <- read.FCS(file.location[1])
-e07 <- read.FCS(file.location[2])
-f06 <- read.FCS(file.location[3])
-@
-\begin{verbatim}
-> fcs.loc <- system.file("extdata", package="flowCore")
-> file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
-> b08 = read.FCS(file.location[1])
-> e07 = read.FCS(file.location[2])
-> f06 = read.FCS(file.location[3])
-\end{verbatim}
-
-<<showB08,eval=FALSE,echo=FALSE>>=
-show(b08)
-@
-
-We can plot the raw data for the FSC-H and SSC-H parameters.
-<<plotRaw1,eval=TRUE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
-plot(b08,plotParameters=c(1,2))
-@
-<<plotRaw2,eval=TRUE,echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
-plot(b08,plotParameters=c(3,4))
-@ 
-\begin{verbatim}
-> plot(b08,plotParameters=c(1,2))
-> plot(b08,plotParameters=c(3,4))
-> plot(e07,plotParameters=c(1,2))
-> plot(f06,plotParameters=c(1,2))
-\end{verbatim}
-
-\begin{figure}
-\centering
-\subfigure{\includegraphics[width=40mm]{plotRaw1} A}
-\subfigure{\includegraphics[width=40mm]{plotRaw2} B}
-\caption{
-  \label{fig:Fig01}{Example of smooth scatter plot.(A) FSC-H
-    \textit{vs.} SSC-H (B) FL2-H \textit{vs.} FL3-H}} 
-\end{figure}
-
-\subsection{Transformation data}
-While analyzing flow cytometry data, various parameter transformations are performed to provide user-friendly visualization and/or to perform statistical analyses and interpret the data.
-Following the standard 8 transformations are pre-defined: linear, quadratic, natural logarithm, logarithm (with a specified base), hyperlog, bi-exponential, logicle, and split-scale.
-
-As example we performed two different tramsformations on "FSC-H" channel of sample b08: a linear tranformation and a log base 2 transformation.
-
-A linear transformation is like parameter= a*parameter +b
-<<linearTransform,echo=FALSE,results=hide>>=
-linear <- linearTransform(transformationId="linear-transformation",a=2, b=0)
-data08Transform <- transform(b08,`FSC-H`=linear(`FSC-H`))
-@ 
-\begin{verbatim}
->linear <- linearTransform(transformationId="linear-transformation",a=2, b=0)
->data08Transform <- transform(b08,`FSC-H`=linear(`FSC-H`))
-\end{verbatim}
-
-<<plotTRansfo,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
-plot(data08Transform,plotParameters=c(1,2))
-@
-
-A log transformation is of type: parameter=log(parameter,logbase) * r/d
-<<logTransform,eval=FALSE,echo=FALSE,results=hide>>=
-logTransform <- logTransform(transformationId="Log-Transformation",logbase=2,r=1,d=1)
-data08LogTransform <- transform(b08,`FSC-H`=logTransform(`FSC-H`))
-@ 
-\begin{verbatim}
->logTransform <- logTransformation(transformationId="Log-Transformation", r=1, d=1,logbase=2)
->data08LogTransform <- transformb08,`FSC-H`=logTransform(`FSC-H`))
-\end{verbatim}
-
-<<ResultTransfoLinear>>=
-b08@exprs[1:5]
-#Linear transformation
-data08Transform@exprs[1:5]
-##Log base 2 transformation
-##data08LogTransform@exprs[1:5]
-@
-
-<<plotTRansfo,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
-##we should find a way to specified if an parameters has been transformed 
-## either in the title or via the label of the axis 
-plot(data08LogTransform,plotParameters=c("FSC-H","SSC-H"))
-@
-
-\subsection{Filtering data (gating)}
-\subsubsection{Rectangular gate}
-We will define rectangular gates and apply them to filter the bO8 data.
-Figure \ref{fig:Fig02} presents the results of such filtering. 
-The plot in Figure \ref{fig:Fig02}A shows the result of 
-filtering out the data points that are below 200 in the FSC-H channel.
-Figure \ref{fig:Fig02}B shows the result of 
-filtering out the data points are between 200 
-and 500  in the FSC-H channel.
-
-<<define and apply rectangleGate1,eval=TRUE, echo=FALSE,results=hide>>=
-min1 <- 200
-max1 <- Inf
-fil1 <- new("rectangleGate",filterId="nonDebris",
-            parameters="FSC-H",min=min1,max=max1)
-b08fil1 <- applyFilter(fil1,b08)
-@ 
-<<plotfiltered, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08fil1,pch="*")
-@ 
-<<define and apply polygonGate2,eval=TRUE, echo=FALSE,results=hide>>=
-min2 <- 200
-max2 <- 500
-centerFilter <- new("rectangleGate",filterId="center",
-            parameters=c("FSC-H"), min=min2, max=max2)
-b08centerFilter <- applyFilter(centerFilter,b08)
-@ 
-<<plotCenterfiltered, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08centerFilter,pch="*")
-@
-
-\begin{figure}
-\centering
-\subfigure{\includegraphics[width=40mm]{plotfiltered} A}
-\subfigure{\includegraphics[width=40mm]{plotCenterfiltered} B} 
-\caption{
-  \label{fig:Fig02}{Example of the effect of filtering with a rectangular gate.}} 
-\end{figure}
-
-\subsubsection{polygon gate}
-%%We will define a rectangular gate and apply the filter it to the bO8 data.
-<<define and apply rectangleGate,eval=TRUE,echo=FALSE>>=
-sqrcut <- matrix(c(300,300,600,600,50,300,300,50),ncol=2,nrow=4)
-sqrfilter <- new("polygonGate",filterId="nonDebris",
-            parameters=c("FSC-H","SSC-H"),boundaries= sqrcut)
-b08sqrfilter <- applyFilter(sqrfilter,b08)
-@
-
-<<plotRectanglefilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08sqrfilter,pch="*")
-@
-
-<<define and apply polygonGate,eval=TRUE,echo=FALSE>>=
-triangleCut <- matrix(c(300,400,600,300,400,300),ncol=2,nrow=3)
-trianglefilter <- new("polygonGate",filterId="nonDebris",
-            parameters=c("FSC-H","SSC-H"),boundaries= triangleCut)
-b08Trgfilter <- applyFilter(trianglefilter,b08)
-@
-
-<<plotTrianglefilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(x=b08,y=b08Trgfilter,filter=trianglefilter,colSubSet="red",pch="*")
-@
-
-\begin{figure}
-\centering
-\subfigure{\includegraphics[width=40mm]{plotRectanglefilter} A}
-\subfigure{\includegraphics[width=40mm]{plotTrianglefilter} B} 
-\caption{
-  \label{fig:Fig03}{Example of the effect of filtering with a
-    polygon gate.}} 
-\end{figure}
-
-\clearpage
-\section{Case studies}
-\subsection{BD}
-
-<<rawdata1, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,plotParameters=c("FSC-H","SSC-H"),main="B08")
-@
-<<rawdata2, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>= 
-plot(e07,plotParameters=c("FSC-H","SSC-H"),main="E07")
-@
-<<rawdata3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(f06,plotParameters=c("FSC-H","SSC-H"),main="F06")
-@ 
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{rawdata1}}
-%%\subfigure{\includegraphics[width=40mm]{rawdata2}} 
-%%\subfigure{\includegraphics[width=40mm]{rawdata3}} 
-%%\caption{
-%%  \label{fig:Fig04}{Example of the effect of filtering with a
-%%    polygon gate.}} 
-%%\end{figure}
-
-The first gate is a range gate to filter out debris
-
-<<SelectNonDebris,eval=TRUE,echo=FALSE,results=hide>>=
-min1 <- c(200,0)
-max1 <- c(800,800)
-filter1 <- new("rectangleGate",filterId="Nondebris",parameters=c("FSC-H","SSC-H"),
-  min=min1,max=max1)
-
-b08.result1 <-  applyFilter(filter1,b08)
-e07.result1 <- applyFilter(filter1,e07)
-f06.result1 <- applyFilter(filter1,f06)
-
-tab <- c("B08"=sum(b08.result1@subSet),"E07"=sum(e07.result1@subSet),"F06"=sum(f06.result1@subSet))
-@ 
-<<plotResFilter1, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result1,main="B08",pch="*")
-@ 
-<<plotResFilter2, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(e07,y=e07.result1,main="E07",pch="*")
-@ 
-<<plotResFilter3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(f06,y=f06.result1,main="F06",pch="*")
-@ 
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{plotResFilter1}}
-%%\subfigure{\includegraphics[width=40mm]{plotResFilter2}} 
-%%\subfigure{\includegraphics[width=40mm]{plotResFilter3}} 
-%%\caption{
-%%  \label{fig:Fig05}{Example of the effect of filtering with a
-%%    polygon gate.}} 
-%%\end{figure}
-
-The second gate gets the live cells
-
-<<FilterLiveCell,eval=TRUE,echo=FALSE,results=hide>>=
-filter2 = new("norm2Filter",filterId="Live Cells",scale.factor=2,method="covMcd",parameters=c("FSC-H","SSC-H"))
-
-b08.result2 = applyFilter(filter2,b08,parent=b08.result1)
-e07.result2 = applyFilter(filter2,e07,parent=e07.result1)
-f06.result2 = applyFilter(filter2,f06,parent=f06.result1)
-
-tab <- rbind(tab,c(sum(b08.result2@subSet),sum(e07.result2@subSet),sum(f06.result2@subSet)))
-
-@
-
-<<plotLivingCell1, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result2,parent=b08.result1,xlim=c(0,1024),ylim=c(0,1024),main="B08",pch="*")
-@ 
-<<plotLivingCell2, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(e07,y=e07.result2,parent=e07.result1,xlim=c(0,1024),ylim=c(0,1024),main="E07",pch="*")
-@ 
-<<plotLivingCell3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(f06,y=f06.result2,parent=f06.result1, xlim=c(0,1024),ylim=c(0,1024),main="F06",pch="*")
-@ 
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{plotLivingCell1}}
-%%\subfigure{\includegraphics[width=40mm]{plotLivingCell2}} 
-%%\subfigure{\includegraphics[width=40mm]{plotLivingCell3}} 
-%%\caption{
-%%  \label{fig:Fig06}{Example of the effect of filtering with a
-%%    polygon gate.}} 
-%%\end{figure}
-
-The third-fifth gates get the positive cells for the marker in FL1-H. 
-This is a really interesting example because it illustrates that there
-are two subpopulations. Naturally we would like to automatically find them
-In this case we want to now what percent the positive population in FL1-H 
-is of the total population.
-
-<<FL1-H, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),ylim=c(0,1024),xlim=c(0,1024))
-@ 
-%%\begin{figure}
-%%\centering
-%%\includegraphics{FL1-H}
-%%\end{figure}
-
-<<FL1Filter,eval=TRUE,echo=FALSE,results=hide>>=
-filter3 <- new("rectangleGate",filterId="FL1-H+",parameters="FL1-H",min=400,max=Inf)
-b08.result3 <- applyFilter(filter3,b08,b08.result2)
-
-filter4 <-new("norm2Filter",filterId="FL1-H+",scale.factor=2,method="covMcd",parameters=c("FSC-H","FL1-H"))
-b08.result4 <- applyFilter(filter4,b08,b08.result2)
-
-##b08.result5 <- applyFilter(filter4,b08,b08.result2@subSet-b08.result4@subSet)
-
-##ans <- c(sum(b08.result3@subSet)/sum(b08.result2@subSet),
-##sum(b08.result4@subSet)/(sum(b08.result4@subSet)+sum(b08.result5@subSet)))
-@
-
-<<FL1Filtered1, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result3,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
-          xlim=c(0,1024),ylim=c(0,1024),pch="*")
-@ 
-<<FL1Filtered2, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result4,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
-          xlim=c(0,1024),ylim=c(0,1024),pch="*")
-@ 
-<<FL1Filtered3, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result5,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
-          xlim=c(0,1024),ylim=c(0,1024),pch="*")
-@
-
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{FL1Filtered1}}
-%%\subfigure{\includegraphics[width=40mm]{FL1Filtered2}} 
-%%\subfigure{\includegraphics[width=40mm]{FL1Filtered3}} 
-%%\caption{
-%%  \label{fig:Fig07}{Example of the effect of filtering with a
-%%    polygon gate.}} 
-%%\end{figure}
-
-\clearpage
-\small{
-<<sessionInfo>>=
-sessionInfo()
-@ 
-}
 \end{document}

### 2006-12-09 R/AllClasses.R

@@ -13,13 +13,11 @@ require("Biobase")
 ## on microtiter plate from experiment.  
 ## 
 ## ---------------------------------------------------------------------------
-setClass("parameterDescription",
-	representation(parameterId="character",
-                       name="character",description="character",max.value="numeric"))
 setClass("flowFrame",                
   representation(exprs="matrix",
+				 parameters="list",
                  description="vector"),
-  prototype=list(exprs=matrix(numeric(0), nrow=0, ncol=0),
+  prototype=list(exprs=matrix(numeric(0), nrow=0, ncol=0),parameters=list(),
                  description=c(note="empty")),
  validity=function(object){
    msg <- TRUE
@@ -215,7 +213,7 @@ setClass("filter",
              msg <- TRUE
              if(!is.character(object@filterId) ||
                 length(object@filterId)!=1)
-               msg <- "\nslot 'name' must be character vector of length 1"
+               msg <- "\nslot 'filterId' must be character vector of length 1"
              if(!is.character(object@parameters))
              msg <- "\nslot 'parameters' must be  vector"
              print(object@parameters)
@@ -330,6 +328,12 @@ setClass("filterTree",
          }
          )
 
+setClass("unionFilter",representation("filter",filters="list"))
+setClass("intersectFilter",representation("filter",filters="list"))
+setClass("subsetFilter",representation("filter",left="filter",right="filter"))
+setClass("complementFilter",representation("filter",filter="filter"))
+setClass("fixedFilter",representation("filter",subset="logical"))
+
 
 ## ===========================================================================
 ## Decision Tree - listed in the standard
@@ -363,9 +367,10 @@ setClass("filterResult",
          ## a single FCS object
          ## the subset is the filter applied to the original data
          ## filterDetails is any output calculated by an algorithmic
-         representation(subSet="numeric",
-                        filterDetails="list")
-         )
+         representation("filter",
+						subSet="numeric",
+						frameId="character",
+                        filterDetails="list"))
 ## ===========================================================================
 
 

### 2006-12-09 R/readFCS-functions.R
BIG set of changes! This changes the implementation of the "filter" method by moving most of the work out of the functions and into the %in% method, which is exposed to the user. This is designed to be easier for programmers and future filter implementors.
filter() itself is now a single function that dispatches the appropriate %in%. Also, we're now carrying around some extra filterResult data to help us make filterResult a type of filter class that can be used in conjunction with other unresolved filters or with sets of filter results using the usual sorts of logical operators. 

We also have

@@ -12,7 +12,6 @@ read.FCS <- function(filename, transformation="linearize", debug=FALSE,alter.nam
 {
   stopifnot(is.character(filename), length(filename)==1, filename!="")
   con <- file(filename, open="rb")
-
   if(is.logical(transformation) && transformation || !is.null(transformation) && transformation == "linearize") {
           transformation <- TRUE
           scale <- FALSE
@@ -20,7 +19,6 @@ read.FCS <- function(filename, transformation="linearize", debug=FALSE,alter.nam
           transformation <- FALSE
           scale <- TRUE
       }
-  
   offsets <- readFCSheader(con)
   txt     <- readFCStext(con, offsets, debug)
   mat     <- readFCSdata(con, offsets, txt, transformation, debug, scale, alter.names)

### 2006-12-11 inst/doc/HowTo-flowCore.Rnw
Work on vignette
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21422 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
-b08 <- read.FCS(file.location[1])
-e07 <- read.FCS(file.location[2])
-f06 <- read.FCS(file.location[3])
+b08 <- read.FCS(file.location[1],transformation="scale")
+e07 <- read.FCS(file.location[2],transformation="scale")
+f06 <- read.FCS(file.location[3],transformation="scale")
 @
@@
 ## either in the title or via the label of the axis 
-plot(data08LogTransform,plotParameters=c("FSC-H","SSC-H"))
+plot(data08LogTransform,plotParameters=c("FSC-H","SSC-H"),main="Log2 transformed data")
 @
@@
 The plot in Figure \ref{fig:Fig02}A shows the result of 
-filtering out the data points that are below 200 in the FSC-H channel.
+filtering out the data points that are below 0.2 in the FSC-H channel.
 Figure \ref{fig:Fig02}B shows the result of 
-filtering out the data points are between 200 
-and 500  in the FSC-H channel.
+filtering out the debris that are between 0.2 and 0.8 in the FSC-H channel and 
+and 0 and 0.8 in the SSC-H channel.
 
-<<define and apply rectangleGate1,eval=TRUE, echo=FALSE,results=hide>>=
-min1 <- 200
-max1 <- Inf
-fil1 <- new("rectangleGate",filterId="nonDebris",
-            parameters="FSC-H",min=min1,max=max1)
-b08fil1 <- applyFilter(fil1,b08)
+<<define and apply rangeGate1,eval=TRUE, echo=FALSE,results=hide>>=
+rangeGate <- rectangleGate(filterId="rangeGate","FSC-H"=c(.2,Inf))
+b08rangeGate <- filter(b08,rangeGate)
 @ 
 <<plotfiltered, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08fil1,pch="*")
+plot(b08,y=b08rangeGate, pch="*",main="B08, lower debris")
 @ 
-<<define and apply polygonGate2,eval=TRUE, echo=FALSE,results=hide>>=
-min2 <- 200
-max2 <- 500
-centerFilter <- new("rectangleGate",filterId="center",
-            parameters=c("FSC-H"), min=min2, max=max2)
-b08centerFilter <- applyFilter(centerFilter,b08)
+<<define and apply ploygoneGate2,eval=TRUE, echo=FALSE,results=hide>>=
+rectGate <- rectangleGate(filterId="nonDebris","FSC-H"=c(0.2,0.8),"SSC-H"=c(0,0.8))
+b08rectGate <- filter(b08,rectGate)
+b08Filtered <- b08[b08rectGate,]
 @ 
 <<plotCenterfiltered, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08centerFilter,pch="*")
+plot(b08,y=b08rectGate,pch="*",main="B08 - Nondebris")
 @
@@
 <<define and apply rectangleGate,eval=TRUE,echo=FALSE>>=
-sqrcut <- matrix(c(300,300,600,600,50,300,300,50),ncol=2,nrow=4)
-sqrfilter <- new("polygonGate",filterId="nonDebris",
-            parameters=c("FSC-H","SSC-H"),boundaries= sqrcut)
-b08sqrfilter <- applyFilter(sqrfilter,b08)
+##sqrcut <- matrix(c(300,300,600,600,50,300,300,50),ncol=2,nrow=4)
+##sqrfilter <- new("polygonGate",filterId="nonDebris",
+##            parameters=c("FSC-H","SSC-H"),boundaries= sqrcut)
+sqrfilter <- polygonGate(filterId="nonDebris","FSC-H"=c(0.3,0.3,0.6,0.6),"SSC-H"=c(0.05,0.3,0.3,.05))
+b08sqrfilter <- filter(b08,sqrfilter)
 @
@@
 <<define and apply polygonGate,eval=TRUE,echo=FALSE>>=
-triangleCut <- matrix(c(300,400,600,300,400,300),ncol=2,nrow=3)
-trianglefilter <- new("polygonGate",filterId="nonDebris",
-            parameters=c("FSC-H","SSC-H"),boundaries= triangleCut)
-b08Trgfilter <- applyFilter(trianglefilter,b08)
+triangleCut <- matrix(c(0.3,0.4,0.6,0.3,0.4,0.3),ncol=2,nrow=3)
+colnames(triangleCut) <- c("FSC-H","SSC-H")
+trianglefilter <- polygonGate(filterId="nonDebris",boundaries= triangleCut)
+b08Trgfilter <- filter(b08,trianglefilter)
 @
@@
 
-The first gate is a range gate to filter out debris
+The first filter is a serie of rectangle gates to filter out debris
 
 <<SelectNonDebris,eval=TRUE,echo=FALSE,results=hide>>=
-min1 <- c(200,0)
-max1 <- c(800,800)
-filter1 <- new("rectangleGate",filterId="Nondebris",parameters=c("FSC-H","SSC-H"),
-  min=min1,max=max1)
-
-b08.result1 <-  applyFilter(filter1,b08)
-e07.result1 <- applyFilter(filter1,e07)
-f06.result1 <- applyFilter(filter1,f06)
-
-tab <- c("B08"=sum(b08.result1@subSet),"E07"=sum(e07.result1@subSet),"F06"=sum(f06.result1@subSet))
+filter1 <- rectangleGate(filterId="nonDebris","FSC-H"=c(0.2,0.8),"SSC-H"=c(0,0.8))
+b08.result1 <-  filter(b08,filter1)
+##e07.result1 <- filter(e07,filter1)
+##f06.result1 <- filter(f06,filter1)
+summary(b08.result1)
+##tab <- c("B08"=sum(b08.result1@subSet),"E07"=sum(e07.result1@subSet),"F06"=sum(f06.result1@subSet))
 @ 
 <<plotResFilter1, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result1,main="B08",pch="*")
+plot(b08,y=b08.result1,main="B08 nonDebris",pch="*")
 @ 
 <<plotResFilter2, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(e07,y=e07.result1,main="E07",pch="*")
+plot(e07,y=e07.result1,main="E07 nonDebris",pch="*")
 @ 
 <<plotResFilter3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(f06,y=f06.result1,main="F06",pch="*")
+plot(f06,y=f06.result1,main="F06 nonDebris",pch="*")
 @ 
@@
 <<FilterLiveCell,eval=TRUE,echo=FALSE,results=hide>>=
-filter2 = new("norm2Filter",filterId="Live Cells",scale.factor=2,method="covMcd",parameters=c("FSC-H","SSC-H"))
+##filter2 <- norm2Filter(filterId="Live Cells",scale.factor=2,method="covMcd",parameters=c("FSC-H","SSC-H"))
+filter2 <- norm2Filter("FSC-H","SSC-H",filterId="Live Cells",scale.factor=2,method="covMcd")
+b08.result2 <-  filter(b08, filter2 %subset% b08.result1)
 
-b08.result2 = applyFilter(filter2,b08,parent=b08.result1)
-e07.result2 = applyFilter(filter2,e07,parent=e07.result1)
-f06.result2 = applyFilter(filter2,f06,parent=f06.result1)
+##e07.result2 <- filter(e07,filter2,parent=e07.result1)
+##f06.result2 <- filter(f06,filter2,parent=f06.result1)
 
-tab <- rbind(tab,c(sum(b08.result2@subSet),sum(e07.result2@subSet),sum(f06.result2@subSet)))
+summary(b08.result2)
+##tab <- rbind(tab,c(sum(b08.result2@subSet),sum(e07.result2@subSet),sum(f06.result2@subSet)))
 
@@
 <<FL1Filter,eval=TRUE,echo=FALSE,results=hide>>=
-filter3 <- new("rectangleGate",filterId="FL1-H+",parameters="FL1-H",min=400,max=Inf)
-b08.result3 <- applyFilter(filter3,b08,b08.result2)
+filter3 <- rectangleGate(filterId="FL1-H+","FL1-H"=c(400,Inf))
+b08.result3 <- filter(b08,filter3 %subset%b08.result2)
 
-filter4 <-new("norm2Filter",filterId="FL1-H+",scale.factor=2,method="covMcd",parameters=c("FSC-H","FL1-H"))
-b08.result4 <- applyFilter(filter4,b08,b08.result2)
+filter4 <- norm2Filter("FSC-H","FL1-H",filterId="FL1-H+",scale.factor=2,method="covMcd")
+b08.result4 <- filter(b08,filter4 %subset% b08.result2)
 

### 2006-12-11 R/filter-methods.R

@@ -74,10 +74,11 @@ setMethod("show","rectangleGate",function(object) {
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("%in%",signature("flowFrame","polygonGate"),function(x,table) {
 	ndim = length(table@parameters)
-	#If there is only a single dimension then we have a degenerate case.
+        ##If there is only a single dimension then we have a degenerate case.
 	if(ndim==1) 
 		!is.na(cut(exprs(x)[,table@parameters[[1]]],range(table@boundaries[,1]),labels=FALSE))
 	else if(ndim==2) {
+            print(table@boundaries)
 	 	as.logical(.Call(inPolygon,exprs(x)[,table@parameters],table@boundaries))
 	} else 
 		stop("Polygonal gates only support 1 or 2 dimensional gates (for now).")

### 2006-12-11 inst/doc/filtering-flowCore.Rnw
* allow filter to return different types of class based on the input type. This will eventually allow the return of randomResultType for soft clustering operations.
* started adding split() and Subset() (the default R subset is not currently compatible with S4) methods for flowFrame to obtain flowSets based on gates.

* added a compensate() method for flowFrames and flowSets that returns a new frame or set adjusted by a spillover matrix.

* added a spillover method that tries to automatically calculation a spillover matrix from a set of comp beads/cells. This is a work in progress.

git-svn-id

@@
 
+\section{Implementing New Filters}
+
+Implementing a new filter is a fairly straightforward process requiring the definition of a filtering class, derived from the \Rclass{filter} base class, and, at the minimum, the definition of an \Rfunction{%in%} method.
+
+\subsection{Customzing a \Robject{filter} class}
+
+A \Robject{filter} subclass can be customized in one of two ways.
+First, the class returned by the \Rfunction{filter} method can be changed by definiting a \Rfunction{filterResultType} method for your filter class.
+This allows filters that return non-deterministic results (e.g.
+soft-clustering filters) to use the \Rclass{randomFilterResult}.
+
 \end{document}

### 2006-12-11 R/AllGeneric.R

@@ -44,9 +44,16 @@ if(!isGeneric("ncol"))
 
 
 ## ===========================================================================
-## Generic for applyTransformation
+## Generic for transformation
 ## ---------------------------------------------------------------------------
   setGeneric("transform")
+## ---------------------------------------------------------------------------
+
+## ===========================================================================
+## Generic for split
+## ---------------------------------------------------------------------------
+setGeneric("split")
+setGeneric("Subset",function(x,subset,...) standardGeneric("Subset"))
 ## ---------------------------------------------------------------------------
 
 
@@ -55,6 +62,7 @@ if(!isGeneric("ncol"))
 ## ---------------------------------------------------------------------------
 setGeneric("filter", function(flowObject, filter) standardGeneric("filter"),useAsDefault=FALSE)
 setGeneric("filterDetails",function(flowObject,filter,result) standardGeneric("filterDetails"))
+setGeneric("filterResultType",function(filter) standardGeneric("filterResultType"))
 ## ---------------------------------------------------------------------------
 
 

### 2006-12-11 R/flowFrame-methods.R
* allow filter to return different types of class based on the input type. This will eventually allow the return of randomResultType for soft clustering operations.
* started adding split() and Subset() (the default R subset is not currently compatible with S4) methods for flowFrame to obtain flowSets based on gates.

* added a compensate() method for flowFrames and flowSets that returns a new frame or set adjusted by a spillover matrix.

* added a spillover method that tries to automatically calculation a spillover matrix from a set of comp beads/cells. This is a work in progress.

git-svn-id

@@ -131,3 +131,14 @@ setMethod("show",signature=signature("flowFrame"),definition=function(object) {
     return(msg)
 })
 ## ==========================================================================
+
+setMethod("Subset",signature("flowFrame","filter"),function(x,subset,select,...) {
+	result = subset %in% x
+	if(!missing(select)) x[result & !is.na(result),select] else x[result & !is.na(result),]
+})
+
+setMethod("split",signature("flowFrame","filter"),function(x,f,drop=FALSE,population=NULL,...) {
+	result = x %in% f
+	if(is.null(population)) population=f@filterId
+	as(structure(list(x[result,],x[!result,]),names=c(paste(population,"+",sep=""),paste(population,"-",sep=""))),"flowSet")
+})

### 2006-12-11 R/readFCS-functions.R

@@ -8,7 +8,7 @@
 ## http://www.isac-net.org and the file
 ## fcs3.html in the doc directory
 
-read.FCS <- function(filename, transformation="linearize", debug=FALSE,alter.names=FALSE)
+read.FCS <- function(filename, transformation="linearize", debug=FALSE,alter.names=FALSE,column.pattern=NULL)
 {
   stopifnot(is.character(filename), length(filename)==1, filename!="")
   con <- file(filename, open="rb")
@@ -22,9 +22,9 @@ read.FCS <- function(filename, transformation="linearize", debug=FALSE,alter.nam
   offsets <- readFCSheader(con)
   txt     <- readFCStext(con, offsets, debug)
   mat     <- readFCSdata(con, offsets, txt, transformation, debug, scale, alter.names)
-
   close(con)
-  
+  if(!is.null(column.pattern))
+	mat = mat[,grep(column.pattern,colnames(mat))]
   if(as.integer(readFCSgetPar(txt, "$TOT"))!=nrow(mat))
     stop(paste("file", filename, "seems to corrupted."))
   if(transformation==TRUE){

### 2006-12-13 R/filter-constructors.R
* Fixed (I think) bug in filtering-flowCore.Rnw that was causing it to fail R CMD build * Improved and fixed some aspects of multipleFilterResult to get more natural usage * Implemented 1-d k-means to split a well-defined mixture of populations into 2 or more named populations. I was thinking about FCB during implementation. * Improved implementation of "split" to handle multiple population case.
MILESTONE-- I can now complete all preprocessing for my use case except for transformation and some outlier removal associated with the transformation, which I had to do "by hand."  

git-svn-id: file

@@ -78,3 +78,5 @@ norm2Filter <- function(x,y,method="covMcd",scale.factor=1,filterId="norm2Gate",
 	}
 	new("norm2Filter",parameters=c(x,y),method=method,scale.factor=scale.factor,filterId=filterId,...)
 }
+
+

### 2006-12-13 R/kmeansFilter-implementation.R

@@ -1,0 +1,31 @@
+## Constructors
+kmeansFilter = function(...,filterId="kmeans") {
+	l = length(list(...))
+	if(l>1)
+		stop("Barcode filters only operate on a single parameter.")
+	x = ..1
+	if(is.list(x)) {
+	} else
+		new("kmeansFilter",parameters=names(list(...))[1],populations=x,filterId=filterId)
+}
+
+## Filtering Methods -- we are not a logical filter so we return a vector
+## of indices indicating a population.
+setMethod("filterResultType",signature("kmeansFilter"),function(filter) "multipleFilterResult")
+setMethod("filterDetails",signature("flowFrame","kmeansFilter","ANY"),function(flowObject,filter,result) {
+	msg = paste(class(filter),"applied on",
+		deparse(substitute(flowObject))," (file:",
+		basename(flowObject@description["$FIL"]),") a",
+		class(flowObject),"object")
+	list(message=msg,filter=filter,populations=filter@populations)	
+})
+setMethod("%in%",signature("flowFrame","kmeansFilter"),function(x,table) {
+	##We accomplish the actual filtering via K-means
+	param  = table@parameters[1]
+	values = exprs(x)[,param]
+	npop   = length(table@populations)
+	km     = kmeans(values,centers=quantile(values,(1:npop)/(npop+1)))
+	#Ensure that the populations are sorted according to their center,
+	#which matches the assumption of the population vector.
+	order(km$centers)[km$cluster]
+})

### 2006-12-13 inst/doc/filtering-flowCore.Rnw
* Fixed (I think) bug in filtering-flowCore.Rnw that was causing it to fail R CMD build * Improved and fixed some aspects of multipleFilterResult to get more natural usage * Implemented 1-d k-means to split a well-defined mixture of populations into 2 or more named populations. I was thinking about FCB during implementation. * Improved implementation of "split" to handle multiple population case.
MILESTONE-- I can now complete all preprocessing for my use case except for transformation and some outlier removal associated with the transformation, which I had to do "by hand."  

git-svn-id: http

@@
 
-Implementing a new filter is a fairly straightforward process requiring the definition of a filtering class, derived from the \Rclass{filter} base class, and, at the minimum, the definition of an \Rfunction{%in%} method.
+Implementing a new filter is a fairly straightforward process requiring the definition of a filtering class, derived from the \Rclass{filter} base class, and, at the minimum, the definition of an \Rfunction{\%in\%} method.
 
-\subsection{Customzing a \Robject{filter} class}
+\subsection{Customizing a \Rclass{filter} class}
 

### 2006-12-16 R/filter-functions.R
* Converted multipleFilterResult to use factors rather than keeping the two separate. This makes more sense vis a vis %in% operators and makes the initialization of the multipleFilterResult object unnecessary.
* Added the ability to %subset% a multipleFilterResult such a way that the subsetting filter is applied to each population to construct a narrower population.

git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21517 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1,3 +1,11 @@
+print.filterSummary <- function(x,...) {
+	if(length(x$name) == 1) {
+		with(x,cat(sprintf("%s: %d of %d (%.2f%%)\n",name,true,n,100*p)))
+	} else {
+		for(i in seq(along=x$name))
+			with(x,cat(sprintf("%s: %d of %d (%.2f%%)\n",name[i],true[i],n[i],100*p[i])))
+	}
+}
 
 ## ==========================================================================
 ## filtering function for polygonal gates

### 2006-12-16 R/flowFrame-methods.R

@@ -1,5 +1,5 @@
 setMethod("identifier","flowFrame",function(object) if(is.null(object@description["GUID"])) object@description["$FIL"] else object@description["GUID"])
-setMethod("featureNames","flowFrame",function(object) object@description[gsub("N","S",names(colnames(exprs(object))))])
+
 
 ## ==========================================================================
 ## accessor method for slot exprs
@@ -45,6 +45,18 @@ setReplaceMethod("description",
 setMethod("colnames",
   signature="flowFrame", definition=function(x, do.NULL="missing",
   prefix="missing") colnames(exprs(x)), valueClass="character")
+setMethod("featureNames","flowFrame",function(object) object@parameters$desc)
+setMethod("names","flowFrame",function(x) {
+	cn = colnames(x)
+	fn = featureNames(x)
+	if(length(fn) == length(cn)) {
+		for(i in seq(along=fn)) {
+			if(!is.na(fn[i])) cn[i] = paste("<",cn[i],"> ",fn[i],sep="")
+		}
+	}
+	cn
+})
+
 ## ==========================================================================
 
 
@@ -88,6 +100,8 @@ setMethod("[",
          { exprs(x)[ , j, ..., drop=drop] },
          { exprs(x)[i,  , ..., drop=drop] },
          { exprs(x)[ ,  , ..., drop=drop] } )
+	if(!missing(j)) 
+		x@parameters = x@parameters[j,]
     x
   },
   valueClass="flowFrame")
@@ -137,7 +151,7 @@ setMethod("ncol",
 setMethod("show",signature=signature("flowFrame"),definition=function(object) {
    dm <- dim(exprs(object))
     msg <- paste("flowFrame object with ", dm[1], " cells and ", 
-        dm[2], " observables:\n", paste(colnames(exprs(object)), 
+        dm[2], " observables:\n", paste(names(object), 
             collapse = " "), "\nslot 'description' has ", length(description(object)), 
         " elements\n", sep = "")
     cat(msg)
@@ -146,9 +160,10 @@ setMethod("show",signature=signature("flowFrame"),definition=function(object) {
 ## ==========================================================================
 
 setMethod("Subset",signature("flowFrame","filter"),function(x,subset,select,...) {
-	result = subset %in% x
+	result = x %in% subset
[…]

### 2006-12-16 R/kmeansFilter-implementation.R

@@ -11,7 +11,6 @@ kmeansFilter = function(...,filterId="kmeans") {
 
 ## Filtering Methods -- we are not a logical filter so we return a vector
 ## of indices indicating a population.
-setMethod("filterResultType",signature("kmeansFilter"),function(filter) "multipleFilterResult")
 setMethod("filterDetails",signature("flowFrame","kmeansFilter","ANY"),function(flowObject,filter,result) {
 	msg = paste(class(filter),"applied on",
 		deparse(substitute(flowObject))," (file:",
@@ -27,5 +26,13 @@ setMethod("%in%",signature("flowFrame","kmeansFilter"),function(x,table) {
 	km     = kmeans(values,centers=quantile(values,(1:npop)/(npop+1)))
 	#Ensure that the populations are sorted according to their center,
 	#which matches the assumption of the population vector.
-	order(km$centers)[km$cluster]
+	structure(as.integer(order(km$centers)[km$cluster]),levels=table@populations)
+})
+setMethod("length",signature("kmeansFilter"),function(x) length(x@populations))
+setMethod("show",signature("kmeansFilter"),function(object) {
+	msg = paste("A k-means filter named '",object@filterId,"': ",table@parameters[1]," (",
+		paste(table@populations,sep=","),")")
+	cat(msg)
+	cat("\n")
+	invisible(msg)
 })

### 2006-12-16 R/AllClasses.R
* Converted multipleFilterResult to use factors rather than keeping the two separate. This makes more sense vis a vis %in% operators and makes the initialization of the multipleFilterResult object unnecessary.
* Added the ability to %subset% a multipleFilterResult such a way that the subsetting filter is applied to each population to construct a narrower population.

git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21517 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -15,9 +15,11 @@ require("Biobase")
 ## ---------------------------------------------------------------------------
 setClass("flowFrame",                
   representation(exprs="matrix",
-				 parameters="list",
+				 parameters="AnnotatedDataFrame",
                  description="vector"),
-  prototype=list(exprs=matrix(numeric(0), nrow=0, ncol=0),parameters=list(),
+  prototype=list(exprs=matrix(numeric(0), nrow=0, ncol=0),
+				 parameters=new("AnnotatedDataFrame",
+					data=data.frame(name=I(character(0))),varMetadata=data.frame(labelDescription="Name in frame",row.names="name")),
                  description=c(note="empty")),
  validity=function(object){
    msg <- TRUE
@@ -370,7 +372,7 @@ setClass("filterResult",
 setClass("randomFilterResult",
 	representation("filter",subSet="numeric",frameId="character",filterDetails="list"))
 setClass("multipleFilterResult",
-	representation("filter",subSet="integer",populations="character",frameId="character",filterDetails="list"))
+	representation("filter",subSet="factor",frameId="character",filterDetails="list"))
 
 
 ## ===========================================================================

### 2006-12-16 R/filter-methods.R

@@ -11,7 +11,7 @@
 ## AND associating a filterResult with a particular flowFrame if possible.
 setMethod("filter",signature("flowFrame","filter"),function(flowObject,filter) {
 	result = flowObject %in% filter
-	resultType = filterResultType(filter)
+	resultType = filterResultType(filter,result)
 	details= filterDetails(flowObject,filter,result)
 	new(resultType,
 		parameters=filter@parameters,
@@ -22,6 +22,9 @@ setMethod("filter",signature("flowFrame","filter"),function(flowObject,filter) {
 ## Printing out a filter should give us something at least mildly sensible.
 setMethod("show","filter",function(object) 
 	cat(paste("A filter named '",object@filterId,"'\n",sep="")))
+## Most filters define only a single population
+setMethod("length","filter",function(x) 1)
+setMethod("identifier","filter",function(object) object@filterId)
 ## ==========================================================================
 ## filterDetails constructs the metadata carried around by the filterResult
 ## this should generally be specific enough, but you may want to override your
@@ -33,7 +36,11 @@ setMethod("filterDetails",signature("flowFrame","filter","ANY"),function(flowObj
 		class(flowObject),"object")
 	list(message=msg,filter=filter)
 })
-setMethod("filterResultType","filter",function(filter) "filterResult")
+#By default we know some different types of results we can get from an %in% operation.
+setMethod("filterResultType",signature("filter","logical"),function(filter,result) "filterResult")
+setMethod("filterResultType",signature("filter","factor"),function(filter,result)  "multipleFilterResult")
+setMethod("filterResultType",signature("filter","numeric"),function(filter,result) "randomFilterResult")
+
 
 ## ==========================================================================
 ## support methods for filtering flowFrame Object using rectangleGate
@@ -147,69 +154,6 @@ setMethod("filterDetails",signature("flowFrame","norm2Filter","ANY"),function(fl
 	list(message=msg,filter=filter,cov=attr(result,'cov'),center=attr(result,'center'))	
 })
 
-## ==========================================================================
-## The | operator returns an object representing the union of two filters. A simple
-## optimization would be to linearize the union of a filter and another union filter.
[…]

### 2007-01-22 inst/doc/HowTo-flowCore.Rnw
* Converted multipleFilterResult to use factors rather than keeping the two separate. This makes more sense vis a vis %in% operators and makes the initialization of the multipleFilterResult object unnecessary.
* Added the ability to %subset% a multipleFilterResult such a way that the subsetting filter is applied to each population to construct a narrower population.

git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@21517 bc3139a8-67e5-0310-9ffc-ced21a209358
fix vignette remains documentation error with check
git-svn-id: file:///home/git/hedgehog.fhcrc.o

@@
 @ 
+<<plotTOfilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+plot(b08,y="FSC-H", pch="*",main="B08, with debris")
+@
 <<plotfiltered, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08rangeGate, pch="*",main="B08, lower debris")
+plot(b08[b08rangeGate],y="FSC-H", pch="*",main="B08, lower debris removed")
 @
+
+\begin{figure}
+\centering
+\subfigure{\includegraphics[width=40mm]{plotTOfilter} A}
+\subfigure{\includegraphics[width=40mm]{plotCenterfiltered} B} 
+\caption{
+  \label{fig:Fig02}{Example of the effect of filtering out low intensities debris with a rectangular gate.}} 
+\end{figure}
+
 <<define and apply ploygoneGate2,eval=TRUE, echo=FALSE,results=hide>>=
@@
 @ 
-<<plotCenterfiltered, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08rectGate,pch="*",main="B08 - Nondebris")
+<<plot3, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+plot(b08,y=c("FL1-H","FL2-H"),pch="*",main="B08 - Nondebris")
+@
+<<plot4, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+plot(b08Filtered,y=c("FL1-H","FL2-H"),pch="*",main="B08 - Nondebris")
 @
@@
 \centering
-\subfigure{\includegraphics[width=40mm]{plotfiltered} A}
-\subfigure{\includegraphics[width=40mm]{plotCenterfiltered} B} 
+\subfigure{\includegraphics[width=40mm]{plot3} A}
+\subfigure{\includegraphics[width=40mm]{plot4} B} 
 \caption{
-  \label{fig:Fig02}{Example of the effect of filtering with a rectangular gate.}} 
+  \label{fig:Fig03}{Second example of the effect of filtering out low intensities debris with a rectangular gate.}} 
 \end{figure}
@@
 <<plotRectanglefilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08sqrfilter,pch="*")
+plot(b08[b08sqrfilter],pch="*")
 @
 
-<<define and apply polygonGate,eval=TRUE,echo=FALSE>>=
+<<define and apply polygonGate,eval=FALSE,echo=FALSE>>=
 triangleCut <- matrix(c(0.3,0.4,0.6,0.3,0.4,0.3),ncol=2,nrow=3)
@@
 
-<<plotTrianglefilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+<<plotTrianglefilter, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
 plot(x=b08,y=b08Trgfilter,filter=trianglefilter,colSubSet="red",pch="*")
@@
 
-\begin{figure}
-\centering
-\subfigure{\includegraphics[width=40mm]{plotRectanglefilter} A}
-\subfigure{\includegraphics[width=40mm]{plotTrianglefilter} B} 
-\caption{
-  \label{fig:Fig03}{Example of the effect of filtering with a
-    polygon gate.}} 
-\end{figure}
+%%\begin{figure}
+%%\\centering
+%%\\subfigure{\includegraphics[width=40mm]{plotRectanglefilter} A}
+%%\\subfigure{\includegraphics[width=40mm]{plotTrianglefilter} B} 
+%%\\caption{
+%%\  \label{fig:Fig03}{Example of the effect of filtering with a
+%%\    polygon gate.}} 
+%%\\end{figure}
 

### 2007-01-22 R/transformation-functions.R

@@ -15,6 +15,12 @@ linearTransform <- function(transformationId,a=1,b=0){
 ## Quadratic transformation function
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 quadraticTransform <- function(transformationId,a,b,c){
+  if(!is.double(a)) 
+      stop("a must be numeric")
+    if(!is.double(b))
+       stop("b must be numeric")
+  if(!is.double(c))
+       stop("c must be numeric")
     function(x){
         x <- a*x^2 + b*x + c
     }
@@ -24,6 +30,10 @@ quadraticTransform <- function(transformationId,a,b,c){
 ## Natural logarithm transformation function
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 lnTransform <- function(transformationId,r,d){
+    if(!is.double(r) || r <= 0)
+       stop("r must be numeric and positive")
+    if(!is.double(d) || d <=0)
+       stop("d must be numeric")
     function(x){
      x<-log(x)*(r/d)
  }
@@ -32,7 +42,15 @@ lnTransform <- function(transformationId,r,d){
 ## ==========================================================================
 ## Logarithm transformation function
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-logTransform <- function(transformationId,logbase,r,d){
+logTransform <- function(transformationId,logbase=10,r,d){
+     if(!is.double(r) || r <= 0)
+       stop("r must be numeric and positive")
+    if(!is.double(d) || d <=0)
+       stop("d must be numeric")
+      if(!is.double(r) || r <=0)
+       stop("r must be numeric and positive")
+    if(!is.double(logbase) || logbase <= 1)
+       stop("logabse must be a pnumeric greater than 1")
     function(x){
         x <- log(x,logbase)*(r/d)
     }
@@ -41,7 +59,8 @@ logTransform <- function(transformationId,logbase,r,d){
 ## ==========================================================================
 ## General biexponential transformation function
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
[…]

### 2007-01-29 inst/doc/HowTo-flowCore.Rnw
correct vignette add man page
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@22286 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-<<plotTRansfo,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
+<<plotTransfo1,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
 plot(data08Transform,plotParameters=c(1,2))
@@
 
-<<plotTRansfo,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
+<<plotTransfo2,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
 ##we should find a way to specified if an parameters has been transformed 
@@
 @ 
-<<plotTOfilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+<<plotTofilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
 plot(b08,y="FSC-H", pch="*",main="B08, with debris")
@@
 \centering
-\subfigure{\includegraphics[width=40mm]{plotTOfilter} A}
-\subfigure{\includegraphics[width=40mm]{plotCenterfiltered} B} 
+\subfigure{\includegraphics[width=40mm]{plotTofilter} A}
+\subfigure{\includegraphics[width=40mm]{plotfiltered} B} 
 \caption{
@@
 @ 
-<<FL1Filtered2, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
+<<FL1Filtered2, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
 plot(b08,y=b08.result4,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
@@
 @ 
-<<FL1Filtered3, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
+<<FL1Filtered3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
 plot(b08,y=b08.result5,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
@@
 \clearpage
+\bibliographystyle{cyto} 
+\bibliography{cytoref}
+\clearpage
 \small{

### 2007-02-05 inst/doc/HowTo-flowCore.Rnw
Fix vignette
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@22337 bc3139a8-67e5-0310-9ffc-ced21a209358
vignette andd man pages fixes
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@22434 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \newcommand{\Rclass}[1]{{\textit{#1}}}
-
-\bibliographystyle{plainnat}
 
@@
 We can plot the raw data for the FSC-H and SSC-H parameters.
-<<plotRaw1,eval=TRUE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
+<<plotRaw1,eval=TRUE, echo=FALSE,fig=TRUE, prefix=FALSE,include=FALSE>>=
 plot(b08,plotParameters=c(1,2))
 @
-<<plotRaw2,eval=TRUE,echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
+<<plotRaw2,eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE,include=FALSE>>=
 plot(b08,plotParameters=c(3,4))
@@
 \centering
-\subfigure{\includegraphics[width=40mm]{plotRaw1} A}
-\subfigure{\includegraphics[width=40mm]{plotRaw2} B}
+\subfigure{\includegraphics[width=0.4\textwidth]{plotRaw1} A}
+\subfigure{\includegraphics[width=0.4\textwidth]{plotRaw2} B}
 \caption{
@@
 \centering
-\subfigure{\includegraphics[width=40mm]{plotTofilter} A}
-\subfigure{\includegraphics[width=40mm]{plotfiltered} B} 
+\subfigure{\includegraphics[width=0.4\textwidth]{plotTofilter} A}
+\subfigure{\includegraphics[width=0.4\textwidth]{plotfiltered} B} 
 \caption{
@@
 \centering
-\subfigure{\includegraphics[width=40mm]{plot3} A}
-\subfigure{\includegraphics[width=40mm]{plot4} B} 
+\subfigure{\includegraphics[width=0.4\textwidth]{plot3} A}
+\subfigure{\includegraphics[width=0.4\textwidth]{plot4} B} 
 \caption{
@@
 %%\begin{figure}
-%%\\centering
-%%\\subfigure{\includegraphics[width=40mm]{plotRectanglefilter} A}
-%%\\subfigure{\includegraphics[width=40mm]{plotTrianglefilter} B} 
-%%\\caption{
-%%\  \label{fig:Fig03}{Example of the effect of filtering with a
-%%\    polygon gate.}} 
-%%\\end{figure}
+%%\centering
+%%\subfigure{\includegraphics[width=0.4\textwidth]{plotRectanglefilter} A}
+%%\subfigure{\includegraphics[width=0.4\textwidth]{plotTrianglefilter} B} 
+%%\caption{
+%%  \label{fig:Fig03}{Example of the effect of filtering with a
+%%    polygon gate.}} 
+%%\end{figure}
 
@@
 %%\centering
-%%\subfigure{\includegraphics[width=40mm]{rawdata1}}
-%%\subfigure{\includegraphics[width=40mm]{rawdata2}} 
-%%\subfigure{\includegraphics[width=40mm]{rawdata3}} 
+%%\subfigure{\includegraphics[width=0.4\textwidth]{rawdata1}}
+%%\subfigure{\includegraphics[width=0.4\textwidth]{rawdata2}} 
+%%\subfigure{\includegraphics[width=0.4\textwidth]{rawdata3}} 
 %%\caption{
@@
 %%\centering
-%%\subfigure{\includegraphics[width=40mm]{plotResFilter1}}
-%%\subfigure{\includegraphics[width=40mm]{plotResFilter2}} 
-%%\subfigure{\includegraphics[width=40mm]{plotResFilter3}} 
+%%\subfigure{\includegraphics[width=0.4\textwidth]{plotResFilter1}}
+%%\subfigure{\includegraphics[width=0.4\textwidth]{plotResFilter2}} 
+%%\subfigure{\includegraphics[width=0.4\textwidth]{plotResFilter3}} 
 %%\caption{
@@
 %%\centering
-%%\subfigure{\includegraphics[width=40mm]{plotLivingCell1}}
-%%\subfigure{\includegraphics[width=40mm]{plotLivingCell2}} 
-%%\subfigure{\includegraphics[width=40mm]{plotLivingCell3}} 
+%%\subfigure{\includegraphics[width=0.4\textwidth]{plotLivingCell1}}
+%%\subfigure{\includegraphics[width=0.4\textwidth]{plotLivingCell2}} 
+%%\subfigure{\includegraphics[width=0.4\textwidth]{plotLivingCell3}} 
 %%\caption{
@@
 \clearpage
-\bibliographystyle{cyto} 
+\bibliographystyle{plainnat} 
 \bibliography{cytoref}

### 2007-02-05 R/filter-constructors.R
vignette andd man pages fixes
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@22434 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -43,7 +43,7 @@ polytopeGate <- function(filterId="polytopeGate", .gate, ...) {
       ##nrowGate <- max(unlist(lapply(list(...),length)))
       .gate <- sapply(if(missing(.gate)) list(...) else .gate, function(x) x)
          
-    new("polytopeGate",filterId=filterId,parameters=colnames(.gate), boundaries=.gate)
+    new("polytopeGate", filterId=filterId, parameters=colnames(.gate), boundaries=.gate)
 }
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 
@@ -55,7 +55,7 @@ ellipsoidGate <- function(filterId="ellipsoidGate", .gate, distance,...) {
     if(missing(.gate) || !is.matrix(.gate))
       .gate <- sapply(if(missing(.gate)) list(...) else .gate, function(x) x)
       
-    new("ellipsoidGate",filterId=filterId,parameters=colnames(.gate), focus=.gate,distance=distance)
+    new("ellipsoidGate", filterId=filterId, parameters=colnames(.gate), focus=.gate, distance=distance)
 }
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 

### 2007-02-10 inst/doc/HowTo-flowCore.Rnw
add man page fix read.FCS function (hopefully) work on the vignette
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@22523 bc3139a8-67e5-0310-9ffc-ced21a209358

--- 2007-01-29 inst/doc/HowTo-flowCore.Rnw
@@
 
-\bibliographystyle{plainnat}
-
-\title{How to use flowCore package}
+\title{flowCore: data structures for flow cytometry data}
 \author{N. Le Meur F. Hahne B. Ellis P. Haaland}
@@
 We believe that the Open Source statistical software R in conjunction with the Bioconductor Project fills all of these requirements.
-Consequently we have developed a Bioconductor package that we call flowcore.
-The flowcore package is not intended to be a complete analysis package for FCM data, rather, we see it as providing a clear object model and a collection of standard tools that enable R as an informatics research platform for flow cytometry.
-One of the important issues that we have addressed in the flowcore package is that of using a standardized representation that will insure compatibility with existing technologies for data analysis and will support collaboration and interoperability of new methods as they are developed.
+Consequently we have developed a Bioconductor package that we call flowCore.
+The flowCore package is not intended to be a complete analysis package for FCM data, rather, we see it as providing a clear object model and a collection of standard tools that enable R as an informatics research platform for flow cytometry.
+One of the important issues that we have addressed in the flowCore package is that of using a standardized representation that will insure compatibility with existing technologies for data analysis and will support collaboration and interoperability of new methods as they are developed.
 In order to do this, we have followed the current standardized descriptions of FCM data analysis as being developed under NIH Grant xxxx [n].
-We believe that researchers will find flowcore to be a solid foundation for future development of new methods to attack the many interesting open research questions in FCM data analysis.\\
-\textbf{Methods} We propose a variety different data structures. We have implemented the classes and methods in the Bioconductor package flowcore. We illustrate their use with X case studies.\\
+We believe that researchers will find flowCore to be a solid foundation for future development of new methods to attack the many interesting open research questions in FCM data analysis.\\
+\textbf{Methods} We propose a variety different data structures. We have implemented the classes and methods in the Bioconductor package flowCore. 
+We illustrate their use with X case studies.\\
 \textbf{Results} We hope that those proposed data structures will the base for the development of many tools for the analysis of high throughput flowcytometry.\\
@@
 Traditionnally, flow cytometry has been a tube-based technique limited to small-scale laboratory and clinical studies. 
-High throughput methods for flow cytometry have recently been developed for drug discovery and advanced research methods \cite{Gasparetto2004, Young2004}.
-As an example, the flow cytometry high content screening (FC-HCS) can process up to a thousand samples daily at a single workstation, and the results have been equivalent or superior to traditional manual multiparameter staining and analysis techniques.
-\\
-The amount of information generated by high throughput technologies such as FC-HCS need to be transformed into executive summaries (which are brief enough) for creative studies by a human researcher \cite{Brazma2001}.
-Standardization is critical when developing new high throughput technologies and their associated information services \cite{Brazma2001, Chicurel2002, Boguski2003}.
-Standardization efforts have been made in clinical cell analysis by flow cytometry \cite{Keeney2004}, however data interpretation has not been standardized for even low throughput FCM.
-It is one of the most difficult and time consuming aspects of the entire analytical process as well as a primary source of variation in clinical tests, and investigators have traditionally relied on intuition rather than standardized statistical inference \cite{Bagwell2004,Braylan2004,Parks1997,Suni2003}.
-In the development of standards in high throughput FCM, few progress has been done in term of Open Source software.
-In this article we propose data structures to handle flow cytometry data through the main steps of prprocessing: compensation, transformation, filtering.\\
+High throughput methods for flow cytometry have recently been developed for drug discovery and advanced research methods 
+\cite{Gasparetto2004, Young2004}. As an example, the flow cytometry high content screening (FC-HCS) can process up 
+to a thousand samples daily at a single workstation, and the results have been equivalent or superior to traditional 
+manual multiparameter staining and analysis techniques. \\
 
-\section{Methodology}
-\subsection{Reading data}
-We first load the libraries and read the data into R as FCS object 
-(rflowcyt data format). Note than this small example works for cytoFrame object. 
-<<load library>>=
-library("flowCore")
-@ 
-<<load data,echo=FALSE,results=hide>>=
-fcs.loc <- system.file("extdata", package="flowCore")
-file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
-b08 <- read.FCS(file.location[1],transformation="scale")
-e07 <- read.FCS(file.location[2],transformation="scale")
-f06 <- read.FCS(file.location[3],transformation="scale")
-@
-\begin{verbatim}
-> fcs.loc <- system.file("extdata", package="flowCore")
-> file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
-> b08 = read.FCS(file.location[1])
-> e07 = read.FCS(file.location[2])
-> f06 = read.FCS(file.location[3])
-\end{verbatim}
+The amount of information generated by high throughput technologies such as FC-HCS need to be transformed into 
+executive summaries (which are brief enough) for creative studies by a human researcher \cite{Brazma2001}. 
+Standardization is critical when developing new high throughput technologies and their associated information services 
+\cite{Brazma2001, Chicurel2002, Boguski2003}. Standardization efforts have been made in clinical cell analysis 
+by flow cytometry \cite{Keeney2004}, however data interpretation has not been standardized for even low throughput FCM. 
+It is one of the most difficult and time consuming aspects of the entire analytical process as well as a primary source 
+of variation in clinical tests, and investigators have traditionally relied on intuition rather than standardized statistical inference 
+\cite{Bagwell2004,Braylan2004,Parks1997,Suni2003}. In the development of standards in high throughput FCM, 
+few progress has been done in term of Open Source software. In this article we propose R data structures to handle 
+flow cytometry data through the main steps of preprocessing: compensation, transformation, filtering.\\
 
-<<showB08,eval=FALSE,echo=FALSE>>=
-show(b08)
+The aim is to merge both prada  and rflowcyt \cite{LeMeur2006} into one core package wich is compliant 
+with the data exchange standards that are currently developed in the community \cite{Spidlen2006}.
+
+Visualization as well as quality control will than be part of the utility packages that depend on the data structures defined in the flowCore package.
+
+\section{Data model}
+The package \Rpackage{flowCore} offers the possibility to store single measurement (FCS file) or combine several individual
+measurements in the confines of a single experiment and to include all the necessary metadata. Its object
+model tries to stay close to the familiar micro-array data structures (expressionSet) making use of already
+defined generic functions.\\
+
+\subsection{flowFrame}
+The objects of class \Robject{flowFrame} are the containers for storing individual 
+cytometry measurements originated from one FCS file. \\
+%%need description of the slots and maybe methods associated
+
+\subsection{flowSet}
+An flow cytometry experiment usually include several sample and several measurements. 
+A data structure is thus needed to store those collections of measurements.  
+The package \Rpackage{flowCore} stores the measurement and the associated data 
+information in objects of class \Robject{flowSet}. 
+A\Robject{flowSet} object contains two main slot exprs, an enviroment data astructure 
+that stores the flow cytometry measurements and the phenoData. The phenoData slot 
+contains all the relevant experiment-wide meta-data, \textit{e.g.} 
+phenotypic and experimental information relevant to the management and/or the analysis of the data.\\
+
+\section{Preprocessing Methods}
+\subsection{Compensation}
+Fluorescence compensation is the process by which the amounts of spectral overlap are 
+estimated and subtracted from the total detected signals to yield an estimate of the 
+actual amount of each dye \cite{Herzenberg2006}. Single-laser, two-color FACS experiments 
+in the mid-1970s showed that fluorescence compensation is important. 
+It is now recognized as an obligate step in most FACS analyses.
+%%definition: Compensation vs spillover
+
+\subsection{Transformation}
+While analyzing flow cytometry data, various parameter transformations are performed 
+to provide user-friendly visualization and/or to perform statistical analyses and interpret the data. 
+Following the standard \cite{Spidlen2006} 8 transformations are pre-defined: linear, quadratic, natural logarithm, 
+logarithm (with a specified base), hyperlog, bi-exponential, logicle, and split-scale.\\
+
+%%Can we define other?   
+%%can be applied when reading in the data.
+
+\subsection{Filter}
+The role of filtering strategies are to identify and characterize diffrent cells population in sample.
+%% Filtering vs Gating
+%%Byron's vignette?
+%%several approach from basic to DAG ...\\
+
+\section{Case study}
+\subsection{dataset}
+Our experiment is composed of 5 96-well plates\\
+%%Perry can we use that example or do you prefere an other one? 
+%% Can you add a paragraph describing the experiment?
+
+First we load the library and read in some experimental data files.
+<<loadlib, echo=TRUE, results= hide>>=
+library(flowCore)
 @
 
-We can plot the raw data for the FSC-H and SSC-H parameters.
-<<plotRaw1,eval=TRUE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
-plot(b08,plotParameters=c(1,2))
+<<readData, echo=TRUE, results=hide>>=
+fcs.loc <- system.file("extdata/compdata/data", package="flowCore")
+file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
+experiment <- read.flowSet(file.location, transformation = "linearize")
+experiment
 @ 
-<<plotRaw2,eval=TRUE,echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
-plot(b08,plotParameters=c(3,4))
-@ 
-\begin{verbatim}
-> plot(b08,plotParameters=c(1,2))
-> plot(b08,plotParameters=c(3,4))
-> plot(e07,plotParameters=c(1,2))
-> plot(f06,plotParameters=c(1,2))
-\end{verbatim}
-
-\begin{figure}
-\centering
-\subfigure{\includegraphics[width=40mm]{plotRaw1} A}
-\subfigure{\includegraphics[width=40mm]{plotRaw2} B}
-\caption{
-  \label{fig:Fig01}{Example of smooth scatter plot.(A) FSC-H
-    \textit{vs.} SSC-H (B) FL2-H \textit{vs.} FL3-H}} 
-\end{figure}
-
-\subsection{Transformation data}
-While analyzing flow cytometry data, various parameter transformations are performed to provide user-friendly visualization and/or to perform statistical analyses and interpret the data.
-Following the standard 8 transformations are pre-defined: linear, quadratic, natural logarithm, logarithm (with a specified base), hyperlog, bi-exponential, logicle, and split-scale.
-
-As example we performed two different tramsformations on "FSC-H" channel of sample b08: a linear tranformation and a log base 2 transformation.
-
-A linear transformation is like parameter= a*parameter +b
-<<linearTransform,echo=FALSE,results=hide>>=
-linear <- linearTransform(transformationId="linear-transformation",a=2, b=0)
-data08Transform <- transform(b08,`FSC-H`=linear(`FSC-H`))
-@ 
-\begin{verbatim}
->linear <- linearTransform(transformationId="linear-transformation",a=2, b=0)
->data08Transform <- transform(b08,`FSC-H`=linear(`FSC-H`))
-\end{verbatim}
-
-<<plotTransfo1,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
-plot(data08Transform,plotParameters=c(1,2))
+\subsection{Quality Assessment}
+Quickly look at the range of the intensities between samples 
+other statistics sur as ... cf Janet paper.
+<<summaryRange>>=
+rang <- round(fsApply(experiment, each_col, range), 2)
+rownames(rang) <- paste(rep(dir(fcs.loc), each=2), c("lower","upper") ,sep="")
+rang
 @
 
-A log transformation is of type: parameter=log(parameter,logbase) * r/d
-<<logTransform,eval=FALSE,echo=FALSE,results=hide>>=
-logTransform <- logTransform(transformationId="Log-Transformation",logbase=2,r=1,d=1)
-data08LogTransform <- transform(b08,`FSC-H`=logTransform(`FSC-H`))
-@ 
-\begin{verbatim}
->logTransform <- logTransformation(transformationId="Log-Transformation", r=1, d=1,logbase=2)
->data08LogTransform <- transformb08,`FSC-H`=logTransform(`FSC-H`))
-\end{verbatim}
-
-<<ResultTransfoLinear>>=
-b08@exprs[1:5]
-#Linear transformation
-data08Transform@exprs[1:5]
-##Log base 2 transformation
-##data08LogTransform@exprs[1:5]
+\subsection{Preprocessing}
+<<compensationMatrix, echo=FALSE, results= hide>>=
+comp.loc <- system.file("extdata/compdata", package="flowCore")
+comp.location <- paste(comp.loc, dir(comp.loc), sep="/")
+comp.mat <- as.matrix(read.table(comp.location[1], header=TRUE, skip=2, check.names = FALSE))
 @
 
-<<plotTransfo2,eval=FALSE, echo=FALSE,fig=TRUE,prefix=FALSE,include=FALSE>>=
-##we should find a way to specified if an parameters has been transformed 
-## either in the title or via the label of the axis 
-plot(data08LogTransform,plotParameters=c("FSC-H","SSC-H"),main="Log2 transformed data")
+Compensation is ....
+<<compensateExp, echo=TRUE>>=
+expCompensate <- compensate(experiment,comp.mat)
 @
 
-\subsection{Filtering data (gating)}
-\subsubsection{Rectangular gate}
-We will define rectangular gates and apply them to filter the bO8 data.
-Figure \ref{fig:Fig02} presents the results of such filtering. 
-The plot in Figure \ref{fig:Fig02}A shows the result of 
-filtering out the data points that are below 0.2 in the FSC-H channel.
-Figure \ref{fig:Fig02}B shows the result of 
-filtering out the debris that are between 0.2 and 0.8 in the FSC-H channel and 
-and 0 and 0.8 in the SSC-H channel.
+Transformation
+<<scaleScatter, echo=TRUE>>=
+scaleTrans <- scaleTransform("scatter-scale", a=0,  b=1023)
+expCompensate <- transform(expCompensate,`FSC-H`= scaleTrans(`FSC-H`), `SSC-H`= scaleTrans(`SSC-H`))
+@
 
-<<define and apply rangeGate1,eval=TRUE, echo=FALSE,results=hide>>=
-rangeGate <- rectangleGate(filterId="rangeGate","FSC-H"=c(.2,Inf))
-b08rangeGate <- filter(b08,rangeGate)
+Filter to leave out debris:
+<<rectangleFilter, echo=TRUE>>=
+filterDebris <- rectangleGate("FSC-H"=c(0.2, 1), "SSC-H"=c(0.04, 1))
+expResult <- fsApply(expCompensate, function(x) filter(x, filterDebris))
+
+cell <- Subset(expCompensate[[1]],expResult[[1]])
 @
-<<plotTofilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y="FSC-H", pch="*",main="B08, with debris")
+
+<<plotWithDebris, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
+print(plot(expCompensate[[1]], xlim=c(0, 1), ylim=c(0, 1), 
+main="File1 (compensated, scaled)"))
 @
-<<plotfiltered, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08[b08rangeGate],y="FSC-H", pch="*",main="B08, lower debris removed")
+
+<<plotCell,echo=False, fig=TRUE, prefix=FALSE, include=FALSE>>=
+print(plot(cell, xlim=c(0, 1), ylim=c(0, 1), 
+     main="File1 (compensated, scaled, debris removed)"))
 @
@@
 \centering
-\subfigure{\includegraphics[width=40mm]{plotTofilter} A}
-\subfigure{\includegraphics[width=40mm]{plotfiltered} B} 
+\subfigure{\includegraphics[width=40mm]{plotWithDebris} A}
+\subfigure{\includegraphics[width=40mm]{plotCell} B}
 \caption{
-  \label{fig:Fig02}{Example of the effect of filtering out low intensities debris with a rectangular gate.}} 
+  \label{fig:Fig01}{Example of smooth scatter plot of FSC-H
+    \textit{vs.} SSC-H data (A) before and (B) after a applying a rectangle 
+on the  measurements (compensated and scaled) extracted from File1}} 
 \end{figure}
 
-<<define and apply ploygoneGate2,eval=TRUE, echo=FALSE,results=hide>>=
-rectGate <- rectangleGate(filterId="nonDebris","FSC-H"=c(0.2,0.8),"SSC-H"=c(0,0.8))
-b08rectGate <- filter(b08,rectGate)
-b08Filtered <- b08[b08rectGate,]
-@ 
-<<plot3, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=c("FL1-H","FL2-H"),pch="*",main="B08 - Nondebris")
-@
-<<plot4, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08Filtered,y=c("FL1-H","FL2-H"),pch="*",main="B08 - Nondebris")
-@
+\section{Discussion and Conclusion}
+%% Explain the usefullness of a commun data structure maybe in comparaison with the ExpressionSet developed for
+%% microarray - count how manay package in BioC depend on the  ExpressionSet data structure?
+%%flowViz
+%%flowQ etc
 
-\begin{figure}
-\centering
-\subfigure{\includegraphics[width=40mm]{plot3} A}
-\subfigure{\includegraphics[width=40mm]{plot4} B} 
-\caption{
-  \label{fig:Fig03}{Second example of the effect of filtering out low intensities debris with a rectangular gate.}} 
-\end{figure}
-
-\subsubsection{polygon gate}
-%%We will define a rectangular gate and apply the filter it to the bO8 data.
-<<define and apply rectangleGate,eval=TRUE,echo=FALSE>>=
-##sqrcut <- matrix(c(300,300,600,600,50,300,300,50),ncol=2,nrow=4)
-##sqrfilter <- new("polygonGate",filterId="nonDebris",
-##            parameters=c("FSC-H","SSC-H"),boundaries= sqrcut)
-sqrfilter <- polygonGate(filterId="nonDebris","FSC-H"=c(0.3,0.3,0.6,0.6),"SSC-H"=c(0.05,0.3,0.3,.05))
-b08sqrfilter <- filter(b08,sqrfilter)
-@
-
-<<plotRectanglefilter, eval=TRUE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08[b08sqrfilter],pch="*")
-@
-
-<<define and apply polygonGate,eval=FALSE,echo=FALSE>>=
-triangleCut <- matrix(c(0.3,0.4,0.6,0.3,0.4,0.3),ncol=2,nrow=3)
-colnames(triangleCut) <- c("FSC-H","SSC-H")
-trianglefilter <- polygonGate(filterId="nonDebris",boundaries= triangleCut)
-b08Trgfilter <- filter(b08,trianglefilter)
-@
-
-<<plotTrianglefilter, eval=FALSE, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(x=b08,y=b08Trgfilter,filter=trianglefilter,colSubSet="red",pch="*")
-@
-
-%%\begin{figure}
-%%\\centering
-%%\\subfigure{\includegraphics[width=40mm]{plotRectanglefilter} A}
-%%\\subfigure{\includegraphics[width=40mm]{plotTrianglefilter} B} 
-%%\\caption{
-%%\  \label{fig:Fig03}{Example of the effect of filtering with a
-%%\    polygon gate.}} 
-%%\\end{figure}
+\section*{Acknowledgements}
+We would like to thank R. Gentleman RR. Brinkman
+%%or should they be Authors
 
 \clearpage
-\section{Case studies}
-\subsection{BD}
-
-<<rawdata1, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,plotParameters=c("FSC-H","SSC-H"),main="B08")
-@
-<<rawdata2, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>= 
-plot(e07,plotParameters=c("FSC-H","SSC-H"),main="E07")
-@
-<<rawdata3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(f06,plotParameters=c("FSC-H","SSC-H"),main="F06")
-@ 
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{rawdata1}}
-%%\subfigure{\includegraphics[width=40mm]{rawdata2}} 
-%%\subfigure{\includegraphics[width=40mm]{rawdata3}} 
-%%\caption{
-%%  \label{fig:Fig04}{Example of the effect of filtering with a
-%%    polygon gate.}} 
-%%\end{figure}
-
-The first filter is a serie of rectangle gates to filter out debris
-
-<<SelectNonDebris,eval=TRUE,echo=FALSE,results=hide>>=
-filter1 <- rectangleGate(filterId="nonDebris","FSC-H"=c(0.2,0.8),"SSC-H"=c(0,0.8))
-b08.result1 <-  filter(b08,filter1)
-##e07.result1 <- filter(e07,filter1)
-##f06.result1 <- filter(f06,filter1)
-summary(b08.result1)
-##tab <- c("B08"=sum(b08.result1@subSet),"E07"=sum(e07.result1@subSet),"F06"=sum(f06.result1@subSet))
-@ 
-<<plotResFilter1, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result1,main="B08 nonDebris",pch="*")
-@ 
-<<plotResFilter2, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(e07,y=e07.result1,main="E07 nonDebris",pch="*")
-@ 
-<<plotResFilter3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(f06,y=f06.result1,main="F06 nonDebris",pch="*")
-@ 
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{plotResFilter1}}
-%%\subfigure{\includegraphics[width=40mm]{plotResFilter2}} 
-%%\subfigure{\includegraphics[width=40mm]{plotResFilter3}} 
-%%\caption{
-%%  \label{fig:Fig05}{Example of the effect of filtering with a
-%%    polygon gate.}} 
-%%\end{figure}
-
-The second gate gets the live cells
-
-<<FilterLiveCell,eval=TRUE,echo=FALSE,results=hide>>=
-##filter2 <- norm2Filter(filterId="Live Cells",scale.factor=2,method="covMcd",parameters=c("FSC-H","SSC-H"))
-filter2 <- norm2Filter("FSC-H","SSC-H",filterId="Live Cells",scale.factor=2,method="covMcd")
-b08.result2 <-  filter(b08, filter2 %subset% b08.result1)
-
-##e07.result2 <- filter(e07,filter2,parent=e07.result1)
-##f06.result2 <- filter(f06,filter2,parent=f06.result1)
-
-summary(b08.result2)
-##tab <- rbind(tab,c(sum(b08.result2@subSet),sum(e07.result2@subSet),sum(f06.result2@subSet)))
-
-@
-
-<<plotLivingCell1, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result2,parent=b08.result1,xlim=c(0,1024),ylim=c(0,1024),main="B08",pch="*")
-@ 
-<<plotLivingCell2, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(e07,y=e07.result2,parent=e07.result1,xlim=c(0,1024),ylim=c(0,1024),main="E07",pch="*")
-@ 
-<<plotLivingCell3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(f06,y=f06.result2,parent=f06.result1, xlim=c(0,1024),ylim=c(0,1024),main="F06",pch="*")
-@ 
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{plotLivingCell1}}
-%%\subfigure{\includegraphics[width=40mm]{plotLivingCell2}} 
-%%\subfigure{\includegraphics[width=40mm]{plotLivingCell3}} 
-%%\caption{
-%%  \label{fig:Fig06}{Example of the effect of filtering with a
-%%    polygon gate.}} 
-%%\end{figure}
-
-The third-fifth gates get the positive cells for the marker in FL1-H. 
-This is a really interesting example because it illustrates that there
-are two subpopulations. Naturally we would like to automatically find them
-In this case we want to now what percent the positive population in FL1-H 
-is of the total population.
-
-<<FL1-H, echo=FALSE, eval=FALSE,fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),ylim=c(0,1024),xlim=c(0,1024))
-@ 
-%%\begin{figure}
-%%\centering
-%%\includegraphics{FL1-H}
-%%\end{figure}
-
-<<FL1Filter,eval=TRUE,echo=FALSE,results=hide>>=
-filter3 <- rectangleGate(filterId="FL1-H+","FL1-H"=c(400,Inf))
-b08.result3 <- filter(b08,filter3 %subset%b08.result2)
-
-filter4 <- norm2Filter("FSC-H","FL1-H",filterId="FL1-H+",scale.factor=2,method="covMcd")
-b08.result4 <- filter(b08,filter4 %subset% b08.result2)
-
-##b08.result5 <- applyFilter(filter4,b08,b08.result2@subSet-b08.result4@subSet)
-
-##ans <- c(sum(b08.result3@subSet)/sum(b08.result2@subSet),
-##sum(b08.result4@subSet)/(sum(b08.result4@subSet)+sum(b08.result5@subSet)))
-@
-
-<<FL1Filtered1, echo=FALSE, eval=FALSE,  fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result3,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
-          xlim=c(0,1024),ylim=c(0,1024),pch="*")
-@ 
-<<FL1Filtered2, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result4,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
-          xlim=c(0,1024),ylim=c(0,1024),pch="*")
-@ 
-<<FL1Filtered3, echo=FALSE, eval=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-plot(b08,y=b08.result5,parent=b08.result2,plotParameters=c("FSC-H","FL1-H"),
-          xlim=c(0,1024),ylim=c(0,1024),pch="*")
-@
-
-%%\begin{figure}
-%%\centering
-%%\subfigure{\includegraphics[width=40mm]{FL1Filtered1}}
-%%\subfigure{\includegraphics[width=40mm]{FL1Filtered2}} 
-%%\subfigure{\includegraphics[width=40mm]{FL1Filtered3}} 
-%%\caption{
-%%  \label{fig:Fig07}{Example of the effect of filtering with a
-%%    polygon gate.}} 
-%%\end{figure}
-
-\clearpage
-\bibliographystyle{cyto} 
+\bibliographystyle{plainnat} 
 \bibliography{cytoref}

### 2007-02-10 inst/doc/filtering-flowCore.Rnw
add man page fix read.FCS function (hopefully) work on the vignette
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@22523 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \begin{abstract}
-\noindent One of the more common operations in flow cytometry data analysis is a process of filtering, also known as gating.
-This is often either the target of analysis in the case of surface staining or a necessary pre-processing step for further analysis on the cell population of interest.
-For example, a researcher may be interested in CD4+ T cells, which typically requires a lymphocyte gate and a CD4 gate to obtain a relatively clean population and other cell types might be even more complex.
-In the case where the number of cells within a gate is of interest, it is not uncommon to require a fairly complex composition of gates for analysis.
-This document describes the rich filter construction and composition facilities in the flowCore package as well as what is needed for developing your own filter objects.
-\\ \\
+\noindent One of the more common operations in flow cytometry data analysis is a process of filtering, 
+also known as gating. This is often either the target of analysis in the case of surface staining or a 
+necessary pre-processing step for further analysis on the cell population of interest. For example, 
+a researcher may be interested in CD4+ T cells, which typically requires a lymphocyte gate and a CD4 gate 
+to obtain a relatively clean population and other cell types might be even more complex. 
+In the case where the number of cells within a gate is of interest, it is not uncommon to require a fairly 
+complex composition of gates for analysis. This document describes the rich filter construction and composition 
+facilities in the flowCore package as well as what is needed for developing your own filter objects. \\ \\
 \textbf{keywords} Flow cytometry, high throughtput, software, standard
@@
 
-Implementing a new filter is a fairly straightforward process requiring the definition of a filtering class, derived from the \Rclass{filter} base class, and, at the minimum, the definition of an \Rfunction{\%in\%} method.
+Implementing a new filter is a fairly straightforward process requiring the definition of a filtering class, 
+derived from the \Rclass{filter} base class, and, at the minimum, the definition of an \Rfunction{\%in\%} method.
 
@@
 
-A \Robject{filter} subclass can be customized in one of two ways.
-First, the class returned by the \Rfunction{filter} method can be changed by definiting a \Rfunction{filterResultType} method for your filter class.
-This allows filters that return non-deterministic results (e.g.
-soft-clustering filters) to use the \Rclass{randomFilterResult}.
+A \Robject{filter} subclass can be customized in one of two ways. First, the class returned by the 
+\Rfunction{filter} method can be changed by definiting a \Rfunction{filterResultType} method for your filter class. 
+This allows filters that return non-deterministic results (e.g. soft-clustering filters) to use the 
+\Rclass{randomFilterResult}.
 

### 2007-02-10 R/filter-operations.R
add man page fix read.FCS function (hopefully) work on the vignette
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@22523 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -73,6 +73,7 @@ setMethod("summary","subsetFilter",function(object,...) {
 
 setMethod("%in%",c("flowFrame","filterResult"),function(x,table) {
 	frameId = identifier(x)
+        print(class(frameId))
 	if(all(!is.na(c(frameId,table@frameId))) && frameId != table@frameId)
 		warning("Frame identifiers do not match. It is possible that this filter is not compatible with this frame.")
 	if(nrow(x) != length(table@subSet))

### 2007-02-10 R/flowFrame-methods.R

@@ -4,9 +4,9 @@
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("identifier", signature="flowFrame",
           definition=function(object){
-            oid <- object@description["GUID"]
+            oid <- object@description[["GUID"]]
             if(is.null(oid) || is.na(oid))
-              as.vector(object@description["$FIL"])
+              as.vector(object@description[["$FIL"]])
             else
               as.vector(oid)
           })

### 2007-02-10 R/readFCS-functions.R

@@ -39,7 +39,11 @@ read.FCS <- function(filename, transformation="linearize", debug=FALSE,alter.nam
   if(transformation==TRUE){
       txt[["transformation"]] <-"applied" 
   }
-  return(new("flowFrame", exprs=mat, description=c(txt), parameters=params))
+
+  
+  description <- strsplit(txt,split=" ")
+  names(description) <- names(txt)
+  return(new("flowFrame", exprs=mat, description= description, parameters=params))
 }
 
 makeFCSparameters <- function(cn,txt) {
@@ -74,8 +78,8 @@ readFCSheader <- function(con) {
  
   ioffs <- c(as.double(version),as.integer(coffs))
   names(ioffs) <- c("FCSversion", "textstart", "textend", "datastart", "dataend", "anastart", "anaend")
-  stopifnot(all(!is.na(ioffs) | ioffs==""), !any(ioffs[1:5]==0))
-
+  ##stopifnot(all(!is.na(ioffs) | ioffs==""), !any(ioffs[1:5]==0))
+  stopifnot(all(!is.na(ioffs) | ioffs==""))
   return(ioffs)
 }
 
@@ -116,6 +120,26 @@ readFCSdata <- function(con, offsets, x, transformation, debug, scale, alter.nam
   if(length(bitwidth)!=1)
     stop("Sorry, I am expecting the bitwidth to be the same for all parameters")
 
+  ##for DATA segment exceeding 99,999,999 byte.
+  if(offsets["FCSversion"] == 3){
+      datastart <- as.numeric(readFCSgetPar(x, "$BEGINDATA"))
+      dataend <- as.numeric(readFCSgetPar(x, "$ENDDATA"))
+      if(offsets["datastart"] != datastart && offsets["datastart"]== 0){
+          offsets["datastart"] <-  datastart
+      }
+      if(offsets["datastart"] != datastart && offsets["datastart"]!= 0){
+          print(datastart)
+          print(offsets["datastart"])
+          stop("The HEADER and the TEXT segment define different starting point to read the data.")
+      }
+      
+      if(offsets["dataend"] != dataend && offsets["dataend"]== 0){
+          offsets["dataend"] <-  dataend
+      }
+      if(offsets["dataend"] != dataend && offsets["dataend"]!= 0){
+          stop("The HEADER and the TEXT segment define different ending point to read the data.")
+      }
+  }
   seek(con, offsets["datastart"])
 
   size <- bitwidth/8

### 2007-03-15 R/AllClasses.R
Doc updates and a couple of minor bug fixes (like adding the '!' back into complement filters...)
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23212 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -194,18 +194,6 @@ ellipsoidGate <- function(filterId="ellipsoidGate", .gate, distance,...) {
         focus=.gate, distance=distance)
 }
 
-
-## ===========================================================================
-## mode gate
-## ---------------------------------------------------------------------------
-## rectangle gates that automatically select their dominant region.
-## ---------------------------------------------------------------------------
-setClass("modeGate",
-         representation(bw="ANY",n="numeric"),
-         contains="filter",
-         prototype=list(bw="nrd0",n=512))
-
-
 ## ===========================================================================
 ## norm2Filter
 ## ---------------------------------------------------------------------------
@@ -279,6 +267,25 @@ sampleFilter = function(filterId="sample",size) {
 setClass("multiFilter",
          representation(populations="character",filters="list"),
          contains="filter")
+
+
+
+## =================================================================
+## filterCollection
+## ----------------------------------------------------------------
+## A collection of Filters organized as a DAG. The DAG specifies the
+## Filters and the Populations. The Populations are names only because
+## they get created only when a FilterTree is applied to a dataset.
+## -----------------------------------------------------------------
+setClass("filterCollection",representation("environment"))
+
+## ===========================================================================
+## promisedFilter 
+## ---------------------------------------------------------------------------
+## A reference to a filter in an environment.
+## ---------------------------------------------------------------------------
+setClass("promisedFilter",representation("filter",name="character",where="environment"))
+
 
 
 ## =================================================================

### 2007-03-15 R/AllGeneric.R

@@ -19,6 +19,7 @@ if(!isGeneric("hist"))
 if(!isGeneric("lines"))
   setGeneric("lines",          useAsDefault=lines)
 
+setGeneric("draw",function(x,data,...) standardGeneric("draw"))
 
 ## ===========================================================================
 ## Generics for our operators

### 2007-03-15 R/complementFilter-accessors.R

@@ -3,7 +3,7 @@
 ## --------------------------------------------------------------------------
 setMethod("%in%",c("flowFrame","complementFilter"),function(x,table) {
 	r = filter(x,table@filters[[1]])
-	z = as(r,"logical")
+	z = !as(r,"logical")
 	attr(z,'filterDetails') = filterDetails(r)
 	z
 })

### 2007-03-15 R/filter-accessors.R
Doc updates and a couple of minor bug fixes (like adding the '!' back into complement filters...)
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23212 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -81,3 +81,26 @@ setMethod("!",signature("filter"),function(e1)
 ## ==========================================================================
 ## --------------------------------------------------------------------------
 setMethod("%&%",signature("ANY","ANY"),function(e1,e2) e1 %subset% e2)
+
+## ==========================================================================
+## Plotting method for filters lets us get a basic plot of the data and a 
+## filter. Uses the draw method to actually display the filter on some data.
+## --------------------------------------------------------------------------
+setMethod("plot",signature(x="flowFrame",y="filter"),function(x,y,z=NULL,results=NULL,...) {
+	if(is.null(z))	z = parameters(y)
+	if(length(z)>2) {
+		warning("Only plotting the first two parameters.")
+		z = z[1:2]
+	}
+	#First we plot the actual data.
+	plot(x,z,...)
+	#Then draw the filter on top of it which takes the form (filter,data,params,results (if any))
+	draw(y,x,z,results,...)
+})
+
+setMethod("draw",signature("filter","flowFrame"),function(x,data,params=NULL,results=NULL) {
+	if(is.null(params)) parameters=x@params
+	if(length(params)>2) {
+		params = params[1:2]
+	}
+})

### 2007-03-16 inst/doc/HowTo-flowCore.Rnw
Doc updates and a couple of minor bug fixes (like adding the '!' back into complement filters...)
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23212 bc3139a8-67e5-0310-9ffc-ced21a209358
fix definition of generic methods fix definition of filterCollection, can not extend environment
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23255 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \section{Introduction}
-Traditionnally, flow cytometry has been a tube-based technique limited to small-scale laboratory and clinical studies. 
+Traditionally, flow cytometry has been a tube-based technique limited to small-scale laboratory and clinical studies. 
 High throughput methods for flow cytometry have recently been developed for drug discovery and advanced research methods 
@@
 information in objects of class \Robject{flowSet}. 
-A\Robject{flowSet} object contains two main slot exprs, an enviroment data astructure 
-that stores the flow cytometry measurements and the phenoData. The phenoData slot 
+A \Robject{flowSet} object contains two main slots \Robject{exprs}, an environment data structure 
+that stores the flow cytometry measurements and the \Robject{phenoData}. The \Robject{phenoData} slot 
 contains all the relevant experiment-wide meta-data, \textit{e.g.} 

### 2007-03-16 R/AllClasses.R

@@ -277,14 +277,16 @@ setClass("multiFilter",
 ## Filters and the Populations. The Populations are names only because
 ## they get created only when a FilterTree is applied to a dataset.
 ## -----------------------------------------------------------------
-setClass("filterCollection",representation("environment"))
+## Environment can't be extended
+## setClass("filterCollection",representation("environment"))
 
+setClass("filterCollection",representation("list"))
 ## ===========================================================================
 ## promisedFilter 
 ## ---------------------------------------------------------------------------
 ## A reference to a filter in an environment.
 ## ---------------------------------------------------------------------------
-setClass("promisedFilter",representation("filter",name="character",where="environment"))
+setClass("promisedFilter",representation("filter",name="character"))
 
 
 

### 2007-03-16 R/AllGeneric.R

@@ -1,24 +1,8 @@
 ##  Generic methods definition - last changes NLM January 17, 2006
-##      add parameters
 
 ## ===========================================================================
 ## These are already defined as generic functions:
 ## ---------------------------------------------------------------------------
-if(!isGeneric("colnames<-"))
-  setGeneric("colnames<-",    function(x, value) standardGeneric("colnames<-"))
-
-if(!isGeneric("colnames"))
-  setGeneric("colnames",      function(x, do.NULL=TRUE, prefix="col") standardGeneric("colnames"))
-
-if(!isGeneric("plot"))
-  setGeneric("plot",          useAsDefault=plot)
-
-if(!isGeneric("hist"))
-  setGeneric("hist",          useAsDefault=hist)
-
-if(!isGeneric("lines"))
-  setGeneric("lines",          useAsDefault=lines)
-
 setGeneric("draw",function(x,data,...) standardGeneric("draw"))
 
 ## ===========================================================================
@@ -32,6 +16,12 @@ setGeneric("%on%",function(e1,e2) standardGeneric("%on%"))
 
 
 ## ===========================================================================
+## Generic for ncol
+## ---------------------------------------------------------------------------
+##if(!isGeneric("ncol"))
+##  setGeneric("ncol", function(x) standardGeneric("ncol"))
+
+## ===========================================================================
 ## Generic for keyword
 ## ---------------------------------------------------------------------------
 setGeneric("keyword",function(object,keyword) standardGeneric("keyword"))
@@ -41,20 +31,6 @@ setGeneric("keyword",function(object,keyword) standardGeneric("keyword"))
 ## Generic to access parameters in a flowframe
 ## ---------------------------------------------------------------------------
 setGeneric("parameters",function(object) standardGeneric("parameters"))
-
-
-## ===========================================================================
-## Generic for nrow
-## ---------------------------------------------------------------------------
-if(!isGeneric("nrow"))
-  setGeneric("nrow", function(x) standardGeneric("nrow"))
-
-
-## ===========================================================================
-## Generic for ncol
-## ---------------------------------------------------------------------------
-if(!isGeneric("ncol"))
-  setGeneric("ncol", function(x) standardGeneric("ncol"))
 
 
 ## ===========================================================================
@@ -74,7 +50,6 @@ setGeneric("identifier",function(object) standardGeneric("identifier"))
 ## ---------------------------------------------------------------------------
 setGeneric("split")
[…]

### 2007-04-05 R/AllGeneric.R
work on man page still warnings for summary methods and %&%
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23624 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1,4 +1,4 @@
-##  Generic methods definition - last changes NLM January 17, 2006
+##  Generic methods definition
 
 ## ===========================================================================
 ## These are already defined as generic functions:
@@ -61,22 +61,27 @@ setGeneric("filter", function(x, filter,...) standardGeneric("filter"),useAsDefa
 ## ===========================================================================
 ## Generics for accessing certain aspects of the filtering process.
 ## ---------------------------------------------------------------------------
-setGeneric("filterDetails<-",function(result,filterId,...,value)  standardGeneric("filterDetails<-"))
-setGeneric("filterDetails",function(result,filterId,...) standardGeneric("filterDetails"))
+setGeneric("filterDetails<-", function(result, filterId,...,value)  standardGeneric("filterDetails<-"))
+setGeneric("filterDetails", function(result, filterId,...) standardGeneric("filterDetails"))
 
 
 ## ===========================================================================
 ## Used to summarize the operation of a filter on a frame. Used for the implementation
 ## of filters, not for the use of filters.
 ## ---------------------------------------------------------------------------
-setGeneric("summarizeFilter",function(result,filter) standardGeneric("summarizeFilter"))
+setGeneric("summarizeFilter", function(result,filter) standardGeneric("summarizeFilter"))
 
+## ===========================================================================
+## Used to summarize the operation of a filter on a frame. Used for the implementation
+## of filters, not for the use of filters.
+## ---------------------------------------------------------------------------
+setGeneric("summary", function(object,...) standardGeneric("summary"))
 
 ## ===========================================================================
 ## Generics for spillover and compensate
 ## ---------------------------------------------------------------------------
-setGeneric("spillover",function(x,...) standardGeneric("spillover"))
-setGeneric("compensate",function(x,spillover) standardGeneric("compensate"))
+setGeneric("spillover", function(x,...) standardGeneric("spillover"))
+setGeneric("compensate", function(x,spillover) standardGeneric("compensate"))
 
 
 ## ===========================================================================

### 2007-04-05 R/filter-accessors.R

@@ -7,6 +7,15 @@
 
 
 ## ==========================================================================
+## accessor method for slot parameters
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("parameters", signature="filter",
+          definition=function(object)
+            object@parameters
+          )
+
+
+## ==========================================================================
 ## summary method for filters
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("summary",signature("filter"),function(object,...) {
@@ -86,21 +95,30 @@ setMethod("%&%",signature("ANY","ANY"),function(e1,e2) e1 %subset% e2)
 ## Plotting method for filters lets us get a basic plot of the data and a 
 ## filter. Uses the draw method to actually display the filter on some data.
 ## --------------------------------------------------------------------------
-setMethod("plot",signature(x="flowFrame",y="filter"),function(x,y,z=NULL,results=NULL,...) {
-	if(is.null(z))	z = parameters(y)
-	if(length(z)>2) {
-		warning("Only plotting the first two parameters.")
-		z = z[1:2]
-	}
-	#First we plot the actual data.
-	plot(x,z,...)
-	#Then draw the filter on top of it which takes the form (filter,data,params,results (if any))
-	draw(y,x,z,results,...)
+setMethod("plot",signature(x="flowFrame", y="filter"),function(x, y, z=NULL, results=NULL, ...) {
+    if(is.null(z))	z = parameters(y)
+    if(length(z)>2) {
+        warning("Only plotting the first two parameters.")
+        z = z[1:2]
+    }
+    ##First we plot the actual data.
+    plot(x, z, ...)
+    ##Then draw the filter on top of it which takes the form (filter, data, params, results (if any))
+    draw(y, x, z, results, ...)
 })
 
-setMethod("draw",signature("filter","flowFrame"),function(x,data,params=NULL,results=NULL) {
-	if(is.null(params)) parameters=x@params
+setMethod("draw", signature("filter","flowFrame"), function(x, data, params=NULL, results=NULL, ...) {
+	if(is.null(params))
+          params = parameters(x)
 	if(length(params)>2) {
 		params = params[1:2]
 	}
+        if(is(x,"rectangleGate")){
[…]

### 2007-04-05 inst/doc/HowTo-flowCore.Rnw
work on man page still warnings for summary methods and %&%
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23624 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\title{flowCore: data structures for flow cytometry data}
+\title{flowCore: data structures pacakge for flow cytometry data}
 \author{N. Le Meur F. Hahne B. Ellis P. Haaland}
@@
 The objects of class \Robject{flowFrame} are the containers for storing individual 
-cytometry measurements originated from one FCS file. \\
+cytometry measurements originated from one FCS file. The information available are:\\
 %%need description of the slots and maybe methods associated
+\begin{itemize}
+\item exprs, containing the measured intensities. Rows correspond to cells, columns to the
+      different channels. 
+\item parameters, containing information about each column of the  flowFrame. This will generally
+      be filled in by \Rfunction{read.FCS} or similar functions using data from
+      the \Robject{description} slot that describes the parameters.
+\item description is list containing the meta data included in the FCS file.
+\end{itemize}
 
@@
 The package \Rpackage{flowCore} stores the measurement and the associated data 
-information in objects of class \Robject{flowSet}. 
-A \Robject{flowSet} object contains two main slots \Robject{exprs}, an environment data structure 
-that stores the flow cytometry measurements and the \Robject{phenoData}. The \Robject{phenoData} slot 
-contains all the relevant experiment-wide meta-data, \textit{e.g.} 
+information in objects of class \Robject{flowSet}, a collection of \Robject{flowFrame}. 
+A \Robject{flowSet} object contains two main slots \Robject{frames}, an environment data structure 
+that stores the flow cytometry measurements as \Robject{flowFrame} and the \Robject{phenoData}. 
+The \Robject{phenoData} slot contains all the relevant experiment-wide meta-data, \textit{e.g.} 
 phenotypic and experimental information relevant to the management and/or the analysis of the data.\\
@@
 to provide user-friendly visualization and/or to perform statistical analyses and interpret the data. 
-Following the standard \cite{Spidlen2006} 8 transformations are pre-defined: linear, quadratic, natural logarithm, 
-logarithm (with a specified base), hyperlog, bi-exponential, logicle, and split-scale.\\
+The proposed standard \cite{Spidlen2006} defined 8 transformations communly used: linear, quadratic, natural logarithm, 
+logarithm (with a specified base), hyperlog, bi-exponential, logicle, split-scale. All of them are 
+implemented in \Rpackage{flowcore} and a few other have been added. Each transformation has specific properties:\\
 
-%%Can we define other?   
-%%can be applied when reading in the data.
+%%could be a table
+\begin{itemize}
+\item linear
+\item logarithm
+\item biexponential
+\item logicle transformation help to visualize cells with minimal fluorescence 
+  (cells usually pilled up on the axis with logarithmic displays) \citep{Herzenberg2006}
+\item arcsinh 
+\end{itemize}
 
 \subsection{Filter}
-The role of filtering strategies are to identify and characterize diffrent cells population in sample.
+The role of filtering strategies are to identify and characterize different cells population in sample.
 %% Filtering vs Gating

### 2007-04-11 inst/doc/HowTo-flowCore.Rnw
man page and vignette
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23749 bc3139a8-67e5-0310-9ffc-ced21a209358

--- 2007-03-16 inst/doc/HowTo-flowCore.Rnw
@@
 \newcommand{\Rfunction}[1]{{\texttt{#1}}}
+\newcommand{\Rcode}[1]{{\texttt{#1}}}
 \newcommand{\Robject}[1]{{\texttt{#1}}}
@@
 We believe that researchers will find flowCore to be a solid foundation for future development of new methods to attack the many interesting open research questions in FCM data analysis.\\
-\textbf{Methods} We propose a variety different data structures. We have implemented the classes and methods in the Bioconductor package flowCore. 
+
+\noindent \textbf{Methods} We propose a variety different data structures. We have implemented the classes and methods in the Bioconductor package flowCore. 
 We illustrate their use with X case studies.\\
-\textbf{Results} We hope that those proposed data structures will the base for the development of many tools for the analysis of high throughput flowcytometry.\\
-\textbf{keywords} Flow cytometry, high throughtput, software, standard
+
+\noindent \textbf{Results} We hope that those proposed data structures will the base for the development of many tools for the analysis of high throughput flowcytometry.\\
+
+\noindent \textbf{keywords} Flow cytometry, high throughtput, software, standard
 \end{abstract}
@@
 High throughput methods for flow cytometry have recently been developed for drug discovery and advanced research methods 
-\cite{Gasparetto2004, Young2004}. As an example, the flow cytometry high content screening (FC-HCS) can process up 
+\citep{Gasparetto2004}. As an example, the flow cytometry high content screening (FC-HCS) can process up 
 to a thousand samples daily at a single workstation, and the results have been equivalent or superior to traditional 
@@
 The amount of information generated by high throughput technologies such as FC-HCS need to be transformed into 
-executive summaries (which are brief enough) for creative studies by a human researcher \cite{Brazma2001}. 
+executive summaries (which are brief enough) for creative studies by a human researcher \citep{Brazma2001}. 
 Standardization is critical when developing new high throughput technologies and their associated information services 
-\cite{Brazma2001, Chicurel2002, Boguski2003}. Standardization efforts have been made in clinical cell analysis 
-by flow cytometry \cite{Keeney2004}, however data interpretation has not been standardized for even low throughput FCM. 
+\citep{Brazma2001, Chicurel2002, Boguski2003}. Standardization efforts have been made in clinical cell analysis 
+by flow cytometry \citep{Keeney2004}, however data interpretation has not been standardized for even low throughput FCM. 
 It is one of the most difficult and time consuming aspects of the entire analytical process as well as a primary source 
 of variation in clinical tests, and investigators have traditionally relied on intuition rather than standardized statistical inference 
-\cite{Bagwell2004,Braylan2004,Parks1997,Suni2003}. In the development of standards in high throughput FCM, 
+\citep{Bagwell2004,Braylan2004,Parks1997,Suni2003}. In the development of standards in high throughput FCM, 
 few progress has been done in term of Open Source software. In this article we propose R data structures to handle 
@@
 
-The aim is to merge both prada  and rflowcyt \cite{LeMeur2006} into one core package wich is compliant 
-with the data exchange standards that are currently developed in the community \cite{Spidlen2006}.
+The aim is to merge both prada  and rflowcyt \citep{LeMeur2006} into one core package wich is compliant 
+with the data exchange standards that are currently developed in the community \citep{Spidlen2006}.
 
@@
 The objects of class \Robject{flowFrame} are the containers for storing individual 
-cytometry measurements originated from one FCS file. \\
-%%need description of the slots and maybe methods associated
+cytometry measurements originated from one FCS file. The information available are:\\
+\begin{description}
+\item[exprs] containing the measured intensities. Rows correspond to cells, columns to the
+      different channels. 
+\item[parameters] containing information about each column of the  \Robject{flowFrame}. This will generally
+      be filled in by \Rfunction{read.FCS} or similar functions using data from
+      the \Robject{description} slot that describes the parameters.
+\item[description] a list containing the meta data included in the header of the FCS file.
+\end{description}
+
+Several method have been defined to access and manipulate the data of a \Robject{flowFrame}. 
+As example, the function \Rfunction{keyword} retrieve one or all the keywords saved in 
+the header of a FCS file and \Rfunction{compensate} and \Rfunction{transform} apply 
+on a \Robject{flowFrame} a compensation matrix and a transformation method, 
+respectively (see the \Robject{flowFrame} man page  and the section~\ref{sec:case} "Case Study" for more details.)
 
 \subsection{flowSet}
-An flow cytometry experiment usually include several sample and several measurements. 
+An flow cytometry experiment usually include several samples and several measurements. 
 A data structure is thus needed to store those collections of measurements.  
-The package \Rpackage{flowCore} stores the measurement and the associated data 
-information in objects of class \Robject{flowSet}. 
-A \Robject{flowSet} object contains two main slots \Robject{exprs}, an environment data structure 
-that stores the flow cytometry measurements and the \Robject{phenoData}. The \Robject{phenoData} slot 
-contains all the relevant experiment-wide meta-data, \textit{e.g.} 
-phenotypic and experimental information relevant to the management and/or the analysis of the data.\\
+The package \Rpackage{flowCore} stores the measurements and the associated data 
+information in objects of class \Robject{flowSet}, a collection of \Robject{flowFrame}. 
+A \Robject{flowSet} object contains three slots \Robject{frames}, \Robject{phenoData} and \Robject{colnames}. 
+The slot \Robject{frames} is an environment data structure that stores the flow cytometry measurements as \Robject{flowFrame}. 
+The \Robject{phenoData} slot contains all the relevant experiment-wide meta-data, \textit{e.g.} 
+phenotypic and experimental information relevant to the management and/or the analysis of the data. 
+The  \Robject{colnames} is a vector containing the names of the recorded parameters.
+
+Numerous methods have been developed to manipulate and access the  \Robject{flowSet} data, 
+see the \Robject{flowSet} man page and the section~\ref{sec:case} "Case Study" for more details.
 
@@
 estimated and subtracted from the total detected signals to yield an estimate of the 
-actual amount of each dye \cite{Herzenberg2006}. Single-laser, two-color FACS experiments 
+actual amount of each dye \citep{Herzenberg2006}. Single-laser, two-color FACS experiments 
 in the mid-1970s showed that fluorescence compensation is important. 
@@
 to provide user-friendly visualization and/or to perform statistical analyses and interpret the data. 
-Following the standard \cite{Spidlen2006} 8 transformations are pre-defined: linear, quadratic, natural logarithm, 
-logarithm (with a specified base), hyperlog, bi-exponential, logicle, and split-scale.\\
-
-%%Can we define other?   
-%%can be applied when reading in the data.
+The proposed standard \citep{Spidlen2006} defined 8 communly used transformations: linear, quadratic, natural logarithm, 
+logarithm (with a specified base), hyperlog, bi-exponential, logicle, split-scale. All of them are 
+implemented in \Rpackage{flowcore} and a few other have been added (\textit{e.g.}, asinh).
 
 \subsection{Filter}
-The role of filtering strategies are to identify and characterize diffrent cells population in sample.
+One of the more common operations in flow cytometry data analysis is a process of filtering, 
+also known as gating. The role of filtering strategies is to identify and characterize different cells population in sample. 
+This is often either the target of analysis in the case of surface staining or a 
+necessary pre-processing step for further analysis on the cell population of interest. For example, 
+a researcher may be interested in CD4+ T cells, which typically requires a lymphocyte gate and a CD4 gate 
+to obtain a relatively clean population and other cell types might be even more complex. 
+In the case where the number of cells within a gate is of interest, it is not uncommon to require a fairly 
+complex composition of gates for analysis.
+
+Following the proposed standard \citep{Spidlen2006}, we currently propose 4 type of the most commun filters: 
+rectangleGate, polygonGate, polytopeGate, ellipsoidGate. For more details on those methods, see the man pages 
+and section~\ref{sec:case} "Case Study". we also added a few more approaches such as norm2Filter 
+(originally available in \Rpackage{prada}) and the kmeansFilter.
+
 %% Filtering vs Gating
-%%Byron's vignette?
 %%several approach from basic to DAG ...\\
@@
 \section{Case study}
+\label{sec:case}
 \subsection{dataset}
@@
 \subsection{Quality Assessment}
-Quickly look at the range of the intensities between samples 
-other statistics sur as ... cf Janet paper.
+Data quality control and quality assessment are crucial steps in processing 
+and analyzing high throughput flow cytometry data \citep{LeMeur2007} .
+Graphical exploratory data analysis and descripticve statistics are the two major approaches to asses the quality of the data
+
+In \Rpackage{flowCore} the function \Rfunction{fsApply} allows to apply methods to each element of a flowSet. 
+For example, one might want to quickly look at the range of the intensities between samples: 
+ %cf Janet paper?.
 <<summaryRange>>=
@@
 @ 
+Other quality control and quality assessment methods are being developed in the dedicated \Rpackage{flowQ} package.
 
 \subsection{Preprocessing}
+\subsubsection{compensation}
 <<compensationMatrix, echo=FALSE, results= hide>>=
@@
 
-Transformation
+\subsubsection{Transformation}
 <<scaleScatter, echo=TRUE>>=
@@
 
+\subsubsection{Filtering}
 Filter to leave out debris:
@@
 \section{Discussion and Conclusion}
+The application of flow cytometry in modern cell biology is diverse and so are the demands on data analysis.
+The multitude of packages within R and Bioconductor already provides for many tools that are also useful in the analysis of FCM data.
+The package \Rpackage{flowCore}  try to close the gap between data acquisition and data analysis by enabling the researches
+to take their data into the powerful R environment and to make use of the statistical and graphical solutions already available there.
+
+The package \Rpackage{flowCore} is the result of the collaboration  of several groups involved in high-throughput FCM together 
+with instrument manufacturers and members of the flow cytometry standards initiative \citep{Spidlen2006}. In a way \Rpackage{flowCore} 
+is the merging of the \Rpackage{prada}, \Rpackage{rflowcyt}  object model. Our goal was to have a common S4 object model compliant 
+with  the data exchange standards that are currently developed and which will be the base the development of 
+all subsequent flow cytometry based packages. Utility packages such as \Rpackage{flowQ} (quality control and quality assessment), 
+\Rpackage{flowViz} (Visualization) or \Rpackage{flowUtils} (various tools sure as XML parser) already use the \Rpackage{flowCore} data structure.
+
 %% Explain the usefullness of a commun data structure maybe in comparaison with the ExpressionSet developed for

### 2007-04-13 R/AllClasses.R
* Big changes to HowTo vignette. * Minor changes to coercion and read.flowSet to address Errol's troubles. * Removed as()<-. I guess Seth will be happy. * Minor documentation changes * A couple of small bugs with split * A $ accessor for flowSets
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23837 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -18,15 +18,7 @@ setClass("flowFrame",
              data=data.frame(name=I(character(0))),
              varMetadata=data.frame(labelDescription="Name in frame",
                row.names="name")),
-           description=list(note="empty")),
-         validity=function(object){
-             msg <- TRUE
-             if(!is.matrix(object@exprs))
-               msg <- "\nslot 'exprs' must be matrix"
-             if(!is.list(object@description))
-               msg <- "\nslot 'description' must be a list"
-         })
-
+           description=list(note="empty")))
 
 ## ===========================================================================
 ##  flowSet
@@ -51,9 +43,7 @@ setClass("flowSet",
              all.names=TRUE)))
            if(any(name.check)) {
              name.list <- paste(sampleNames(object)[name.check],sep=",")
-             warning(paste("These objects are not in the data environment:",
-                           name.list))
-             return(FALSE)
+             return(paste("These objects are not in the data environment:",name.list))
            }
            ##Ensure that all frames match our colnames
            if(!all(sapply(sampleNames(object),function(i) {

### 2007-04-13 inst/doc/HowTo-flowCore.Rnw
* Big changes to HowTo vignette. * Minor changes to coercion and read.flowSet to address Errol's troubles. * Removed as()<-. I guess Seth will be happy. * Minor documentation changes * A couple of small bugs with split * A $ accessor for flowSets
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23837 bc3139a8-67e5-0310-9ffc-ced21a209358

--- 2007-03-16 inst/doc/HowTo-flowCore.Rnw
@@
 We believe that the Open Source statistical software R in conjunction with the Bioconductor Project fills all of these requirements.
-Consequently we have developed a Bioconductor package that we call flowCore.
-The flowCore package is not intended to be a complete analysis package for FCM data, rather, we see it as providing a clear object model and a collection of standard tools that enable R as an informatics research platform for flow cytometry.
-One of the important issues that we have addressed in the flowCore package is that of using a standardized representation that will insure compatibility with existing technologies for data analysis and will support collaboration and interoperability of new methods as they are developed.
+Consequently we have developed a Bioconductor package that we call \Rpackage{flowCore}.
+The \Rpackage{flowCore} package is not intended to be a complete analysis package for FCM data, rather, we see it as providing a clear object model and a collection of standard tools that enable R as an informatics research platform for flow cytometry.
+One of the important issues that we have addressed in the \Rpackage{flowCore} package is that of using a standardized representation that will insure compatibility with existing technologies for data analysis and will support collaboration and interoperability of new methods as they are developed.
 In order to do this, we have followed the current standardized descriptions of FCM data analysis as being developed under NIH Grant xxxx [n].
 We believe that researchers will find flowCore to be a solid foundation for future development of new methods to attack the many interesting open research questions in FCM data analysis.\\
-\textbf{Methods} We propose a variety different data structures. We have implemented the classes and methods in the Bioconductor package flowCore. 
+
+\noindent \textbf{Methods} We propose a variety different data structures. We have implemented the classes and methods in the Bioconductor package flowCore. 
 We illustrate their use with X case studies.\\
-\textbf{Results} We hope that those proposed data structures will the base for the development of many tools for the analysis of high throughput flowcytometry.\\
-\textbf{keywords} Flow cytometry, high throughtput, software, standard
+
+\noindent \textbf{Results} We hope that those proposed data structures will the base for the development of many tools for the analysis of high throughput flowcytometry.\\
+
+\noindent \textbf{keywords} Flow cytometry, high throughtput, software, standard
 \end{abstract}
+
+<<echo=false,results=hide>>=
+library(flowCore)
+@
 
@@
 
-\section{Data model}
-The package \Rpackage{flowCore} offers the possibility to store single measurement (FCS file) or combine several individual
-measurements in the confines of a single experiment and to include all the necessary metadata. Its object
-model tries to stay close to the familiar micro-array data structures (expressionSet) making use of already
-defined generic functions.\\
+\section{Representing Flow Cytometry Data}
 
-\subsection{flowFrame}
-The objects of class \Robject{flowFrame} are the containers for storing individual 
-cytometry measurements originated from one FCS file. \\
-%%need description of the slots and maybe methods associated
+\Rpackage{flowCore}'s primary task is the representation and basic manipulation of flow cytometry (or similar) data.
+This is accomplished through a data model very similar to that adopted by other Bioconductor packages using the \Robject{expressionSet} and \Robject{AnnotatedDataFrame} structures familiar to most Bioconductor users.
 
-\subsection{flowSet}
-An flow cytometry experiment usually include several sample and several measurements. 
-A data structure is thus needed to store those collections of measurements.  
-The package \Rpackage{flowCore} stores the measurement and the associated data 
-information in objects of class \Robject{flowSet}. 
-A \Robject{flowSet} object contains two main slots \Robject{exprs}, an environment data structure 
-that stores the flow cytometry measurements and the \Robject{phenoData}. The \Robject{phenoData} slot 
-contains all the relevant experiment-wide meta-data, \textit{e.g.} 
-phenotypic and experimental information relevant to the management and/or the analysis of the data.\\
+\subsection{The \Rclass{flowFrame} Class}
 
-\section{Preprocessing Methods}
-\subsection{Compensation}
-Fluorescence compensation is the process by which the amounts of spectral overlap are 
-estimated and subtracted from the total detected signals to yield an estimate of the 
-actual amount of each dye \cite{Herzenberg2006}. Single-laser, two-color FACS experiments 
-in the mid-1970s showed that fluorescence compensation is important. 
-It is now recognized as an obligate step in most FACS analyses.
-%%definition: Compensation vs spillover
+The basic unit of manipulation in \Rpackage{flowCore} is the \Rclass{flowFrame}, which corresponds roughly with a single ``FCS'' file exported from the flow cytometer's acquisition software.
+At the moment we support FCS file versions 1.0 through 3.0, and we expect to support FCS4/ACS1 as soon as the specification has been ratified.
 
-\subsection{Transformation}
-While analyzing flow cytometry data, various parameter transformations are performed 
-to provide user-friendly visualization and/or to perform statistical analyses and interpret the data. 
-Following the standard \cite{Spidlen2006} 8 transformations are pre-defined: linear, quadratic, natural logarithm, 
-logarithm (with a specified base), hyperlog, bi-exponential, logicle, and split-scale.\\
+\subsubsection{Data elements}
 
-%%Can we define other?   
-%%can be applied when reading in the data.
+The primary elements of the \Robject{flowFrame} are the \Robject{exprs} and \Robject{parameters} slots, which contain the event-level information and column metadata respectively.
+The event information, stored as a single matrix, is accessed and manipulated via the \Rfunction{exprs()} and \Rfunction{exprs<-} methods, allowing \Robject{flowFrame}s to be stitched together if necessary (for example, if the same tube has been collected in two acquisition files for memory reasons).
 
-\subsection{Filter}
-The role of filtering strategies are to identify and characterize diffrent cells population in sample.
-%% Filtering vs Gating
-%%Byron's vignette?
-%%several approach from basic to DAG ...\\
+The \Robject{parameters} slot is an \Rclass{AnnotatedDataFrame} that contains information derived from an FCS file's ``\$P<n>'' keywords, which describe the detector and stain information.
+The entire list is available via the \Rfunction{parameter()} method, but more commonly this information is accessed through the \Rfunction{names}, \Rfunction{featureNames} and \Rfunction{colnames} methods.
+The \Rfunction{names} function returns a concatenated version of \Rfunction{names} and \Rfunction{featureNames} using a format similar to the one employed by most flow cytometry analysis software.
+The \Rfunction{colnames} method returns the detector names, often named for the fluorochrome detected, while the \Rfunction{featureNames} methods returns the description field of the parameters, which will typically be an identifier for the antibody.
 
-\section{Case study}
-\subsection{dataset}
-Our experiment is composed of 5 96-well plates\\
-%%Perry can we use that example or do you prefere an other one? 
-%% Can you add a paragraph describing the experiment?
+The \Rfunction{keyword} method allows access to the raw FCS keywords, which are a mix of standard entries such as ``SAMPLE ID,'' vendor specific keywords and user-defined keywords that add more information about an experiment.
+In the case of plate-based experiments, there are also one or more keywords that identify the specific well on the plate.
 
-First we load the library and read in some experimental data files.
-<<loadlib, echo=TRUE, results= hide>>=
-library(flowCore)
+Most vendor software also include some sort of unique identifier for the file itself.
+The specialized methods \Rfunction{identifier} attempts to locate an appropriate globally unique identifier that can be used to uniquely identify a frame.
+Failing that, this method will return the original file name offering some assurance that this frame is at least unique to a particular session.
+
+\subsubsection{Reading a \Robject{flowFrame}}
+
+FCS files are read into the R environment via the \Rfunction{read.FCS} function using the standard connection interface---allowing for the possibility of accessing FCS files hosted on a remote resource as well as those that have been compressed or even retrieved as a blob from a database interface.
+
+There are also several immediate processing options available in this function, the most important of which is the \Rfunction{transformation} parameter, which can either ``linearize'' (the default) or ``standardize'' our data.
+To see how this works, first we will examine an FCS file without any transformation at all:
+
+<<echo=true,results=verbatim>>=
+file.name <- system.file("extdata","0877408774.B08", package="flowCore")
+x = read.FCS(file.name,transformation=FALSE)
+summary(x)
 @
 
-<<readData, echo=TRUE, results=hide>>=
-fcs.loc <- system.file("extdata/compdata/data", package="flowCore")
-file.location <- paste(fcs.loc, dir(fcs.loc), sep="/")
-experiment <- read.flowSet(file.location, transformation = "linearize")
-experiment
-@ 
-\subsection{Quality Assessment}
-Quickly look at the range of the intensities between samples 
-other statistics sur as ... cf Janet paper.
-<<summaryRange>>=
-rang <- round(fsApply(experiment, each_col, range), 2)
-rownames(rang) <- paste(rep(dir(fcs.loc), each=2), c("lower","upper") ,sep="")
-rang
+As we can see, in this case the values from each parameter seem to run from $0$ to $1023$ ($2^{10}-1$).
+However, inspection of the ``exponentiation'' keyword (\$P<n>E) reveals that some of the parameters (3 and 4) have been stored in the format of the form $a\times{10^{x/R}}$ where $a$ is given by the first element of the string.
+
+<<echo=true,results=verbatim>>=
+keyword(x,c("$P1E","$P2E","$P3E","$P4E"))
 @
 
-\subsection{Preprocessing}
-<<compensationMatrix, echo=FALSE, results= hide>>=
-comp.loc <- system.file("extdata/compdata", package="flowCore")
-comp.location <- paste(comp.loc, dir(comp.loc), sep="/")
-comp.mat <- as.matrix(read.table(comp.location[1], header=TRUE, skip=2, check.names = FALSE))
+The default ``linearize'' transformation option will convert these to, effectively, have a ``\$P<n>E'' of ``0,0'':
+
+<<echo=true,results=verbatim>>=
+summary(read.FCS(file.name))
 @
 
-Compensation is ....
-<<compensateExp, echo=TRUE>>=
-expCompensate <- compensate(experiment,comp.mat)
+Finally, the ``scale'' option will both linearize values as well as ensure that output values are contained in $[0,1]$, which is the proposed method of data storage for the ACS1.0/FCS4.0 specification:
+
+<<echo=true,results=verbatim>>=
+summary(read.FCS(file.name,transformation="scale"))
 @
 
-Transformation
-<<scaleScatter, echo=TRUE>>=
-scaleTrans <- scaleTransform("scatter-scale", a=0,  b=1023)
-expCompensate <- transform(expCompensate,`FSC-H`= scaleTrans(`FSC-H`), `SSC-H`= scaleTrans(`SSC-H`))
+Another parameter of interest is the \Rfunction{alter.names} parameter, which will convert the parameter names into more ``R friendly'' equivalents, usually by replacing ``-'' with ``.'':
+
+<<echo=true,results=verbatim>>=
+read.FCS(file.name,alter.names=TRUE)
 @
 
-Filter to leave out debris:
-<<rectangleFilter, echo=TRUE>>=
-filterDebris <- rectangleGate("FSC-H"=c(0.2, 1), "SSC-H"=c(0.04, 1))
-expResult <- fsApply(expCompensate, function(x) filter(x, filterDebris))
+Finally, when only a particular subset of parameters is desired the \Rfunction{column.pattern} parameter allows for the specification of a regular expression and only parameters that match the regular expression will be included in the frame.
+For example, to include on the the Height parameters:
 
-cell <- Subset(expCompensate[[1]],expResult[[1]])
+<<echo=true,results=verbatim>>=
+x = read.FCS(file.name,column.pattern="-H")
+x
 @
 
-<<plotWithDebris, echo=FALSE, fig=TRUE, prefix=FALSE, include=FALSE>>=
-print(plot(expCompensate[[1]], xlim=c(0, 1), ylim=c(0, 1), 
-main="File1 (compensated, scaled)"))
+\textit{Note that \Rfunction{column.pattern} is applied after \Rfunction{alter.names} if it is used.}
+
+\subsubsection{Visualizing a \Rclass{flowFrame}}
+
+Much of the more sophisticated visualization of \Rclass{flowFrame} and \Rclass{flowSet} objects, including an interface to the \Rpackage{lattice} graphics system is implemented by the \Rpackage{flowViz} package, also included as part of Bioconductor.
+However, \Rpackage{flowCore} does implement some basic plotting facilities using the standard \Rfunction{plot} function.
+The basic plot provides a simple bivariate density plot of the first two parameters:
+
+<<fig=true,echo=true>>=
+plot(x)
 @
 
-<<plotCell,echo=False, fig=TRUE, prefix=FALSE, include=FALSE>>=
-print(plot(cell, xlim=c(0, 1), ylim=c(0, 1), 
-     main="File1 (compensated, scaled, debris removed)"))
+To control the parameters being plotted we can supply a \Robject{y} value in the form of a character vector:
+
+<<fig=true,echo=true>>=
+plot(x,c("FL1-H","FL2-H"))
 @
 
-\begin{figure}
-\centering
-\subfigure{\includegraphics[width=40mm]{plotWithDebris} A}
-\subfigure{\includegraphics[width=40mm]{plotCell} B}
-\caption{
-  \label{fig:Fig01}{Example of smooth scatter plot of FSC-H
-    \textit{vs.} SSC-H data (A) before and (B) after a applying a rectangle 
-on the  measurements (compensated and scaled) extracted from File1}} 
-\end{figure}
+However, if we only supply a single parameter we instead get a univariate histogram, which also accepts the usual histogram arguments:
 
-\section{Discussion and Conclusion}
-%% Explain the usefullness of a commun data structure maybe in comparaison with the ExpressionSet developed for
-%% microarray - count how manay package in BioC depend on the  ExpressionSet data structure?
-%%flowViz
-%%flowQ etc
+<<fig=true,echo=true>>=
+plot(x,"FL1-H",breaks=256)
+@
 
-\section*{Acknowledgements}
-We would like to thank R. Gentleman RR. Brinkman
-%%or should they be Authors
+\subsection{The \Rclass{flowSet} Class}
+
+Most experiments consist of several \Rclass{flowFrame} objects, which are organized using a \Rclass{flowSet} object.
+This class provides a mechanism for efficiently hosting the \Rclass{flowFrame} objects with minimal copying, reducing memory requirements, as well as ensuring that experimental metadata stays properly to the appropriate \Rclass{flowFrame} objects.
+
+\subsubsection{Creating a \Rclass{flowSet}}
+
+To facilitate the creation of \Rclass{flowSet} objects from a variety of sources, we provide a means to coerce \Rclass{list} and \Rclass{environment} objects to a \Rclass{flowSet} object using the usual coercion mechanisms.
+For example, if we have a directory containing FCS files we can read in a list of those files and create a \Rclass{flowSet} out of them:
+
+<<echo=true,results=verbatim>>=
+frames = lapply(dir(system.file("extdata","compdata","data",package="flowCore"),full.names=TRUE),read.FCS)
+as(frames,"flowSet")
+@
+
+Note that the original list is unnamed and that the resulting sample names are not particularly meaningful.
+If the list is named, the list constructed is much more meaningful.
+One such approach is to employ the \Rfunction{keyword} method for \Rclass{flowFrame} objects to extract the ``SAMPLE ID'' keyword from each frame:
+
+<<echo=true,results=verbatim>>=
+names(frames) = sapply(frames,keyword,"SAMPLE ID")
+fs = as(frames,"flowSet")
+fs
+@
+
+\subsubsection{Working with experimental metadata}
+
+Like most Bioconductor organizational classes, the \Rclass{flowSet} has an associated \Rclass{AnnotatedDataFrame} that provides metadata not contained within the \Rclass{flowFrame} objects themselves.
+This data frame is accessed and modified via the usual \Rfunction{phenoData} and \Rfunction{phenoData<-} methods.
+You can also generally treat the phenotypic data as a normal data frame to add new descriptive columns.
+For example, we might want to track the original filename of the frames from above in the phenotypic data for easier access:
+
+<<echo=true,results=verbatim>>=
+phenoData(fs)$Filename = fsApply(fs,keyword,"$FIL")
+pData(phenoData(fs))
+@
+
+Note that we have used the \Rclass{flowSet}-specific iterator, \Rfunction{fsApply}, which acts much like \Rfunction{sapply} or \Rfunction{lapply}.
+Additionally, we should also note that the \Rfunction{phenoData} data frame \textbf{must} have row names that correspond to the original names used to create the \Rclass{flowSet}.
+
+\subsubsection{Bringing it all together: \Rfunction{read.flowSet}}
+
+Much of the functionality described above has been packaged into the \Rfunction{read.flowSet} convenience function.
+In it's simplest incarnation, this function takes a \Rfunction{path}, that defaults to the current working directory, and an optional \Rfunction{pattern} argument that allows only a subset of files contained within the working directory to be selected.
+For example, to read a \Rclass{flowSet} of the files read in by \Robject{frame} above:
+
+<<echo=true,results=verbatim>>=
+read.flowSet(path=system.file("extdata","compdata","data",package="flowCore"))
+@
+
+\Rfunction{read.flowSet} will pass on additional arguments meant for the underlying \Rfunction{read.FCS} function, such as \Rfunction{alter.names} and \Rfunction{column.pattern}, but also supports several other interesting arguments for conducting initial processing:
+\begin{description}
+\item[files]{ An alternative to the \Rfunction{pattern} argument, you may also supply a vector of filenames to read.}
+\item[name.keyword]{ Like the example in the previous section, you may specify a particular keyword to use in place of the filename when creating the \Rclass{flowSet}.}
+\item[phenoData]{ If this is an \Rclass{AnnotatedDataFrame}, then this will be used in place of the data frame that is ordinarily created.
+Additionally, the row names of this object will be taken to be the filenames of the FCS files in the directory specified by \Robject{path}.
+This argument may also be a named list made up of a combination of \Robject{character} and \Robject{function} objects that specify a keyword to extract from the FCS file or a function to apply to each frame that will return a result.
+}
+\end{description}
+
+To recreate the \Rclass{flowSet} that we created by hand from the last section we can use \Rfunction{read.flowSet}s advanced functionality:
+
+<<echo=true,results=verbatim>>=
+fs = read.flowSet(path=system.file("extdata","compdata","data",package="flowCore"),name.keyword="SAMPLE ID",phenoData=list(name="SAMPLE ID",Filename="$FIL"))
+fs
+pData(phenoData(fs))
+@
+
+\subsubsection{Manipulating a \Rclass{flowSet}}
+
+You can extract a \Rclass{flowFrame} from a \Rclass{flowSet} object in the usual way using the \Rfunction{[[} or \Rfunction{\$} extraction operators.
+On the other hand using he \Rfunction{[} extraction operator returns a new \Rclass{flowSet} by \textbf{copying} the environment.
+However, simply assigning the \Rclass{flowFrame} to a new variable will \textbf{not} copying the contained frames.
+
+The primary iterator method for a \Rclass{flowSet} is the \Rfunction{fsApply} method, which works more-or-less like \Rfunction{sapply} or \Rfunction{lapply} with two extra options.
+The first argument, \Robject{simplify}, which defaults to \Robject{TRUE}, instructs \Rfunction{fsApply} to attempt to simplify it's results much in the same way as \Rfunction{sapply}.
+The primary difference is that if all of the return values of the iterator are \Rclass{flowFrame} objects, \Rfunction{fsApply} will create a new \Rclass{flowSet} object to hold them.
+The second argument, \Robject{use.exprs}, which defaults to \Robject{FALSE} instructs \Rfunction{fsApply} to pass the expression matrix of each frame rather than the \Rclass{flowFrame} object itself.
+This allows functions to operate directly on the intensity information without first having to extract it.
+
+As an aid to this sort of operation we also introduce the \Rfunction{each\_row} and \Rfunction{each\_col} convenience functions that take the place of \Rfunction{apply} in the \Rfunction{fsApply} call.
+For example, if we wanted the median value of each parameter of each \Rclass{flowFrame} we might write:
+
+<<echo=true,results=verbatim>>=
+fsApply(fs,each_col,median)
+@
+
+which is equivalent to the less readable
+
+<<echo=true,results=verbatim>>=
+fsApply(fs,function(x) apply(x,2,median),use.exprs=TRUE)
+@
+
+In this case, the \Robject{use.exprs} argument is not required in the first case because \Rfunction{each\_col} and \Rfunction{each\_row} are methods and have been defined to work on \Rclass{flowFrame} objects by first extracting the intensity data.
+
+\section{Transformation}
+
+\Rpackage{flowCore} features two methods of transforming parameters within a \Rclass{flowFrame}: inline and out-of-line.
+The inline method, discussed in the next section has been developed primarily to support filtering features and is strictly more limited than the out-of-line transformation method, which uses R's \Rfunction{transform} function to accomplish the filtering.
+Like the normal \Rfunction{transform} function, the \Rclass{flowFrame}is considered to be a data frame with columns named for parameters of the FCS file.
+For example, if we wished to plot our first \Rclass{flowFrame}'s first two fluorescence parameters on the log scale we might write:
+
+<<fig=true,echo=true>>=
+plot(transform(fs[[1]],`FL1-H`=log(`FL1-H`),`FL2-H`=log(`FL2-H`)),c("FL1-H","FL2-H"))
+@
+
+Like the usual \Rfunction{transform} function, we can also create new parameters based on the old parameters, without destroying the old
+
+<<fig=true,echo=true>>=
+plot(transform(fs[[1]],log.FL1.H=log(`FL1-H`),log.FL2.H=log(`FL2-H`)),c("log.FL1.H","log.FL2.H"))
+@
+
+\subsection{Standard Transforms}
+
+Though any function can be used as a transform in both the out-of-line and inline transformation techniques, \Rpackage{flowCore} provides a number of parameterized transform generators that correspond to the transforms commonly found in flow cytometry and defined in the Transformation Markup Language (Transformation-ML).
+Briefly, the predefined transforms are:
+
+\begin{description}
+\item[truncateTransform]{ $y = \left\{\begin{array}{ll}
+a & x < a \\
+x & x \geq a \\
+\end{array}\right.$ }
+\item[scaleTransform]{$f(x) = \frac{x-a}{b-a}$}
+\item[linearTransform]{ $f(x) = a+bx$ }
+\item[quadraticTransform]{ $f(x) = ax^2 + bx + c$ }
+\item[lnTransform]{$ f(x) = \log\left(x\right)\frac{r}{d} $}
+\item[logTransform]{$ f(x) = \log_b\left(x\right)\frac{r}{d} $}
+\item[biexponentialTransform]{ $f^{-1}(x) = ae^{bx}-ce^{dx}+f$}
+\item[logicleTransform]{ A special form of the biexponential transform with parameters selected by the data.}
+\item[arcsinhTransform]{ $f(x) = \asinh\left(a+bx\right)+c$}
+\end{description}
+
+To use a standard transform, first we create a transform function via the constructors supplied by \Rpackage{flowCore}:
+
+<<echo=true,results=verbatim>>=
+aTrans = truncateTransform("truncate at 1",a=1)
+aTrans
+@
+which we can then use on the parameter of interest in the usual way
+<<echo=true,results=verbatim>>=
+transform(fs,`FL1-H`=aTrans(`FL1-H`))
+@
+
+\section{Filtering}
+
+The most common task in the analysis of flow cytometry data is usual some form of filtering operation, either to obtain summary statistics about the number of events that meet a certain criteria or to perform further analysis on a subset of the data.
+
+\subsection{Standard Filters}
+
+Most filtering operations are a composition of one or more common filtering operations. Like transformations, \Rpackage{flowCore} includes a number of built-in common flow cytometry filters.
+
+The simplest of these filters are the geometric filters, which correspond to those typically found in interactive flow cytometry software:
+\begin{description}
+\item[rectangleGate]{Describes a cubic shape in one or more dimensions--a rectangle in one dimension is simply an interval gate.}
+\item[polygonGate]{Describes an arbitrary two dimensional polygonal gate.}
+\item[polytopeGate]{Describes a region that is the convex hull of the given points. This gate can exist in dimensions higher than $2$, unlike the \Robject{polygonGate}.}
+\item[ellipsoidGate]{Describes an ellipsoidal region in two or more dimensions}
+\end{description}
+
+These gates are all described in more or less the same manner:
+
+<<echo=true,results=verbatim>>=
+rectGate = rectangleGate(filterId="Fluorescence Region","FL1-H"=c(50,100),"FL2-H"=c(50,100))
+@
+
+We also include some data-driven filters not usually found in flow cytometry software:
+
+\begin{description}
+\item[norm2Filter]{ A robust method for finding a region that most resembles a bivariate Normal distribution. }
+\item[kmeansFilter]{ Identifies populations based on a one dimensional k-means clustering operation. Allows the specification of \textbf{multiple} populations. }
+\end{description}
+
+\subsection{Count Statistics}
+
+When we have constructed a filter, we can apply it in two basic ways.
+The first is to collect simple summary statistics on the number and proportion of events considered to be contained within the gate or filter.
+This is done using the \Rfunction{filter} method.
+The first step is to apply our filter to some data
+
+<<echo=true,results=verbatim>>=
+result = filter(fs[[1]],rectGate)
+result
+@
+
+As we can see, we have returned a \Rclass{filterResult} object, which is in turn a filter allowing for reuse in, for example, subsetting operations.
+To obtain count and proportion statistics, we take the summary of this \Rclass{filterResult}, which returns a list of summary values:
+
+<<echo=true,results=verbatim>>=
+summary(result)
+summary(result)$n
+summary(result)$true
+summary(result)$p
+@
+
+A filter which contains multiple populations, such as the \Rclass{kmeansFilter}, can return a list of summary lists:
+
+<<echo=true,results=verbatim>>=
+summary(filter(fs[[1]],kmeansFilter("myKMeans","FSC-H"=c("Low","Medium","High"))))
+@
+
+A filter may also be applied to an entire \Rclass{flowSet}, in which case it returns a list of \Rclass{filterResult} objects:
+
+<<echo=true,results=verbatim>>=
+filter(fs,rectGate)
+@
+
+\subsection{Subsetting}
+
+To subset or split a \Rclass{flowFrame} or \Rclass{flowSet}, we use the \Rfunction{Subset} and \Rfunction{split} methods respectively.
+The first, \Rfunction{Subset}, behaves similarly to the standard R \Rfunction{subset} function, which unfortunately could not used.
+For example, recall from our initial plots of this data that the morphology parameters, Forward Scatter and Side Scatter contain a large more-or-less ellipse shaped population.
+If we wished to deal only with that population, we might use \Rfunction{Subset} along with a \Rclass{norm2Filter} object as follows:
+
+<<echo=true,results=verbatim>>=
+morphGate = norm2Filter(filterId="MorphologyGate","FSC-H","SSC-H",scale=2)
+smaller = Subset(fs,morphGate)
+fs[[1]]
+smaller[[1]]
+@
+
+Notice how the \Robject{smaller} \Rclass{flowFrame} objects contain fewer events.
+Now imagine we wanted to use a \Rclass{kmeansFilter} as before to split our first fluorescence parameter into three populations.
+To do this we employ the \Rfunction{split} function:
+
+<<echo=true,results=verbatim>>=
+split(smaller[[1]],kmeansFilter("myKMeans","FSC-H"=c("Low","Medium","High")))
+@
+
+or for an entire \Rclass{flowSet}
+
+<<echo=true,results=verbatim>>=
+split(smaller,kmeansFilter("myKMeans","FSC-H"=c("Low","Medium","High")))
+@
+
+\subsection{Combining Filters}
+
+Of course, most filtering operations consist of more than one gate.
+To combine gates and filters we use the standard R Boolean operators: \&, \| and !
+to construct an intersection, union and complement respectively:
+
+<<echo=true,results=verbatim>>=
+
+rectGate & morphGate
+rectGate | morphGate
+!morphGate
+
+@
+
+we also introduce the notion of the subset operation, denoted by either \%subset\% or \%\&\%.
+This combination of two gates first performs a subsetting operation on the input \Rclass{flowFrame} using the right-hand filter and then applies the left-hand filter.
+For example,
+
+<<echo=true,results=verbatim>>=
+summary(filter(smaller[[1]],rectGate %&% morphGate))
+@
+
+first calculates a subset based on the \Robject{morphGate} filter and then applies the \Robject{rectGate}.
+
+\subsection{Transformation Filters}
+
+Finally, it is sometimes desirable to construct a filter with respect to transformed parameters.
+To allow for this in our filtering constructs we introduce a special form of the \Rfunction{transform} method along with another filter combination operator \%on\%, which can be applied to both filters and \Rclass{flowFrame} or \Rclass{flowSet} objects.
+To specify our transform filter we must first construct a transform list using a simplified version of the \Rfunction{transform} function:
+
+<<echo=true,results=verbatim>>=
+tFilter = transform("FL1-H"=log,"FL2-H"=log)
+tFilter
+@
+
+Note that this version of the transform filter does not take parameters on the right-hand side--the functions can only take a single vector that is specified by the parameter on the left-hand side.
+In this case those parameters are ``FL1-H'' and ``FL2-H.'' The function also does not take a specific \Rclass{flowFrame} or \Rclass{flowSet} allowing us to use this with any appropriate data.
+We can then construct a filter with respect to the transform as follows:
+
+<<echo=true,results=verbatim>>=
+rect2 = rectangleGate(filterId="Another Rect","FL1-H"=c(1,2),"FL2-H"=c(2,3)) %on% tFilter
+rect2
+@
+
+Additionally, we can use this construct directly on a \Rclass{flowFrame} or \Rclass{flowSet} by moving the transform to the left-hand side and placing the data on the right-hand side:
+
+<<fig=true,echo=true>>=
+plot(tFilter %on% smaller[[1]],c("FL1-H","FL2-H"))
+@
+
+which has the same effect as the $\log$ transform used earlier.
 
@@
 \bibliography{cytoref}
-\clearpage
-\small{
-<<sessionInfo>>=
-sessionInfo()
-@ 
-}
 \end{document}

### 2007-04-16 inst/doc/HowTo-flowCore.Rnw
Fix vignette and man page
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23904 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\title{flowCore: data structures pacakge for flow cytometry data}
+\title{flowCore: data structures package for flow cytometry data}
 \author{N. Le Meur F. Hahne B. Ellis P. Haaland}
@@
 
-<<echo=false,results=hide>>=
+<<loadPackage, echo=false,results=hide>>=
 library(flowCore)
@@
 
-<<echo=true,results=verbatim>>=
+<<ReadFiles, echo=true, results=verbatim>>=
 file.name <- system.file("extdata","0877408774.B08", package="flowCore")
-x = read.FCS(file.name,transformation=FALSE)
+x <- read.FCS(file.name, transformation=FALSE)
 summary(x)
@@
 
-<<echo=true,results=verbatim>>=
-keyword(x,c("$P1E","$P2E","$P3E","$P4E"))
+<<SearchKeywords, echo=true,results=verbatim>>=
+keyword(x,c("$P1E", "$P2E", "$P3E", "$P4E"))
 @
@@
 
-<<echo=true,results=verbatim>>=
+<<PrintSummary, echo=true, results=verbatim>>=
 summary(read.FCS(file.name))
@@
 
-<<echo=true,results=verbatim>>=
+<<PrintSummary2, echo=true,results=verbatim>>=
 summary(read.FCS(file.name,transformation="scale"))
@@
 
-<<echo=true,results=verbatim>>=
+<<ReadFiles2, echo=true,results=verbatim>>=
 read.FCS(file.name,alter.names=TRUE)
@@
 
-<<echo=true,results=verbatim>>=
-x = read.FCS(file.name,column.pattern="-H")
+<<RedFiles3, echo=true, results=verbatim>>=
+x <- read.FCS(file.name,column.pattern="-H")
 x
@@
 
-<<fig=true,echo=true>>=
+<<Plot1, fig=true, echo=true>>=
 plot(x)
@@
 
-<<fig=true,echo=true>>=
-plot(x,c("FL1-H","FL2-H"))
+<<Plot2, fig=true, echo=true>>=
+plot(x,c("FL1-H", "FL2-H"))
 @
@@
 
-<<fig=true,echo=true>>=
-plot(x,"FL1-H",breaks=256)
+<<plot3, fig=true, echo=true>>=
+plot(x, "FL1-H", breaks=256)
 @
@@
 
-<<echo=true,results=verbatim>>=
-frames = lapply(dir(system.file("extdata","compdata","data",package="flowCore"),full.names=TRUE),read.FCS)
-as(frames,"flowSet")
+<<Frames, echo=true, results=verbatim>>=
+frames <- lapply(dir(system.file("extdata", "compdata", "data", package="flowCore"), full.names=TRUE), read.FCS)
+as(frames, "flowSet")
 @
@@
 
-<<echo=true,results=verbatim>>=
-names(frames) = sapply(frames,keyword,"SAMPLE ID")
-fs = as(frames,"flowSet")
+<<Frames, echo=true,results=verbatim>>=
+names(frames) <- sapply(frames, keyword, "SAMPLE ID")
+fs <- as(frames, "flowSet")
 fs
@@
 
-<<echo=true,results=verbatim>>=
-phenoData(fs)$Filename = fsApply(fs,keyword,"$FIL")
+<<metaData, echo=true, results=verbatim>>=
+phenoData(fs)$Filename <- fsApply(fs,keyword, "$FIL")
 pData(phenoData(fs))
@@
 
-<<echo=true,results=verbatim>>=
-read.flowSet(path=system.file("extdata","compdata","data",package="flowCore"))
+<<ReadFlowSet, echo=true,results=verbatim>>=
+read.flowSet(path = system.file("extdata", "compdata", "data", package="flowCore"))
 @
@@
 
-<<echo=true,results=verbatim>>=
-fs = read.flowSet(path=system.file("extdata","compdata","data",package="flowCore"),name.keyword="SAMPLE ID",phenoData=list(name="SAMPLE ID",Filename="$FIL"))
+<<ReadFowSet2, echo=true,results=verbatim>>=
+fs <- read.flowSet(path=system.file("extdata", "compdata", "data", package="flowCore"), name.keyword="SAMPLE ID", phenoData=list(name="SAMPLE ID", Filename="$FIL"))
 fs
@@
 
-<<echo=true,results=verbatim>>=
-fsApply(fs,each_col,median)
+<<fsApply1, echo=true, results=verbatim>>=
+fsApply(fs, each_col, median)
 @
@@
 
-<<echo=true,results=verbatim>>=
-fsApply(fs,function(x) apply(x,2,median),use.exprs=TRUE)
+<<fsApply2, echo=true,results=verbatim>>=
+fsApply(fs,function(x) apply(x, 2, median), use.exprs=TRUE)
 @
@@
 
-<<fig=true,echo=true>>=
-plot(transform(fs[[1]],`FL1-H`=log(`FL1-H`),`FL2-H`=log(`FL2-H`)),c("FL1-H","FL2-H"))
+<<Transfo1, fig=true, echo=true>>=
+plot(transform(fs[[1]], `FL1-H`=log(`FL1-H`), `FL2-H`=log(`FL2-H`)), c("FL1-H","FL2-H"))
 @
@@
 
-<<fig=true,echo=true>>=
-plot(transform(fs[[1]],log.FL1.H=log(`FL1-H`),log.FL2.H=log(`FL2-H`)),c("log.FL1.H","log.FL2.H"))
+<<Transfo2, fig=true, prefix=FALSE, include=FALSE, echo=true>>=
+plot(transform(fs[[1]], log.FL1.H=log(`FL1-H`), log.FL2.H=log(`FL2-H`)), c("log.FL1.H", "log.FL2.H"))
 @
+
+\begin{figure}
+\centering
+\includegraphic{Transfo2}
+\end{figure}
 
@@
 \end{array}\right.$ }
-\item[scaleTransform]{$f(x) = \frac{x-a}{b-a}$}
-\item[linearTransform]{ $f(x) = a+bx$ }
-\item[quadraticTransform]{ $f(x) = ax^2 + bx + c$ }
+\item[scaleTransform]{$ f(x) = \frac{x-a}{b-a}$}
+\item[linearTransform]{ $ f(x) = a+bx$ }
+\item[quadraticTransform]{ $ f(x) = ax^2 + bx + c$ }
 \item[lnTransform]{$ f(x) = \log\left(x\right)\frac{r}{d} $}
 \item[logTransform]{$ f(x) = \log_b\left(x\right)\frac{r}{d} $}
-\item[biexponentialTransform]{ $f^{-1}(x) = ae^{bx}-ce^{dx}+f$}
+\item[biexponentialTransform]{ $ f^{-1}(x) = ae^{bx}-ce^{dx}+f$}
 \item[logicleTransform]{ A special form of the biexponential transform with parameters selected by the data.}
-\item[arcsinhTransform]{ $f(x) = \asinh\left(a+bx\right)+c$}
+\item[arcsinhTransform]{ $ f(x) = asinh\left(a+bx\right)+c$}
 \end{description}
@@
 
-<<echo=true,results=verbatim>>=
-aTrans = truncateTransform("truncate at 1",a=1)
+<<Transfo3, echo=true>>=
+aTrans <- truncateTransform("truncate at 1", a=1)
 aTrans
@@
 which we can then use on the parameter of interest in the usual way
-<<echo=true,results=verbatim>>=
+<<Transfo4, echo=true,results=verbatim>>=
 transform(fs,`FL1-H`=aTrans(`FL1-H`))
@@
 
-<<echo=true,results=verbatim>>=
-rectGate = rectangleGate(filterId="Fluorescence Region","FL1-H"=c(50,100),"FL2-H"=c(50,100))
+<<rectGate, echo=true, results=verbatim>>=
+rectGate <- rectangleGate(filterId="Fluorescence Region", "FL1-H"=c(50, 100), "FL2-H"=c(50, 100))
 @
@@
 
-<<echo=true,results=verbatim>>=
+<<Summary3, echo=true, results=verbatim>>=
 summary(result)
@@
 
-<<echo=true,results=verbatim>>=
-summary(filter(fs[[1]],kmeansFilter("myKMeans","FSC-H"=c("Low","Medium","High"))))
+<<SummarFilter, echo=true, results=verbatim>>=
+summary(filter(fs[[1]], kmeansFilter("myKMeans","FSC-H"=c("Low", "Medium", "High"))))
 @
@@
 
-<<echo=true,results=verbatim>>=
-morphGate = norm2Filter(filterId="MorphologyGate","FSC-H","SSC-H",scale=2)
-smaller = Subset(fs,morphGate)
+<<Norm2Filter, echo=true, results=verbatim>>=
+morphGate <- norm2Filter(filterId="MorphologyGate", "FSC-H", "SSC-H", scale=2)
+smaller <- Subset(fs, morphGate)
 fs[[1]]
@@
 
-<<echo=true,results=verbatim>>=
-split(smaller[[1]],kmeansFilter("myKMeans","FSC-H"=c("Low","Medium","High")))
+<<Split, echo=true, results=verbatim>>=
+split(smaller[[1]], kmeansFilter("myKMeans","FSC-H"=c("Low","Medium","High")))
 @
@@
 
-<<echo=true,results=verbatim>>=
-split(smaller,kmeansFilter("myKMeans","FSC-H"=c("Low","Medium","High")))
+<<Split2, echo=true, results=verbatim>>=
+split(smaller, kmeansFilter("myKMeans", "FSC-H"=c("Low", "Medium", "High")))
 @
@@
 Of course, most filtering operations consist of more than one gate.
-To combine gates and filters we use the standard R Boolean operators: \&, \| and !
+To combine gates and filters we use the standard R Boolean operators: \&, | and !
 to construct an intersection, union and complement respectively:
 
-<<echo=true,results=verbatim>>=
+<<CombineFilter, echo=true, results=verbatim>>=
 
@@
 
-<<echo=true,results=verbatim>>=
+<<Summary5, echo=true, results=verbatim>>=
 summary(filter(smaller[[1]],rectGate %&% morphGate))
@@
 
-<<echo=true,results=verbatim>>=
-tFilter = transform("FL1-H"=log,"FL2-H"=log)
+<<Transfo5, echo=true,results=verbatim>>=
+tFilter <- transform("FL1-H"=log,"FL2-H"=log)
 tFilter
@@
 
-<<echo=true,results=verbatim>>=
-rect2 = rectangleGate(filterId="Another Rect","FL1-H"=c(1,2),"FL2-H"=c(2,3)) %on% tFilter
+<<TectGate3, echo=true, results=verbatim>>=
+rect2 <- rectangleGate(filterId="Another Rect", "FL1-H"=c(1,2), "FL2-H"=c(2,3)) %on% tFilter
 rect2
@@
 
-<<fig=true,echo=true>>=
+<<Plot6, fig=true,echo=true>>=
 plot(tFilter %on% smaller[[1]],c("FL1-H","FL2-H"))

### 2007-04-16 inst/doc/HowTo-flowCore.Rnw
vignette
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@23923 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-<<Transfo2, fig=true, prefix=FALSE, include=FALSE, echo=true>>=
+<<Transfo2, fig=true, prefix=FALSE>>=
 plot(transform(fs[[1]], log.FL1.H=log(`FL1-H`), log.FL2.H=log(`FL2-H`)), c("log.FL1.H", "log.FL2.H"))
 @
-
-\begin{figure}
-\centering
-\includegraphic{Transfo2}
-\end{figure}
 

### 2007-05-30 inst/doc/HowTo-flowCore.Rnw
fix read.FCS function to support NA under the keywords ANASTART ANAEND of the header section
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@24495 bc3139a8-67e5-0310-9ffc-ced21a209358
Add splitScaleTransformation (for transformML) Modify read.FCS to remove random sampling but leave which.lines Add parentId argument in class Filter Fix rectangle gate boundary allow Min>=Max like in GateML standard, return empty gate
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@25068 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-The basic unit of manipulation in \Rpackage{flowCore} is the \Rclass{flowFrame}, which corresponds roughly with a single ``FCS'' file exported from the flow cytometer's acquisition software.
-At the moment we support FCS file versions 1.0 through 3.0, and we expect to support FCS4/ACS1 as soon as the specification has been ratified.
+The basic unit of manipulation in \Rpackage{flowCore} is the
+\Rclass{flowFrame}, which corresponds roughly with a single
+``FCS'' file exported from the flow cytometer's acquisition
+software. At the moment we support FCS file versions 1.0
+through 3.0, and we expect to support FCS4/ACS1 as soon as
+the specification has been ratified.
 
@@
 
-The primary elements of the \Robject{flowFrame} are the \Robject{exprs} and \Robject{parameters} slots, which contain the event-level information and column metadata respectively.
-The event information, stored as a single matrix, is accessed and manipulated via the \Rfunction{exprs()} and \Rfunction{exprs<-} methods, allowing \Robject{flowFrame}s to be stitched together if necessary (for example, if the same tube has been collected in two acquisition files for memory reasons).
+The primary elements of the \Robject{flowFrame} are the
+\Robject{exprs} and \Robject{parameters} slots, which
+contain the event-level information and column metadata
+respectively. The event information, stored as a single
+matrix, is accessed and manipulated via the
+\Rfunction{exprs()} and \Rfunction{exprs<-} methods,
+allowing \Robject{flowFrame}s to be stitched together if
+necessary (for example, if the same tube has been collected
+in two acquisition files for memory reasons).
 
-The \Robject{parameters} slot is an \Rclass{AnnotatedDataFrame} that contains information derived from an FCS file's ``\$P<n>'' keywords, which describe the detector and stain information.
-The entire list is available via the \Rfunction{parameter()} method, but more commonly this information is accessed through the \Rfunction{names}, \Rfunction{featureNames} and \Rfunction{colnames} methods.
-The \Rfunction{names} function returns a concatenated version of \Rfunction{names} and \Rfunction{featureNames} using a format similar to the one employed by most flow cytometry analysis software.
-The \Rfunction{colnames} method returns the detector names, often named for the fluorochrome detected, while the \Rfunction{featureNames} methods returns the description field of the parameters, which will typically be an identifier for the antibody.
+The \Robject{parameters} slot is an
+\Rclass{AnnotatedDataFrame} that contains information
+derived from an FCS file's ``\$P<n>'' keywords, which
+describe the detector and stain information. The entire list
+is available via the \Rfunction{parameter()} method, but
+more commonly this information is accessed through the
+\Rfunction{names}, \Rfunction{featureNames} and
+\Rfunction{colnames} methods. The \Rfunction{names} function
+returns a concatenated version of \Rfunction{names} and
+\Rfunction{featureNames} using a format similar to the one
+employed by most flow cytometry analysis software. The
+\Rfunction{colnames} method returns the detector names,
+often named for the fluorochrome detected, while the
+\Rfunction{featureNames} methods returns the description
+field of the parameters, which will typically be an
+identifier for the antibody.
 
-The \Rfunction{keyword} method allows access to the raw FCS keywords, which are a mix of standard entries such as ``SAMPLE ID,'' vendor specific keywords and user-defined keywords that add more information about an experiment.
-In the case of plate-based experiments, there are also one or more keywords that identify the specific well on the plate.
+The \Rfunction{keyword} method allows access to the raw FCS
+keywords, which are a mix of standard entries such as
+``SAMPLE ID,'' vendor specific keywords and user-defined
+keywords that add more information about an experiment. In
+the case of plate-based experiments, there are also one or
+more keywords that identify the specific well on the plate.
 
-Most vendor software also include some sort of unique identifier for the file itself.
-The specialized methods \Rfunction{identifier} attempts to locate an appropriate globally unique identifier that can be used to uniquely identify a frame.
-Failing that, this method will return the original file name offering some assurance that this frame is at least unique to a particular session.
+Most vendor software also include some sort of unique
+identifier for the file itself. The specialized methods
+\Rfunction{identifier} attempts to locate an appropriate
+globally unique identifier that can be used to uniquely
+identify a frame. Failing that, this method will return the
+original file name offering some assurance that this frame
+is at least unique to a particular session.
 
@@
 
-FCS files are read into the R environment via the \Rfunction{read.FCS} function using the standard connection interface---allowing for the possibility of accessing FCS files hosted on a remote resource as well as those that have been compressed or even retrieved as a blob from a database interface.
+FCS files are read into the R environment via the
+\Rfunction{read.FCS} function using the standard connection
+interface---allowing for the possibility of accessing FCS
+files hosted on a remote resource as well as those that have
+been compressed or even retrieved as a blob from a database
+interface. FCS files (version 2.0 and 3.0) and LMD (List
+Mode Data) extensions are currently supported.
 
-There are also several immediate processing options available in this function, the most important of which is the \Rfunction{transformation} parameter, which can either ``linearize'' (the default) or ``standardize'' our data.
-To see how this works, first we will examine an FCS file without any transformation at all:
+There are also several immediate processing options
+available in this function, the most important of which is
+the \Rfunction{transformation} parameter, which can either
+``linearize'' (the default) or ``standardize'' our data. To
+see how this works, first we will examine an FCS file
+without any transformation at all:
 
@@
 
-As we can see, in this case the values from each parameter seem to run from $0$ to $1023$ ($2^{10}-1$).
-However, inspection of the ``exponentiation'' keyword (\$P<n>E) reveals that some of the parameters (3 and 4) have been stored in the format of the form $a\times{10^{x/R}}$ where $a$ is given by the first element of the string.
+As we can see, in this case the values from each parameter
+seem to run from $0$ to $1023$ ($2^{10}-1$).  However,
+inspection of the ``exponentiation'' keyword (\$P<n>E)
+reveals that some of the parameters (3 and 4) have been
+stored in the format of the form $a\times{10^{x/R}}$ where
+$a$ is given by the first element of the string.
 
@@
 
-Finally, the ``scale'' option will both linearize values as well as ensure that output values are contained in $[0,1]$, which is the proposed method of data storage for the ACS1.0/FCS4.0 specification:
+Finally, the ``scale'' option will both linearize values as
+well as ensure that output values are contained in $[0,1]$,
+which is the proposed method of data storage for the
+ACS1.0/FCS4.0 specification:
 
@@
 
-Another parameter of interest is the \Rfunction{alter.names} parameter, which will convert the parameter names into more ``R friendly'' equivalents, usually by replacing ``-'' with ``.'':
+Another parameter of interest is the \Rfunction{alter.names}
+parameter, which will convert the parameter names into more
+``R friendly'' equivalents, usually by replacing ``-'' with
+``.'':
 
@@
 
-Finally, when only a particular subset of parameters is desired the \Rfunction{column.pattern} parameter allows for the specification of a regular expression and only parameters that match the regular expression will be included in the frame.
-For example, to include on the the Height parameters:
+When only a particular subset of parameters is
+desired the \Rfunction{column.pattern} parameter allows for
+the specification of a regular expression and only
+parameters that match the regular expression will be
+included in the frame. For example, to include on the the
+Height parameters:
 
 <<RedFiles3, echo=true, results=verbatim>>= 
-x <- read.FCS(file.name,column.pattern="-H")
+x <- read.FCS(file.name, column.pattern="-H") 
 x 
@@
 \textit{Note that \Rfunction{column.pattern} is applied after \Rfunction{alter.names} if it is used.}
+
+Finally, only a sample of lines can be read in case you need a quick overview of a large serie of files. 
+<<RedFiles3, echo=true, results=verbatim>>= 
+lines <- sample(100:500, 50)
+y <- read.FCS(file.name, which.lines = lines) 
+y 
+@
 
@@
 
-However, if we only supply a single parameter we instead get a univariate histogram, which also accepts the usual histogram arguments:
+However, if we only supply a single parameter we instead get
+a univariate histogram, which also accepts the usual
+histogram arguments:
 
@@
 
-These gates are all described in more or less the same manner:
+These gates are all described in more or less the same manner (see man pages for more details):
 

### 2007-05-30 R/rectangleGate-accessors.R

@@ -3,12 +3,17 @@
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("%in%",signature(x="flowFrame",table="rectangleGate"),
           function(x,table) {
+             
             e <- if(length(table@parameters)==1) as.matrix(exprs(x)[,table@parameters]) else
             exprs(x)[,table@parameters]
             apply(sapply(seq(along=table@parameters), function(i) {
+                if(table@min[i] >= table@max[i]){
+                    print(i)
+                  e[,i]
+                } else {
               !is.na(cut(e[,i],c(table@min[i],table@max[i]),labels=FALSE,
-                         right=FALSE))
-            }),1,all)
+                         right=FALSE))}
+            }), 1, all)
           })
 
 

### 2007-08-19 R/AllClasses.R
remove parenId from class definition of gates
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@26510 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -73,7 +73,7 @@ setClass("flowSet",
 ## ---------------------------------------------------------------------------
 setClass("filter", 
          representation("VIRTUAL", filterId="character",
-                        parentId="character",
+                        ##parentId="character",
                         parameters="ANY"),
          validity=function(object){
              msg <- TRUE
@@ -95,13 +95,13 @@ setClass("rectangleGate",
            min=0,max=Inf)
          )
 
-rectangleGate <- function(filterId="rectangleGate", .gate,parentId="",...) {
+rectangleGate <- function(filterId="rectangleGate", .gate,...) {
     if(missing(.gate) || !is.matrix(.gate))
       	.gate <- sapply(if(missing(.gate)) list(...) else .gate,function(x) {
 			x = sort(x);c("min"=x[1],"max"=x[2])
 		})
 	new("rectangleGate", filterId=filterId, parameters=colnames(.gate),
-            min=.gate[1,], max=.gate[2,],parentId=parentId)
+            min=.gate[1,], max=.gate[2,])
 }
 
 
@@ -119,11 +119,11 @@ setClass("polygonGate",
              return(msg)
          })
 
-polygonGate <- function(filterId="polygonGate", boundaries, parentId="",...) {
+polygonGate <- function(filterId="polygonGate", boundaries,...) {
 	if(missing(boundaries) || !is.matrix(boundaries)) 
 		boundaries = as.matrix(if(missing(boundaries)) do.call("cbind",list(...)) else boundaries)
     new("polygonGate",filterId=filterId, parameters=colnames(boundaries),
-        boundaries=boundaries,parentId=parentId)
+        boundaries=boundaries)
 }
 
 
@@ -142,13 +142,13 @@ setClass("polytopeGate",
              return(msg)
          })
 
-polytopeGate <- function(filterId="polytopeGate", .gate,parentId="", ...) {
+polytopeGate <- function(filterId="polytopeGate", .gate,...) {
     if(missing(.gate) || !is.matrix(.gate))
       ##nrowGate <- max(unlist(lapply(list(...),length)))
       .gate <- sapply(if(missing(.gate)) list(...) else .gate, function(x) x)
          
     new("polytopeGate", filterId=filterId, parameters=colnames(.gate),
-        boundaries=.gate,parentId=parentId)
+        boundaries=.gate)
 }
 
 
@@ -171,12 +171,12 @@ setClass("ellipsoidGate",
              return(msg)
          })
 
[…]

### 2007-08-19 inst/doc/HowTo-flowCore.Rnw
remove parenId from class definition of gates
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@26510 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 ``FCS'' file exported from the flow cytometer's acquisition
-software. At the moment we support FCS file versions 1.0
+software. At the moment we support FCS file versions 2.0
 through 3.0, and we expect to support FCS4/ACS1 as soon as
@@
 the \Rfunction{transformation} parameter, which can either
-``linearize'' (the default) or ``standardize'' our data. To
+``linearize'' (the default) or ``scale'' our data. To
 see how this works, first we will examine an FCS file
@@
 parameters that match the regular expression will be
-included in the frame. For example, to include on the the
-Height parameters:
+included in the frame. For example, to include on the Height parameters:
 
@@
 On the other hand using he \Rfunction{[} extraction operator returns a new \Rclass{flowSet} by \textbf{copying} the environment.
-However, simply assigning the \Rclass{flowFrame} to a new variable will \textbf{not} copying the contained frames.
+However, simply assigning the \Rclass{flowFrame} to a new variable will \textbf{not} copy the contained frames.
 

### 2007-10-01 R/flowSet-accessors.R
updated vignette to include filterSet examples. removed extraneous print statement and fixed two method signatures for filterSet operations.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@27678 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -297,6 +297,7 @@ setMethod("filter",signature=signature(x="flowSet",filter="filter"),
           function(x,filter) {
             fsApply(x,function(x) filter(x,filter))
 })
+setMethod("filter",signature(x="flowSet",filter="filterSet"),function(x,filter) fsApply(x,function(x) filter(x,filter)))
 
 setMethod("filter",signature(x="flowSet",filter="list"),function(x,filter) {
 	if(is.null(names(filter)))

### 2007-10-01 inst/doc/HowTo-flowCore.Rnw
updated vignette to include filterSet examples. removed extraneous print statement and fixed two method signatures for filterSet operations.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@27678 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
+\section{filterSet: Organizing Filtering Strategies}
+
+\subsection{Building Filter Sets}
+
+In addition to allowing for sequential filtering, \Rpackage{flowCore} also provides a \Rclass{filterSet} object that serves as an analogue to \Rclass{flowSets} for filters.
+The primary use of this object is to organize and manipulate complex gating strategies.
+For example, recall from the filtering section we had two gates, a ``morphologyGate'' and a ``Fluorescence Region'' gate that we used to demonstrate the various logical operators available for gates.
+To recreate this example as a \Rclass{filterSet}, we would do the following:
+
+<<FilterSet1, echo=true, results=verbatim>>=
+fset1 = filterSet(
+	rectangleGate(filterId="Fluorescence Region","FL1-H"=c(50,100),"FL2-H"=c(50,100)),
+	norm2Filter(filterId="Morphology Gate","FSC-H","SSC-H",scale=2),
+	~ `Fluorescence Region` & `Morphology Gate`,
+	~ `Fluorescence Region` | `Morphology Gate`,
+	Debris ~ ! `Morphology Gate`,
+	~ `Fluorescence Region` %&% `Morphology Gate`
+)
+fset1
+
+@
+
+There are two features of note in \Rfunction{filterSet}, which can also take a single list argument for programmatic creation of \Rclass{filterSet} objects.
+The first is that there is a formula interface for the creation of \Rclass{filter} objects from operators.
+The formula interface can be used with or without a left-hand side, which specifies an optional filter identifier.
+If the filter identifier is not specified, as is the case with all but the ``Debris'' gate in the example above the usual default filter identifier is constructed.
+Non-formula gates can also have an optional name specified which overrides the filter identifier specified in the gate constructor.
+This is discouraged at this time as it leads to confusion when creating new filters.
+
+\subsection{Manipulating Filter Sets}
+
+Manipulating a \Rclass{filterSet} is done using the normal list element replacement methods, though we do provide a special case where replacing the ``""'' element or the ``NULL'' element causes the \Rclass{filterSet} to use the ``filterId'' slot of the incoming filter to choose a name.
+Similarly, specifying an identifier will override any ``filterId'' slot in the original filter.
+
+Additionally, there are several convenience functions and methods available for manipulating a \Rclass{filterSet}.
+The \Rfunction{names} function provides a list of filter identifiers in an arbitrary order.
+The \Rfunction{sort} lists the filter identifiers according to a topological sort of the filter dependencies such that parent filters are always evaluated before their children.
+Filters without dependencies are provided in an arbitrary relative ordering.
+To obtain the adjacency matrix for this sorting, the option ``dependencies=TRUE'' supplied to the \Rfunction{sort} function will attach an ``AdjM'' attribute to the resulting character vector.
+
+\subsection{Using Filter Sets}
+
+Though they are not explicitly filters, meaning that they cannot play a role in the construction of subfilters via operators, \Rclass{filterSet} objects do define \Rfunction{filter} and \Rfunction{split} methods that allow for the two most common use cases.
+Additionally, a filter can be selected from the \Rclass{filterSet} using ``[['' to perform a \Rfunction{Subset} operation if desired.
+
+For standard filtering operations, the \Rclass{filterSet} can yield better performance than a manual filtering operation because it ensures that each parent filter is only evaluated once per call to \Rfunction{filter} so that if many subfilters rely on a particular, likely expensive, filtering operation they use the original result rather than recalculating each time.
+The individual \Rclass{filterResult} objects may be later extracted for further analysis of desire, making the \Rclass{filterSet} the logical choice for most filtering operations.
+Additionally, though not presently implemented, it can allow visualization methods to make more intelligent choices about display when rendering gating strategies because the disposition of the entire filtering graph is known and the filter results are already available.
+
+Like any other filtering operation, \Rfunction{summary} methods are also available to collect the usual sorts of count statistics for a filtering operation:
+
+<<FilterSet2, echo=TRUE, verbatim=TRUE>>=
+
+f = filter(fs,fset1)
+f
+@
+
+When splitting a \Rclass{flowFrame} using \Rfunction{split}, we have two options. The first is to include all of the resultant subsets in the usual way:
+
+<<FilterSet3, echo=TRUE, verbatim=TRUE>>=
+
+split(fs[[1]],fset1,flowSet=TRUE)
+
+@
+
+While this provides access to all subsets, for automated analysis we will often only wish to work with the ``leaf'' gates (i.e.
+those without children of their own).
+In this case, we can specify ``drop=TRUE'' to obtain only the leaf gates:
+
+<<FilterSet4, echo=TRUE, verbatim=TRUE>>=
+
+split(fs[[1]],fset1,drop=TRUE,flowSet=TRUE)
+
+@
+
+Note that in both cases, we use the ``flowSet=TRUE'' option that has been added to the \Rfunction{split} function to automatically create a \Rclass{flowSet} object from the resulting \Rclass{flowFrame} objects.
+For \Rfunction{filterSet} objects, the \Rfunction{split} function can take advantage of the added structure in this case to provide some information about the filtering operation itself in the \Rclass{flowSet}'s metadata as well.
+
 \clearpage

### 2007-10-01 R/coerce.R

@@ -44,7 +44,7 @@ setAs("complementFilter","logical",function(from)
 ## Allows for the resolution of filterReferences and formulas
 setAs("filterReference","concreteFilter",function(from) {
 	x = from@env[[from@name]]
-	if(is.null(x)) stop("Unable to resolve filter reference")
+	if(is.null(x)) stop(paste("Unable to resolve filter reference:",from@name))
 	x
 })
 setAs("formula","filter",function(from) {
@@ -108,9 +108,10 @@ setAs("list","filterSet",function(from) {
 	fs = filterSet()
 	e  = new.env()
 	n    = names(from)
+	if(is.null(n) || length(n)==0) n = rep("",length(from))
 	for(i in 1:length(from)) {
 		filter = from[[i]]
-		name   = if(nzchar(n[i])) n[i] else NULL
+		name   = if(!is.null(n[i]) && nzchar(n[i])) n[i] else NULL
 		fs[[name]] = eval(filter,e)
 	}
 	fs

### 2007-12-25 paper/outline.tex
added paper outline
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@29311 bc3139a8-67e5-0310-9ffc-ced21a209358

\documentclass[12pt]{article}
\usepackage{url}

\title{Outline flowCore paper}

\author{Florian Hahne*\\
  Nolwenn LeMeur*\\
  Byron Ellis\\
  Ryan Brinkman\\
  Perry Haaland\\
  Errol Strain\\
  Deepayan Sarkar\\
  Joseph Spidlen\\
  Robert Gentleman
 }

\begin{document}
\maketitle

\vspace{2ex} The paper will deal with the capabilities offered by
flowCore for handling flow cytometry data in R. The main focus should
be on high-throughput and how our infrastructure can be used to deal
with such data. We don't want to write a user manual but rather
highlight the problems that arise from high throughput flow
applications and the solutions we offer. A strong point should be on
integration of our infrastructure in the flow world. flowCore is
intended to be a development platform for other scientists, we don't
want to provide full-fledged solutions but rather tools for others to
devise such solutions.

It would be nice to show these things on one data set in the form of a
use case, but that is open to discussion. The problem with such an
approach is that it often looks like a software manual which is not
what should be aiming for.

The author list is also still up for discussion, listed here are all
the names that came up during the last telephone conference. NLM and
FH are supposed to share first authorship.

Target journal should be Cytometry, and there are two flavors:
Cytometry A deals with more general topics while Cytometry B is more
geared towards clinical applications. Florian suggests Cytometry A,
here is a link to the author instructions:
\url{http://www3.interscience.wiley.com/journal/33945/home/ForAuthors.html}.
They require the standard Materials and Methods, Results and
Discussion structure which might not be suited for our needs. FH
hasn't found out about alternative methods-type articles yet.  We can
suggest to the editor new format for software/methods oriented papaer?
(NLM)

Also it is not clear from their web page whether they support LaTeX
manuscript submissions but FH would strongly argue for using
LaTeX. They do accept pdf submission.

There is an svn repository set up at
\url{https://hedgehog.fhcrc.org/gentleman/bioconductor/trunk/madman/Rpacks/flowCore/paper}
which is also the repository for the flowCore source code. All
contributors should already have usernames and passwords, if not
please contact FH.

FH put the following very rough structure up for discussion:

\section{Introduction}

\subsection{High throughput flow}
A brief introduction to the flow technology and high throughput flow
in particular.

\subsection{R and Bioconductor}
A couple of sentences what Bioconductor and R is all
about.

\subsection{Existing data standards and conventions}
How does the flow data world look like
FCS
Other flow software (what other software are handling multiple FCS files all together)
(NLM: should we mention prada and rflowcyt?) FH say: Maybe not, I
think it is time to move away from history and establish flowCore as
the one software for flow in Bioconductor.

\section{Basic data structures}

\subsection{demands on flow data structures}
numeric raw data matrices
instrument-specific meta data
experiment meta data
memory issues, scale of data (tubes, plates)

\subsection{flowFrame}
to represent individual FCS files

\subsection{flowSet} 
experiments, multiple FCS files, plates

\section{Standard flow operations}
\subsection{transformations}
Different flavors, R's own transformation infrastructure allows for
arbitrary transforms

\subsection{gating}
gate definitions
capturing gating schemes, workflow
gate summary statistics

\section{Application, Outlook}
Challenges in high throughput cytometry
(QA, automated gating, class discovery)

flowCore as toolbox

already in devel 'plug-in' packages (Visualization, QA, stats, ...)

\end{document}

### 2007-12-25 R/flowFrame-accessors.R

@@ -9,7 +9,9 @@ setMethod("identifier", signature="flowFrame",
             if(is.null(oid) || is.na(oid))
                oid <- as.vector(object@description[["$FIL"]])
             if(is.null(oid) || is.na(oid))
-                as.vector(object@description[["FILENAME"]])
+                oid <- as.vector(object@description[["FILENAME"]])
+            if(is.null(oid) || is.na(oid))
+                "anonymous"
             else
               as.vector(oid)
         })

### 2007-12-28 paper/outline.tex
fxed typos
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@29340 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
   Deepayan Sarkar\\
-  Joseph Spidlen\\
+  Josef Spidlen\\
   Robert Gentleman

### 2008-01-09 paper/outline.tex
included discussion from emails
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@29459 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\vspace{2ex} The paper will deal with the capabilities offered by
-flowCore for handling flow cytometry data in R. The main focus should
-be on high-throughput and how our infrastructure can be used to deal
-with such data. We don't want to write a user manual but rather
-highlight the problems that arise from high throughput flow
-applications and the solutions we offer. A strong point should be on
-integration of our infrastructure in the flow world. flowCore is
-intended to be a development platform for other scientists, we don't
-want to provide full-fledged solutions but rather tools for others to
-devise such solutions.
+\vspace{2ex}
+
+The paper will deal with the capabilities offered by flowCore for
+handling flow cytometry data in R. The main focus should be on
+high-throughput and how our infrastructure can be used to deal with
+such data. We don't want to write a user manual but rather highlight
+the problems that arise from high throughput flow applications and the
+solutions we offer. A strong point should be on integration of our
+infrastructure in the flow world. flowCore is intended to be a
+development platform for other scientists, we don't want to provide
+full-fledged solutions but rather tools for others to devise such
+solutions.
 
@@
 approach is that it often looks like a software manual which is not
-what should be aiming for.
+what we should be aiming for.
 
+\bigskip
 The author list is also still up for discussion, listed here are all
@@
 
+\bigskip
 Target journal should be Cytometry, and there are two flavors:
@@
 Discussion structure which might not be suited for our needs. FH
-hasn't found out about alternative methods-type articles yet.  We can
-suggest to the editor new format for software/methods oriented papaer?
-(NLM)
+hasn't found out about alternative methods-type articles yet.
+
+{\bf NLM:} We can suggest to the editor new format for software/methods
+oriented paper?
+
+{\bf RB:} I know the editor and would be happy to speak to him
+regarding methods-type articles or other issues if needed.
 
@@
 
+{\bf RB:} Cytometry does not accept LaTeX (we've tried). While they say
+they will accept PDF they made us convert PDF to Word in the past. I
+was able to do this reasonably well with a third-party software tool.
+
+\bigskip
 There is an svn repository set up at
@@
 
+\bigskip
+There have been discussions on how much technical detail we want to
+provide and how and where example should be given:
+
+{\bf PH:} Do we have a simple illustration of why we need to be able to
+program with flow data in order to do more meaningful science? Could
+we say something about finding populations or automated gating as
+being enabled by flowCore?  We have found the lattice plots and the
+ability to make overlays very helpful. Could we include an example
+where these capabilities would be helpful?
+
+{\bf RB:} Enabling automated gating is certainly true as we have just had a
+manuscript accepted for automated gating that relies on flowCore.
+
+{\bf PH:} It seems to me some examples of what flowCore makes possible needs
+to be at the heart of the paper. That doesn't mean that we need to
+show a detailed analysis.
+
+{\bf ES:} I also think that use cases/examples are very important for selling
+the software.  I don't believe we need a complete analysis vignette in
+the paper, but specific examples that I've found useful include
+\begin{itemize}
+\item Correcting for the effects of cell size on autofluoresence using
+  Florian's regression,
+\item Automating report generation (using Sweave), including making
+  overlay plots (supposed hard to batch this job in FlowJo?),
+\item automated gating,
+\item population finding (using mclust, other R density tools, etc).
+\end{itemize}
+
+{\bf RB:} The code examples in the supplementary material, but could
+be highlighted in the main text with the pretty results (the selling
+point to biologists), something like: “For example, with a short
+section of user-configurable code (see Supplementary Material) it is
+possible to automate the processing of 1000s of individual FCS data
+files and generate high quality reports (Figures 1, 2) including such
+steps as quality checking plots, correcting for cell size on auto
+fluorescence, followed by automated gating.  This code can be freely
+extended and modified by users as the underlying code is all available
+at no cost under an open source license ”.  I think it’s important
+to remember the audience. The people reading Cytometry we need to
+convince on the importance of flowCore are the biologists, and putting
+code examples in the main text won’t help, and will likely
+hinder. The programmers/statisticians/bioinformaticians out there will
+go to the code in the Supplementary Material/Vignettes section
+already. All the biologists need to see are the pretty pictures that
+let them interpret their data. Highlighting the support network
+available (Bioconductor list) will be important in terms of hand
+holding.  In terms of making biologists happy, it might be worth
+talking to Adam about plans to call R from FlowJo. At the least it
+should be highlighted that this is possible in the generic sense.
+
+{\bf PH:} agree that the code examples should go in the supplemental
+materials, and I like the examples that Ryan gives below. One thing
+that we have found to be very challenging to the biologists is that
+they no longer do the primary analysis of their data. We have
+addressed that by highlighting the ability to programmatically analyze
+lots of data, most of which has fairly straightforward results and
+then to focus the biologists attention only on the data that needs
+expert interpretation/tweaking.
+
+{\bf BE:} This is, unfortunately, somewhat of a problem---the biologist (as
+a community, I realize that there are specific individuals to whom
+this does not apply and that I'm likely especially sensitive to this
+phenomenon) doesn't believe that data can be analyzed apart from
+direct inspection. It might be worth stressing that flow* allows for
+not only automated analysis, but provides a platform for improving
+collaboration. Actually, it occurs to me that this is probably "the
+important bit" for the data structures, which are not particularly
+profound in and of themselves---they mostly take the form they do
+because that's the way Bioconductor does things in general. However,
+what they buy you is a standard mechanism for the organization of
+annotation of experiments that is suitable for programmatic
+manipulation rather than the usual ad hoc collection of 30 randomly
+formatted Excel spreadsheets that may or may not have anything to do
+with the actual flow data you have.
+
+{\bf PH:} I agree with that. We need to explain to the biologists that they
+"can't get there from here" unless they adopt better data organization
+and automated analysis and work collaboratively. It would be nice to
+think about how we would make that argument.
+
+\bigskip
 FH put the following very rough structure up for discussion:
@@
 
+{\bf RB:} Given the proposed journal, an introduction to flow technology isn't
+required.
+
 \subsection{R and Bioconductor}
@@
 \subsection{Existing data standards and conventions}
-How does the flow data world look like
-FCS
-Other flow software (what other software are handling multiple FCS files all together)
-(NLM: should we mention prada and rflowcyt?) FH say: Maybe not, I
-think it is time to move away from history and establish flowCore as
-the one software for flow in Bioconductor.
+How does the flow data world look like FCS Other flow software (what
+other software are handling multiple FCS files all together )
+
+{\bf NLM:} should we mention prada and rflowcyt?
+
+{\bf FH:} Maybe not, I think it is time to move away from history and
+establish flowCore as the one software for flow in Bioconductor.
+
+{\bf RB:} Since rflowcyt laid the groundwork for this work we might consider
+inviting Tony Rossini to contribute to the manuscript, in which case we
+wouldn't need to mention rflowcyt. It would be important to highlight
+the differences between flowCore and commercial software packages,
+including in philosophy (open vs. closed source) and interaction
+(programmatic data analysis vs. GUI-driven).
+
+{\bf PH:} I think the Frame/Set concept is pretty interesting and
+powerful. Was it originated for this project? The ability to program
+sets flowCore aside from all other flow software applications, I
+think. I think there is something really important and valuable about
+how the data structures are ogranized. I'm not sure how to write that
+in a way to appeal to Cytometry A readers, but I think we could work
+this out. Second, I think we have to consider what would be the
+motiviation for 
+\begin{itemize}
+\item why the journal would publish the paper 
+\item why someone would read the paper, and
+\item why someone would want to try the software
+\end{itemize}
 
 \section{Basic data structures}
+
+{\bf BE:} For this section I think we'd want to describe the philosophy that
+underlies the data structure rather than a technical description of
+the data structure since they weren't chosen arbitrarily. Like any
+other flow tool we start with a collection of events, a flowFrame.
+However, we would like to do tasks consistently across several frames
+so we introduce a flowSet (a set because there is generally no
+inherent ordering) that manages the consistent application of
+operations and so on.
 
@@
 \section{Standard flow operations}
+{\bf BE:} We would also like to abstractly describe these operations. This is
+somewhat more difficult, especially since we'd like other packages to
+be able to extend these operations and introduce their own as
+first-class citizens in the analysis. Additionally, we would like to
+factor out as much bookkeeping as possible to allow authors to focus
+on the actual operation rather than interfacing with flowSets versus
+flowFrames and so on. We can then describe the general classes of
+operations and how they interact.
+
 \subsection{transformations}

### 2008-01-16 paper/outline.tex
I start to add text to the introduction and different section.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@29548 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\subsection{High throughput flow}
+%%\subsection{High throughput flow}
 A brief introduction to the flow technology and high throughput flow
 in particular.
-
 {\bf RB:} Given the proposed journal, an introduction to flow technology isn't
@@
 
+\textit{introduction or abstract (first draft)} \\
+
+Traditionally, flow cytometry has been a tube-based technique limited
+to small-scale laboratory and clinical studies.  High throughput
+methods for flow cytometry have recently been developed and are now
+used in both basic and clinical research. Today, the need for flexible
+and well structured informatic tools to efficiently handle and analyze
+FC-HCS has considerably increased. Indeed, the amount of information
+generated by these technology must be stored, managed, and needs to be
+summarized in order to be accessible to the researcher. However, the
+absence of any research platform which bioinformaticians, computer
+scientists, and statisticians can use to develop standard or novel
+methods for flow cytometric has become problematic. We believe that
+the Open Source statistical software R in conjuction with the
+Bioconductor Project can fill this gap.  In this paper, we propose R
+data structures to handle flow cytometry data through the main steps
+of importing, storing, assessing and preprocessing data from FCM
+experiments. These methods, along with a number of specialized
+functions for compensation, transformation and filtering preprocessing
+steps, are implemented in the new developped Bioconductor package
+flowCore.  The aim is that researchers will find flowCore to be a
+solid foundation for efficiently handling FC-HCS data and for future
+development of new methods and tools for the analysis of flow
+cytometry.
+
+\textit{end --- introduction (first draft)}
+
 \subsection{R and Bioconductor}
-A couple of sentences what Bioconductor and R is all
-about.
+
+R is a solid statistical framework/environment, particularly desirable
+when analyzing the large numeric datasets arising from High throughput
+experiments such as FC-HCS. Indeed, this open source software already
+offer lots of statistical and visualization methods developpement for
+various field of application but potentially interesting for FC-HCS
+data analysis. R is also a research platform which bioinformaticians,
+computer scientists, and statisticians can use to develop new methods
+for FC-HCS. In biolgy the Bioconductor project....
 
@@
 
+\textit{example Text}
+
+\Rpackage{flowCore}'s primary task is the representation and basic
+manipulation of flow cytometry (or similar) data. This is accomplished
+through a data model very similar to that adopted by other
+Bioconductor packages used for gene expression analysis and thus
+familiar to most Bioconductor users \cite{}.  The data structure is
+designed to combine several differents sources of information into a
+single convenient structure: (\textit{e.g.,} measurements from a
+combination of samples, metadata describing samples in the
+experiments).
+
+\textit{end Text}
+
 {\bf BE:} For this section I think we'd want to describe the philosophy that
@@
 \subsection{demands on flow data structures}
+
 numeric raw data matrices
@@
 \subsection{flowFrame}
+
+\textit{example Text}
+
+The basic unit of manipulation in \Rpackage{flowCore} is the
+\Rclass{flowFrame}, which corresponds roughly with a single ``FCS''
+file exported from the flow cytometer's acquisition software. The
+\Robject{flowFrame} object store the event-level information and the
+parameters, which contain metadata. The event information, stored as a
+single matrix, is easily accessed and manipulated (see
+\textit{Supplementary Materials}). The parameters slot contains
+information derived from an FCS file's ``\$P<n>'' keywords, which
+describe the detector and stain information.
+
+\textit{end Text}
+
+{\bf NLM:} Should be more abstract ?
+
 to represent individual FCS files
@@
 Different flavors, R's own transformation infrastructure allows for
-arbitrary transforms
+arbitrary transforms, we have implemented the commonly used one.
 

### 2008-01-16 paper/outline.tex
some updates of the paper manuscript
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@29559 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\textit{introduction or abstract (first draft)} \\
+\textit{introduction or abstract (first draft)}
 
@@
 used in both basic and clinical research. Today, the need for flexible
-and well structured informatic tools to efficiently handle and analyze
-FC-HCS has considerably increased. Indeed, the amount of information
-generated by these technology must be stored, managed, and needs to be
-summarized in order to be accessible to the researcher. However, the
-absence of any research platform which bioinformaticians, computer
-scientists, and statisticians can use to develop standard or novel
-methods for flow cytometric has become problematic. We believe that
-the Open Source statistical software R in conjuction with the
-Bioconductor Project can fill this gap.  In this paper, we propose R
-data structures to handle flow cytometry data through the main steps
-of importing, storing, assessing and preprocessing data from FCM
-experiments. These methods, along with a number of specialized
-functions for compensation, transformation and filtering preprocessing
-steps, are implemented in the new developped Bioconductor package
-flowCore.  The aim is that researchers will find flowCore to be a
-solid foundation for efficiently handling FC-HCS data and for future
-development of new methods and tools for the analysis of flow
-cytometry.
+and well structured computatioonal tools to efficiently handle and
+analyze FC-HCS has considerably increased. Indeed, the amount of
+information generated by these technology must be stored, managed, and
+needs to be summarized in order to be accessible to the
+researcher. However, the absence of any shared research platform which
+bioinformaticians, computer scientists, and statisticians can use to
+develop standard or novel methods for flow cytometric has become
+problematic. We believe that the Open Source statistical software R in
+conjuction with the Bioconductor Project can fill this gap.  In this
+paper, we propose R data structures to handle flow cytometry data
+through the main steps of importing, storing, assessing and
+preprocessing data from FCM experiments. These methods, along with a
+number of specialized functions for compensation, transformation and
+filtering and other preprocessing steps, are implemented in the newly
+developped Bioconductor package flowCore.  We envision flowCore to be
+a solid foundation for researchers to efficiently handle FC-HCS data
+and for future development of new methods and tools for the analysis
+of flow cytometry.
 
@@
 
-R is a solid statistical framework/environment, particularly desirable
-when analyzing the large numeric datasets arising from High throughput
-experiments such as FC-HCS. Indeed, this open source software already
-offer lots of statistical and visualization methods developpement for
-various field of application but potentially interesting for FC-HCS
-data analysis. R is also a research platform which bioinformaticians,
-computer scientists, and statisticians can use to develop new methods
-for FC-HCS. In biolgy the Bioconductor project....
+\textit{ example text}
+
+R is a solid statistical framework/environment, particularly geared
+towards the analysis the large numeric datasets like those arising
+from high throughput experiments such as FC-HCS. Indeed, this open
+source software already offers a wide range of statistical and
+visualization methods developed for various field of application, many
+of them relavant also to FC-HCS data analysis. In addition, R is a
+research platform which bioinformaticians, computer scientists, and
+statisticians are using to develop new tools for statistical data
+analyis. For biological application, the Bioconductor project has
+become the standard toolset...
+
+\textit{end --- example text}
 
 \subsection{Existing data standards and conventions}
+
+\textit{ example text}
+
 How does the flow data world look like FCS Other flow software (what
-other software are handling multiple FCS files all together )
+other software are handling multiple FCS files all together
+). Somebody more familiar with flowJo and friends should add to
+this...
+
+FCS -> we can read these files
+
+\textit{end --- example text}
 
@@
 
+What are the parts we need to keep together
+
+flowCore does the maintenance, the user doesn't have to care about
+that
+
+The concept of sets
+
+Figure: A diagram of the data structures
+
 \textit{end Text}
@@
 file exported from the flow cytometer's acquisition software. The
-\Robject{flowFrame} object store the event-level information and the
-parameters, which contain metadata. The event information, stored as a
+\Robject{flowFrame} object store the event-level information along
+with a number of additional metadata. The event information, stored as a
 single matrix, is easily accessed and manipulated (see
-\textit{Supplementary Materials}). The parameters slot contains
-information derived from an FCS file's ``\$P<n>'' keywords, which
-describe the detector and stain information.
+\textit{Supplementary Materials}). All available information regarding
+detectors and stains is drawn from the FCS files and attached to the frame.
 
@@
 {\bf NLM:} Should be more abstract ?
-
-to represent individual FCS files
 

### 2008-02-04 paper/outline.tex
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@29821 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \usepackage{url}
+\usepackage[authoryear,round]{natbib}
 
@@
 visualization methods developed for various field of application, many
-of them relavant also to FC-HCS data analysis. In addition, R is a
-research platform which bioinformaticians, computer scientists, and
-statisticians are using to develop new tools for statistical data
-analyis. For biological application, the Bioconductor project has
-become the standard toolset...
+of them relevant also to FC-HCS data analysis. 
+%% FIXME: either cite some of our paper or drop this sentence. 
+%% I think it it too redundant with the one above it.
+In addition, R is a research platform which
+bioinformaticians, computer scientists, and statisticians are using to
+develop new tools for statistical data analyis. For biological
+application, the Bioconductor project has become the standard toolset
+\citep{gentleman2006bos}...
 
-\textit{end --- example text}
+NLM: reference to microarray ?
+\textit{end--- example text}
 
@@
 
-{\bf NLM:} should we mention prada and rflowcyt?
-
-{\bf FH:} Maybe not, I think it is time to move away from history and
-establish flowCore as the one software for flow in Bioconductor.
-
-{\bf RB:} Since rflowcyt laid the groundwork for this work we might consider
-inviting Tony Rossini to contribute to the manuscript, in which case we
-wouldn't need to mention rflowcyt. It would be important to highlight
+{\bf RB:}  It would be important to highlight
 the differences between flowCore and commercial software packages,
@@
 
+\bibliographystyle{plainnat}  
+\bibliography{flowCoreRef} 
 \end{document}

### 2008-02-20 paper/outline.tex
updated paper draft
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@30063 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \begin{document}
-\maketitle
 
@@
 \bigskip
-FH put the following very rough structure up for discussion:
+
+\maketitle
 
@@
 
-%%\subsection{High throughput flow}
-A brief introduction to the flow technology and high throughput flow
-in particular.
-{\bf RB:} Given the proposed journal, an introduction to flow technology isn't
-required.
-
-\textit{introduction or abstract (first draft)}
-
-Traditionally, flow cytometry has been a tube-based technique limited
-to small-scale laboratory and clinical studies.  High throughput
-methods for flow cytometry have recently been developed and are now
-used in both basic and clinical research. Today, the need for flexible
-and well structured computatioonal tools to efficiently handle and
-analyze FC-HCS has considerably increased. Indeed, the amount of
-information generated by these technology must be stored, managed, and
-needs to be summarized in order to be accessible to the
-researcher. However, the absence of any shared research platform which
-bioinformaticians, computer scientists, and statisticians can use to
-develop standard or novel methods for flow cytometric has become
-problematic. We believe that the Open Source statistical software R in
-conjuction with the Bioconductor Project can fill this gap.  In this
-paper, we propose R data structures to handle flow cytometry data
-through the main steps of importing, storing, assessing and
-preprocessing data from FCM experiments. These methods, along with a
-number of specialized functions for compensation, transformation and
-filtering and other preprocessing steps, are implemented in the newly
-developped Bioconductor package flowCore.  We envision flowCore to be
-a solid foundation for researchers to efficiently handle FC-HCS data
-and for future development of new methods and tools for the analysis
-of flow cytometry.
-
-\textit{end --- introduction (first draft)}
+Traditionally, flow cytometry (FC) has been a tube-based technique
+limited to small-scale laboratory and clinical studies.  High
+throughput methods for flow cytometry (FC-HCS) have recently been
+developed and are now used in both basic and clinical research. Today,
+the need for flexible and well structured computational tools to
+efficiently handle and analyze FC-HCS has increased
+considerably. Indeed, the amount of information generated by these
+technologies must be stored, managed, and needs to be summarized in
+order to be accessible to the researcher. However, the absence of any
+shared research platform which bioinformaticians, computer scientists,
+and statisticians can utilze to develop novel or standard methods for
+flow cytometric has become problematic. We believe that the open
+source statistical software R in conjuction with the Bioconductor
+Project can fill this gap.  In this paper, we propose R data
+structures to handle flow cytometry data through the main steps of
+importing, storing, assessing and preprocessing data from FC-HCS
+experiments. These structures, along with a range of specialized
+methods for compensation, transformation, gating and additional
+preprocessing steps, are implemented in the newly developed
+Bioconductor package flowCore.  We envision flowCore to be a solid
+foundation for researchers to efficiently handle FC-HCS data and for
+future development of new methods and tools for their coherent
+analysis.
 
 \subsection{R and Bioconductor}
-
-\textit{ example text}
-
-R is a solid statistical framework/environment, particularly geared
-towards the analysis the large numeric datasets like those arising
-from high throughput experiments such as FC-HCS. Indeed, this open
-source software already offers a wide range of statistical and
-visualization methods developed for various field of application, many
-of them relevant also to FC-HCS data analysis. 
-%% FIXME: either cite some of our paper or drop this sentence. 
-%% I think it it too redundant with the one above it.
-In addition, R is a research platform which
-bioinformaticians, computer scientists, and statisticians are using to
-develop new tools for statistical data analyis. For biological
-application, the Bioconductor project has become the standard toolset
-\citep{gentleman2006bos}...
-
-NLM: reference to microarray ?
-\textit{end--- example text}
+R is a solid statistical programing environment, particularly geared
+towards the analysis of large numeric datasets typical to arise from
+high throughput experiments such as microarray gene expression anlysis
+or FC-HCS. Today, the R open source project offers a wide range of
+statistical and visualization methods developed for various fields of
+application. In addition, R is a research platform which
+bioinformaticians, computer scientists, and statisticians routinely
+use to develop novel tools and methodology for statistical data
+analyis. For biological applications, the Bioconductor project has
+become the standard toolset \citep{gentleman2006bos} to process data
+spanning a diversity of research fields from genomics to proteomics to
+complex cell biology. This shared platform proved to be especially
+beneficial for the development of analyis routines for microaray gene
+expression data, and by now many of these innovations have been
+incoorporated into comercial software products, thus greatly improving
+the quality of gene expression data analysis. Shared data structures
+have been shown to be essential for this kind of collaborative effort.
 
 \subsection{Existing data standards and conventions}
+Currently, data from flow cytometry experiments are stored in single
+files following a well-defined data standard (FCS). This standard
+tries to capture as much of the accompanying meta information as
+possible, most of which is produced by the measurement instrument or
+has to be supplied manually by the operator upon experiment
+setup. However, the recent developments in high-throughput flow
+cytometry are shifting the focus of interest away from single-tube
+based measurements towards large and complex experimental designs with
+dozens of co-variates and influencing factors. Self-contained data
+structures are a prerequesite to allow for coordinated analysis of
+such data sets. Most of the currently available software solutions
+offer only limited support for such structures, or make use of
+containers like XML or binary storage formats that are designed
+specifically for the mostly GUI-driven user interfaces and hence are
+not easily amenable to programatical access. In addition, the
+closed-source nature of most of these products makes it virtually
+impossible to connect and integrated these data.
 
-\textit{ example text}
+Traditionally, the majority of flow cytometry experiments have been
+analysed manually by direct data inspection or very basic comparisons
+of summary statistics. We believe that these approaches do not give
+justice to the highly complex nature of FC data, essentially
+discarding many of the fundamental aspects such as data distribution
+or high-dimensionality. In addition, the subjective character of
+manual analyses are a major obstacle for reproducibility. For FC-HCS
+data, unassisted manual inspection is no longer feasible, and robust
+statistical methods need to be developed to point the investigator to
+the intersting aspects of the data or to potential problems. While the
+expert knowledge of immunologists and researchers remains crucial for
+the understanding of FC data, we believe that tight colaboration
+with other reserach fields such as statistics and computer science can
+greatly improve the relevance of flow cytomtry in today's
+high-throughput biological sciences.
 
-How does the flow data world look like FCS Other flow software (what
-other software are handling multiple FCS files all together
-). Somebody more familiar with flowJo and friends should add to
-this...
+\textit{How does the flow data world look like? Other flow software
+  (what other software are handling multiple FCS files at
+  all). Somebody more familiar with flowJo and friends should add to
+  this... Also somebody should fill in a couple of sentences about the
+  gating-ML etc. stuff here}
 
-FCS -> we can read these files
-
-\textit{end --- example text}
+The flowCore framework presented here offers import functionality for
+both the raw data FCS files along with the complete set of
+file-specific meta information as well as all the information that is
+stored in one of the gating-, compensation- or transformation-ML
+standards. Thus, it is possible to intergrate flowCore in existing
+work flows and to communicate with virtually every other FC tool that
+adheres to these standards.
 
@@
 (programmatic data analysis vs. GUI-driven).
+
+\section{Basic data structures}
+\subsection{flowFrame}
+flowCore's primary task is the representation and basic manipulation
+of flow cytometry (or similar) data. This is accomplished through a
+data model very close to that adopted by other Bioconductor packages
+used for gene expression analysis and thus familiar to most
+Bioconductor users. Similar to any other flow tool, the basic unit of
+organisation in flowCore is a collection of events. We call the
+structure that hold this data a flowFrame. All relevant information
+for the collection are stored along with the matrix of raw data values
+and can be accessed programatically. Most prominently, these are
+descriptors of the stains used in the experiment and the respective
+measurement channels, information about compensation performed at the
+instrument side and any additional keywords the user deems to be
+important to annotate the data. A number of quality checks are
+performed during the creation of a flowFrame to ensure data
+integrety. There is an extended interface that allows for the
+standardized interaction with the flow data, thus enabling the
+development of more specialized tools for specific tasks of the data
+analysis process.
+
+\subsection{flowSet}
+In high-throughput flow cytometry, many of the analysis tasks need to
+be performed consistently across multiple flow measurements, hence we
+introduce the concept of a collection of flowFrames called a
+flowSet. flowSets store the relevant information associated with each
+individual frame such as descriptions of the samples or the treatment
+to which a sample was subjected, and they manage the consistent
+application of operations on the individual flowFrames. The burden of
+keeping score of the annotation is shifted from the user to the
+infrastructure, thus reducing the risk of mixups and errors that are
+all too common when dealing with complex data in the ususal way of
+more or less ad hoc spreadsheats that lack a controlled structure. In
+general, there is no native ordering to a flowSet, which in a way
+reflects the still predominant application of tube-based experiments
+rather than experiments following a plate-design. However, the flowSet
+structure can be readily extended to also accomodate such spacial
+information \textit{(reference)}
+
+\textit{Figure: A diagram of the data structures and the basic
+  operations (Figure1.ppt)}
 
@@
 in a way to appeal to Cytometry A readers, but I think we could work
-this out. Second, I think we have to consider what would be the
-motiviation for 
-\begin{itemize}
-\item why the journal would publish the paper 
-\item why someone would read the paper, and
-\item why someone would want to try the software
-\end{itemize}
-
-\section{Basic data structures}
-
-\textit{example Text}
-
-\Rpackage{flowCore}'s primary task is the representation and basic
-manipulation of flow cytometry (or similar) data. This is accomplished
-through a data model very similar to that adopted by other
-Bioconductor packages used for gene expression analysis and thus
-familiar to most Bioconductor users \cite{}.  The data structure is
-designed to combine several differents sources of information into a
-single convenient structure: (\textit{e.g.,} measurements from a
-combination of samples, metadata describing samples in the
-experiments).
-
-What are the parts we need to keep together
-
-flowCore does the maintenance, the user doesn't have to care about
-that
-
-The concept of sets
-
-Figure: A diagram of the data structures
-
-\textit{end Text}
-
-{\bf BE:} For this section I think we'd want to describe the philosophy that
-underlies the data structure rather than a technical description of
-the data structure since they weren't chosen arbitrarily. Like any
-other flow tool we start with a collection of events, a flowFrame.
-However, we would like to do tasks consistently across several frames
-so we introduce a flowSet (a set because there is generally no
-inherent ordering) that manages the consistent application of
-operations and so on.
-
-\subsection{demands on flow data structures}
-
-numeric raw data matrices
-instrument-specific meta data
-experiment meta data
-memory issues, scale of data (tubes, plates)
-
-\subsection{flowFrame}
-
-\textit{example Text}
-
-The basic unit of manipulation in \Rpackage{flowCore} is the
-\Rclass{flowFrame}, which corresponds roughly with a single ``FCS''
-file exported from the flow cytometer's acquisition software. The
-\Robject{flowFrame} object store the event-level information along
-with a number of additional metadata. The event information, stored as a
-single matrix, is easily accessed and manipulated (see
-\textit{Supplementary Materials}). All available information regarding
-detectors and stains is drawn from the FCS files and attached to the frame.
-
-\textit{end Text}
-
-{\bf NLM:} Should be more abstract ?
-
-\subsection{flowSet} 
-experiments, multiple FCS files, plates
+this out.
 
 \section{Standard flow operations}
-{\bf BE:} We would also like to abstractly describe these operations. This is
-somewhat more difficult, especially since we'd like other packages to
-be able to extend these operations and introduce their own as
-first-class citizens in the analysis. Additionally, we would like to
-factor out as much bookkeeping as possible to allow authors to focus
-on the actual operation rather than interfacing with flowSets versus
-flowFrames and so on. We can then describe the general classes of
-operations and how they interact.
+The basic operations in flow cytometry analysis are almost always the
+same: the data need to be compensated (if not already done on the
+instrument side), often transformed and sup-population of interest
+need to be selected based on a set of (predominantly sequential)
+gates. All software solutions for flow data analysis offer more or
+less elaborate support for these kinds of operation, most of them in
+an interactive, GUI-driven spirit. The approach we have taken in
+flowCore is to abstractly describe these operations and build a set of
+tools to perform them on both flowFrames as well as on entire
+flowSets. While transformation and to a certain extent also
+compensation are fairly routine operations with only limited potential
+for improvement, being able to implement new methodology for gating of
+flow cytometry data and extending the capabilies of flowCore through
+the usual ways of object oriented programming is a feature, that
+clearly sets our framework apart from all the other flow toolsets
+around. By factoring out as much of the bookkeeping as possible,
+authors can focus on the actual operations rather then having to deal
+with the teadious details of data integration and access. Third party
+methods implemented by others can act on their own as first-class
+citizens in the analysis, without breaking the workflow or the basic
+infrastructure. This design allows for the straight forward extension
+of flowCore's capabilities and already fostered the development of a
+number of valuable improvements. \textit{(cite stuff here)}
 
 \subsection{transformations}
-Different flavors, R's own transformation infrastructure allows for
-arbitrary transforms, we have implemented the commonly used one.
+All the major transformations that are routinely used in flow
+cytometry analysis have been implemented in flowCore. Furthermore, the
+design of the R language makes it easy to define arbitrary functions
+to apply to the data of individual flowFrames or entrire flowSets,
+respectively. Proper transformation have been proven beneficial for
+both data visualization and modelling \textit{cite flowClust paper}.
 
 \subsection{gating}
-gate definitions
-capturing gating schemes, workflow
-gate summary statistics
+In flowCore, gates are represented as individual objects that can be
+extended in an object oriented manner. All basic gate types such as
+rectangular gates, ellipses and polygons are implemented as part of
+the base framework. In addition, we introduce the notion of
+data-driven gates, for which the necessary parameters are computed
+based on the properties of the underlying data, for instance
+distribution or density estimates. The ability to programatically
+access gates is a prerequesite for automated or semi-automated
+gating. By utilizing a unified interface for all different types of
+gates, the user is able to subset data sets as well as create summary
+statistics, for instance of proportions of events falling in single
+gates or combinations of multiple gates. The definition of gates in
+flowCore follows the proposed gating-ML standard, thus any flowCore
+gating strategy can be reproduced by any other software that also
+adheres to the standard.
+
+\section{Related flow packages}
+In addition to the flowCore package that offers basic infrastructure
+we have implemented a range of additional Bioconductor packages that
+are dedicated to more specific tasks of flow cytomtery data
+analysis. The flowViz package provides sophisticated data
+visualization tools that make use of multivariate trellis
+plotting. These functions can be used to quickly generate customized
+plots for extended cytometry data sets for both direct data inspection
+and quality control. The flowQ package offers more advanced quality
+assurance (QA) methodology and a framework to create interactive
+reports of QA results. Most of the flowUtils content deals with data
+import and output from and to flowCore and the flow-cytometry specific
+standard markup languages. Finally, flowStats provides statistical
+methods that are relevant in the context of flow cytometry data
+analysis.
+
+\textit{A diagram of the packages and what they offer}
 
 \section{Application, Outlook}
-Challenges in high throughput cytometry
-(QA, automated gating, class discovery)
-
-flowCore as toolbox
-
-already in devel 'plug-in' packages (Visualization, QA, stats, ...)
+\textit{This is the most important part, we need to show nice examples
+  of use cases, limit ourselfs to the basics and show what we can do
+  that nobody else can do at the moment. Need at lot of input here
+  from everybody. Also for the things that go in here we should
+  provide further details in the Supplementary Materials.}
 

### 2008-03-05 inst/doc/HowTo-flowCore.Rnw
update to documentation and addition of methods
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@30380 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-Much of the more sophisticated visualization of \Rclass{flowFrame} and \Rclass{flowSet} objects, including an interface to the \Rpackage{lattice} graphics system is implemented by the \Rpackage{flowViz} package, also included as part of Bioconductor.
-However, \Rpackage{flowCore} does implement some basic plotting facilities using the standard \Rfunction{plot} function.
-The basic plot provides a simple bivariate density plot of the first two parameters:
+Much of the more sophisticated visualization of \Rclass{flowFrame} and
+\Rclass{flowSet} objects, including an interface to the
+\Rpackage{lattice} graphics system is implemented by the
+\Rpackage{flowViz} package, also included as part of
+Bioconductor. However, \Rpackage{flowCore} does implement some basic
+plotting facilities using the standard \Rfunction{plot} function. The
+basic plot provides a simple pairs plot of all parameters:
 
@@
 
-To control the parameters being plotted we can supply a \Robject{y} value in the form of a character vector:
+To control the parameters being plotted we can supply a \Robject{y}
+value in the form of a character vector. If we choose exactly two
+parameters this will create a bivariate density plot.
 

### 2008-03-05 R/IO.R

@@ -303,8 +303,6 @@ readFCSdata <- function(con, offsets, x, transformation,  which.lines, debug,
     stopifnot(length(dat)%%nrpar==0)
 
     ## apply bitmask for integer data
-    ## FIXME: We should think about doing this in C for
-    ## efficiency reasons
     if(dattype=="integer"){
         usedBits <- unique(log2(range))
         if(usedBits<bitwidth)
@@ -409,10 +407,10 @@ read.flowSet <- function(files=NULL, path=".", pattern=NULL, phenoData,
     }
     
     flowSet = lapply(files, read.FCS, ...)
-    ##Allows us to specify a particular keyword to use as our sampleNames
-    ##rather than requiring the filename be used. This is handy when something
-    ##like SAMPLE ID is a more reasonable choice. Sadly reading the flowSet is
-    ##a lot more insane now.
+    ## Allows us to specify a particular keyword to use as our sampleNames
+    ## rather than requiring the filename be used. This is handy when something
+    ## like SAMPLE ID is a more reasonable choice. Sadly reading the flowSet is
+    ## a lot more insane now.
     if(!missing(name.keyword))
         names(flowSet) <- sapply(flowSet,keyword,name.keyword)
     else

### 2008-04-08 inst/doc/HowTo-flowCore.Rnw
making sure that the samplesare in the same order after splitting. Also added a couple of updates to the documentation
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@30903 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \section{Introduction}
-Traditionally, flow cytometry has been a tube-based technique limited to small-scale laboratory and clinical studies. 
-High throughput methods for flow cytometry have recently been developed for drug discovery and advanced research methods 
-\citep{Gasparetto2004}. As an example, the flow cytometry high content screening (FC-HCS) can process up 
-to a thousand samples daily at a single workstation, and the results have been equivalent or superior to traditional 
-manual multiparameter staining and analysis techniques. \\
+Traditionally, flow cytometry has been a tube-based technique limited
+to small-scale laboratory and clinical studies.  High throughput
+methods for flow cytometry have recently been developed for drug
+discovery and advanced research methods \citep{Gasparetto2004}. As an
+example, the flow cytometry high content screening (FC-HCS) can
+process up to a thousand samples daily at a single workstation, and
+the results have been equivalent or superior to traditional manual
+multiparameter staining and analysis techniques.
 
-The amount of information generated by high throughput technologies such as FC-HCS need to be transformed into 
-executive summaries (which are brief enough) for creative studies by a human researcher \citep{Brazma2001}. 
-Standardization is critical when developing new high throughput technologies and their associated information services 
-\citep{Brazma2001, Chicurel2002, Boguski2003}. Standardization efforts have been made in clinical cell analysis 
-by flow cytometry \citep{Keeney2004}, however data interpretation has not been standardized for even low throughput FCM. 
-It is one of the most difficult and time consuming aspects of the entire analytical process as well as a primary source 
-of variation in clinical tests, and investigators have traditionally relied on intuition rather than standardized statistical inference 
-\citep{Bagwell2004,Braylan2004,Parks1997,Suni2003}. In the development of standards in high throughput FCM, 
-few progress has been done in term of Open Source software. In this article we propose R data structures to handle 
-flow cytometry data through the main steps of preprocessing: compensation, transformation, filtering.\\
+The amount of information generated by high throughput technologies
+such as FC-HCS need to be transformed into executive summaries (which
+are brief enough) for creative studies by a human researcher
+\citep{Brazma2001}.  Standardization is critical when developing new
+high throughput technologies and their associated information services
+\citep{Brazma2001, Chicurel2002, Boguski2003}. Standardization efforts
+have been made in clinical cell analysis by flow cytometry
+\citep{Keeney2004}, however data interpretation has not been
+standardized for even low throughput FCM.  It is one of the most
+difficult and time consuming aspects of the entire analytical process
+as well as a primary source of variation in clinical tests, and
+investigators have traditionally relied on intuition rather than
+standardized statistical inference
+\citep{Bagwell2004,Braylan2004,Parks1997,Suni2003}. In the development
+of standards in high throughput FCM, few progress has been done in
+term of Open Source software. In this article we propose R data
+structures to handle flow cytometry data through the main steps of
+preprocessing: compensation, transformation, filtering.
 
-The aim is to merge both prada  and rflowcyt \citep{LeMeur2006} into one core package wich is compliant 
-with the data exchange standards that are currently developed in the community \citep{Spidlen2006}.
+The aim is to merge both prada and rflowcyt \citep{LeMeur2006} into
+one core package wich is compliant with the data exchange standards
+that are currently developed in the community \citep{Spidlen2006}.
 
-Visualization as well as quality control will than be part of the utility packages that depend on the data structures defined in the flowCore package.
+Visualization as well as quality control will than be part of the
+utility packages that depend on the data structures defined in the
+flowCore package.
 
@@
 
-\Rpackage{flowCore}'s primary task is the representation and basic manipulation of flow cytometry (or similar) data.
-This is accomplished through a data model very similar to that adopted by other Bioconductor packages using the \Robject{expressionSet} and \Robject{AnnotatedDataFrame} structures familiar to most Bioconductor users.
+\Rpackage{flowCore}'s primary task is the representation and basic
+manipulation of flow cytometry (or similar) data. This is accomplished
+through a data model very similar to that adopted by other
+Bioconductor packages using the \Robject{expressionSet} and
+\Robject{AnnotatedDataFrame} structures familiar to most Bioconductor
+users.
 
@@
 
-The default ``linearize'' transformation option will convert these to, effectively, have a ``\$P<n>E'' of ``0,0'':
+The default ``linearize'' transformation option will convert these to,
+effectively, have a ``\$P<n>E'' of ``0,0'':
 
@@
 
-When only a particular subset of parameters is
-desired the \Rfunction{column.pattern} parameter allows for
-the specification of a regular expression and only
-parameters that match the regular expression will be
-included in the frame. For example, to include on the Height parameters:
+When only a particular subset of parameters is desired the
+\Rfunction{column.pattern} parameter allows for the specification of a
+regular expression and only parameters that match the regular
+expression will be included in the frame. For example, to include on
+the Height parameters:
 
@@
 
-\textit{Note that \Rfunction{column.pattern} is applied after \Rfunction{alter.names} if it is used.}
+\textit{Note that \Rfunction{column.pattern} is applied after
+  \Rfunction{alter.names} if it is used.}
 
-Finally, only a sample of lines can be read in case you need a quick overview of a large serie of files. 
+Finally, only a sample of lines can be read in case you need a quick
+overview of a large serie of files.
 <<RedFiles3, echo=true, results=verbatim>>= 
@@
 
-Most experiments consist of several \Rclass{flowFrame} objects, which are organized using a \Rclass{flowSet} object.
-This class provides a mechanism for efficiently hosting the \Rclass{flowFrame} objects with minimal copying, reducing memory requirements, as well as ensuring that experimental metadata stays properly to the appropriate \Rclass{flowFrame} objects.
+Most experiments consist of several \Rclass{flowFrame} objects, which
+are organized using a \Rclass{flowSet} object. This class provides a
+mechanism for efficiently hosting the \Rclass{flowFrame} objects with
+minimal copying, reducing memory requirements, as well as ensuring
+that experimental metadata stays properly to the appropriate
+\Rclass{flowFrame} objects.
 
@@
 
-To facilitate the creation of \Rclass{flowSet} objects from a variety of sources, we provide a means to coerce \Rclass{list} and \Rclass{environment} objects to a \Rclass{flowSet} object using the usual coercion mechanisms.
-For example, if we have a directory containing FCS files we can read in a list of those files and create a \Rclass{flowSet} out of them:
+To facilitate the creation of \Rclass{flowSet} objects from a variety
+of sources, we provide a means to coerce \Rclass{list} and
+\Rclass{environment} objects to a \Rclass{flowSet} object using the
+usual coercion mechanisms. For example, if we have a directory
+containing FCS files we can read in a list of those files and create a
+\Rclass{flowSet} out of them:
 
 <<Frames, echo=true, results=verbatim>>=
-frames <- lapply(dir(system.file("extdata", "compdata", "data", package="flowCore"), full.names=TRUE), read.FCS)
+frames <- lapply(dir(system.file("extdata", "compdata", "data",
+                                 package="flowCore"), full.names=TRUE), 
+                 read.FCS)
 as(frames, "flowSet")
@@
 
-Note that the original list is unnamed and that the resulting sample names are not particularly meaningful.
-If the list is named, the list constructed is much more meaningful.
-One such approach is to employ the \Rfunction{keyword} method for \Rclass{flowFrame} objects to extract the ``SAMPLE ID'' keyword from each frame:
+Note that the original list is unnamed and that the resulting sample
+names are not particularly meaningful. If the list is named, the list
+constructed is much more meaningful. One such approach is to employ
+the \Rfunction{keyword} method for \Rclass{flowFrame} objects to
+extract the ``SAMPLE ID'' keyword from each frame:
 
@@
 
-Like most Bioconductor organizational classes, the \Rclass{flowSet} has an associated \Rclass{AnnotatedDataFrame} that provides metadata not contained within the \Rclass{flowFrame} objects themselves.
-This data frame is accessed and modified via the usual \Rfunction{phenoData} and \Rfunction{phenoData<-} methods.
-You can also generally treat the phenotypic data as a normal data frame to add new descriptive columns.
-For example, we might want to track the original filename of the frames from above in the phenotypic data for easier access:
+Like most Bioconductor organizational classes, the \Rclass{flowSet}
+has an associated \Rclass{AnnotatedDataFrame} that provides metadata
+not contained within the \Rclass{flowFrame} objects themselves. This
+data frame is accessed and modified via the usual
+\Rfunction{phenoData} and \Rfunction{phenoData<-} methods. You can
+also generally treat the phenotypic data as a normal data frame to add
+new descriptive columns. For example, we might want to track the
+original filename of the frames from above in the phenotypic data for
+easier access:
 
@@
 
-Note that we have used the \Rclass{flowSet}-specific iterator, \Rfunction{fsApply}, which acts much like \Rfunction{sapply} or \Rfunction{lapply}.
-Additionally, we should also note that the \Rfunction{phenoData} data frame \textbf{must} have row names that correspond to the original names used to create the \Rclass{flowSet}.
+Note that we have used the \Rclass{flowSet}-specific iterator,
+\Rfunction{fsApply}, which acts much like \Rfunction{sapply} or
+\Rfunction{lapply}. Additionally, we should also note that the
+\Rfunction{phenoData} data frame \textbf{must} have row names that
+correspond to the original names used to create the \Rclass{flowSet}.
 
@@
 
-Much of the functionality described above has been packaged into the \Rfunction{read.flowSet} convenience function.
-In it's simplest incarnation, this function takes a \Rfunction{path}, that defaults to the current working directory, and an optional \Rfunction{pattern} argument that allows only a subset of files contained within the working directory to be selected.
-For example, to read a \Rclass{flowSet} of the files read in by \Robject{frame} above:
+Much of the functionality described above has been packaged into the
+\Rfunction{read.flowSet} convenience function. In it's simplest
+incarnation, this function takes a \Rfunction{path}, that defaults to
+the current working directory, and an optional \Rfunction{pattern}
+argument that allows only a subset of files contained within the
+working directory to be selected. For example, to read a
+\Rclass{flowSet} of the files read in by \Robject{frame} above:
 
 <<ReadFlowSet, echo=true,results=verbatim>>=
-read.flowSet(path = system.file("extdata", "compdata", "data", package="flowCore"))
+read.flowSet(path = system.file("extdata", "compdata", "data", 
+             package="flowCore"))
 @
 
-\Rfunction{read.flowSet} will pass on additional arguments meant for the underlying \Rfunction{read.FCS} function, such as \Rfunction{alter.names} and \Rfunction{column.pattern}, but also supports several other interesting arguments for conducting initial processing:
+\Rfunction{read.flowSet} will pass on additional arguments meant for
+the underlying \Rfunction{read.FCS} function, such as
+\Rfunction{alter.names} and \Rfunction{column.pattern}, but also
+supports several other interesting arguments for conducting initial
+processing:
 \begin{description}
-\item[files]{ An alternative to the \Rfunction{pattern} argument, you may also supply a vector of filenames to read.}
-\item[name.keyword]{ Like the example in the previous section, you may specify a particular keyword to use in place of the filename when creating the \Rclass{flowSet}.}
-\item[phenoData]{ If this is an \Rclass{AnnotatedDataFrame}, then this will be used in place of the data frame that is ordinarily created.
-Additionally, the row names of this object will be taken to be the filenames of the FCS files in the directory specified by \Robject{path}.
-This argument may also be a named list made up of a combination of \Robject{character} and \Robject{function} objects that specify a keyword to extract from the FCS file or a function to apply to each frame that will return a result.
-}
+\item[files]{ An alternative to the \Rfunction{pattern} argument, you
+    may also supply a vector of filenames to read.}
+\item[name.keyword]{ Like the example in the previous section, you may
+    specify a particular keyword to use in place of the filename when
+    creating the \Rclass{flowSet}.}
+\item[phenoData]{ If this is an \Rclass{AnnotatedDataFrame}, then this
+    will be used in place of the data frame that is ordinarily
+    created. Additionally, the row names of this object will be taken
+    to be the filenames of the FCS files in the directory specified by
+    \Robject{path}. This argument may also be a named list made up of
+    a combination of \Robject{character} and \Robject{function}
+    objects that specify a keyword to extract from the FCS file or a
+    function to apply to each frame that will return a result. }
 \end{description}
 
-To recreate the \Rclass{flowSet} that we created by hand from the last section we can use \Rfunction{read.flowSet}s advanced functionality:
+To recreate the \Rclass{flowSet} that we created by hand from the last
+section we can use \Rfunction{read.flowSet}s advanced functionality:
 
 <<ReadFowSet2, echo=true,results=verbatim>>=
-fs <- read.flowSet(path=system.file("extdata", "compdata", "data", package="flowCore"), name.keyword="SAMPLE ID", phenoData=list(name="SAMPLE ID", Filename="$FIL"))
+fs <- read.flowSet(path=system.file("extdata", "compdata", "data",
+                   package="flowCore"), name.keyword="SAMPLE ID",
+                   phenoData=list(name="SAMPLE ID", Filename="$FIL"))
 fs
@@
 
-You can extract a \Rclass{flowFrame} from a \Rclass{flowSet} object in the usual way using the \Rfunction{[[} or \Rfunction{\$} extraction operators.
-On the other hand using he \Rfunction{[} extraction operator returns a new \Rclass{flowSet} by \textbf{copying} the environment.
-However, simply assigning the \Rclass{flowFrame} to a new variable will \textbf{not} copy the contained frames.
+You can extract a \Rclass{flowFrame} from a \Rclass{flowSet} object in
+the usual way using the \Rfunction{[[} or \Rfunction{\$} extraction
+  operators. On the other hand using he \Rfunction{[} extraction
+  operator returns a new \Rclass{flowSet} by \textbf{copying} the
+  environment. However, simply assigning the \Rclass{flowFrame} to a
+  new variable will \textbf{not} copy the contained frames.
 
-The primary iterator method for a \Rclass{flowSet} is the \Rfunction{fsApply} method, which works more-or-less like \Rfunction{sapply} or \Rfunction{lapply} with two extra options.
-The first argument, \Robject{simplify}, which defaults to \Robject{TRUE}, instructs \Rfunction{fsApply} to attempt to simplify it's results much in the same way as \Rfunction{sapply}.
-The primary difference is that if all of the return values of the iterator are \Rclass{flowFrame} objects, \Rfunction{fsApply} will create a new \Rclass{flowSet} object to hold them.
-The second argument, \Robject{use.exprs}, which defaults to \Robject{FALSE} instructs \Rfunction{fsApply} to pass the expression matrix of each frame rather than the \Rclass{flowFrame} object itself.
-This allows functions to operate directly on the intensity information without first having to extract it.
+  The primary iterator method for a \Rclass{flowSet} is the
+  \Rfunction{fsApply} method, which works more-or-less like
+  \Rfunction{sapply} or \Rfunction{lapply} with two extra options. The
+  first argument, \Robject{simplify}, which defaults to
+  \Robject{TRUE}, instructs \Rfunction{fsApply} to attempt to simplify
+  it's results much in the same way as \Rfunction{sapply}. The primary
+  difference is that if all of the return values of the iterator are
+  \Rclass{flowFrame} objects, \Rfunction{fsApply} will create a new
+  \Rclass{flowSet} object to hold them. The second argument,
+  \Robject{use.exprs}, which defaults to \Robject{FALSE} instructs
+  \Rfunction{fsApply} to pass the expression matrix of each frame
+  rather than the \Rclass{flowFrame} object itself. This allows
+  functions to operate directly on the intensity information without
+  first having to extract it.
 
-As an aid to this sort of operation we also introduce the \Rfunction{each\_row} and \Rfunction{each\_col} convenience functions that take the place of \Rfunction{apply} in the \Rfunction{fsApply} call.
-For example, if we wanted the median value of each parameter of each \Rclass{flowFrame} we might write:
+  As an aid to this sort of operation we also introduce the
+  \Rfunction{each\_row} and \Rfunction{each\_col} convenience
+  functions that take the place of \Rfunction{apply} in the
+  \Rfunction{fsApply} call. For example, if we wanted the median value
+  of each parameter of each \Rclass{flowFrame} we might write:
 
@@
 
-In this case, the \Robject{use.exprs} argument is not required in the first case because \Rfunction{each\_col} and \Rfunction{each\_row} are methods and have been defined to work on \Rclass{flowFrame} objects by first extracting the intensity data.
+In this case, the \Robject{use.exprs} argument is not required in the
+first case because \Rfunction{each\_col} and \Rfunction{each\_row} are
+methods and have been defined to work on \Rclass{flowFrame} objects by
+first extracting the intensity data.
 
@@
 
-\Rpackage{flowCore} features two methods of transforming parameters within a \Rclass{flowFrame}: inline and out-of-line.
-The inline method, discussed in the next section has been developed primarily to support filtering features and is strictly more limited than the out-of-line transformation method, which uses R's \Rfunction{transform} function to accomplish the filtering.
-Like the normal \Rfunction{transform} function, the \Rclass{flowFrame}is considered to be a data frame with columns named for parameters of the FCS file.
-For example, if we wished to plot our first \Rclass{flowFrame}'s first two fluorescence parameters on the log scale we might write:
+\Rpackage{flowCore} features two methods of transforming parameters
+within a \Rclass{flowFrame}: inline and out-of-line. The inline
+method, discussed in the next section has been developed primarily to
+support filtering features and is strictly more limited than the
+out-of-line transformation method, which uses R's
+\Rfunction{transform} function to accomplish the filtering. Like the
+normal \Rfunction{transform} function, the \Rclass{flowFrame}is
+considered to be a data frame with columns named for parameters of the
+FCS file. For example, if we wished to plot our first
+\Rclass{flowFrame}'s first two fluorescence parameters on the log
+scale we might write:
 
 <<Transfo1, fig=true, echo=true>>=
-plot(transform(fs[[1]], `FL1-H`=log(`FL1-H`), `FL2-H`=log(`FL2-H`)), c("FL1-H","FL2-H"))
+plot(transform(fs[[1]], `FL1-H`=log(`FL1-H`), `FL2-H`=log(`FL2-H`)), 
+     c("FL1-H","FL2-H"))
 @
 
-Like the usual \Rfunction{transform} function, we can also create new parameters based on the old parameters, without destroying the old
+Like the usual \Rfunction{transform} function, we can also create new
+parameters based on the old parameters, without destroying the old
 
 <<Transfo2, fig=true, prefix=FALSE>>=
-plot(transform(fs[[1]], log.FL1.H=log(`FL1-H`), log.FL2.H=log(`FL2-H`)), c("log.FL1.H", "log.FL2.H"))
+plot(transform(fs[[1]], log.FL1.H=log(`FL1-H`), 
+               log.FL2.H=log(`FL2-H`)), c("log.FL1.H", "log.FL2.H"))
 @
@@
 
-Though any function can be used as a transform in both the out-of-line and inline transformation techniques, \Rpackage{flowCore} provides a number of parameterized transform generators that correspond to the transforms commonly found in flow cytometry and defined in the Transformation Markup Language (Transformation-ML).
-Briefly, the predefined transforms are:
+Though any function can be used as a transform in both the out-of-line
+and inline transformation techniques, \Rpackage{flowCore} provides a
+number of parameterized transform generators that correspond to the
+transforms commonly found in flow cytometry and defined in the
+Transformation Markup Language (Transformation-ML). Briefly, the
+predefined transforms are:
 
@@
 
-The most common task in the analysis of flow cytometry data is usual some form of filtering operation, either to obtain summary statistics about the number of events that meet a certain criteria or to perform further analysis on a subset of the data.
+The most common task in the analysis of flow cytometry data is usual
+some form of filtering operation, either to obtain summary statistics
+about the number of events that meet a certain criteria or to perform
+further analysis on a subset of the data.
 
@@
 
-Most filtering operations are a composition of one or more common filtering operations. Like transformations, \Rpackage{flowCore} includes a number of built-in common flow cytometry filters.
+Most filtering operations are a composition of one or more common
+filtering operations. Like transformations, \Rpackage{flowCore}
+includes a number of built-in common flow cytometry filters.
 
-The simplest of these filters are the geometric filters, which correspond to those typically found in interactive flow cytometry software:
+The simplest of these filters are the geometric filters, which
+correspond to those typically found in interactive flow cytometry
+software:
 \begin{description}
-\item[rectangleGate]{Describes a cubic shape in one or more dimensions--a rectangle in one dimension is simply an interval gate.}
-\item[polygonGate]{Describes an arbitrary two dimensional polygonal gate.}
-\item[polytopeGate]{Describes a region that is the convex hull of the given points. This gate can exist in dimensions higher than $2$, unlike the \Robject{polygonGate}.}
-\item[ellipsoidGate]{Describes an ellipsoidal region in two or more dimensions}
+\item[rectangleGate]{Describes a cubic shape in one or more
+    dimensions--a rectangle in one dimension is simply an interval
+    gate.}
+\item[polygonGate]{Describes an arbitrary two dimensional polygonal
+    gate.}
+\item[polytopeGate]{Describes a region that is the convex hull of the
+    given points. This gate can exist in dimensions higher than $2$,
+    unlike the \Robject{polygonGate}.}
+\item[ellipsoidGate]{Describes an ellipsoidal region in two or more
+    dimensions}
 \end{description}
 
-These gates are all described in more or less the same manner (see man pages for more details):
+These gates are all described in more or less the same manner (see man
+pages for more details):
 
 <<rectGate, echo=true, results=verbatim>>=
-rectGate <- rectangleGate(filterId="Fluorescence Region", "FL1-H"=c(50, 100), "FL2-H"=c(50, 100))
+rectGate <- rectangleGate(filterId="Fluorescence Region", 
+                          "FL1-H"=c(50, 100), "FL2-H"=c(50, 100))
 @
 
-We also include some data-driven filters not usually found in flow cytometry software:
+We also include some data-driven filters not usually found in flow
+cytometry software:
 
 \begin{description}
-\item[norm2Filter]{ A robust method for finding a region that most resembles a bivariate Normal distribution. }
-\item[kmeansFilter]{ Identifies populations based on a one dimensional k-means clustering operation. Allows the specification of \textbf{multiple} populations. }
+\item[norm2Filter]{ A robust method for finding a region that most
+    resembles a bivariate Normal distribution. }
+\item[kmeansFilter]{ Identifies populations based on a one dimensional
+    k-means clustering operation. Allows the specification of
+    \textbf{multiple} populations. }
 \end{description}
@@
 
-When we have constructed a filter, we can apply it in two basic ways.
-The first is to collect simple summary statistics on the number and proportion of events considered to be contained within the gate or filter.
-This is done using the \Rfunction{filter} method.
-The first step is to apply our filter to some data
+When we have constructed a filter, we can apply it in two basic
+ways. The first is to collect simple summary statistics on the number
+and proportion of events considered to be contained within the gate or
+filter. This is done using the \Rfunction{filter} method. The first
+step is to apply our filter to some data
 
@@
 
-As we can see, we have returned a \Rclass{filterResult} object, which is in turn a filter allowing for reuse in, for example, subsetting operations.
-To obtain count and proportion statistics, we take the summary of this \Rclass{filterResult}, which returns a list of summary values:
+As we can see, we have returned a \Rclass{filterResult} object, which
+is in turn a filter allowing for reuse in, for example, subsetting
+operations. To obtain count and proportion statistics, we take the
+summary of this \Rclass{filterResult}, which returns a list of summary
+values:
 
@@
 
-A filter which contains multiple populations, such as the \Rclass{kmeansFilter}, can return a list of summary lists:
+A filter which contains multiple populations, such as the
+\Rclass{kmeansFilter}, can return a list of summary lists:
 
@@
 
-A filter may also be applied to an entire \Rclass{flowSet}, in which case it returns a list of \Rclass{filterResult} objects:
+A filter may also be applied to an entire \Rclass{flowSet}, in which
+case it returns a list of \Rclass{filterResult} objects:
 
@@
 
-To subset or split a \Rclass{flowFrame} or \Rclass{flowSet}, we use the \Rfunction{Subset} and \Rfunction{split} methods respectively.
-The first, \Rfunction{Subset}, behaves similarly to the standard R \Rfunction{subset} function, which unfortunately could not used.
-For example, recall from our initial plots of this data that the morphology parameters, Forward Scatter and Side Scatter contain a large more-or-less ellipse shaped population.
-If we wished to deal only with that population, we might use \Rfunction{Subset} along with a \Rclass{norm2Filter} object as follows:
+To subset or split a \Rclass{flowFrame} or \Rclass{flowSet}, we use
+the \Rfunction{Subset} and \Rfunction{split} methods respectively. The
+first, \Rfunction{Subset}, behaves similarly to the standard R
+\Rfunction{subset} function, which unfortunately could not used. For
+example, recall from our initial plots of this data that the
+morphology parameters, Forward Scatter and Side Scatter contain a
+large more-or-less ellipse shaped population. If we wished to deal
+only with that population, we might use \Rfunction{Subset} along with
+a \Rclass{norm2Filter} object as follows:
 
@@
 
-Notice how the \Robject{smaller} \Rclass{flowFrame} objects contain fewer events.
-Now imagine we wanted to use a \Rclass{kmeansFilter} as before to split our first fluorescence parameter into three populations.
-To do this we employ the \Rfunction{split} function:
+Notice how the \Robject{smaller} \Rclass{flowFrame} objects contain
+fewer events. Now imagine we wanted to use a \Rclass{kmeansFilter} as
+before to split our first fluorescence parameter into three
+populations. To do this we employ the \Rfunction{split} function:
 
@@
 
-Of course, most filtering operations consist of more than one gate.
-To combine gates and filters we use the standard R Boolean operators: \&, | and !
-to construct an intersection, union and complement respectively:
+Of course, most filtering operations consist of more than one gate. To
+combine gates and filters we use the standard R Boolean operators: \&,
+| and ! to construct an intersection, union and complement
+respectively:
 
@@
 
-we also introduce the notion of the subset operation, denoted by either \%subset\% or \%\&\%.
-This combination of two gates first performs a subsetting operation on the input \Rclass{flowFrame} using the right-hand filter and then applies the left-hand filter.
-For example,
+we also introduce the notion of the subset operation, denoted by
+either \%subset\% or \%\&\%. This combination of two gates first
+performs a subsetting operation on the input \Rclass{flowFrame} using
+the right-hand filter and then applies the left-hand filter. For
+example,
 
@@
 
-first calculates a subset based on the \Robject{morphGate} filter and then applies the \Robject{rectGate}.
+first calculates a subset based on the \Robject{morphGate} filter and
+then applies the \Robject{rectGate}.
 
@@
 
-Finally, it is sometimes desirable to construct a filter with respect to transformed parameters.
-To allow for this in our filtering constructs we introduce a special form of the \Rfunction{transform} method along with another filter combination operator \%on\%, which can be applied to both filters and \Rclass{flowFrame} or \Rclass{flowSet} objects.
-To specify our transform filter we must first construct a transform list using a simplified version of the \Rfunction{transform} function:
+Finally, it is sometimes desirable to construct a filter with respect
+to transformed parameters. To allow for this in our filtering
+constructs we introduce a special form of the \Rfunction{transform}
+method along with another filter combination operator \%on\%, which
+can be applied to both filters and \Rclass{flowFrame} or
+\Rclass{flowSet} objects. To specify our transform filter we must
+first construct a transform list using a simplified version of the
+\Rfunction{transform} function:
 
@@
 
-Note that this version of the transform filter does not take parameters on the right-hand side--the functions can only take a single vector that is specified by the parameter on the left-hand side.
-In this case those parameters are ``FL1-H'' and ``FL2-H.'' The function also does not take a specific \Rclass{flowFrame} or \Rclass{flowSet} allowing us to use this with any appropriate data.
-We can then construct a filter with respect to the transform as follows:
+Note that this version of the transform filter does not take
+parameters on the right-hand side--the functions can only take a
+single vector that is specified by the parameter on the left-hand
+side. In this case those parameters are ``FL1-H'' and ``FL2-H.'' The
+function also does not take a specific \Rclass{flowFrame} or
+\Rclass{flowSet} allowing us to use this with any appropriate data. We
+can then construct a filter with respect to the transform as follows:
 
@@
 
-Additionally, we can use this construct directly on a \Rclass{flowFrame} or \Rclass{flowSet} by moving the transform to the left-hand side and placing the data on the right-hand side:
+Additionally, we can use this construct directly on a
+\Rclass{flowFrame} or \Rclass{flowSet} by moving the transform to the
+left-hand side and placing the data on the right-hand side:
 
@@
 
-In addition to allowing for sequential filtering, \Rpackage{flowCore} also provides a \Rclass{filterSet} object that serves as an analogue to \Rclass{flowSets} for filters.
-The primary use of this object is to organize and manipulate complex gating strategies.
-For example, recall from the filtering section we had two gates, a ``morphologyGate'' and a ``Fluorescence Region'' gate that we used to demonstrate the various logical operators available for gates.
-To recreate this example as a \Rclass{filterSet}, we would do the following:
+In addition to allowing for sequential filtering, \Rpackage{flowCore}
+also provides a \Rclass{filterSet} object that serves as an analogue
+to \Rclass{flowSets} for filters. The primary use of this object is to
+organize and manipulate complex gating strategies. For example, recall
+from the filtering section we had two gates, a ``morphologyGate'' and
+a ``Fluorescence Region'' gate that we used to demonstrate the various
+logical operators available for gates. To recreate this example as a
+\Rclass{filterSet}, we would do the following:
 
@@
 
-There are two features of note in \Rfunction{filterSet}, which can also take a single list argument for programmatic creation of \Rclass{filterSet} objects.
-The first is that there is a formula interface for the creation of \Rclass{filter} objects from operators.
-The formula interface can be used with or without a left-hand side, which specifies an optional filter identifier.
-If the filter identifier is not specified, as is the case with all but the ``Debris'' gate in the example above the usual default filter identifier is constructed.
-Non-formula gates can also have an optional name specified which overrides the filter identifier specified in the gate constructor.
-This is discouraged at this time as it leads to confusion when creating new filters.
+There are two features of note in \Rfunction{filterSet}, which can
+also take a single list argument for programmatic creation of
+\Rclass{filterSet} objects. The first is that there is a formula
+interface for the creation of \Rclass{filter} objects from
+operators. The formula interface can be used with or without a
+left-hand side, which specifies an optional filter identifier. If the
+filter identifier is not specified, as is the case with all but the
+``Debris'' gate in the example above the usual default filter
+identifier is constructed. Non-formula gates can also have an optional
+name specified which overrides the filter identifier specified in the
+gate constructor. This is discouraged at this time as it leads to
+confusion when creating new filters.
 
@@
 
-Manipulating a \Rclass{filterSet} is done using the normal list element replacement methods, though we do provide a special case where replacing the ``""'' element or the ``NULL'' element causes the \Rclass{filterSet} to use the ``filterId'' slot of the incoming filter to choose a name.
-Similarly, specifying an identifier will override any ``filterId'' slot in the original filter.
+Manipulating a \Rclass{filterSet} is done using the normal list
+element replacement methods, though we do provide a special case where
+replacing the ``""'' element or the ``NULL'' element causes the
+\Rclass{filterSet} to use the ``filterId'' slot of the incoming filter
+to choose a name. Similarly, specifying an identifier will override
+any ``filterId'' slot in the original filter.
 
-Additionally, there are several convenience functions and methods available for manipulating a \Rclass{filterSet}.
-The \Rfunction{names} function provides a list of filter identifiers in an arbitrary order.
-The \Rfunction{sort} lists the filter identifiers according to a topological sort of the filter dependencies such that parent filters are always evaluated before their children.
-Filters without dependencies are provided in an arbitrary relative ordering.
-To obtain the adjacency matrix for this sorting, the option ``dependencies=TRUE'' supplied to the \Rfunction{sort} function will attach an ``AdjM'' attribute to the resulting character vector.
+Additionally, there are several convenience functions and methods
+available for manipulating a \Rclass{filterSet}. The \Rfunction{names}
+function provides a list of filter identifiers in an arbitrary
+order. The \Rfunction{sort} lists the filter identifiers according to
+a topological sort of the filter dependencies such that parent filters
+are always evaluated before their children. Filters without
+dependencies are provided in an arbitrary relative ordering. To obtain
+the adjacency matrix for this sorting, the option
+``dependencies=TRUE'' supplied to the \Rfunction{sort} function will
+attach an ``AdjM'' attribute to the resulting character vector.
 
@@
 
-Though they are not explicitly filters, meaning that they cannot play a role in the construction of subfilters via operators, \Rclass{filterSet} objects do define \Rfunction{filter} and \Rfunction{split} methods that allow for the two most common use cases.
-Additionally, a filter can be selected from the \Rclass{filterSet} using ``[['' to perform a \Rfunction{Subset} operation if desired.
+Though they are not explicitly filters, meaning that they cannot play
+a role in the construction of subfilters via operators,
+\Rclass{filterSet} objects do define \Rfunction{filter} and
+\Rfunction{split} methods that allow for the two most common use
+cases. Additionally, a filter can be selected from the
+\Rclass{filterSet} using ``[['' to perform a \Rfunction{Subset}
+operation if desired.
 
-For standard filtering operations, the \Rclass{filterSet} can yield better performance than a manual filtering operation because it ensures that each parent filter is only evaluated once per call to \Rfunction{filter} so that if many subfilters rely on a particular, likely expensive, filtering operation they use the original result rather than recalculating each time.
-The individual \Rclass{filterResult} objects may be later extracted for further analysis of desire, making the \Rclass{filterSet} the logical choice for most filtering operations.
-Additionally, though not presently implemented, it can allow visualization methods to make more intelligent choices about display when rendering gating strategies because the disposition of the entire filtering graph is known and the filter results are already available.
+For standard filtering operations, the \Rclass{filterSet} can yield
+better performance than a manual filtering operation because it
+ensures that each parent filter is only evaluated once per call to
+\Rfunction{filter} so that if many subfilters rely on a particular,
+likely expensive, filtering operation they use the original result
+rather than recalculating each time. The individual
+\Rclass{filterResult} objects may be later extracted for further
+analysis of desire, making the \Rclass{filterSet} the logical choice
+for most filtering operations. Additionally, though not presently
+implemented, it can allow visualization methods to make more
+intelligent choices about display when rendering gating strategies
+because the disposition of the entire filtering graph is known and the
+filter results are already available.
 
-Like any other filtering operation, \Rfunction{summary} methods are also available to collect the usual sorts of count statistics for a filtering operation:
+Like any other filtering operation, \Rfunction{summary} methods are
+also available to collect the usual sorts of count statistics for a
+filtering operation:
 
@@
 
-When splitting a \Rclass{flowFrame} using \Rfunction{split}, we have two options. The first is to include all of the resultant subsets in the usual way:
+When splitting a \Rclass{flowFrame} using \Rfunction{split}, we have
+two options. The first is to include all of the resultant subsets in
+the usual way:
 
@@
 
-While this provides access to all subsets, for automated analysis we will often only wish to work with the ``leaf'' gates (i.e.
-those without children of their own).
-In this case, we can specify ``drop=TRUE'' to obtain only the leaf gates:
+While this provides access to all subsets, for automated analysis we
+will often only wish to work with the ``leaf'' gates (i.e. those
+without children of their own). In this case, we can specify
+``drop=TRUE'' to obtain only the leaf gates:
 
@@
 
-Note that in both cases, we use the ``flowSet=TRUE'' option that has been added to the \Rfunction{split} function to automatically create a \Rclass{flowSet} object from the resulting \Rclass{flowFrame} objects.
-For \Rfunction{filterSet} objects, the \Rfunction{split} function can take advantage of the added structure in this case to provide some information about the filtering operation itself in the \Rclass{flowSet}'s metadata as well.
+Note that in both cases, we use the ``flowSet=TRUE'' option that has
+been added to the \Rfunction{split} function to automatically create a
+\Rclass{flowSet} object from the resulting \Rclass{flowFrame}
+objects. For \Rfunction{filterSet} objects, the \Rfunction{split}
+function can take advantage of the added structure in this case to
+provide some information about the filtering operation itself in the
+\Rclass{flowSet}'s metadata as well.
 

### 2008-04-08 R/AllClasses.R
making sure that the samplesare in the same order after splitting. Also added a couple of updates to the documentation
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@30903 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -117,7 +117,7 @@ setClass("flowSet",
          })
 
 ## constructor
-flowSet <- function(...,phenoData) {
+flowSet <- function(..., phenoData) {
     x <- list(...)
     if(length(x) == 1 && is.list(x[[1]]))
         x <- x[[1]]

### 2008-04-08 R/flowSet-accessors.R

@@ -67,9 +67,9 @@ setMethod("colnames","flowSet",function(x, do.NULL="missing",prefix="missing")
           x@colnames)
 
 
-## fixme: should we make sure that parameter names are changed for each frame?
+## FIXME: should we make sure that parameter names are changed for each frame?
 setReplaceMethod("colnames","flowSet",function(x,value) {
-	x@colnames = value
+	x@colnames <- value
 	x
 })
 
@@ -99,7 +99,15 @@ setMethod("phenoData<-","flowSet",function(object,value) {
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("pData","flowSet",function(object) pData(object@phenoData))
 
-
+setReplaceMethod("pData",
+                 signature("flowSet", "data.frame"),
+                 function(object,value)
+             {
+                 pd <- phenoData(object)
+                 pData(pd) <- value
+                 phenoData(object) <- pd
+                 object
+             })
 
 ## ==========================================================================
 ## set and extract the varLabels of the phenoData
@@ -126,16 +134,24 @@ setReplaceMethod("varLabels",
 setMethod("sampleNames", "flowSet", function(object) 
        sampleNames(phenoData(object)))
 
+## Note that the replacement method also replaces the GUID for each flowFrame
 setReplaceMethod("sampleNames","flowSet",function(object,value)
              {
                  oldNames <- sampleNames(object)
+                 value <- as.character(value)
                  if(length(oldNames)!=length(value) ||
                     !is.character(value))
                      stop(" replacement values must be character vector ",
-                          "of length equal to number of frames in the set'")
+                          "of length equal to number of frames in the set'",
+                          call.=FALSE)
+                 if(any(duplicated(value)))
+                     stop("Replacement values are not unique.", call.=FALSE)
                  env <- new.env(hash=TRUE,parent=emptyenv())
-                 for(f in seq_along(oldNames))
-                     assign(value[f], get(oldNames[f], object@frames), env)
+                 for(f in seq_along(oldNames)){
+                     tmp <- get(oldNames[f], object@frames)
+                     identifier(tmp) <- value[f]
+                     assign(value[f], tmp, env)
+                 }
                  pd <- phenoData(object)
                  sampleNames(pd) <- value
[…]

### 2008-04-08 R/split-methods.R

@@ -1,0 +1,240 @@
+## ==========================================================================
+## split methods for flowFrame
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+## We actually split on filterResults and multipleFilterResults
+setMethod("split",
+          signature("flowFrame", "filter"),
+          function(x, f, drop=FALSE, ...)
+          split(x, filter(x, f), drop, ...))
+
+setMethod("split",
+          signature("flowFrame","filterSet"),
+          function(x, f, drop=FALSE, ...)
+          split(x, filter(x, f), drop, ...))
+
+## split on logicalFilterResults
+setMethod("split",
+          signature("flowFrame", "logicalFilterResult"),
+          function(x, f, drop=FALSE, population=NULL, prefix=NULL,
+                   flowSet=FALSE, ...)
+      {
+          if(is.null(population))
+              population <- f@filterId
+          if(!is.null(prefix))
+              population <- paste(prefix,population, sep="")
+          out <- structure(list(x[f@subSet, ], x[!f@subSet, ]),
+                           names=c(paste(population, "+", sep=""),
+                           paste(population,"-", sep="")))
+          if(length(flowSet) > 0 && flowSet)
+              out <- flowSet(out)
+          return(if(is.list(out) && length(out)==1) out[[1]] else out)
+      })
+
+## split on multipleFilterResults, argument population can be used to
+## select only certain subpopulations
+setMethod("split",
+          signature("flowFrame", "multipleFilterResult"),
+          function(x, f, drop=FALSE, prefix=NULL, flowSet=FALSE,
+                   population=NULL, ...)
+      {
+          if(is.null(population))
+              population <- names(f)
+          else if(!all(population %in% names(f)))
+              stop("Population(s) not valid in this filter", call.=FALSE)
+          if(is.null(prefix))
+              nn <- population
+          else
+              nn <- paste(prefix, population, sep="")
+          tmp <- lapply(population, function(i) x[f[[i]], ])
+          out <- structure(tmp, names=nn)
+          if(length(flowSet) > 0 && flowSet)
+              out <- flowSet(out)
+          return(if(is.list(out) && length(out)==1) out[[1]] else out)
+      })
+
+
+## Split on manyFilterResults. FIXME: Need to take a closer look at this
+setMethod("split", signature("flowFrame","manyFilterResult"),
+          function(x, f, drop=FALSE, prefix=NULL, flowSet=FALSE, ...)
[…]

### 2008-04-08 paper/outline.tex
Added a few sentences about Gating-ML and fixed some typos in the text.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@30924 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-Traditionally, flow cytometry (FC) has been a tube-based technique
+Traditionally, flow cytometry (FCM) has been a tube-based technique
 limited to small-scale laboratory and clinical studies.  High
@@
 order to be accessible to the researcher. However, the absence of any
-shared research platform which bioinformaticians, computer scientists,
-and statisticians can utilze to develop novel or standard methods for
-flow cytometric has become problematic. We believe that the open
-source statistical software R in conjuction with the Bioconductor
+shared research platform, which bioinformaticians, computer scientists,
+and statisticians can utilize to develop novel or standard methods for
+flow cytometric analysis has become problematic. We believe that the open
+source statistical software R in conjunction with the Bioconductor
 Project can fill this gap.  In this paper, we propose R data
@@
 towards the analysis of large numeric datasets typical to arise from
-high throughput experiments such as microarray gene expression anlysis
+high throughput experiments such as microarray gene expression analysis
 or FC-HCS. Today, the R open source project offers a wide range of
@@
 use to develop novel tools and methodology for statistical data
-analyis. For biological applications, the Bioconductor project has
+analysis. For biological applications, the Bioconductor project has
 become the standard toolset \citep{gentleman2006bos} to process data
@@
 complex cell biology. This shared platform proved to be especially
-beneficial for the development of analyis routines for microaray gene
+beneficial for the development of analysis routines for microarray gene
 expression data, and by now many of these innovations have been
-incoorporated into comercial software products, thus greatly improving
+incorporated into commercial software products, thus greatly improving
 the quality of gene expression data analysis. Shared data structures
@@
 dozens of co-variates and influencing factors. Self-contained data
-structures are a prerequesite to allow for coordinated analysis of
+structures are a prerequisite to allow for coordinated analysis of
 such data sets. Most of the currently available software solutions
@@
 specifically for the mostly GUI-driven user interfaces and hence are
-not easily amenable to programatical access. In addition, the
+not easily amenable to programmatical access. In addition, the
 closed-source nature of most of these products makes it virtually
@@
 of summary statistics. We believe that these approaches do not give
-justice to the highly complex nature of FC data, essentially
+justice to the highly complex nature of FCM data, essentially
 discarding many of the fundamental aspects such as data distribution
@@
 statistical methods need to be developed to point the investigator to
-the intersting aspects of the data or to potential problems. While the
+the interesting aspects of the data or to potential problems. While the
 expert knowledge of immunologists and researchers remains crucial for
-the understanding of FC data, we believe that tight colaboration
-with other reserach fields such as statistics and computer science can
-greatly improve the relevance of flow cytomtry in today's
+the understanding of FCM data, we believe that tight collaboration
+with other research fields such as statistics and computer science can
+greatly improve the relevance of flow cytometry in today's
 high-throughput biological sciences.
+
+These issues are currently being addressed by emerging new flow cytometry 
+data standards developed in collaboration with the ISAC Data Standards Task Force. 
+The Gating-ML Candidate Recommendation (CR) represents a proposal on how to form 
+unambiguous XML-based gate definitions, which can facilitate the interchange and 
+validation of data between different software packages with the potential of 
+significant increase of hardware and software interoperability. Gates may be 
+ordered into a hierarchical structure to describe a gating strategy. They may be 
+applied on parameters as in FCS files or on transformed parameters as described 
+by an explicit parameter transformation. This allows to encode a simple analytical 
+work flow, which is sufficient in order to reconstruct the analysis programmatically.
 
@@
   all). Somebody more familiar with flowJo and friends should add to
-  this... Also somebody should fill in a couple of sentences about the
-  gating-ML etc. stuff here}
+  this...}
 
 The flowCore framework presented here offers import functionality for
-both the raw data FCS files along with the complete set of
-file-specific meta information as well as all the information that is
-stored in one of the gating-, compensation- or transformation-ML
-standards. Thus, it is possible to intergrate flowCore in existing
-work flows and to communicate with virtually every other FC tool that
-adheres to these standards.
+raw data FCS files along with a complete set of file-specific meta information.
+Moreover, it is one of the very first software implementations of the Gating-ML CR, 
+which opens the possibility to integrate flowCore in existing work flows and to 
+communicate with virtually every other FCM tool that adheres to this standard.
 
@@
 flowCore's primary task is the representation and basic manipulation
-of flow cytometry (or similar) data. This is accomplished through a
+of flow cytometry data. This is accomplished through a
 data model very close to that adopted by other Bioconductor packages
@@
 Bioconductor users. Similar to any other flow tool, the basic unit of
-organisation in flowCore is a collection of events. We call the
+organization in flowCore is a collection of events. We call the
 structure that hold this data a flowFrame. All relevant information
 for the collection are stored along with the matrix of raw data values
-and can be accessed programatically. Most prominently, these are
+and can be accessed programmatically. Most prominently, these are
 descriptors of the stains used in the experiment and the respective
@@
 performed during the creation of a flowFrame to ensure data
-integrety. There is an extended interface that allows for the
+integrity. There is an extended interface that allows for the
 standardized interaction with the flow data, thus enabling the
@@
 all too common when dealing with complex data in the ususal way of
-more or less ad hoc spreadsheats that lack a controlled structure. In
+more or less ad hoc spreadsheets that lack a controlled structure. In
 general, there is no native ordering to a flowSet, which in a way
@@
 rather than experiments following a plate-design. However, the flowSet
-structure can be readily extended to also accomodate such spacial
+structure can be readily extended to also accommodate such spacial
 information \textit{(reference)}
@@
 for improvement, being able to implement new methodology for gating of
-flow cytometry data and extending the capabilies of flowCore through
+flow cytometry data and extending the capabilities of flowCore through
 the usual ways of object oriented programming is a feature, that
@@
 authors can focus on the actual operations rather then having to deal
-with the teadious details of data integration and access. Third party
+with the tedious details of data integration and access. Third party
 methods implemented by others can act on their own as first-class
@@
 based on the properties of the underlying data, for instance
-distribution or density estimates. The ability to programatically
-access gates is a prerequesite for automated or semi-automated
+distribution or density estimates. The ability to programmatically
+access gates is a prerequisite for automated or semi-automated
 gating. By utilizing a unified interface for all different types of
@@
 gates or combinations of multiple gates. The definition of gates in
-flowCore follows the proposed gating-ML standard, thus any flowCore
+flowCore follows the Gating-ML CR, thus any flowCore
 gating strategy can be reproduced by any other software that also
@@
 we have implemented a range of additional Bioconductor packages that
-are dedicated to more specific tasks of flow cytomtery data
+are dedicated to more specific tasks of flow cytometry data
 analysis. The flowViz package provides sophisticated data

### 2008-04-12 R/AllGeneric.R
moved all plotting stuff to flowViz, flowCore now issues a message to attach flowViz for all plotting
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31029 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -7,13 +7,6 @@
 
 
 
-
-## ===========================================================================
-## These are already defined as generic functions:
-## ---------------------------------------------------------------------------
-setGeneric("draw",function(x,data,...) standardGeneric("draw"))
-
-
 ## ===========================================================================
 ## Generics for our operators
 ## %in% asks a question whereas %subset% forms a set.

### 2008-04-12 R/filter-accessors.R

@@ -116,34 +116,3 @@ setMethod("!",signature("filter"),function(x) {
 ## --------------------------------------------------------------------------
 "%&%" = function(e1,e2) e1 %subset% e2
 
-## ==========================================================================
-## Plotting method for filters lets us get a basic plot of the data and a 
-## filter. Uses the draw method to actually display the filter on some data.
-## --------------------------------------------------------------------------
-setMethod("plot",signature(x="flowFrame", y="filter"),function(x, y, z=NULL, results=NULL, ...) {
-    if(is.null(z))	z = parameters(y)
-    if(length(z)>2) {
-        warning("Only plotting the first two parameters.")
-        z = z[1:2]
-    }
-    ##First we plot the actual data.
-    plot(x, z, ...)
-    ##Then draw the filter on top of it which takes the form (filter, data, params, results (if any))
-    draw(y, x, z, results, ...)
-})
-
-setMethod("draw", signature("filter","flowFrame"), function(x, data, params=NULL, results=NULL, ...) {
-	if(is.null(params))
-          params = parameters(x)
-	if(length(params)>2) {
-		params = params[1:2]
-	}
-        if(is(x,"rectangleGate")){
-          lines(c(x@min[params[2]], x@min[params[2]]), c(x@min[params[1]], x@max[params[1]]), col="red", ...)
-          lines(c(x@min[params[1]], x@min[params[1]]), c(x@min[params[2]], x@max[params[2]]), col="red", ...)
-      }
-        if(is(x,"polygonGate") | is(x,"polytopeGate"))
-          lines(x@boundaries[,params[1]], x@boundaries[,params[2]], col="red", ...)
-        ##if(is(x, ellipsoidGate))
-          
-})

### 2008-04-12 R/flowFrame-accessors.R

@@ -227,48 +227,16 @@ setMethod("show",signature=signature("flowFrame"),
 
 
 
-## ==========================================================================
-## a simple plot method without strange plot parameter and friends. It does
-## the most intuitive thing: take a flowFrame and do a pairs plot if there
-## are more than 2 parameters. Do a smoothScatter plot for exactly 2
-## parameters and a histogram for exactly one.
-## If you want to plot specific columns, subset the frame before
-## plotting or specify them in the y argument.
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-## only one argument: the flowFrame
-setMethod("plot", signature(x="flowFrame", y="missing"),
-          definition=function(x, smooth=TRUE, pch, ...)
-      {
-          l = ncol(x)
-          values=exprs(x)
-          if(missing(pch))
-              pch="."
-          if(l==1)
-              hist(values, xlab=colnames(x), ...)
-          else if (l==2){
-              if(smooth)
-                  smoothScatter(values, pch=pch, ...)
-              else
-                  plot(values, pch=pch, ...)
-          }else{
-              sel <- tolower(colnames(x)) != "time"
-              if(smooth){
-                  require(flowViz)
-                  plot(splom(x[,sel], pch=pch, ...))
-              }
-              else
-                  pairs(values[,sel], pch=pch, ...)
-          }
-      })
 
-## second argument contains the parameters(s) to plot
-setMethod("plot",signature(x="flowFrame",y="character"),
-          function(x,y, smooth=TRUE, ...){
-              if(!all(y %in% colnames(x)))
-                  stop("subset out of bounds", call.=FALSE)
-              callGeneric(x[,y], smooth=smooth, ...)
+## ==========================================================================
+## plot method: We actually attach flowViz to do the plotting
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("plot", signature(x="flowFrame", y="ANY"),
+          function(x, y, ...)
+      {
+          message("For plotting, please attach the 'flowViz' package.\n",
+                  "   e.g. 'library(flowViz)'")
       })
-         
 
 
 

### 2008-04-15 paper/outline.tex
comments and notes
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31090 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
+{\bf NLM:} Only a couple of commercial software (e.g., FlowJo and
+DeNovo) allows the user to analyse batch of FCS files.  Only a few
+allows export in XML (e.g., FlowJo). Most commercial software require
+a Microsoft Windows platform whereas flowCore runs on multiple
+operating systems including Windows, Mac OS X and Unix.
+
 {\bf RB:}  It would be important to highlight
@@
 information \textit{(reference)}
+
+{\bf NLM:} How about mentioning PlateCore here?
 
@@
 
+{\bf NLM:} example: cite flowClust and floViz paper?
+
 \section{Application, Outlook}
@@
 
+{\bf NLM:} We could illustrate/summarize:
+\begin{itemize}
+\item the power of subsetting: for example to manipulate 
+patient below a certain age or of the same sex...
+\item Gating: filterSet strategy, kmeanFilter
+\item visualization adding metadata as parameter
+example: cite flowViz paper
+\end{itemize}
+
 \bibliographystyle{plainnat}  

### 2008-04-17 paper/outline.tex
more on the outline
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31140 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\subsection{R and Bioconductor}
+\subsection*{R and Bioconductor}
 R is a solid statistical programing environment, particularly geared
 towards the analysis of large numeric datasets typical to arise from
-high throughput experiments such as microarray gene expression analysis
-or FC-HCS. Today, the R open source project offers a wide range of
-statistical and visualization methods developed for various fields of
-application. In addition, R is a research platform which
+high throughput experiments such as microarray gene expression
+analysis or FC-HCS. Today, the R open source project offers a wide
+range of statistical and visualization methods developed for various
+fields of application. In addition, R is a research platform which
 bioinformaticians, computer scientists, and statisticians routinely
@@
 spanning a diversity of research fields from genomics to proteomics to
-complex cell biology. This shared platform proved to be especially
-beneficial for the development of analysis routines for microarray gene
-expression data, and by now many of these innovations have been
-incorporated into commercial software products, thus greatly improving
-the quality of gene expression data analysis. Shared data structures
-have been shown to be essential for this kind of collaborative effort.
+complex cell biology. This shared platform proved to be particularly
+beneficial for the development of analysis routines for microarray
+gene expression data, and by today many of these early innovations
+have been incorporated into commercial software products, thus greatly
+improving the quality of gene expression data analysis. In this
+context, shared data structures have been shown to be essential for
+such a collaborative effort.
 
-\subsection{Existing data standards and conventions}
+\subsection*{Existing data standards and conventions}
 Currently, data from flow cytometry experiments are stored in single
 files following a well-defined data standard (FCS). This standard
-tries to capture as much of the accompanying meta information as
+attempts to capture as much of the accompanying meta information as
 possible, most of which is produced by the measurement instrument or
@@
 based measurements towards large and complex experimental designs with
-dozens of co-variates and influencing factors. Self-contained data
-structures are a prerequisite to allow for coordinated analysis of
-such data sets. Most of the currently available software solutions
-offer only limited support for such structures, or make use of
-containers like XML or binary storage formats that are designed
-specifically for the mostly GUI-driven user interfaces and hence are
-not easily amenable to programmatical access. In addition, the
-closed-source nature of most of these products makes it virtually
-impossible to connect and integrated these data.
+dozens of co-variates and influencing factors. Self-contained and
+self-reflective data structures are a prerequisite to allow for
+coordinated analysis of such experiments. Most of the currently
+available software solutions offer only limited support for such
+structures, or make use of containers like XML or binary storage
+formats that are designed specifically for the mostly GUI-driven user
+interfaces and hence are not easily amenable to programmatical
+access. In addition, the closed-source nature of most of these
+products makes it virtually impossible to connect and integrated the
+data.
 
 Traditionally, the majority of flow cytometry experiments have been
-analysed manually by direct data inspection or very basic comparisons
-of summary statistics. We believe that these approaches do not give
-justice to the highly complex nature of FCM data, essentially
-discarding many of the fundamental aspects such as data distribution
-or high-dimensionality. In addition, the subjective character of
-manual analyses are a major obstacle for reproducibility. For FC-HCS
-data, unassisted manual inspection is no longer feasible, and robust
+analysed manually by direct data inspection in one ore tow dimensions,
+or by very basic comparisons of summary statistics. We believe that
+these approaches do not give justice to the highly complex nature of
+FCM data, in particualar, they discard many of the fundamental aspects
+such as underlying distribution of the data or the
+high-dimensionality. In addition, the subjective character of manual
+analyses are a major obstacle for reproducibility. For FC-HCS data,
+unassisted manual inspection is no longer feasible, and robust
 statistical methods need to be developed to point the investigator to
-the interesting aspects of the data or to potential problems. While the
-expert knowledge of immunologists and researchers remains crucial for
-the understanding of FCM data, we believe that tight collaboration
+the interesting aspects of the data or to potential problems. While
+the expert knowledge of immunologists and researchers remains crucial
+for the understanding of FCM data, we believe that tight collaboration
 with other research fields such as statistics and computer science can
@@
 
-These issues are currently being addressed by emerging new flow cytometry 
-data standards developed in collaboration with the ISAC Data Standards Task Force. 
-The Gating-ML Candidate Recommendation (CR) represents a proposal on how to form 
-unambiguous XML-based gate definitions, which can facilitate the interchange and 
-validation of data between different software packages with the potential of 
-significant increase of hardware and software interoperability. Gates may be 
-ordered into a hierarchical structure to describe a gating strategy. They may be 
-applied on parameters as in FCS files or on transformed parameters as described 
-by an explicit parameter transformation. This allows to encode a simple analytical 
-work flow, which is sufficient in order to reconstruct the analysis programmatically.
+These issues are currently being addressed by emerging new flow
+cytometry data standards developed in collaboration with the ISAC Data
+Standards Task Force.  The Gating-ML Candidate Recommendation (CR)
+represents a proposal on how to form unambiguous XML-based gate
+definitions, which can facilitate the interchange and validation of
+data between different software packages with the potential of
+significant increase of hardware and software interoperability. Gates
+may be ordered into a hierarchical structure to describe a gating
+strategy. They may be applied on parameters as in FCS files or on
+transformed parameters as described by an explicit parameter
+transformation. This allows to encode a simple analytical work flow,
+which is sufficient in order to reconstruct the analysis
+programmatically.
+
+The flowCore framework presented here offers import functionality for
+raw data FCS files along with their complete set of file-specific meta
+information.  Moreover, it is one of the very first software
+implementations of the Gating-ML CR, which opens the possibility to
+integrate flowCore in existing work flows and to communicate with
+virtually every other FCM tool that adheres to the proposed
+standard. In addition, the software is platform independent, running
+on various operating systems including Windows, Mac OS X and
+Linux/Unix.
+
+flowCore is not a GUI-driven software, and all operations are
+currently done using a command line interface. Conceptually, it is
+possible to add a more elaborate user interface on top of that,
+however we strongly focus on programmatic access to enable the
+convenient development of novel analysis methods. By taking the burden
+of data management from the programmer and by providing well-defined
+APIs and a structured class hierarchy, it is possible to readily test
+new ideas and to easily extend the frameworks functionality. Both the
+underlying R and Bioconductor software and flowCore itself are open
+source and open development projects, thus community involvement in the
+developing process is highly appreciated.
 
@@
 
-The flowCore framework presented here offers import functionality for
-raw data FCS files along with a complete set of file-specific meta information.
-Moreover, it is one of the very first software implementations of the Gating-ML CR, 
-which opens the possibility to integrate flowCore in existing work flows and to 
-communicate with virtually every other FCM tool that adheres to this standard.
-
 {\bf NLM:} Only a couple of commercial software (e.g., FlowJo and
 DeNovo) allows the user to analyse batch of FCS files.  Only a few
-allows export in XML (e.g., FlowJo). Most commercial software require
-a Microsoft Windows platform whereas flowCore runs on multiple
-operating systems including Windows, Mac OS X and Unix.
+allows export in XML (e.g., FlowJo).
 
@@
 \section{Basic data structures}
-\subsection{flowFrame}
+\subsection*{flowFrame}
 flowCore's primary task is the representation and basic manipulation
-of flow cytometry data. This is accomplished through a
-data model very close to that adopted by other Bioconductor packages
-used for gene expression analysis and thus familiar to most
-Bioconductor users. Similar to any other flow tool, the basic unit of
-organization in flowCore is a collection of events. We call the
-structure that hold this data a flowFrame. All relevant information
-for the collection are stored along with the matrix of raw data values
-and can be accessed programmatically. Most prominently, these are
-descriptors of the stains used in the experiment and the respective
-measurement channels, information about compensation performed at the
-instrument side and any additional keywords the user deems to be
-important to annotate the data. A number of quality checks are
-performed during the creation of a flowFrame to ensure data
-integrity. There is an extended interface that allows for the
-standardized interaction with the flow data, thus enabling the
-development of more specialized tools for specific tasks of the data
-analysis process.
+of flow cytometry data. This is accomplished through a data model very
+close to that adopted by other Bioconductor packages used for gene
+expression analysis, and thus familiar to most Bioconductor
+users. Similar to any other flow tool, the basic unit of organization
+in flowCore is a collection of events. We call the structure that hold
+this data a flowFrame. All relevant information for the collection are
+stored along with the matrix of raw data values and can be accessed
+programmatically. Most prominently, these are descriptors of the
+stains used in the experiment and the respective measurement channels,
+information about compensation performed at the instrument side and
+any additional keywords the user deems to be important to annotate the
+data. A number of quality checks are performed during the creation of
+a flowFrame to ensure data integrity. There is an extended interface
+that allows for the standardized interaction with the flow data, thus
+enabling the development of more specialized tools for specific tasks
+of the data analysis process.
 
-\subsection{flowSet}
+\subsection*{flowSet}
 In high-throughput flow cytometry, many of the analysis tasks need to
@@
 infrastructure, thus reducing the risk of mixups and errors that are
-all too common when dealing with complex data in the ususal way of
-more or less ad hoc spreadsheets that lack a controlled structure. In
-general, there is no native ordering to a flowSet, which in a way
-reflects the still predominant application of tube-based experiments
-rather than experiments following a plate-design. However, the flowSet
-structure can be readily extended to also accommodate such spacial
-information \textit{(reference)}
-
-{\bf NLM:} How about mentioning PlateCore here?
+all too common when dealing with complex data in the ususal way:
+namely, more or less ad hoc spreadsheets that lack a controlled
+structure. In general, there is no native ordering to a flowSet, which
+in a way reflects the still predominant application of tube-based
+experiments rather than experiments following a plate-design. However,
+the flowSet structure can be readily extended to also accommodate such
+spatial information \citep{(plateCore)}
 
@@
   operations (Figure1.ppt)}
-
-{\bf PH:} I think the Frame/Set concept is pretty interesting and
-powerful. Was it originated for this project? The ability to program
-sets flowCore aside from all other flow software applications, I
-think. I think there is something really important and valuable about
-how the data structures are ogranized. I'm not sure how to write that
-in a way to appeal to Cytometry A readers, but I think we could work
-this out.
 
@@
 around. By factoring out as much of the bookkeeping as possible,
-authors can focus on the actual operations rather then having to deal
-with the tedious details of data integration and access. Third party
-methods implemented by others can act on their own as first-class
-citizens in the analysis, without breaking the workflow or the basic
-infrastructure. This design allows for the straight forward extension
-of flowCore's capabilities and already fostered the development of a
-number of valuable improvements. \textit{(cite stuff here)}
+programmers can focus on the actual operations rather then having to
+deal with the tedious details of data integration and access. Third
+party methods implemented by others can act on their own as
+first-class citizens in the analysis framework, without breaking the
+workflow or the basic infrastructure. This design allows for the
+straight forward extension of flowCore's capabilities and already
+fostered the development of a number of valuable
+improvements. \citep{(flowClust, plateCore)}
 
-\subsection{transformations}
-All the major transformations that are routinely used in flow
-cytometry analysis have been implemented in flowCore. Furthermore, the
-design of the R language makes it easy to define arbitrary functions
-to apply to the data of individual flowFrames or entrire flowSets,
-respectively. Proper transformation have been proven beneficial for
-both data visualization and modelling \textit{cite flowClust paper}.
+\subsection*{Transformation and compensation}
+Proper transformation have been proven beneficial for both data
+visualization and modelling \textit{cite flowClust paper}. All the
+major transformations that are routinely used in flow cytometry
+analysis have been implemented in flowCore. Furthermore, the design of
+the R language makes it easy to define arbitrary functions to apply to
+the data of individual flowFrames or entrire flowSets,
+respectively. Compensation is available for both flowFrames and
+flowSets, and the software also offers functionality to compute
+spillover or compensation matrices from a set of appropriate
+compensation samples.
 
-\subsection{gating}
-In flowCore, gates are represented as individual objects that can be
+\subsection*{Gating}
+In flowCore, gates are represented as individual classes that can be
 extended in an object oriented manner. All basic gate types such as
-rectangular gates, ellipses and polygons are implemented as part of
-the base framework. In addition, we introduce the notion of
-data-driven gates, for which the necessary parameters are computed
-based on the properties of the underlying data, for instance
-distribution or density estimates. The ability to programmatically
-access gates is a prerequisite for automated or semi-automated
-gating. By utilizing a unified interface for all different types of
-gates, the user is able to subset data sets as well as create summary
-statistics, for instance of proportions of events falling in single
-gates or combinations of multiple gates. The definition of gates in
-flowCore follows the Gating-ML CR, thus any flowCore
-gating strategy can be reproduced by any other software that also
-adheres to the standard.
+rectangular gates, ellipses and polygon gates are implemented as part
+of the base framework. In addition, we introduce the notion of
+data-driven gates, or filters, for which the necessary parameters are
+computed based on the properties of the underlying data, for instance
+by modelling underlying data distribution or by density
+estimation. The ability to programmatically access gates is a
+prerequisite for automated or semi-automated gating. By utilizing a
+unified interface for all different types of gates, the user is able
+to subset data sets as well as to create summary statistics, like the
+proportions of events falling in single gates or in the combinations
+of multiple gates. Conmplex combinations and hierarchies of gates can
+be captured in objects of class filterSet, allowing to apply
+multi-step gating strategies. The definition of gates in flowCore
+follows the Gating-ML CR, thus any flowCore gating strategy can be
+reproduced by any other software that also adheres to the standard.
 
 \section{Related flow packages}
-In addition to the flowCore package that offers basic infrastructure
+In addition to the flowCore package that offers basic infrastructure,
 we have implemented a range of additional Bioconductor packages that
@@
 analysis. The flowViz package provides sophisticated data
-visualization tools that make use of multivariate trellis
+visualization tools, that make use of multivariate trellis
 plotting. These functions can be used to quickly generate customized
@@
 assurance (QA) methodology and a framework to create interactive
-reports of QA results. Most of the flowUtils content deals with data
-import and output from and to flowCore and the flow-cytometry specific
-standard markup languages. Finally, flowStats provides statistical
-methods that are relevant in the context of flow cytometry data
-analysis.
+web-based reports of QA results. Most of flowUtil's content deals with
+data import and output from and to flowCore and the flow-cytometry
+specific standard markup languages. Finally, flowStats provides
+statistical methods that are relevant in the context of flow cytometry
+data analysis.
 

### 2008-04-25 paper/outline.tex
paper: first paragraph for the outlook-application section
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31325 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 fostered the development of a number of valuable
-improvements. \citep{(flowClust, plateCore)}
+improvements. \citep{lo2008agf} (+plateCore).
 
@@
 
-{\bf NLM:} example: cite flowClust and floViz paper?
-
 \section{Application, Outlook}
@@
 
-{\bf NLM:} We could illustrate/summarize:
-\begin{itemize}
-\item the power of subsetting: for example to manipulate 
-patient below a certain age or of the same sex...
-\item Gating: filterSet strategy, kmeanFilter
-\item visualization adding metadata as parameter
-example: cite flowViz paper
-\end{itemize}
+{\bf NLM:} Do we really need supplementary material, can you cite the
+paper and packages's vignette?
+
+The flowCore package has already been used to analyze a number of
+datasets \citep{gasparetto2004ice,brinkman2007hcf}. Its concepts have
+also been integrated in other Bioconductor packages. As an example,
+\cite{lo2008agf} have recently proposed an automatic gating approach
+via robust model-based clustering using the flowCore data model as the
+container for the experimental data and some of the concepts as
+clustering methods. The authors have wrapped their clustering method
+in a filter operation and the outcome extends the flowCore
+filterResult class. Most of the methods available for the flowCore
+filterResult class are then available to manipulate the clustering
+result. Similarly, the flowViz package make use of the filter concepts
+to group variables in a display \cite{sarkar2008ufv}. Even more, the
+flowViz plotting functions take advantages of the flowCore data model
+to represent multiple variables within one display. To that aim and to
+list a few, the plotting functions make use of the data structure to
+modify the scale, group, subset, and color the variables.
+
+{\bf NLM:} Should we add some figure or code?
 

### 2008-04-29 paper/outline.tex
Deepayans suggestions and Application section
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31424 bc3139a8-67e5-0310-9ffc-ced21a209358

--- 2008-04-17 paper/outline.tex
@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 
-\textit{How does the flow data world look like? Other flow software
-  (what other software are handling multiple FCS files at
-  all). Somebody more familiar with flowJo and friends should add to
-  this...}
-
-{\bf NLM:} Only a couple of commercial software (e.g., FlowJo and
-DeNovo) allows the user to analyse batch of FCS files.  Only a few
-allows export in XML (e.g., FlowJo).
-
-{\bf RB:}  It would be important to highlight
-the differences between flowCore and commercial software packages,
-including in philosophy (open vs. closed source) and interaction
-(programmatic data analysis vs. GUI-driven).
-
-\section{Basic data structures}
+\section*{Basic data structures}
 \subsection*{flowFrame}
@@
 
-\section{Standard flow operations}
+\section*{Standard flow operations}
 The basic operations in flow cytometry analysis are almost always the
@@
 
-\section{Related flow packages}
+\section*{Related flow packages}
 In addition to the flowCore package that offers basic infrastructure,
@@
 
-{\bf NLM:} example: cite flowClust and floViz paper?
+\section*{Application}
+The flowCore package has already been applied in the analysis of a
+number of datasets \citep{gasparetto2004ice,brinkman2007hcf}, and
+several other Bioconductor packages are making use of its code
+base. As an example, \cite{lo2008agf} have recently proposed an
+automatic gating approach via robust model-based clustering using the
+flowCore data model and infrastructure (implemented in the
+Bioconductor package flowClust). Another package providing more
+specialized support for experiments conducted on microtitre plates is
+currently being developed. The documentation of all of flowCore's
+features is way beyond the scope of this publication. Instead, we
+want to use a single example to demonstrate some of the software's
+hallmarks, that is, the concept of data-driven automated gating, the
+integration of existing software into the framework and the generation
+of publication-quality graphics for data visualization. The data used
+for this example are flow measurements of human T-lymphocytes from 30
+different patients, stained for both the CD4 and CD8 markers.
 
-\section{Application, Outlook}
-\textit{This is the most important part, we need to show nice examples
-  of use cases, limit ourselfs to the basics and show what we can do
-  that nobody else can do at the moment. Need at lot of input here
-  from everybody. Also for the things that go in here we should
-  provide further details in the Supplementary Materials.}
+A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
+where each panel in the matrix represents the data of one individual
+patient. Instead of plotting individual points, the local bivariate
+density of the data is shown as a false color image. For each panel,
+the outlines of the regions identified in the automated gating
+operation described below are added to the plot. The input to the
+plotting function is either a single flow frame or a complete flow set
+object, and all available metadata can be accessed to control the
+appearance and the composition of the plot. Similarily, gate objects
+are directly passed on to the plotting function and are included in
+the plot if possible. Besides the very basic scatterplot shown here,
+there is a wealth of additional data visualizations available in the
+flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
+models allowing to easily arrange the data according to numerous
+experimental factors. Furthermore, the design and the API of the
+visualization software itself is very generic, and the user can
+readily extend its capabilities by providing self-defined plotting
+functions.
 
-{\bf NLM:} We could illustrate/summarize:
-\begin{itemize}
-\item the power of subsetting: for example to manipulate 
-patient below a certain age or of the same sex...
-\item Gating: filterSet strategy, kmeanFilter
-\item visualization adding metadata as parameter
-example: cite flowViz paper
-\end{itemize}
+Static gating for all samples in a high-throughput flow cytometry
+experiment is often not an option, since the measured variables tend
+to vary between different treatments or over time. Automated or
+data-driven gating tries to estimate the gating regions from the
+underlying data, thus providing a fast objective solution to the
+analysis of potentially very large and diverse data sets
+\cite{lo2008agf}. One of the automated gating methods implemented in
+flowCore is based on identifying areas of significant curvature in a
+kernel density estimate of the data (\ref{Wandetal}). Assuming that
+the regions of interest are of high density, the software is able to
+reliably detect them in either the one ore the two-dimensional density
+landscape. Kernel density estimation is a well-known problem in
+statistical computing, and a lot of effort has been invested in the
+development of good software to address it(\ref{Wandetal}). The
+modular design of R, Bioconductor and flowCore allows to easily
+integrate these existing solutions into our framework. Instead of
+re-writing existing code, we are able to include it via the well
+tested distribution mechanism provided by R's software package
+system. This process is bi-directional, and all funtionality
+implemented in flowCore is available to other package authors, as we
+have seen with the afore-mentioned flowClust package.
+
+In developing flowCore, we tried to provided a unified develoment and
+analysis platform for flow-cytometry data. The scope and complexity of
+the technology is evolving rapidly, and a collaborative effort in
+devising new methodolgy has been proven beneficial for a number of
+different biological and computational biology challenges. We hope
+that our framework will be the foundation for fruitful shared research
+by many calloborators from multiple scientific fields, further pushing
+the boundaries of the exiting and still promising flow
+cytometry technology.
 

### 2008-04-30 paper/outline.tex
illustration and citation
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31442 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \documentclass[12pt]{article}
+\usepackage{graphicx}
 \usepackage{url}
@@
 raw data FCS files along with their complete set of file-specific meta
-information.  Moreover, it is one of the very first software
+information (Figure~\ref{fig1:FrameWork}).  Moreover, it is one of the very first software
 implementations of the Gating-ML CR, which opens the possibility to
@@
 the flowSet structure can be readily extended to also accommodate such
-spatial information \citep{(plateCore)}
+spatial information [citing plateCore - Any reference here Perry, Errol?]
 
-\textit{Figure: A diagram of the data structures and the basic
-  operations (Figure1.ppt)}
+%%Figure1 Diagram of the data structures and the basic operations
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
+\caption{\label{fig1:FrameWork}{The flowCore frame work.For each experiement,
+  FCS files, phenotypic and meta data can be store in a flowSet. Each
+  flowFrame of a flowSet corresponds to one FCS file.  Any basic
+  operation (compensate, filter, ...) can be applied on a flowSet or a
+  flowFrame.Different summary output can be generated with flowCore or
+  the associated packages.}}
+\end{figure}
 
@@
 fostered the development of a number of valuable
-improvements. \citep{lo2008agf} (+plateCore).
+improvements \citep{lo2008agf, sarkar2008ufv}.
 
@@
 Proper transformation have been proven beneficial for both data
-visualization and modelling \textit{cite flowClust paper}. All the
+visualization and modelling \citep{lo2008agf}. All the
 major transformations that are routinely used in flow cytometry
@@
 we have implemented a range of additional Bioconductor packages that
-are dedicated to more specific tasks of flow cytometry data
-analysis. The flowViz package provides sophisticated data
-visualization tools, that make use of multivariate trellis
-plotting. These functions can be used to quickly generate customized
-plots for extended cytometry data sets for both direct data inspection
-and quality control. The flowQ package offers more advanced quality
-assurance (QA) methodology and a framework to create interactive
-web-based reports of QA results. Most of flowUtil's content deals with
-data import and output from and to flowCore and the flow-cytometry
-specific standard markup languages. Finally, flowStats provides
-statistical methods that are relevant in the context of flow cytometry
-data analysis.
+are dedicated to more specific tasks of flow cytometry data analysis
+(Figure~\ref{fig2:flowWheel}). The flowViz package provides
+sophisticated data visualization tools, that make use of multivariate
+trellis plotting. These functions can be used to quickly generate
+customized plots for extended cytometry data sets for both direct data
+inspection and quality control. The flowQ package offers more advanced
+quality assurance (QA) methodology and a framework to create
+interactive web-based reports of QA results. Most of flowUtil's
+content deals with data import and output from and to flowCore and the
+flow-cytometry specific standard markup languages. Finally, flowStats
+provides statistical methods that are relevant in the context of flow
+cytometry data analysis.
 
-\textit{A diagram of the packages and what they offer}
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{Figure2-flowWheel-version4.jpg}
+\caption{\label{fig2:flowWheel}{The flowCore Suite: visualization and analysis
+  packages relying on the flowCore data structure.  Dedicated packages
+  have been developped using the flowCore data structure for
+  manipulating the experiemental data: flowViz is a data visualization
+  package; flowQ helps at data quality assessment; flowUtils provides
+  xml parser, flowClust proposes population 
+  filtration using model-based clustering}}
+\end{figure}
 
@@
 there is a wealth of additional data visualizations available in the
-flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
+flowViz package \citep{sarkar2008ufv}, all making use of flowCore's data
 models allowing to easily arrange the data according to numerous
@@
 analysis of potentially very large and diverse data sets
-\cite{lo2008agf}. One of the automated gating methods implemented in
+\citep{lo2008agf}. One of the automated gating methods implemented in
 flowCore is based on identifying areas of significant curvature in a
-kernel density estimate of the data (\ref{Wandetal}). Assuming that
-the regions of interest are of high density, the software is able to
-reliably detect them in either the one ore the two-dimensional density
-landscape. Kernel density estimation is a well-known problem in
-statistical computing, and a lot of effort has been invested in the
-development of good software to address it(\ref{Wandetal}). The
-modular design of R, Bioconductor and flowCore allows to easily
-integrate these existing solutions into our framework. Instead of
-re-writing existing code, we are able to include it via the well
-tested distribution mechanism provided by R's software package
-system. This process is bi-directional, and all funtionality
-implemented in flowCore is available to other package authors, as we
-have seen with the afore-mentioned flowClust package.
+kernel density estimate of the data
+\citep{wand2008,wand2008_2}. Assuming that the regions of interest are
+of high density, the software is able to reliably detect them in
+either the one ore the two-dimensional density landscape. Kernel
+density estimation is a well-known problem in statistical computing,
+and a lot of effort has been invested in the development of good
+software to address it \citep{wand2008,wand2008_2}. The modular design
+of R, Bioconductor and flowCore allows to easily integrate these
+existing solutions into our framework. Instead of re-writing existing
+code, we are able to include it via the well tested distribution
+mechanism provided by R's software package system. This process is
+bi-directional, and all funtionality implemented in flowCore is
+available to other package authors, as we have seen with the
+afore-mentioned flowClust package.
 

### 2008-04-30 paper/manuscript.tex
manuscript.tex now holds the actual paper
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31448 bc3139a8-67e5-0310-9ffc-ced21a209358

--- 2008-04-29 paper/outline.tex
@@
 \usepackage[authoryear,round]{natbib}
+\usepackage{graphicx}
 
@@
 
-\vspace{2ex}
-
-The paper will deal with the capabilities offered by flowCore for
-handling flow cytometry data in R. The main focus should be on
-high-throughput and how our infrastructure can be used to deal with
-such data. We don't want to write a user manual but rather highlight
-the problems that arise from high throughput flow applications and the
-solutions we offer. A strong point should be on integration of our
-infrastructure in the flow world. flowCore is intended to be a
-development platform for other scientists, we don't want to provide
-full-fledged solutions but rather tools for others to devise such
-solutions.
-
-It would be nice to show these things on one data set in the form of a
-use case, but that is open to discussion. The problem with such an
-approach is that it often looks like a software manual which is not
-what we should be aiming for.
-
-\bigskip
-The author list is also still up for discussion, listed here are all
-the names that came up during the last telephone conference. NLM and
-FH are supposed to share first authorship.
-
-\bigskip
-Target journal should be Cytometry, and there are two flavors:
-Cytometry A deals with more general topics while Cytometry B is more
-geared towards clinical applications. Florian suggests Cytometry A,
-here is a link to the author instructions:
-\url{http://www3.interscience.wiley.com/journal/33945/home/ForAuthors.html}.
-They require the standard Materials and Methods, Results and
-Discussion structure which might not be suited for our needs. FH
-hasn't found out about alternative methods-type articles yet.
-
-{\bf NLM:} We can suggest to the editor new format for software/methods
-oriented paper?
-
-{\bf RB:} I know the editor and would be happy to speak to him
-regarding methods-type articles or other issues if needed.
-
-Also it is not clear from their web page whether they support LaTeX
-manuscript submissions but FH would strongly argue for using
-LaTeX. They do accept pdf submission.
-
-{\bf RB:} Cytometry does not accept LaTeX (we've tried). While they say
-they will accept PDF they made us convert PDF to Word in the past. I
-was able to do this reasonably well with a third-party software tool.
-
-\bigskip
-There is an svn repository set up at
-\url{https://hedgehog.fhcrc.org/gentleman/bioconductor/trunk/madman/Rpacks/flowCore/paper}
-which is also the repository for the flowCore source code. All
-contributors should already have usernames and passwords, if not
-please contact FH.
-
-\bigskip
-There have been discussions on how much technical detail we want to
-provide and how and where example should be given:
-
-{\bf PH:} Do we have a simple illustration of why we need to be able to
-program with flow data in order to do more meaningful science? Could
-we say something about finding populations or automated gating as
-being enabled by flowCore?  We have found the lattice plots and the
-ability to make overlays very helpful. Could we include an example
-where these capabilities would be helpful?
-
-{\bf RB:} Enabling automated gating is certainly true as we have just had a
-manuscript accepted for automated gating that relies on flowCore.
-
-{\bf PH:} It seems to me some examples of what flowCore makes possible needs
-to be at the heart of the paper. That doesn't mean that we need to
-show a detailed analysis.
-
-{\bf ES:} I also think that use cases/examples are very important for selling
-the software.  I don't believe we need a complete analysis vignette in
-the paper, but specific examples that I've found useful include
-\begin{itemize}
-\item Correcting for the effects of cell size on autofluoresence using
-  Florian's regression,
-\item Automating report generation (using Sweave), including making
-  overlay plots (supposed hard to batch this job in FlowJo?),
-\item automated gating,
-\item population finding (using mclust, other R density tools, etc).
-\end{itemize}
-
-{\bf RB:} The code examples in the supplementary material, but could
-be highlighted in the main text with the pretty results (the selling
-point to biologists), something like: “For example, with a short
-section of user-configurable code (see Supplementary Material) it is
-possible to automate the processing of 1000s of individual FCS data
-files and generate high quality reports (Figures 1, 2) including such
-steps as quality checking plots, correcting for cell size on auto
-fluorescence, followed by automated gating.  This code can be freely
-extended and modified by users as the underlying code is all available
-at no cost under an open source license ”.  I think it’s important
-to remember the audience. The people reading Cytometry we need to
-convince on the importance of flowCore are the biologists, and putting
-code examples in the main text won’t help, and will likely
-hinder. The programmers/statisticians/bioinformaticians out there will
-go to the code in the Supplementary Material/Vignettes section
-already. All the biologists need to see are the pretty pictures that
-let them interpret their data. Highlighting the support network
-available (Bioconductor list) will be important in terms of hand
-holding.  In terms of making biologists happy, it might be worth
-talking to Adam about plans to call R from FlowJo. At the least it
-should be highlighted that this is possible in the generic sense.
-
-{\bf PH:} agree that the code examples should go in the supplemental
-materials, and I like the examples that Ryan gives below. One thing
-that we have found to be very challenging to the biologists is that
-they no longer do the primary analysis of their data. We have
-addressed that by highlighting the ability to programmatically analyze
-lots of data, most of which has fairly straightforward results and
-then to focus the biologists attention only on the data that needs
-expert interpretation/tweaking.
-
-{\bf BE:} This is, unfortunately, somewhat of a problem---the biologist (as
-a community, I realize that there are specific individuals to whom
-this does not apply and that I'm likely especially sensitive to this
-phenomenon) doesn't believe that data can be analyzed apart from
-direct inspection. It might be worth stressing that flow* allows for
-not only automated analysis, but provides a platform for improving
-collaboration. Actually, it occurs to me that this is probably "the
-important bit" for the data structures, which are not particularly
-profound in and of themselves---they mostly take the form they do
-because that's the way Bioconductor does things in general. However,
-what they buy you is a standard mechanism for the organization of
-annotation of experiments that is suitable for programmatic
-manipulation rather than the usual ad hoc collection of 30 randomly
-formatted Excel spreadsheets that may or may not have anything to do
-with the actual flow data you have.
-
-{\bf PH:} I agree with that. We need to explain to the biologists that they
-"can't get there from here" unless they adopt better data organization
-and automated analysis and work collaboratively. It would be nice to
-think about how we would make that argument.
-
-\bigskip
-
 \maketitle
@@
 \section*{Introduction}
-
 Traditionally, flow cytometry (FCM) has been a tube-based technique
@@
 technologies must be stored, managed, and needs to be summarized in
-order to be accessible to the researcher. However, the absence of any
-shared research platform, which bioinformaticians, computer scientists,
+order to be accessible to the researcher. However, the absence of a
+shared research platform that bioinformaticians, computer scientists,
 and statisticians can utilize to develop novel or standard methods for
-flow cytometric analysis has become problematic. We believe that the open
-source statistical software R in conjunction with the Bioconductor
-Project can fill this gap.  In this paper, we propose R data
-structures to handle flow cytometry data through the main steps of
-importing, storing, assessing and preprocessing data from FC-HCS
-experiments. These structures, along with a range of specialized
-methods for compensation, transformation, gating and additional
-preprocessing steps, are implemented in the newly developed
-Bioconductor package flowCore.  We envision flowCore to be a solid
-foundation for researchers to efficiently handle FC-HCS data and for
-future development of new methods and tools for their coherent
-analysis.
+flow cytometric analysis has hindered innovation. We believe that the
+open source statistical software R in conjunction with the
+Bioconductor Project can fill this gap.  In this paper, we propose R
+data structures to handle flow cytometry data through the main steps
+of importing, storing, assessing and preprocessing. These structures,
+along with a range of specialized methods for compensation,
+transformation, gating and additional preprocessing steps, are
+implemented in the newly developed Bioconductor package flowCore.  We
+envision flowCore to be a solid foundation for researchers to
+efficiently handle FC-HCS data and for future development of new
+methods and tools for their coherent analysis.
 
 \subsection*{R and Bioconductor}
-R is a solid statistical programing environment, particularly geared
-towards the analysis of large numeric datasets typical to arise from
-high throughput experiments such as microarray gene expression
-analysis or FC-HCS. Today, the R open source project offers a wide
+R is a robust statistical programing environment which offers a wide
 range of statistical and visualization methods developed for various
-fields of application. In addition, R is a research platform which
-bioinformaticians, computer scientists, and statisticians routinely
-use to develop novel tools and methodology for statistical data
-analysis. For biological applications, the Bioconductor project has
-become the standard toolset \citep{gentleman2006bos} to process data
-spanning a diversity of research fields from genomics to proteomics to
-complex cell biology. This shared platform proved to be particularly
-beneficial for the development of analysis routines for microarray
-gene expression data, and by today many of these early innovations
-have been incorporated into commercial software products, thus greatly
-improving the quality of gene expression data analysis. In this
-context, shared data structures have been shown to be essential for
-such a collaborative effort.
+fields of application. In addition, R is an open source research
+platform which bioinformaticians, computer scientists, and
+statisticians routinely use to develop novel tools and methodology for
+statistical data analysis. For biological applications, the
+Bioconductor project has become the standard toolset
+\citep{gentleman2006bos} to process data spanning a diversity of
+research fields from genomics to proteomics to complex cell
+biology. It is particularly geared towards the analysis of large
+numeric datasets that typically arise from high throughput experiments
+such as microarray gene expression analysis or FC-HCS. A shared
+platform proved to be particularly beneficial for the development of
+analysis routines for microarray gene expression data, and by today
+many of these early innovations have been incorporated into commercial
+software products, thus greatly improving the quality of gene
+expression data analysis. Shared data structures have been shown to be
+essential for such a collaborative effort. We envision a similar
+paradigm for FC-HCS data.
 
@@
 files following a well-defined data standard (FCS). This standard
-attempts to capture as much of the accompanying meta information as
-possible, most of which is produced by the measurement instrument or
-has to be supplied manually by the operator upon experiment
-setup. However, the recent developments in high-throughput flow
-cytometry are shifting the focus of interest away from single-tube
-based measurements towards large and complex experimental designs with
-dozens of co-variates and influencing factors. Self-contained and
-self-reflective data structures are a prerequisite to allow for
-coordinated analysis of such experiments. Most of the currently
-available software solutions offer only limited support for such
-structures, or make use of containers like XML or binary storage
-formats that are designed specifically for the mostly GUI-driven user
-interfaces and hence are not easily amenable to programmatical
-access. In addition, the closed-source nature of most of these
-products makes it virtually impossible to connect and integrated the
-data.
+attempts to capture much of the accompanying meta information, most of
+which is produced by the measurement instrument or has to be supplied
+manually by the operator upon experiment setup. However, the recent
+developments in high-throughput flow cytometry are shifting the focus
+of interest away from single-tube based measurements towards large and
+complex experimental designs with dozens of co-variates and
+influencing factors. Self-contained and self-reflective data
+structures are a prerequisite to allow for coordinated analysis of
+such experiments. Most of the currently available software solutions
+offer only limited support for such structures, or make use of
+containers like XML or binary storage formats that are designed
+specifically for the mostly GUI-driven user interfaces and hence are
+not easily amenable to programmatical access. In addition, the
+closed-source nature of most of these products makes it impratical to
+connect to and integrated the data into an analysis pipeline.
 
 Traditionally, the majority of flow cytometry experiments have been
-analysed manually by direct data inspection in one ore tow dimensions,
+analyzed manually by direct data inspection in one or two dimensions,
 or by very basic comparisons of summary statistics. We believe that
-these approaches do not give justice to the highly complex nature of
-FCM data, in particualar, they discard many of the fundamental aspects
-such as underlying distribution of the data or the
-high-dimensionality. In addition, the subjective character of manual
-analyses are a major obstacle for reproducibility. For FC-HCS data,
-unassisted manual inspection is no longer feasible, and robust
+these approaches do not fully address the highly complex nature of FCM
+data, in particular, they discard many of the fundamental aspects of
+the data, such as its underlying distribution and its high-dimensional
+nature. In addition, the subjective character of manual analyses are a
+major obstacle to reproducibility. For FC-HCS data, unassisted manual
+inspection has become way too time and labor intensive, and robust
 statistical methods need to be developed to point the investigator to
@@
 greatly improve the relevance of flow cytometry in today's
-high-throughput biological sciences.
+high-throughput paradigm.
 
@@
 transformed parameters as described by an explicit parameter
-transformation. This allows to encode a simple analytical work flow,
-which is sufficient in order to reconstruct the analysis
+transformation. This enables the encoding of a simple analytical work
+flow, in a way that allows us to reconstruct the analysis
 programmatically.
@@
 raw data FCS files along with their complete set of file-specific meta
-information.  Moreover, it is one of the very first software
+information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
 implementations of the Gating-ML CR, which opens the possibility to
@@
 virtually every other FCM tool that adheres to the proposed
-standard. In addition, the software is platform independent, running
-on various operating systems including Windows, Mac OS X and
-Linux/Unix.
+standard. In addition, the software is platform agnostic, running on
+Windows, Mac OS X and Linux/Unix operating systems.
 
-flowCore is not a GUI-driven software, and all operations are
-currently done using a command line interface. Conceptually, it is
-possible to add a more elaborate user interface on top of that,
-however we strongly focus on programmatic access to enable the
-convenient development of novel analysis methods. By taking the burden
-of data management from the programmer and by providing well-defined
-APIs and a structured class hierarchy, it is possible to readily test
-new ideas and to easily extend the frameworks functionality. Both the
-underlying R and Bioconductor software and flowCore itself are open
-source and open development projects, thus community involvement in the
-developing process is highly appreciated.
+flowCore is not a GUI-driven software, and all operations are done
+using a command line interface. Conceptually, it is possible to add a
+more elaborate user interface on top of that, however the focus is on
+a programmatic approach to enable the convenient development of novel
+analysis methods. By taking the burden of data management from the
+programmer and by providing well-defined APIs and a structured class
+hierarchy, it is possible to readily test new ideas and to easily
+extend the frameworks functionality. Both the underlying R and
+Bioconductor software and flowCore itself are open source and open
+development projects, thus community involvement in the developing
+process is highly appreciated.
 
 \section*{Basic data structures}
-\subsection*{flowFrame}
+\subsection*{Flow Frame}
 flowCore's primary task is the representation and basic manipulation
@@
 in flowCore is a collection of events. We call the structure that hold
-this data a flowFrame. All relevant information for the collection are
-stored along with the matrix of raw data values and can be accessed
-programmatically. Most prominently, these are descriptors of the
+this data a flow frame. All relevant information for the collection
+are stored along with the matrix of raw data values and can be
+accessed programmatically. Most commonly, these are descriptors of the
 stains used in the experiment and the respective measurement channels,
@@
 data. A number of quality checks are performed during the creation of
-a flowFrame to ensure data integrity. There is an extended interface
+a flow frame to ensure data integrity. There is an extended interface
 that allows for the standardized interaction with the flow data, thus
 enabling the development of more specialized tools for specific tasks
-of the data analysis process.
+of the data analysis process that all share a common API.
 
-\subsection*{flowSet}
+\subsection*{Flow Set}
 In high-throughput flow cytometry, many of the analysis tasks need to
 be performed consistently across multiple flow measurements, hence we
-introduce the concept of a collection of flowFrames called a
-flowSet. flowSets store the relevant information associated with each
+introduce the concept of a collection of flow frames called a flow
+set. Flow sets store the relevant information associated with each
 individual frame such as descriptions of the samples or the treatment
 to which a sample was subjected, and they manage the consistent
-application of operations on the individual flowFrames. The burden of
+application of operations on the individual flow frames. The burden of
 keeping score of the annotation is shifted from the user to the
 infrastructure, thus reducing the risk of mixups and errors that are
-all too common when dealing with complex data in the ususal way:
-namely, more or less ad hoc spreadsheets that lack a controlled
-structure. In general, there is no native ordering to a flowSet, which
-in a way reflects the still predominant application of tube-based
-experiments rather than experiments following a plate-design. However,
-the flowSet structure can be readily extended to also accommodate such
-spatial information \citep{(plateCore)}
+all too common when using more or less ad hoc spreadsheets that lack a
+controlled structure. In general, there is no native ordering to a
+flow set, which in a way reflects the still predominant application of
+tube-based experiments rather than experiments following a
+plate-design. However, the flow set structure can be readily extended
+to also accommodate such spatial information \textit{(plateCore
+  reference, Perry, Errol?)}.
 
-\textit{Figure: A diagram of the data structures and the basic
-  operations (Figure1.ppt)}
+%%Figure1 Diagram of the data structures and the basic operations
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
+\caption{\label{fig1:FrameWork}{The flowCore frame work.For each
+    experiement, FCS files, phenotypic and meta data is store in a
+    flow set. Each flow frame of a flow set corresponds to one FCS
+    file. All basic operation (compensation, transformation, gating,
+    ...) can be applied on both single flow frames or entire flow
+    set.}}
+\end{figure}
 
 \section*{Standard flow operations}
-The basic operations in flow cytometry analysis are almost always the
+The basic operations in flow cytometry analysis are typically the
 same: the data need to be compensated (if not already done on the
@@
 flowCore is to abstractly describe these operations and build a set of
-tools to perform them on both flowFrames as well as on entire
-flowSets. While transformation and to a certain extent also
-compensation are fairly routine operations with only limited potential
-for improvement, being able to implement new methodology for gating of
-flow cytometry data and extending the capabilities of flowCore through
-the usual ways of object oriented programming is a feature, that
-clearly sets our framework apart from all the other flow toolsets
-around. By factoring out as much of the bookkeeping as possible,
-programmers can focus on the actual operations rather then having to
-deal with the tedious details of data integration and access. Third
-party methods implemented by others can act on their own as
-first-class citizens in the analysis framework, without breaking the
-workflow or the basic infrastructure. This design allows for the
-straight forward extension of flowCore's capabilities and already
-fostered the development of a number of valuable
-improvements. \citep{lo2008agf} (+plateCore).
+tools to perform them on both flow frames as well as on entire flow
+sets. While transformation and to a certain extent also compensation
+are fairly routine operations with only limited potential for
+improvement, being able to implement new methodology for gating of
+flow cytometry data, and extending the capabilities of flowCore
+through object oriented programming are features that clearly sets our
+framework apart from other flow toolsets around. By factoring out as
+much of the bookkeeping as possible, programmers can focus on the
+actual operations rather then having to deal with the tedious details
+of data integration and access. Third party methods implemented by
+others can act on their own as first-class citizens in the analysis
+framework, without breaking the workflow or the basic
+infrastructure. This design allows for the straightforward extension
+of flowCore's capabilities and already fostered the development of a
+number of valuable improvements. \citep{lo2008agf,sarkar2008ufv}
+\textit{(plateCore reference, Perry, Errol?)}.
 
@@
 Proper transformation have been proven beneficial for both data
-visualization and modelling \textit{cite flowClust paper}. All the
-major transformations that are routinely used in flow cytometry
-analysis have been implemented in flowCore. Furthermore, the design of
-the R language makes it easy to define arbitrary functions to apply to
-the data of individual flowFrames or entrire flowSets,
-respectively. Compensation is available for both flowFrames and
-flowSets, and the software also offers functionality to compute
-spillover or compensation matrices from a set of appropriate
-compensation samples.
+visualization and modelling \citep{lo2008agf}. The major
+transformations that are routinely used in flow cytometry analysis
+have been implemented in flowCore. Furthermore, the design of the R
+language makes it easy to define arbitrary functions to apply to the
+data of individual flow frames or entire flow sets,
+respectively. Compensation is available for both flow frames and flow
+sets, and the software also offers functionality to compute spillover
+or compensation matrices from a set of appropriate compensation
+samples.
 
 \subsection*{Gating}
-In flowCore, gates are represented as individual classes that can be
-extended in an object oriented manner. All basic gate types such as
+In flowCore, gating operations are represented by classes that can be
+extended in an object-oriented manner. Basic gate types such as
 rectangular gates, ellipses and polygon gates are implemented as part
-of the base framework. In addition, we introduce the notion of
-data-driven gates, or filters, for which the necessary parameters are
-computed based on the properties of the underlying data, for instance
-by modelling underlying data distribution or by density
-estimation. The ability to programmatically access gates is a
-prerequisite for automated or semi-automated gating. By utilizing a
-unified interface for all different types of gates, the user is able
-to subset data sets as well as to create summary statistics, like the
-proportions of events falling in single gates or in the combinations
-of multiple gates. Conmplex combinations and hierarchies of gates can
-be captured in objects of class filterSet, allowing to apply
-multi-step gating strategies. The definition of gates in flowCore
-follows the Gating-ML CR, thus any flowCore gating strategy can be
-reproduced by any other software that also adheres to the standard.
+of the framework. In addition, we introduce the notion of data-driven
+gates, or filters, for which the necessary parameters are computed
+based on the properties of the underlying data, for instance by
+modelling underlying data distribution or by density estimation. The
+ability to programmatically access gates is a prerequisite for
+automated or semi-automated gating. By utilizing a unified interface
+for all different types of gates, the user is able to subset data sets
+as well as to create summary statistics, like the proportions of
+events falling in single gates or in the combinations of multiple
+gates. Complex combinations and hierarchies of gates can be captured
+in objects of class filterSet, allowing to apply multi-step gating
+strategies. The definition of gates in flowCore follows the Gating-ML
+CR, thus any flowCore gating strategy can be reproduced by any other
+software that also adheres to the standard.
 
@@
 we have implemented a range of additional Bioconductor packages that
-are dedicated to more specific tasks of flow cytometry data
-analysis. The flowViz package provides sophisticated data
-visualization tools, that make use of multivariate trellis
-plotting. These functions can be used to quickly generate customized
-plots for extended cytometry data sets for both direct data inspection
-and quality control. The flowQ package offers more advanced quality
-assurance (QA) methodology and a framework to create interactive
-web-based reports of QA results. Most of flowUtil's content deals with
-data import and output from and to flowCore and the flow-cytometry
-specific standard markup languages. Finally, flowStats provides
-statistical methods that are relevant in the context of flow cytometry
-data analysis.
+are dedicated to more specific tasks of flow cytometry data analysis
+(Figure \ref{fig2:flowWheel}). The flowViz package provides
+sophisticated data visualization tools, that make use of multivariate
+trellis plotting. These functions can be used to quickly generate
+customized plots for extended cytometry data sets for both direct data
+inspection and quality control. The flowQ package offers more advanced
+quality assurance (QA) methodology and a framework to create
+interactive web-based reports of QA results. Most of flowUtil's
+content deals with data import and output from and to flowCore and the
+flow-cytometry specific standard markup languages. Finally, flowStats
+provides statistical methods that are relevant in the context of flow
+cytometry data analysis.
 
-\textit{A diagram of the packages and what they offer}
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{Figure2-flowWheel-version4.jpg}
+\caption{\label{fig2:flowWheel}{flowCore and assoziated Bioconductor
+    packages: dedicated packages have been developped that utilize
+    flowCore’s data structures. flowViz is a data visualization
+    package; flowQ provides data quality assessment methods; flowUtils
+    contains interfaces for flow data, including xml parsers:
+    flowClust provides automated gating by model-based clustering.}}
+\end{figure}
 
@@
 currently being developed. The documentation of all of flowCore's
-features is way beyond the scope of this publication. Instead, we
-want to use a single example to demonstrate some of the software's
+features is way beyond the scope of this publication. Instead, we want
+to use a single example to demonstrate some of the software's
 hallmarks, that is, the concept of data-driven automated gating, the
@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=0.75\textwidth]{xyplot}
+\caption{\label{xyplot}%
+Scatterplot matrix of a single Flow Set with outlines of the gating
+regions identified by an automated gating operation.}
+\end{figure}
+
 Static gating for all samples in a high-throughput flow cytometry
@@
 flowCore is based on identifying areas of significant curvature in a
-kernel density estimate of the data (\ref{Wandetal}). Assuming that
+kernel density estimate of the data (\citep{wand2008}). Assuming that
 the regions of interest are of high density, the software is able to
@@
 statistical computing, and a lot of effort has been invested in the
-development of good software to address it(\ref{Wandetal}). The
-modular design of R, Bioconductor and flowCore allows to easily
-integrate these existing solutions into our framework. Instead of
-re-writing existing code, we are able to include it via the well
-tested distribution mechanism provided by R's software package
-system. This process is bi-directional, and all funtionality
-implemented in flowCore is available to other package authors, as we
-have seen with the afore-mentioned flowClust package.
+development of good software to address it. The modular design of R,
+Bioconductor and flowCore allows to easily integrate these existing
+solutions into our framework. Instead of re-writing existing code, we
+are able to include it via the well tested distribution mechanism
+provided by R's software package system. This process is
+bi-directional, and all funtionality implemented in flowCore is
+available to other package authors, as we have seen with the
+afore-mentioned flowClust package.
 
@@
 by many calloborators from multiple scientific fields, further pushing
-the boundaries of the exiting and still promising flow
-cytometry technology.
+the boundaries of the exiting and still promising flow cytometry
+technology.
 

### 2008-04-30 paper/manuscript.tex
more specific about sotware imports now
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31449 bc3139a8-67e5-0310-9ffc-ced21a209358
fixed some typos
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31450 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 flowCore is based on identifying areas of significant curvature in a
-kernel density estimate of the data (\citep{wand2008}). Assuming that
+kernel density estimate of the data \citep{wand2008}. Assuming that
 the regions of interest are of high density, the software is able to
@@
 Bioconductor and flowCore allows to easily integrate these existing
-solutions into our framework. Instead of re-writing existing code, we
-are able to include it via the well tested distribution mechanism
-provided by R's software package system. This process is
-bi-directional, and all funtionality implemented in flowCore is
-available to other package authors, as we have seen with the
-afore-mentioned flowClust package.
+solutions into our framework. In this example, we directly use R code
+from the feature package developed by Wand et al. Instead of
+re-writing existing code, we are able to include it via the well
+tested distribution mechanism provided by R's software package
+system. This process is bi-directional, and all funtionality
+implemented in flowCore is available to other package authors, as we
+have seen with the afore-mentioned flowClust package.
 

### 2008-05-07 paper/manuscript.tex
Changed the introduction a little and added a few sentences about plateCore.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31535 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 limited to small-scale laboratory and clinical studies.  High
-throughput methods for flow cytometry (FC-HCS) have recently been
-developed and are now used in both basic and clinical research. Today,
-the need for flexible and well structured computational tools to
-efficiently handle and analyze FC-HCS has increased
-considerably. Indeed, the amount of information generated by these
-technologies must be stored, managed, and needs to be summarized in
-order to be accessible to the researcher. However, the absence of a
-shared research platform that bioinformaticians, computer scientists,
-and statisticians can utilize to develop novel or standard methods for
-flow cytometric analysis has hindered innovation. We believe that the
-open source statistical software R in conjunction with the
-Bioconductor Project can fill this gap.  In this paper, we propose R
-data structures to handle flow cytometry data through the main steps
-of importing, storing, assessing and preprocessing. These structures,
-along with a range of specialized methods for compensation,
-transformation, gating and additional preprocessing steps, are
-implemented in the newly developed Bioconductor package flowCore.  We
-envision flowCore to be a solid foundation for researchers to
-efficiently handle FC-HCS data and for future development of new
-methods and tools for their coherent analysis.
+throughput methods for flow cytometry high content screening (FC-HCS)
+have recently been developed and are now used in both basic and
+clinical research. As a result, the need for flexible and well
+structured computational tools to efficiently handle and analyze
+FC-HCS has increased considerably.
+However, the absence of a shared research platform that
+bioinformaticians, computer scientists, and statisticians can collaboratively
+use has hindered the development of novel methods 
+for FCM data analysis \citep{lizard2007fca}.
+We believe that the open source statistical
+software R, in conjunction with the Bioconductor Project can fill this
+gap.  
+%% EAS: I think we mention flowCore before talking about data structures and
+% methods.  We should first say explicity that we have a solution for
+% the missing shared platform.
+We have developed the Bioconductor package flowCore as a freely available,
+sophisticated and extensible FCM data analysis platform that enables
+researchers to efficiently handle FC-HCS data and encourages open
+development of tools for their coherent analysis. 
+%In this paper, we propose R data structures to handle FCM data
+%through the main steps of importing, storing, assessing and
+%preprocessing. 
+%These structures, along with a range of specialized
+%methods for compensation, transformation, gating and additional
+%preprocessing steps, are implemented in the newly developed
+%Bioconductor package flowCore.  flowCore is a freely available,
+%sophisticated and extensible FCM data analysis platform that enables
+%researchers to efficiently handle FC-HCS data and enables open
+%development of new methods and tools for their coherent
+%analysis.
+flowCore runs on Windows, Mac OS X and Linux/Unix operating
+systems.
 
@@
 range of statistical and visualization methods developed for various
-fields of application. In addition, R is an open source research
-platform which bioinformaticians, computer scientists, and
+fields of application. In addition, R is an open source research %already mentioned R, and by extention flowCore open source
+platform that bioinformaticians, computer scientists, and
 statisticians routinely use to develop novel tools and methodology for
@@
 software products, thus greatly improving the quality of gene
-expression data analysis. Shared data structures have been shown to be
+expression data analysis. 
+%% NLM: that's probably the important point: platform for innovation rather than for standardization
+Shared data structures have been shown to be
 essential for such a collaborative effort. We envision a similar
@@
 \subsection*{Existing data standards and conventions}
-Currently, data from flow cytometry experiments are stored in single
-files following a well-defined data standard (FCS). This standard
-attempts to capture much of the accompanying meta information, most of
-which is produced by the measurement instrument or has to be supplied
-manually by the operator upon experiment setup. However, the recent
-developments in high-throughput flow cytometry are shifting the focus
+Currently, data from FCM experiments are stored in single files
+according to the Flow Cytometry Standard (FCS) \citep{seamer1997pnd}.
+%This standard attempts to capture much of the accompanying meta information, most of which is produced by the measurement instrument or has to be supplied manually by the operator upon experiment setup.
+- I believe FCS only tries to capture a subset of meta information
+However, the recent
+developments in high-throughput FCM are shifting the focus
 of interest away from single-tube based measurements towards large and
@@
 not easily amenable to programmatical access. In addition, the
-closed-source nature of most of these products makes it impratical to
-connect to and integrated the data into an analysis pipeline.
+closed-source nature of %most of <can you think of an example that is not closed source?>
+these products makes it impractical to
+connect to and integrate the data into an analysis pipeline.
 
-Traditionally, the majority of flow cytometry experiments have been
-analyzed manually by direct data inspection in one or two dimensions,
-or by very basic comparisons of summary statistics. We believe that
-these approaches do not fully address the highly complex nature of FCM
-data, in particular, they discard many of the fundamental aspects of
-the data, such as its underlying distribution and its high-dimensional
+Traditionally, the majority of FCM experiments have been analyzed
+manually by direct data inspection in one or two dimensions, or by
+very basic comparisons of summary statistics. We believe that these
+approaches do not fully address the highly complex nature of FCM data,
+in particular, they discard many of the fundamental aspects of the
+data, such as its underlying distribution and its high-dimensional
 nature. In addition, the subjective character of manual analyses are a
 major obstacle to reproducibility. For FC-HCS data, unassisted manual
-inspection has become way too time and labor intensive, and robust
+inspection has become time consuming and labor intensive, and robust
 statistical methods need to be developed to point the investigator to
@@
 the expert knowledge of immunologists and researchers remains crucial
-for the understanding of FCM data, we believe that tight collaboration
+for the understanding of FCM data, we believe that close collaboration
 with other research fields such as statistics and computer science can
-greatly improve the relevance of flow cytometry in today's
-high-throughput paradigm.
+greatly improve the relevance of FCM in today's high-throughput
+paradigm.
 
@@
 cytometry data standards developed in collaboration with the ISAC Data
-Standards Task Force.  The Gating-ML Candidate Recommendation (CR)
+Standards Task Force.  For example, the Gating-ML Candidate Recommendation (CR)
 represents a proposal on how to form unambiguous XML-based gate
@@
 transformed parameters as described by an explicit parameter
-transformation. This enables the encoding of a simple analytical work
-flow, in a way that allows us to reconstruct the analysis
-programmatically.
+transformation. 
+%% I find the previous sentence redundant - 
+%% They may be applied on raw or transformed parameters
+%% EAS: I don't believe the previous sentence is required.
+This enables the encoding of a simple analytical workflow, in a way
+that allows us to reconstruct the analysis programmatically.
 
@@
 information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
-implementations of the Gating-ML CR, which opens the possibility to
-integrate flowCore in existing work flows and to communicate with
-virtually every other FCM tool that adheres to the proposed
-standard. In addition, the software is platform agnostic, running on
-Windows, Mac OS X and Linux/Unix operating systems.
+implementation of the Gating-ML CR, which opens the possibility to
+integrate flowCore in existing workflows and to communicate with
+every other FCM tool that adheres to the proposed
+standard.
 
@@
 analysis methods. By taking the burden of data management from the
-programmer and by providing well-defined APIs and a structured class
+programmer, and by providing well-defined APIs and a structured class
 hierarchy, it is possible to readily test new ideas and to easily
-extend the frameworks functionality. Both the underlying R and
-Bioconductor software and flowCore itself are open source and open
-development projects, thus community involvement in the developing
-process is highly appreciated.
+extend the frameworks functionality. %Both the underlying R and
+%Bioconductor software and flowCore itself are open source and open %already mentioned its open source
+%development projects, thus community involvement in the developing
+%process is highly appreciated.
 
@@
 flowCore's primary task is the representation and basic manipulation
-of flow cytometry data. This is accomplished through a data model very
+of FCM data. This is accomplished through a data model very
 close to that adopted by other Bioconductor packages used for gene
 expression analysis, and thus familiar to most Bioconductor
-users. Similar to any other flow tool, the basic unit of organization
+users. Similar to any other FCM data analysis tool, the basic unit of organization
 in flowCore is a collection of events. We call the structure that hold
@@
 enabling the development of more specialized tools for specific tasks
-of the data analysis process that all share a common API.
+of the data analysis process that all share a common Application Program Interface (API).
 
 \subsection*{Flow Set}
-In high-throughput flow cytometry, many of the analysis tasks need to
-be performed consistently across multiple flow measurements, hence we
+In high-throughput FCM, many of the analysis tasks need to be
+performed consistently across multiple flow measurements, hence we
 introduce the concept of a collection of flow frames called a flow
@@
 keeping score of the annotation is shifted from the user to the
-infrastructure, thus reducing the risk of mixups and errors that are
-all too common when using more or less ad hoc spreadsheets that lack a
-controlled structure. In general, there is no native ordering to a
-flow set, which in a way reflects the still predominant application of
-tube-based experiments rather than experiments following a
-plate-design. However, the flow set structure can be readily extended
-to also accommodate such spatial information \textit{(plateCore
-  reference, Perry, Errol?)}.
+infrastructure, thus reducing the risk of common errors % (\textit{i.e.}, example of an error ) 
+that occur when using more or less ad hoc spreadsheets that
+lack a controlled structure. There is no native ordering
+to a flow set, which in a way reflects the still predominant
+application of tube-based experiments rather than experiments
+following a plate-design. 
+%% EAS: I don't think the previous sentence is required. Does it 
+%% matter if the flowSet is ordered or not?
+
+%However, the flow set structure can be
+%readily extended to also accommodate such spatial information
+%% EAS: The spatial aspect of plateCore is really secondary to having 
+%% an easy, consistent way to manage and access the mulitplex annotation.
+%% Also, it's important to have a standard format for the annotation if
+%% we're trying to encourage collaboration. I haven't done it yet, but I 
+%% intend to look into
+%% into stealing the annotation implementation from CellHTS2 and
+%% using it in plateCore.
+However, the flow set structure can be readily extended to incorporate the
+potentially complex annotation associated with large FC-HCS experiments. Access
+to flowCore data structures and methods has already enabled the development of
+new Bioconductor pacakges, such as plateCore, that faciliate the analysis of
+plate-based FC-HCS data and make handling the annotation and processing
+the data easier.
+
+%\textit{(plateCore reference, Perry, Errol?)}.
 
@@
 \includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
-\caption{\label{fig1:FrameWork}{The flowCore frame work.For each
-    experiement, FCS files, phenotypic and meta data is store in a
+\caption{\label{fig1:FrameWork}{The flowCore frame work. For each
+    experiment, FCS files, phenotypic and meta data is stored in a
     flow set. Each flow frame of a flow set corresponds to one FCS
-    file. All basic operation (compensation, transformation, gating,
-    ...) can be applied on both single flow frames or entire flow
+    file. All basic operations (e.g., compensation, transformation,
+    gating) can be applied on both single flow frames or entire flow
     set.}}
@@
 \section*{Standard flow operations}
-The basic operations in flow cytometry analysis are typically the
-same: the data need to be compensated (if not already done on the
-instrument side), often transformed and sup-population of interest
-need to be selected based on a set of (predominantly sequential)
-gates. All software solutions for flow data analysis offer more or
-less elaborate support for these kinds of operation, most of them in
-an interactive, GUI-driven spirit. The approach we have taken in
-flowCore is to abstractly describe these operations and build a set of
-tools to perform them on both flow frames as well as on entire flow
-sets. While transformation and to a certain extent also compensation
-are fairly routine operations with only limited potential for
-improvement, being able to implement new methodology for gating of
-flow cytometry data, and extending the capabilities of flowCore
-through object oriented programming are features that clearly sets our
-framework apart from other flow toolsets around. By factoring out as
-much of the bookkeeping as possible, programmers can focus on the
-actual operations rather then having to deal with the tedious details
-of data integration and access. Third party methods implemented by
-others can act on their own as first-class citizens in the analysis
-framework, without breaking the workflow or the basic
-infrastructure. This design allows for the straightforward extension
-of flowCore's capabilities and already fostered the development of a
-number of valuable improvements. \citep{lo2008agf,sarkar2008ufv}
-\textit{(plateCore reference, Perry, Errol?)}.
+The basic operations in FCM analyses are typically the same: the data
+need to be compensated (if not already done on the instrument side)
+and transformed, and sup-populations of interest need to be selected
+based on a set of (predominantly sequential) gates. All software
+solutions for FCM analysis offer more or less elaborate support for
+these kinds of operation, and most of them in an interactive,
+GUI-driven interface. The approach we have taken in flowCore is to
+abstractly describe these operations and build a set of tools to
+perform them on both flow frames and flow sets. While transformation,
+and to a certain extent compensation, are fairly routine operations
+with only limited potential for improvement, being able to implement
+new methodologies for gating of FCM data, and extending the
+capabilities of flowCore through object oriented programming are
+features that clearly sets our framework apart from other FCM analysis
+tools. By factoring out as much of the bookkeeping as possible,
+programmers can focus on the actual operations rather then having to
+deal with the tedious details of data integration and
+access. Third-party methods can act on their own as first-class
+citizens in the analysis framework, without breaking the workflow or
+the basic infrastructure. This design allows for the straightforward
+extension of flowCore's capabilities, and has already fostered the
+development of a number of valuable add-ons
+\citep{lo2008agf,sarkar2008ufv} \textit{(plateCore reference, Perry,
+Errol?)}.
 
 \subsection*{Transformation and compensation}
-Proper transformation have been proven beneficial for both data
-visualization and modelling \citep{lo2008agf}. The major
-transformations that are routinely used in flow cytometry analysis
-have been implemented in flowCore. Furthermore, the design of the R
-language makes it easy to define arbitrary functions to apply to the
-data of individual flow frames or entire flow sets,
-respectively. Compensation is available for both flow frames and flow
-sets, and the software also offers functionality to compute spillover
-or compensation matrices from a set of appropriate compensation
-samples.
+Data transformation is essential for both data visualization and
+modelling \citep{lo2008agf}. The major transformations that are
+routinely used in FCM analysis have been implemented in flowCore
+(\textit{e.g.,} quadratic, log or arcsinh). Furthermore, the design of the R language makes
+it easy to define arbitrary functions to apply to the data of
+individual flow frames or entire flow sets, respectively. Compensation
+is available for both flow frames and flow sets, and the software also
+offers functionality to compute spillover or compensation matrices
+from a set of appropriate compensation samples.
 
@@
 as well as to create summary statistics, like the proportions of
-events falling in single gates or in the combinations of multiple
+events falling in single gates or in the combinations of
 gates. Complex combinations and hierarchies of gates can be captured
@@
 we have implemented a range of additional Bioconductor packages that
-are dedicated to more specific tasks of flow cytometry data analysis
+are dedicated to more specific tasks of FCM data analysis
 (Figure \ref{fig2:flowWheel}). The flowViz package provides
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=0.6\textwidth]{Figure2-flowWheel-version4.jpg}
-\caption{\label{fig2:flowWheel}{flowCore and assoziated Bioconductor
-    packages: dedicated packages have been developped that utilize
-    flowCore’s data structures. flowViz is a data visualization
-    package; flowQ provides data quality assessment methods; flowUtils
-    contains interfaces for flow data, including xml parsers:
-    flowClust provides automated gating by model-based clustering.}}
-\end{figure}
+%\begin{figure}
+%\centering
+%\includegraphics[width=0.6\textwidth]{Figure2-flowWheel-version4.jpg}
+%\caption{\label{fig2:flowWheel}{flowCore and assoziated Bioconductor
+%    packages: dedicated packages have been developed that utilize
+%    flowCore’s data structures. flowViz is a data visualization
+%    package; flowQ provides data quality assessment methods; flowUtils
+%    contains interfaces for flow data, including xml parsers:
+%    flowClust provides automated gating by model-based clustering.}}
+%\end{figure}
 
 \section*{Application}
-The flowCore package has already been applied in the analysis of a
-number of datasets \citep{gasparetto2004ice,brinkman2007hcf}, and
+The flowCore package has successfully been applied in the analysis of
+several datasets \citep{gasparetto2004ice,brinkman2007hcf}, and
 several other Bioconductor packages are making use of its code
-base. As an example, \cite{lo2008agf} have recently proposed an
+base. As an example, \cite{lo2008agf} have recently developed an
 automatic gating approach via robust model-based clustering using the
-flowCore data model and infrastructure (implemented in the
-Bioconductor package flowClust). Another package providing more
-specialized support for experiments conducted on microtitre plates is
-currently being developed. The documentation of all of flowCore's
-features is way beyond the scope of this publication. Instead, we want
-to use a single example to demonstrate some of the software's
-hallmarks, that is, the concept of data-driven automated gating, the
-integration of existing software into the framework and the generation
-of publication-quality graphics for data visualization. The data used
-for this example are flow measurements of human T-lymphocytes from 30
-different patients, stained for both the CD4 and CD8 markers.
+flowCore data model and infrastructure and implemented this in the
+Bioconductor package flowClust. Another package (plateCore) providing
+more specialized support for experiments conducted on microtitre
+plates is under development. The documentation of all of flowCore's
+features is beyond the scope of this publication, however additional
+documentation and users guide information with programmatic examples
+is available <here> . However, in the Supplementary Material to this
+article we use a single example to demonstrate some of the software's
+hallmarks, including the concept of data-driven automated gating, the
+integration of existing software into the framework, and the
+generation of publication-quality graphics for data visualization.
+
+%% not sure about this next section vs.
+the use of Suuplementary material.
+Would it be useful/informative to show the code required to generate the plot?
+Otherwise I'm not sold that the next paragraph adds anything.
+
+%% I am not sure about the use of Supplementary Material - flowCore and all the packages mentioned here 
+%% already have at least 1 vignette with examples and code
 
@@
 object, and all available metadata can be accessed to control the
-appearance and the composition of the plot. Similarily, gate objects
+appearance and the composition of the plot. Similarly, gate objects
 are directly passed on to the plotting function and are included in
@@
 
-Static gating for all samples in a high-throughput flow cytometry
-experiment is often not an option, since the measured variables tend
-to vary between different treatments or over time. Automated or
-data-driven gating tries to estimate the gating regions from the
-underlying data, thus providing a fast objective solution to the
-analysis of potentially very large and diverse data sets
-\cite{lo2008agf}. One of the automated gating methods implemented in
-flowCore is based on identifying areas of significant curvature in a
-kernel density estimate of the data \citep{wand2008}. Assuming that
-the regions of interest are of high density, the software is able to
-reliably detect them in either the one ore the two-dimensional density
+Static gating for all samples in a high-throughput FCM experiment is
+often not an option, since the measured variables tend to vary between
+different treatments or over time. Automated or data-driven gating
+tries to estimate the gating regions from the underlying data, thus
+providing a fast objective solution to the analysis of potentially
+very large and diverse data sets \cite{lo2008agf}. One of the
+automated gating methods implemented in flowCore is based on
+identifying areas of significant curvature in a kernel density
+estimate of the data \citep{wand2008}. Assuming that the regions of
+interest are of high density, the software is able to reliably detect
+them in either the one or the two-dimensional density
 landscape. Kernel density estimation is a well-known problem in
@@
 solutions into our framework. In this example, we directly use R code
-from the feature package developed by Wand et al. Instead of
+from the feature package developed by \cite{wand2008}. Instead of
 re-writing existing code, we are able to include it via the well
 tested distribution mechanism provided by R's software package
-system. This process is bi-directional, and all funtionality
+system. This process is bi-directional, and all functionality
 implemented in flowCore is available to other package authors, as we
@@
 
-In developing flowCore, we tried to provided a unified develoment and
-analysis platform for flow-cytometry data. The scope and complexity of
-the technology is evolving rapidly, and a collaborative effort in
-devising new methodolgy has been proven beneficial for a number of
-different biological and computational biology challenges. We hope
-that our framework will be the foundation for fruitful shared research
-by many calloborators from multiple scientific fields, further pushing
-the boundaries of the exiting and still promising flow cytometry
-technology.
+Through flowCore, we have provided the FCM community a freely
+available, sophisticated, development and analysis platform for high
+throughput data analysis. The scope and complexity of the technology
+is evolving rapidly, and a collaborative effort in devising new
+methodology has been proven beneficial for a number of different
+biological and computational biology challenges. We hope that our
+framework will be the foundation for fruitful shared research by many
+collaborators from multiple scientific fields, further pushing the
+boundaries of the exiting and still promising FCM technology.
+
+\section*{Acknowledgements}
+This work was supported by NIH grant EB005034. RRB is a Michael Smith
+Foundation for Health Research Scholar and a International Society for
+Analytical Cytology Scholar.
 

### 2008-05-09 paper/manuscript.tex
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31536 bc3139a8-67e5-0310-9ffc-ced21a209358
Perry's updates to the introduction and conclusions.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31600 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \section*{Introduction}
-Traditionally, flow cytometry (FCM) has been a tube-based technique
-limited to small-scale laboratory and clinical studies.  High
-throughput methods for flow cytometry high content screening (FC-HCS)
-have recently been developed and are now used in both basic and
-clinical research. As a result, the need for flexible and well
-structured computational tools to efficiently handle and analyze
-FC-HCS has increased considerably.
-However, the absence of a shared research platform that
-bioinformaticians, computer scientists, and statisticians can collaboratively
-use has hindered the development of novel methods 
-for FCM data analysis \citep{lizard2007fca}.
-We believe that the open source statistical
-software R, in conjunction with the Bioconductor Project can fill this
-gap.  
+During the last several years, automation technologies have been developed
+that enable the use of flow cytometry high content screening (FC-HCS)
+to generate large, complex datasets in both basic and clinical research
+applications \citep{references from Ryan's FC-HCS paper}. A
+serious bottleneck in the intepretation of existing studies and the application of
+FC-HCS to even larger, more complex problems is that data management
+and data analysis methods have not advanced sufficiently far from the
+methods developed for 
+traditional applications of flow cytometry (FCM) to small-scale, tube-based
+studies.
+
+Some of the consequences of this lag in development are difficulties in 
+maintaining the integrity and documentation of extremely large datasets,
+assessing measurement quality, developing validated assays, assessing the
+accuracy of gating techniques, automating complex gating strategies, and
+aggregating statistical results across large study sets for further analysis.
+Among the most serious consequences of the current situation is, however, 
+that it is very difficult for new analysis
+approaches to find there way into standard practice. We believe that
+this barrier to the development and dissemination of new analysis methods
+is one of the fundamental restraints on the
+future expansion of FC-HCS in both clinical and research applications.
+
+In this paper, we describe a set of flexible and well
+structured computational tools to efficiently analyze
+FC-HCS. This set of tools is based on the open source statistical
+software R, in conjunction with the Bioconductor Project. Our intent
+is to provide a shared research platform that enables
+bioinformaticians, computer scientists, and statisticians to work
+collaboratively with biologists and clinicians to develop novel methods \citep{lizard2007fca} for FCM
+data analysis. A critical component of this approach is the development and implementation
+of standards that will facilitate the adoption of these methods
+by both the larger research community and commercial software vendors.
+Consequently, we are optimistic that this software platform will play a
+critical role in addressing the bottlenecks described above that limit
+further advances in the application of FC-HCS.
+
 %% EAS: I think we mention flowCore before talking about data structures and
@@
 % the missing shared platform.
-We have developed the Bioconductor package flowCore as a freely available,
-sophisticated and extensible FCM data analysis platform that enables
-researchers to efficiently handle FC-HCS data and encourages open
-development of tools for their coherent analysis. 
-%In this paper, we propose R data structures to handle FCM data
-%through the main steps of importing, storing, assessing and
-%preprocessing. 
-%These structures, along with a range of specialized
-%methods for compensation, transformation, gating and additional
-%preprocessing steps, are implemented in the newly developed
-%Bioconductor package flowCore.  flowCore is a freely available,
-%sophisticated and extensible FCM data analysis platform that enables
-%researchers to efficiently handle FC-HCS data and enables open
-%development of new methods and tools for their coherent
-%analysis.
+The computational tools we have developed are distributed in the R 
+software language (www.r-project.org) as the Bioconductor
+(www.bionductor.org) package flowCore. flowCore is a freely available, highly
+functional and extensible FCM data analysis platform that 
+enables researchers to efficiently handle FC-HCS data and encourages open
+development of tools for their coherent analysis. Among the issues
+addressed by flowCore are computationally efficient data structures and
+a range of specialized methods for compensation, transformation, and gating. 
 flowCore runs on Windows, Mac OS X and Linux/Unix operating
-systems.
+systems.  In this paper, we describe the functionality and methods underlying flowCore, 
+and provide examples of their use.
 
@@
 \citep{gentleman2006bos} to process data spanning a diversity of
-research fields from genomics to proteomics to complex cell
+research fields from genomics to proteomics to cell
 biology. It is particularly geared towards the analysis of large
@@
 such as microarray gene expression analysis or FC-HCS. A shared
-platform proved to be particularly beneficial for the development of
+platform has proven to be particularly beneficial for the development of
 analysis routines for microarray gene expression data, and by today
@@
 every other FCM tool that adheres to the proposed
-standard.
+standard. Adherence to standards also plays a critical role in the ability
+of new methods based on flowCore to find there way back into the
+standard practices for FCM data analysis.
 
@@
 a programmatic approach to enable the convenient development of novel
-analysis methods. By taking the burden of data management from the
-programmer, and by providing well-defined APIs and a structured class
+analysis methods and automation of complex analysis approaches. By taking the
+burden of data management from the programmer, and by providing well-defined 
+APIs and a structured class
 hierarchy, it is possible to readily test new ideas and to easily
-extend the frameworks functionality. %Both the underlying R and
+extend the framework's functionality. %Both the underlying R and
 %Bioconductor software and flowCore itself are open source and open %already mentioned its open source
@@
 application of tube-based experiments rather than experiments
-following a plate-design. 
+that depend on automated sample acquisition using a plate-based platform. 
 %% EAS: I don't think the previous sentence is required. Does it 
@@
 
+\secttion*{Conclusions}
+
 Through flowCore, we have provided the FCM community a freely
-available, sophisticated, development and analysis platform for high
-throughput data analysis. The scope and complexity of the technology
-is evolving rapidly, and a collaborative effort in devising new
-methodology has been proven beneficial for a number of different
-biological and computational biology challenges. We hope that our
+available, highly functional, development and analysis platform for high
+throughput data analysis. Our intent is to provide a platform
+for the collaborative development of new analysis methods that
+will facilitate the transition of these new methods to the larger 
+flow community. Our experience has been that the 
+collaborative effort we enable for devising new methodology has been proven
+beneficial for a number of different biological and computational biology challenges. 
+We hope that our
 framework will be the foundation for fruitful shared research by many
-collaborators from multiple scientific fields, further pushing the
-boundaries of the exiting and still promising FCM technology.
+collaborators from multiple scientific fields and will help resolve
+bottlenecks that currently prevent the further development of and deployment of
+FC-HCS to increasingly complex and important scientific and clinical 
+applications.
 

### 2008-05-14 paper/manuscript.tex
Cleaned up some of my comments.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31601 bc3139a8-67e5-0310-9ffc-ced21a209358
minor fixes: typo and biblio
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31665 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 to generate large, complex datasets in both basic and clinical research
-applications \citep{references from Ryan's FC-HCS paper}. A
+applications \citep{brinkman2007hcf}. A
 serious bottleneck in the intepretation of existing studies and the application of
@@
 
-%% EAS: I think we mention flowCore before talking about data structures and
-% methods.  We should first say explicity that we have a solution for
-% the missing shared platform.
 The computational tools we have developed are distributed in the R 
@@
 development of a number of valuable add-ons
-\citep{lo2008agf,sarkar2008ufv} \textit{(plateCore reference, Perry,
-Errol?)}.
+\citep{lo2008agf,sarkar2008ufv} \textit{(plateCore}.
 
@@
 
-\secttion*{Conclusions}
+\section*{Conclusions}
 

### 2008-06-25 paper/manuscript.tex
updates
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32424 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \section*{Introduction}
-During the last several years, automation technologies have been developed
-that enable the use of flow cytometry high content screening (FC-HCS)
-to generate large, complex datasets in both basic and clinical research
-applications \citep{brinkman2007hcf}. A
-serious bottleneck in the intepretation of existing studies and the application of
-FC-HCS to even larger, more complex problems is that data management
-and data analysis methods have not advanced sufficiently far from the
-methods developed for 
-traditional applications of flow cytometry (FCM) to small-scale, tube-based
+During the last several years, automation technologies have been
+developed that enable the use of flow cytometry high content screening
+(FC-HCS) to generate large, complex datasets in both basic and
+clinical research applications \citep{brinkman2007hcf}. A serious
+bottleneck in the intepretation of existing studies and the
+application of FC-HCS to even larger, more complex problems is that
+data management and data analysis methods have not advanced
+sufficiently far from the methods developed for traditional
+applications of flow cytometry (FCM) to small-scale, tube-based
 studies.
 
-Some of the consequences of this lag in development are difficulties in 
-maintaining the integrity and documentation of extremely large datasets,
-assessing measurement quality, developing validated assays, assessing the
-accuracy of gating techniques, automating complex gating strategies, and
-aggregating statistical results across large study sets for further analysis.
-Among the most serious consequences of the current situation is, however, 
-that it is very difficult for new analysis
-approaches to find there way into standard practice. We believe that
-this barrier to the development and dissemination of new analysis methods
-is one of the fundamental restraints on the
-future expansion of FC-HCS in both clinical and research applications.
+Some of the consequences of this lag in development are difficulties
+in maintaining the integrity and documentation of extremely large
+datasets, assessing measurement quality, developing validated assays,
+assessing the accuracy of gating techniques, automating complex gating
+strategies, and aggregating statistical results across large study
+sets for further analysis.  Among the most serious consequences of the
+current situation is, however, that it is very difficult for new
+analysis approaches to find there way into standard practice. We
+believe that this barrier to the development and dissemination of new
+analysis methods is one of the fundamental restraints on the future
+expansion of FC-HCS in both clinical and research applications.
 
-In this paper, we describe a set of flexible and well
-structured computational tools to efficiently analyze
-FC-HCS. This set of tools is based on the open source statistical
-software R, in conjunction with the Bioconductor Project. Our intent
-is to provide a shared research platform that enables
-bioinformaticians, computer scientists, and statisticians to work
-collaboratively with biologists and clinicians to develop novel methods \citep{lizard2007fca} for FCM
-data analysis. A critical component of this approach is the development and implementation
-of standards that will facilitate the adoption of these methods
-by both the larger research community and commercial software vendors.
-Consequently, we are optimistic that this software platform will play a
-critical role in addressing the bottlenecks described above that limit
-further advances in the application of FC-HCS.
+In this paper, we describe a set of flexible and well structured
+computational tools to efficiently analyze FC-HCS. This set of tools
+is based on the open source statistical software R, in conjunction
+with the Bioconductor Project. Our intent is to provide a shared
+research platform that enables bioinformaticians, computer scientists,
+and statisticians to work collaboratively with biologists and
+clinicians to develop novel methods \citep{lizard2007fca} for FCM data
+analysis. A critical component of this approach is the development and
+implementation of standards that will facilitate the adoption of these
+methods by both the larger research community and commercial software
+vendors.  Consequently, we are optimistic that this software platform
+will play a critical role in addressing the bottlenecks described
+above that limit further advances in the application of FC-HCS.
 
@@
 software language (www.r-project.org) as the Bioconductor
-(www.bionductor.org) package flowCore. flowCore is a freely available, highly
-functional and extensible FCM data analysis platform that 
-enables researchers to efficiently handle FC-HCS data and encourages open
-development of tools for their coherent analysis. Among the issues
-addressed by flowCore are computationally efficient data structures and
-a range of specialized methods for compensation, transformation, and gating. 
-flowCore runs on Windows, Mac OS X and Linux/Unix operating
-systems.  In this paper, we describe the functionality and methods underlying flowCore, 
-and provide examples of their use.
+(www.bionductor.org) package flowCore. flowCore is a freely available,
+highly functional and extensible FCM data analysis platform that
+enables researchers to efficiently handle FC-HCS data and encourages
+open development of tools for their coherent analysis. Among the
+issues addressed by flowCore are computationally efficient data
+structures and a range of specialized methods for compensation,
+transformation, and gating.  flowCore runs on Windows, Mac OS X and
+Linux/Unix operating systems. Conceptually, the underlying data
+structures that build the core of our software can be implemented in
+any programming language, and we chose the R language for its obvious
+advantages in statistical programming. In this paper, we describe the
+functionality and methods underlying flowCore, and provide examples of
+their use.
 
@@
 range of statistical and visualization methods developed for various
-fields of application. In addition, R is an open source research %already mentioned R, and by extention flowCore open source
-platform that bioinformaticians, computer scientists, and
-statisticians routinely use to develop novel tools and methodology for
-statistical data analysis. For biological applications, the
-Bioconductor project has become the standard toolset
-\citep{gentleman2006bos} to process data spanning a diversity of
-research fields from genomics to proteomics to cell
-biology. It is particularly geared towards the analysis of large
-numeric datasets that typically arise from high throughput experiments
-such as microarray gene expression analysis or FC-HCS. A shared
-platform has proven to be particularly beneficial for the development of
-analysis routines for microarray gene expression data, and by today
-many of these early innovations have been incorporated into commercial
-software products, thus greatly improving the quality of gene
-expression data analysis. 
-%% NLM: that's probably the important point: platform for innovation rather than for standardization
-Shared data structures have been shown to be
-essential for such a collaborative effort. We envision a similar
-paradigm for FC-HCS data.
+fields of application. In particular, the Bioconductor project aims to
+provide R software modules for biolocial and clinical data analysis.
+It is widely used to process data in various applications, spanning a
+diversity of research fields from genomics to proteomics to cell
+biology \citep{gentleman2006bos}. The R language and Bioconductor are
+particularly geared towards the analysis of large complex datasets
+that typically arise from high throughput experiments such as
+microarray gene expression analysis or FC-HCS. Using flow cytometry as
+a high-throughput technology requires suitable software infrastructure
+to facilitate the rapid handling of the data obtained. We rely on two
+important lessons learned from the field of gene expression data
+analysis: the first being the importance of data structures that
+reflect the underlying data and facilitate the manipulations that are
+of most interest, while the second lesson being the importance of a
+modular architecture that allows for many developers to extend and use
+the underlying infrastructure and to combine in complex workflows.
 
@@
 according to the Flow Cytometry Standard (FCS) \citep{seamer1997pnd}.
-%This standard attempts to capture much of the accompanying meta information, most of which is produced by the measurement instrument or has to be supplied manually by the operator upon experiment setup.
-- I believe FCS only tries to capture a subset of meta information
-However, the recent
-developments in high-throughput FCM are shifting the focus
-of interest away from single-tube based measurements towards large and
-complex experimental designs with dozens of co-variates and
+However, recent developments in high-throughput FCM are shifting the
+focus of interest away from single-tube based measurements towards
+large and complex experimental designs with dozens of covariates and
 influencing factors. Self-contained and self-reflective data
 structures are a prerequisite to allow for coordinated analysis of
-such experiments. Most of the currently available software solutions
+such experiments. Many of the currently available software solutions
 offer only limited support for such structures, or make use of
 containers like XML or binary storage formats that are designed
-specifically for the mostly GUI-driven user interfaces and hence are
-not easily amenable to programmatical access. In addition, the
-closed-source nature of %most of <can you think of an example that is not closed source?>
-these products makes it impractical to
-connect to and integrate the data into an analysis pipeline.
+specifically for particular user interfaces and hence are not easily
+amenable to programmatical access. In addition, the closed-source
+nature of these products makes it impractical to integrate them into
+analysis pipelines.
 
-Traditionally, the majority of FCM experiments have been analyzed
-manually by direct data inspection in one or two dimensions, or by
-very basic comparisons of summary statistics. We believe that these
-approaches do not fully address the highly complex nature of FCM data,
-in particular, they discard many of the fundamental aspects of the
-data, such as its underlying distribution and its high-dimensional
-nature. In addition, the subjective character of manual analyses are a
-major obstacle to reproducibility. For FC-HCS data, unassisted manual
-inspection has become time consuming and labor intensive, and robust
-statistical methods need to be developed to point the investigator to
-the interesting aspects of the data or to potential problems. While
-the expert knowledge of immunologists and researchers remains crucial
-for the understanding of FCM data, we believe that close collaboration
-with other research fields such as statistics and computer science can
-greatly improve the relevance of FCM in today's high-throughput
-paradigm.
+Traditionally, the majority of FCM experiments have been analyzed by
+manual data inspection in one or two dimensions, or by very basic
+comparisons of summary statistics. We believe that these approaches,
+in addition to being extremely expensive and labor intensive, do not
+fully address the highly complex nature of FCM data; in particular,
+they discard many of the fundamental aspects of the data, such as its
+underlying distribution and its high-dimensional nature. Furthermore,
+the subjective character of manual analyses are a major obstacle to
+reproducibility. For FC-HCS data, unassisted manual inspection is too
+time consuming, and robust statistical methods need to be developed to
+point investigators to interesting aspects of the data, or to
+potential problems. While the expert knowledge of immunologists and
+researchers remains crucial for the understanding of FCM data, we
+believe that collaboration with other research fields such as
+statistics and computer science can greatly improve the relevance of
+FCM in today's high-throughput paradigm.
 
@@
 cytometry data standards developed in collaboration with the ISAC Data
-Standards Task Force.  For example, the Gating-ML Candidate Recommendation (CR)
-represents a proposal on how to form unambiguous XML-based gate
-definitions, which can facilitate the interchange and validation of
-data between different software packages with the potential of
-significant increase of hardware and software interoperability. Gates
-may be ordered into a hierarchical structure to describe a gating
-strategy. They may be applied on parameters as in FCS files or on
-transformed parameters as described by an explicit parameter
-transformation. 
-%% I find the previous sentence redundant - 
-%% They may be applied on raw or transformed parameters
-%% EAS: I don't believe the previous sentence is required.
-This enables the encoding of a simple analytical workflow, in a way
-that allows us to reconstruct the analysis programmatically.
+Standards Task Force.  For example, the Gating-ML Candidate
+Recommendation (CR) represents a proposal on how to form unambiguous
+XML-based gate definitions, which can facilitate the interchange and
+validation of data between different software packages. Gates may be
+ordered into a hierarchical structure to describe a gating
+strategy. This enables the encoding of a simple analytical workflow,
+in a way that allows us to reconstruct the analysis programmatically.
 
@@
 implementation of the Gating-ML CR, which opens the possibility to
-integrate flowCore in existing workflows and to communicate with
-every other FCM tool that adheres to the proposed
-standard. Adherence to standards also plays a critical role in the ability
-of new methods based on flowCore to find there way back into the
-standard practices for FCM data analysis.
+integrate flowCore in existing workflows and to communicate with every
+other FCM tool that adheres to the proposed standard. Adherence to
+standards also plays a critical role in the ability of new methods
+based on flowCore to find there way back into the standard practices
+for FCM data analysis. Note that the data strucures defined in
+flowCore aim to model large sets of FCM data in a natural and coherent
+way; they are not specific to the R language and can be implemented in
+any other programming language. Ideally, different software programs
+would communicate on the level of these data structures rather than a
+high-level markup language.
 
@@
 a programmatic approach to enable the convenient development of novel
-analysis methods and automation of complex analysis approaches. By taking the
-burden of data management from the programmer, and by providing well-defined 
-APIs and a structured class
-hierarchy, it is possible to readily test new ideas and to easily
-extend the framework's functionality. %Both the underlying R and
-%Bioconductor software and flowCore itself are open source and open %already mentioned its open source
-%development projects, thus community involvement in the developing
-%process is highly appreciated.
+analysis methods and automation of complex analysis approaches. By
+taking the burden of data management from the programmer, and by
+providing well-defined APIs and a structured class hierarchy, it is
+possible to readily test new ideas and to easily extend the
+framework's functionality.
 
 \section*{Basic data structures}
-\subsection*{Flow Frame}
+\subsection*{flowFrame}
 flowCore's primary task is the representation and basic manipulation
-of FCM data. This is accomplished through a data model very
-close to that adopted by other Bioconductor packages used for gene
-expression analysis, and thus familiar to most Bioconductor
-users. Similar to any other FCM data analysis tool, the basic unit of organization
-in flowCore is a collection of events. We call the structure that hold
-this data a flow frame. All relevant information for the collection
-are stored along with the matrix of raw data values and can be
-accessed programmatically. Most commonly, these are descriptors of the
-stains used in the experiment and the respective measurement channels,
-information about compensation performed at the instrument side and
-any additional keywords the user deems to be important to annotate the
-data. A number of quality checks are performed during the creation of
-a flow frame to ensure data integrity. There is an extended interface
-that allows for the standardized interaction with the flow data, thus
-enabling the development of more specialized tools for specific tasks
-of the data analysis process that all share a common Application Program Interface (API).
+of FCM data. This is accomplished through a data model very close to
+that adopted by other Bioconductor packages. All information from a
+single FCS file, i.e., the collection of events and the accompanying
+metadata, is stored in one single container. We call the structure
+that hold this data a flowFrame, and raw data values as well as
+associated metadata can be accessed programmatically. Most commonly,
+these are descriptors of the stains used in the experiment and the
+respective measurement channels, information about compensation
+performed at the instrument side and any additional keywords the user
+deems to be important to annotate the data. A number of quality checks
+are performed during the creation of a flowFrame to ensure data
+integrity.
 
-\subsection*{Flow Set}
+\subsection*{flowSet}
 In high-throughput FCM, many of the analysis tasks need to be
 performed consistently across multiple flow measurements, hence we
-introduce the concept of a collection of flow frames called a flow
-set. Flow sets store the relevant information associated with each
-individual frame such as descriptions of the samples or the treatment
-to which a sample was subjected, and they manage the consistent
-application of operations on the individual flow frames. The burden of
-keeping score of the annotation is shifted from the user to the
-infrastructure, thus reducing the risk of common errors % (\textit{i.e.}, example of an error ) 
-that occur when using more or less ad hoc spreadsheets that
-lack a controlled structure. There is no native ordering
-to a flow set, which in a way reflects the still predominant
-application of tube-based experiments rather than experiments
-that depend on automated sample acquisition using a plate-based platform. 
-%% EAS: I don't think the previous sentence is required. Does it 
-%% matter if the flowSet is ordered or not?
-
-%However, the flow set structure can be
-%readily extended to also accommodate such spatial information
-%% EAS: The spatial aspect of plateCore is really secondary to having 
-%% an easy, consistent way to manage and access the mulitplex annotation.
-%% Also, it's important to have a standard format for the annotation if
-%% we're trying to encourage collaboration. I haven't done it yet, but I 
-%% intend to look into
-%% into stealing the annotation implementation from CellHTS2 and
-%% using it in plateCore.
-However, the flow set structure can be readily extended to incorporate the
-potentially complex annotation associated with large FC-HCS experiments. Access
-to flowCore data structures and methods has already enabled the development of
-new Bioconductor pacakges, such as plateCore, that faciliate the analysis of
-plate-based FC-HCS data and make handling the annotation and processing
-the data easier.
-
-%\textit{(plateCore reference, Perry, Errol?)}.
+introduce the concept of a collection of flowFrames called a
+flowSet. One can envision a flowSet to be a container for multiple
+flowFrames along with relevant information associated with each
+individual frame such as descriptions of the cell sample or the
+treatment to which the sample was subjected. They manage the
+consistent application of operations on the individual flow frames and
+shift the burden of keeping score of the annotation from the user to
+the infrastructure, thus reducing the risk of common errors
+(\textit{e.g.}, mixups of sample labels). The flowSet structure can be
+readily extended to incorporate the potentially complex annotation
+associated with large FC-HCS experiments. Access to flowCore data
+structures and methods has already enabled the development of new
+Bioconductor packages, such as plateCore, that faciliate the analysis
+of plate-based FC-HCS data and make handling the annotation and
+processing the data easier.
 
@@
 \includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
-\caption{\label{fig1:FrameWork}{The flowCore frame work. For each
-    experiment, FCS files, phenotypic and meta data is stored in a
-    flow set. Each flow frame of a flow set corresponds to one FCS
+\caption{\label{fig1:FrameWork}{For each experiment, FCS files (as
+    individual flowFrames), phenotypic and metadata are stored in a
+    flowSet. Each flowFrame in a flowSet corresponds to one FCS
     file. All basic operations (e.g., compensation, transformation,
-    gating) can be applied on both single flow frames or entire flow
-    set.}}
+    gating) can be applied to either single flowFrames or a flowSet
+    simulataneously.}}
 \end{figure}
@@
 %    packages: dedicated packages have been developed that utilize
-%    flowCore’s data structures. flowViz is a data visualization
+%    flowCoreœôòùs data structures. flowViz is a data visualization
 %    package; flowQ provides data quality assessment methods; flowUtils

### 2008-07-10 inst/doc/HowTo-flowCore.Rnw
moved all plotting stuff to flowViz, flowCore now issues a message to attach flowViz for all plotting
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31029 bc3139a8-67e5-0310-9ffc-ced21a209358
fixed inconsitencies in the splitting methods
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32646 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \begin{abstract}
-\noindent \textbf{Background} The recent application of modern automation technologies to staining and collecting flow cytometry (FCM) samples has led to many new challenges in data management and analysis.
-We limit our attention here to the associated problems in the analysis of the massive amounts of FCM data now being collected.
-From our viewpoint, see two related but substantially different problems arising.
-On the one hand, there is the problem of adapting existing software to apply standard methods to the increased volume of data.
-The second problem, which we intend to address here, is the absence of any research platform which bioinformaticians, computer scientists, and statisticians can use to develop novel methods that address both the volume and multidimensionality of the mounting tide of data.
-In our opinion, such a platform should be Open Source, be focused on visualization,  support rapid prototyping, have a large existing base of users, and have demonstrated suitability for development of new methods.
-We believe that the Open Source statistical software R in conjunction with the Bioconductor Project fills all of these requirements.
-Consequently we have developed a Bioconductor package that we call \Rpackage{flowCore}.
-The \Rpackage{flowCore} package is not intended to be a complete analysis package for FCM data, rather, we see it as providing a clear object model and a collection of standard tools that enable R as an informatics research platform for flow cytometry.
-One of the important issues that we have addressed in the \Rpackage{flowCore} package is that of using a standardized representation that will insure compatibility with existing technologies for data analysis and will support collaboration and interoperability of new methods as they are developed.
-In order to do this, we have followed the current standardized descriptions of FCM data analysis as being developed under NIH Grant xxxx [n].
-We believe that researchers will find flowCore to be a solid foundation for future development of new methods to attack the many interesting open research questions in FCM data analysis.\\
+\noindent \textbf{Background} The recent application of modern
+automation technologies to staining and collecting flow cytometry
+(FCM) samples has led to many new challenges in data management and
+analysis. We limit our attention here to the associated problems in
+the analysis of the massive amounts of FCM data now being
+collected. From our viewpoint, see two related but substantially
+different problems arising. On the one hand, there is the problem of
+adapting existing software to apply standard methods to the increased
+volume of data. The second problem, which we intend to address here,
+is the absence of any research platform which bioinformaticians,
+computer scientists, and statisticians can use to develop novel
+methods that address both the volume and multidimensionality of the
+mounting tide of data. In our opinion, such a platform should be Open
+Source, be focused on visualization, support rapid prototyping, have a
+large existing base of users, and have demonstrated suitability for
+development of new methods. We believe that the Open Source
+statistical software R in conjunction with the Bioconductor Project
+fills all of these requirements. Consequently we have developed a
+Bioconductor package that we call \Rpackage{flowCore}. The
+\Rpackage{flowCore} package is not intended to be a complete analysis
+package for FCM data, rather, we see it as providing a clear object
+model and a collection of standard tools that enable R as an
+informatics research platform for flow cytometry. One of the important
+issues that we have addressed in the \Rpackage{flowCore} package is
+that of using a standardized representation that will insure
+compatibility with existing technologies for data analysis and will
+support collaboration and interoperability of new methods as they are
+developed. In order to do this, we have followed the current
+standardized descriptions of FCM data analysis as being developed
+under NIH Grant xxxx [n]. We believe that researchers will find
+flowCore to be a solid foundation for future development of new
+methods to attack the many interesting open research questions in FCM
+data analysis.\\
 
-\noindent \textbf{Methods} We propose a variety different data structures. We have implemented the classes and methods in the Bioconductor package flowCore. 
-We illustrate their use with X case studies.\\
+\noindent \textbf{Methods} We propose a variety different data
+structures. We have implemented the classes and methods in the
+Bioconductor package flowCore.  We illustrate their use with X case
+studies.\\
 
-\noindent \textbf{Results} We hope that those proposed data structures will the base for the development of many tools for the analysis of high throughput flowcytometry.\\
+\noindent \textbf{Results} We hope that those proposed data structures
+will the base for the development of many tools for the analysis of
+high throughput flowcytometry.\\
 
-\noindent \textbf{keywords} Flow cytometry, high throughtput, software, standard
+\noindent \textbf{keywords} Flow cytometry, high throughtput,
+software, standard
 \end{abstract}
@@
 \Rpackage{flowViz} package, also included as part of
-Bioconductor. However, \Rpackage{flowCore} does implement some basic
-plotting facilities using the standard \Rfunction{plot} function. The
-basic plot provides a simple pairs plot of all parameters:
+Bioconductor. Here, we will only introduce the standard
+\Rfunction{plot} function. The basic plot provides a simple pairs plot
+of all parameters:
 
 <<Plot1, fig=true, echo=true>>=
+library(flowViz)
 plot(x)
@@
 <<TectGate3, echo=true, results=verbatim>>=
-rect2 <- rectangleGate(filterId="Another Rect", "FL1-H"=c(1,2), "FL2-H"=c(2,3)) %on% tFilter
+rect2 <- rectangleGate(filterId="Another Rect", "FL1-H"=c(1,2), 
+"FL2-H"=c(2,3)) %on% tFilter
 rect2
@@
 fset1 = filterSet(
-	rectangleGate(filterId="Fluorescence Region","FL1-H"=c(50,100),"FL2-H"=c(50,100)),
+	rectangleGate(filterId="Fluorescence Region","FL1-H"=c(50,100),
+                      "FL2-H"=c(50,100)),
 	norm2Filter(filterId="Morphology Gate","FSC-H","SSC-H",scale=2),

### 2008-07-10 R/filterReference-accessors.R

@@ -22,3 +22,11 @@ setMethod("summarizeFilter",signature("filterResult","filterReference"),
 setMethod("filterReference",signature("environment","character"),function(from,name) {
 	new("filterReference",name=name,env=from)
 })
+
+## A filterReference from a filterSet
+setMethod("filterReference",
+          signature("filterSet", "character"),
+          function(from,name) {
+              new("filterReference", env=from@env,
+                  name=name)
+})

### 2008-07-10 R/split-methods.R

@@ -1,36 +1,100 @@
+## split methods for flowFrames and flowSets:
+##
+## The splitting operation in the context of 'flowFrames' and 'flowSets' is
+## the logical extension of subsetting. While the latter only returns
+## the events contained within a gate, the former splits the data into
+## the groups of events cotained within and those not contained within a
+## particular gate. This concept is extremely useful in applications where
+## gates describe the distinction between positivity and negativity for a
+## particual marker.
+## Splitting has a special meaning for gates that result in 
+## 'multipleFilterResults', in which case simple subsetting doesn't make
+## much sense (there are multiple populations that are defined by the gate
+## and it is not clear which of those should be used for the subsetting
+## operation). Accordingly, splitting of multipleFilterResults creates
+## multiple subsets. The argument 'population' can be used to limit the
+## ouput to only one or some of the resulting subsets. It takes as values
+## a character vector of names of the populations of interest. See the
+## documentation of the different filter classes on how population names
+## can be defined and the respective default values. For splitting of
+## 'logicalFilterResults', the 'population' argument can be used to set
+## the population names since there is not reasonable default other than the
+## name of the gate. The content of the argument 'prefix' will be prepended
+## to the population names and '+' or '-' are finally appended allowing for
+## more flexible naming schemes.
+## Further control of the output is provided by the argument 'flowSet',
+## which defines whether individual subsets should be returned in the form
+## of a list (the default) or whether they should be coerced into objects
+## of class 'flowSet'. This only applies when splitting 'flowFrames',
+## splitting of 'flowSets' always results in lists of 'flowSet' objects.
+
+
+
+
+
+
 ## ==========================================================================
 ## split methods for flowFrame
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-## We actually split on filterResults and multipleFilterResults
+## We actually split on filterResults and multipleFilterResults, so filters
+## have to be evaluated first. Note that the 'drop' argument is silently
+## ignored when splitting by filter since it doesn't have a clear meaning in
+## this application. It does have the expected meaning when splitting by
+## factors.
+
+## Evaluate the filter first and split on the filterResult
 setMethod("split",
[…]

### 2008-07-10 R/AllClasses.R
fixed inconsitencies in the splitting methods
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32646 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -90,16 +90,16 @@ setClass("flowSet",
          validity=function(object){
              nc <- length(colnames(object))
              ## Make sure that all of our samples list
-             name.check <- is.na(match(sampleNames(object),ls(object@frames,
+             name.check <- is.na(match(sampleNames(object), ls(object@frames,
                                                               all.names=TRUE)))
              if(any(name.check)) {
-                 name.list <- paste(sampleNames(object)[name.check],sep=",")
+                 name.list <- paste(sampleNames(object)[name.check], sep=",")
                  return(paste("These objects are not in the data environment:",
                               name.list))
              }
              ##Ensure that all frames match our colnames
-             if(!all(sapply(sampleNames(object),function(i) {
-                 x = get(i,env=object@frames)
+             if(!all(sapply(sampleNames(object), function(i) {
+                 x <- get(i, env=object@frames)
                  if(all(object@colnames %in% colnames(x))){
                      TRUE
                  }else{ 
@@ -121,7 +121,7 @@ flowSet <- function(..., phenoData) {
         x <- x[[1]]
     if(!all(sapply(x, is, "flowFrame")))
         stop("All additional arguments must be flowFrames")
-    f <- as(x,"flowSet")
+    f <- as(x, "flowSet")
     if(!missing(phenoData))
         phenoData(f) = phenoData
     f
@@ -163,14 +163,14 @@ setClass("rectangleGate",
                         max="numeric"),
          contains="parameterFilter",
          prototype=list(filterId="Rectangle Gate",
-         min=0,max=Inf)
+         min=0, max=Inf)
          )
 
 ## constructor
 rectangleGate <- function(filterId="rectangleGate", .gate,...) {
     if(missing(.gate) || !is.matrix(.gate))
-      	.gate <- sapply(if(missing(.gate)) list(...) else .gate,function(x) {
-            x = sort(x);c("min"=x[1],"max"=x[2])
+      	.gate <- sapply(if(missing(.gate)) list(...) else .gate, function(x) {
+            x <- sort(x);c("min"=x[1], "max"=x[2])
 		})
     new("rectangleGate", filterId=filterId, parameters=colnames(.gate),
         min=.gate[1,], max=.gate[2,])
@@ -228,12 +228,12 @@ setClass("polygonGate",
          })
 
 ## constructor
-polygonGate <- function(filterId="polygonGate", boundaries,...) {
+polygonGate <- function(filterId="polygonGate", boundaries, ...) {
[…]

### 2008-07-15 inst/doc/HowTo-flowCore.Rnw
upadates to the documentation
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32724 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 <<SummarFilter, echo=true, results=verbatim>>=
-summary(filter(fs[[1]], kmeansFilter("myKMeans","FSC-H"=c("Low", "Medium", "High"))))
+summary(filter(fs[[1]], kmeansFilter("FSC-H"=c("Low", "Medium", "High"),
+                                     filterId="myKMeans")))
 @
@@
 <<Norm2Filter, echo=true, results=verbatim>>=
-morphGate <- norm2Filter(filterId="MorphologyGate", "FSC-H", "SSC-H", scale=2)
+morphGate <- norm2Filter("FSC-H", "SSC-H", filterId="MorphologyGate", 
+                         scale=2)
 smaller <- Subset(fs, morphGate)
@@
 <<Split, echo=true, results=verbatim>>=
-split(smaller[[1]], kmeansFilter("myKMeans","FSC-H"=c("Low","Medium","High")))
+split(smaller[[1]], kmeansFilter("FSC-H"=c("Low","Medium","High"),
+                                 filterId="myKMeans"))
 @
@@
 <<Split2, echo=true, results=verbatim>>=
-split(smaller, kmeansFilter("myKMeans", "FSC-H"=c("Low", "Medium", "High")))
+split(smaller, kmeansFilter("FSC-H"=c("Low", "Medium", "High"),
+                            filterId="myKMeans"))
 @
@@
                       "FL2-H"=c(50,100)),
-	norm2Filter(filterId="Morphology Gate","FSC-H","SSC-H",scale=2),
+	norm2Filter("FSC-H","SSC-H",filterId="Morphology Gate", scale=2),
 	~ `Fluorescence Region` & `Morphology Gate`,

### 2008-07-15 R/AllClasses.R
upadates to the documentation
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32724 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -358,18 +358,27 @@ setClass("kmeansFilter",
          contains="parameterFilter")
 
 ## contructor
-## not sure why but the parameters in list format can not be read.
 kmeansFilter <- function(..., filterId="kmeans")
 {
-    l <- length(list(...))
+    ll <- list(...)
+    l <- length(ll)
     if(l>1)
-	stop("k-means filters only operate on a single parameter.")
+	warning("k-means filters only operate on a single parameter.\n",
+                "Using '", names(ll)[1], "' now.", call.=FALSE)
     x <- ..1
-    if(is.list(x)) {
-        new("kmeansFilter", parameters=names(x)[1], populations=x[[1]],
+    if(is.character(x))
+        new("kmeansFilter", parameters=names(ll)[1], populations=x,
             filterId=filterId)
-    } else {
-        new("kmeansFilter", parameters=names(list(...))[1], populations=x,
+    else if(is.list(x) && ! length(names(x))){
+        new("kmeansFilter", parameters=names(ll)[1], populations=unlist(x),
+            filterId=filterId)
+    }else{
+        if(length(x)>1){
+            warning("k-means filters only operate on a single parameter.\n",
+                    "Using '", names(x)[1], "' now.", call.=FALSE)
+            x <- x[1]
+        }
+        new("kmeansFilter", parameters=names(x[1]), populations=unlist(x),
             filterId=filterId)
     }
 }
@@ -391,7 +400,7 @@ setClass("curv2Filter",
 ##constructor
 curv2Filter <-
     function(x, y, filterId="curv2Filter", bwFac=1.2,
-             gridsize=rep(151, 2), ...)
+             gridsize=rep(151, 2))
 {
     if(!is.numeric(bwFac) || length(bwFac)!=1)
         stop("'bwFac must be numeric skalar.", call.=FALSE)
@@ -414,7 +423,7 @@ curv2Filter <-
         y = y[1]
     }
     new("curv2Filter", parameters=as.character(c(x, y)), bwFac=bwFac,
-        gridsize=gridsize, filterId=as.character(filterId), ...)
+        gridsize=gridsize, filterId=as.character(filterId))
 }
 
 

### 2008-07-16 paper/manuscript.tex
some updates to the paper manuscript
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32728 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \section*{Introduction}
+\subsection*{Computational tools for the analysis of FC-HCS data}
 During the last several years, automation technologies have been
@@
 datasets, assessing measurement quality, developing validated assays,
-assessing the accuracy of gating techniques, automating complex gating
+controling the accuracy of gating techniques, automating complex gating
 strategies, and aggregating statistical results across large study
@@
 provide R software modules for biolocial and clinical data analysis.
-It is widely used to process data in various applications, spanning a
-diversity of research fields from genomics to proteomics to cell
-biology \citep{gentleman2006bos}. The R language and Bioconductor are
+It is widely used to process such data, spanning a diversity of
+research fields from genomics to proteomics to cell biology
+\citep{gentleman2006bos}. The R language and Bioconductor are
 particularly geared towards the analysis of large complex datasets
@@
 modular architecture that allows for many developers to extend and use
-the underlying infrastructure and to combine in complex workflows.
+the underlying infrastructure and to combine its tools in complex
+workflows.
 
@@
 
-\section*{Basic data structures}
-\subsection*{flowFrame}
-flowCore's primary task is the representation and basic manipulation
+\section*{Materials and Methods}
+\subsection*{Basic data structures}
+\subsubsection*{flowFrame}
+FlowCore's primary task is the representation and basic manipulation
 of FCM data. This is accomplished through a data model very close to
@@
 metadata, is stored in one single container. We call the structure
-that hold this data a flowFrame, and raw data values as well as
-associated metadata can be accessed programmatically. Most commonly,
-these are descriptors of the stains used in the experiment and the
-respective measurement channels, information about compensation
-performed at the instrument side and any additional keywords the user
-deems to be important to annotate the data. A number of quality checks
-are performed during the creation of a flowFrame to ensure data
-integrity.
+that hold this data a flowFrame. Raw data values as well as associated
+metadata of a flowFrame can be accessed programmatically. Most
+commonly, these are descriptors of the stains used in the experiment
+and the respective measurement channels, information about
+compensation performed at the instrument side and any additional
+keywords the user deems to be important to annotate the data. A number
+of quality checks are performed during the creation of a flowFrame to
+ensure data integrity.
 
-\subsection*{flowSet}
+\subsubsection*{flowSet}
 In high-throughput FCM, many of the analysis tasks need to be
@@
 individual frame such as descriptions of the cell sample or the
-treatment to which the sample was subjected. They manage the
-consistent application of operations on the individual flow frames and
+treatment to which the sample was subjected. FlowSets manage the
+consistent application of operations on the individual flowFrames and
 shift the burden of keeping score of the annotation from the user to
-the infrastructure, thus reducing the risk of common errors
-(\textit{e.g.}, mixups of sample labels). The flowSet structure can be
-readily extended to incorporate the potentially complex annotation
-associated with large FC-HCS experiments. Access to flowCore data
+the infrastructure, thus reducing the risk of errors (\textit{e.g.},
+mixups of sample labels). The flowSet structure can be readily
+extended to incorporate the potentially complex annotation associated
+with even larger FC-HCS experiments. Access to flowCore data
 structures and methods has already enabled the development of new
@@
 
-\section*{Standard flow operations}
+\subsection*{Standard flow operations}
 The basic operations in FCM analyses are typically the same: the data
@@
 solutions for FCM analysis offer more or less elaborate support for
-these kinds of operation, and most of them in an interactive,
-GUI-driven interface. The approach we have taken in flowCore is to
-abstractly describe these operations and build a set of tools to
-perform them on both flow frames and flow sets. While transformation,
-and to a certain extent compensation, are fairly routine operations
-with only limited potential for improvement, being able to implement
-new methodologies for gating of FCM data, and extending the
-capabilities of flowCore through object oriented programming are
-features that clearly sets our framework apart from other FCM analysis
-tools. By factoring out as much of the bookkeeping as possible,
-programmers can focus on the actual operations rather then having to
-deal with the tedious details of data integration and
-access. Third-party methods can act on their own as first-class
-citizens in the analysis framework, without breaking the workflow or
-the basic infrastructure. This design allows for the straightforward
-extension of flowCore's capabilities, and has already fostered the
-development of a number of valuable add-ons
-\citep{lo2008agf,sarkar2008ufv} \textit{(plateCore}.
+these kinds of operation in an interactive, GUI-driven interface. The
+approach we have taken in flowCore is to abstractly describe these
+operations and build a set of tools to perform them on both flowFrames
+and flowSets. While transformation, and to a certain extent
+compensation, are fairly routine operations with only limited
+potential for improvement, being able to implement new methodologies
+for gating of FCM data, and extending the capabilities of flowCore
+through object oriented programming are features that clearly sets our
+framework apart from other FCM analysis tools. By factoring out as
+much of the bookkeeping as possible, programmers can focus on the
+actual operations rather then having to deal with the tedious details
+of data integration and access. Third-party methods can act on their
+own as first-class citizens in the analysis framework, without
+breaking the workflow or the basic infrastructure. This design allows
+for the straightforward extension of flowCore's capabilities, and has
+already fostered the development of a number of valuable add-ons
+\citep{lo2008agf,sarkar2008ufv} \textit{(plateCore)}.
 
-\subsection*{Transformation and compensation}
+\subsubsection*{Transformation and compensation}
 Data transformation is essential for both data visualization and
@@
 routinely used in FCM analysis have been implemented in flowCore
-(\textit{e.g.,} quadratic, log or arcsinh). Furthermore, the design of the R language makes
-it easy to define arbitrary functions to apply to the data of
-individual flow frames or entire flow sets, respectively. Compensation
-is available for both flow frames and flow sets, and the software also
+(\textit{e.g.,} quadratic, log or arcsinh, see Table~\ref{table1} for a
+complete list). Furthermore, the design of the R language makes it
+easy to define arbitrary functions to apply to the data of individual
+flowFrames or entire flowSets, respectively. Compensation is
+available for both flowFrames and flowSets, and the software also
 offers functionality to compute spillover or compensation matrices
@@
 
-\subsection*{Gating}
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\multicolumn{2}{|c|}{Data Transformations} \\
+\hline
+linear & $ax + b$ \\
+quadratic & $ax^2 + bx + c$ \\
+natural logarithm & $ln(x)(r/d)$ \\
+logarithm & log(x)(r/d) \\
+biexponential & need equation \\
+logicle& need equation \\
+truncate & $x_{x<=a} = a$ \\
+scale & (x-a)/(b-a) \\
+split-scale & need equation \\
+arcsinh & $asinh(a + bx)+c$ \\
+\hline
+\end{tabular}
+\caption{\label{table1}Data transformations implemented in flowCore.}
+\end{center}
+\end{table}
+
+\subsubsection*{Gating}
 In flowCore, gating operations are represented by classes that can be
-extended in an object-oriented manner. Basic gate types such as
-rectangular gates, ellipses and polygon gates are implemented as part
-of the framework. In addition, we introduce the notion of data-driven
-gates, or filters, for which the necessary parameters are computed
-based on the properties of the underlying data, for instance by
-modelling underlying data distribution or by density estimation. The
-ability to programmatically access gates is a prerequisite for
-automated or semi-automated gating. By utilizing a unified interface
-for all different types of gates, the user is able to subset data sets
-as well as to create summary statistics, like the proportions of
-events falling in single gates or in the combinations of
-gates. Complex combinations and hierarchies of gates can be captured
-in objects of class filterSet, allowing to apply multi-step gating
-strategies. The definition of gates in flowCore follows the Gating-ML
-CR, thus any flowCore gating strategy can be reproduced by any other
-software that also adheres to the standard.
+extended in an object-oriented manner (Table~\ref{table2}). Basic gate
+types such as rectangular gates, ellipses and polygon gates are
+implemented as part of the framework. In addition, we introduce the
+notion of data-driven gates, or filters, for which the necessary
+parameters are computed based on the properties of the underlying
+data, for instance by modelling data distribution or by density
+estimation. The ability to programmatically access gates is a
+prerequisite for automated or semi-automated gating. By utilizing a
+unified interface for all different types of gates, the user is able
+to subset data sets as well as to create summary statistics, for
+instance the proportions of events falling in single gates or in the
+combinations of gates. Complex combinations and hierarchies of gates
+can be captured in objects of class filterSet, allowing to apply
+multi-step gating strategies. The definition of gates in flowCore
+follows the Gating-ML CR, thus any flowCore gating strategy can be
+reproduced by any other software that also adheres to the standard.
 
-\section*{Related flow packages}
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+\multicolumn{2}{|c|}{Gates} \\
+\hline
+rectangleGate & n-dinensional rectangular regions \\
+quadGate & quadrant regions in two dimensions \\
+polygonGate & polygonal regions in two dimensions \\
+polytopeGate & generalization of polygon in n dimensions \\
+ellipsoidGate & n-dimesnional ellipsoid region \\
+\hline
+\multicolumn{2}{|c|}{Filters} \\
+\hline
+sampleFilter & random subsampling\\
+expressionFilter & filter by resuts of a boolean expression \\
+kmeansFilter & based on K-means clustering \\
+norm2Filter & based on bivariate normal distribution \\
+curv1Filter & based on local density regions in 1D \\
+curv2Filter & based on local density regions in 2D \\
+timeFilter & based on abnormal data acquisition over time \\
+\hline
+filterSet & gating strategy \\
+\hline
+\end{tabular}
+\caption{\label{table2}Filter and gate classes implemented in flowCore.}
+\end{center}
+\end{table}
+
+\subsection*{Related flow packages}
 In addition to the flowCore package that offers basic infrastructure,
@@
 are dedicated to more specific tasks of FCM data analysis
-(Figure \ref{fig2:flowWheel}). The flowViz package provides
+%(Figure \ref{fig2:flowWheel}) 
+. The flowViz package provides
 sophisticated data visualization tools, that make use of multivariate
@@
 operation described below are added to the plot. The input to the
-plotting function is either a single flow frame or a complete flow set
+plotting function is either a single flowFrame or a complete flowSet
 object, and all available metadata can be accessed to control the
@@
 \caption{\label{xyplot}%
-Scatterplot matrix of a single Flow Set with outlines of the gating
+Scatterplot matrix of a single FlowSet with outlines of the gating
 regions identified by an automated gating operation.}

### 2008-07-22 paper/manuscript.tex
some updates to the paper
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32839 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 The computational tools we have developed are distributed in the R
-software language (www.r-project.org) as the Bioconductor
-(www.bionductor.org) package flowCore. flowCore is a freely available,
+software language \citep{Rmain} as the Bioconductor
+\citep{BIOC} package flowCore. flowCore is a freely available,
 highly functional and extensible FCM data analysis platform that
@@
 \hline
-filterSet & gating strategy \\
+filterSet & gating strategies \\
 \hline
@@
 we have implemented a range of additional Bioconductor packages that
-are dedicated to more specific tasks of FCM data analysis
-%(Figure \ref{fig2:flowWheel}) 
-. The flowViz package provides
-sophisticated data visualization tools, that make use of multivariate
-trellis plotting. These functions can be used to quickly generate
-customized plots for extended cytometry data sets for both direct data
-inspection and quality control. The flowQ package offers more advanced
-quality assurance (QA) methodology and a framework to create
-interactive web-based reports of QA results. Most of flowUtil's
-content deals with data import and output from and to flowCore and the
-flow-cytometry specific standard markup languages. Finally, flowStats
-provides statistical methods that are relevant in the context of flow
+are dedicated to more specific tasks of FCM data analysis. The flowViz
+package \citep{sarkar2008ufv} provides sophisticated data
+visualization tools, that make use of multivariate trellis
+plotting. These functions can be used to quickly generate customized
+plots for extended cytometry data sets for both direct data inspection
+and quality control. The objects annotation information can be used to
+arange the layout and composition of the plots. Furthermore, the
+design and the API of the visualization software is very generic, and
+users can readily extend its capabilities by providing self-defined
+plotting functions.  The flowQ package offers more advanced quality
+assurance (QA) methodology and a framework to create interactive
+web-based reports of QA results. Most of flowUtil's content deals with
+data import and output from and to flowCore and the flow-cytometry
+specific standard markup languages. Finally, flowStats provides more
+elaborate statistical methods that are relevant in the context of flow
 cytometry data analysis.
 
-%\begin{figure}
-%\centering
-%\includegraphics[width=0.6\textwidth]{Figure2-flowWheel-version4.jpg}
-%\caption{\label{fig2:flowWheel}{flowCore and assoziated Bioconductor
-%    packages: dedicated packages have been developed that utilize
-%    flowCoreœôòùs data structures. flowViz is a data visualization
-%    package; flowQ provides data quality assessment methods; flowUtils
-%    contains interfaces for flow data, including xml parsers:
-%    flowClust provides automated gating by model-based clustering.}}
-%\end{figure}
-
-\section*{Application}
-The flowCore package has successfully been applied in the analysis of
+\section*{Results}
+The flowCore package has been successfully applied in the analysis of
 several datasets \citep{gasparetto2004ice,brinkman2007hcf}, and
@@
 base. As an example, \cite{lo2008agf} have recently developed an
-automatic gating approach via robust model-based clustering using the
-flowCore data model and infrastructure and implemented this in the
-Bioconductor package flowClust. Another package (plateCore) providing
+automatic gating approach via robust model-based clustering using
+flowCore's data model and infrastructure and implemented this in the
+Bioconductor package flowClust. Another package, plateCore, providing
 more specialized support for experiments conducted on microtitre
-plates is under development. The documentation of all of flowCore's
-features is beyond the scope of this publication, however additional
-documentation and users guide information with programmatic examples
-is available <here> . However, in the Supplementary Material to this
-article we use a single example to demonstrate some of the software's
-hallmarks, including the concept of data-driven automated gating, the
-integration of existing software into the framework, and the
-generation of publication-quality graphics for data visualization.
+plates is under development.
 
-%% not sure about this next section vs.
-the use of Suuplementary material.
-Would it be useful/informative to show the code required to generate the plot?
-Otherwise I'm not sold that the next paragraph adds anything.
+A complete documentation of the flowCore software is way beyond the
+scope of this publication. Much more comprehensive documentation and
+users guide information with programmatic examples is available online
+(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
+and as part of the package distribution. Here, we only want to briefly
+exemplify some of the software's key features, that is, the coherent
+treatment of all samples in a potentially large expreriment, the
+concept of data-driven automated gating, the integration of existing
+software into the framework, and the generation of publication-quality
+graphics for data visualization.
 
-%% I am not sure about the use of Supplementary Material - flowCore and all the packages mentioned here 
-%% already have at least 1 vignette with examples and code
+Data analysis for most experiments usually begins with a quality
+assurance step, and we can use functionality from the flowQ package to
+create an HTML report that highlights potential quality
+issues. Assuming that the data has already been imported as the
+flowSet object ``dat'', the following simple line of code produces the
+quite complex output:
 
-A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
-where each panel in the matrix represents the data of one individual
-patient. Instead of plotting individual points, the local bivariate
-density of the data is shown as a false color image. For each panel,
-the outlines of the regions identified in the automated gating
-operation described below are added to the plot. The input to the
-plotting function is either a single flowFrame or a complete flowSet
-object, and all available metadata can be accessed to control the
-appearance and the composition of the plot. Similarly, gate objects
-are directly passed on to the plotting function and are included in
-the plot if possible. Besides the very basic scatterplot shown here,
-there is a wealth of additional data visualizations available in the
-flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
-models allowing to easily arrange the data according to numerous
-experimental factors. Furthermore, the design and the API of the
-visualization software itself is very generic, and the user can
-readily extend its capabilities by providing self-defined plotting
-functions.
+\begin{verbatim}
+> library(flowCore)
+> library(flowQ)
+> qaReport(dat, c("qaProcess.timeline", "qaProcess.timeflow", 
+                  "qaProcess.cellnumber"))
+\end{verbatim}
+
+The report is interactive and provides drilldown to more detailed
+aspects of the analysis, starting from a concise overview
+(Figure~\ref{flowQ}). The design of flowCore's data model allows for a
+coherent treatment of all the samples, hence we are able to compare
+features between individuals, or between groups of individuals, based
+on the available annotation information.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=0.75\textwidth]{flowQ.jpg}
+\caption{\label{flowQ}% 
+  HTML quality assesment report generated by
+  the flowQ package. Rows correspond to samples, columns to different
+  quality checks.}
+\end{figure}
+
+Static gating for all samples in a high-throughput FCM experiment is
+often not an option, since the measured variables tend to vary between
+different treatments, over time or between different experiment
+batches. Automated or data-driven gating tries to estimate the gating
+regions from the underlying data, thus providing a fast objective
+solution to the analysis of potentially very large and diverse data
+sets \cite{lo2008agf}. One of the automated gating methods implemented
+in flowCore is based on identifying areas of significant curvature in
+a kernel density estimate of the data \citep{wand2008}. Assuming that
+the regions of interest are of high density, the software is able to
+reliably detect them in a one or two-dimensional density landscape.
+
+\begin{verbatim}
+> cf <- curv2Filter("FL1-H", "FL3-H")
+> fres <- filter(dat, cf)
+\end{verbatim}
+
+Kernel density estimation is a well-known problem in statistical
+computing, and a lot of effort has been invested in the development of
+good software to address it. The modular design of flowCore allows to
+easily integrate these existing solutions into our framework. In this
+example, we directly use R code from the feature package developed by
+\cite{wand2008}. Instead of re-writing existing code, we are able to
+include it via the well tested distribution mechanism provided by R's
+software package system. This process is bi-directional, and all
+functionality implemented in flowCore is available to other package
+authors, as we have seen with the afore-mentioned flowClust package.
+
+We can chose one of the many visualization option from the flowViz
+package to plot the results of the recent filtering operation. A very
+basic matrix of density plots is shown in Figure \ref{xyplot}, where
+each panel in the matrix represents the data of one individual
+patient.
+
+\begin{verbatim}
+> xyplot(`FL1-H` ~ `FL3-H` | SampleID, data=dat, filter=fres)
+\end{verbatim}
 
@@
 \caption{\label{xyplot}%
-Scatterplot matrix of a single FlowSet with outlines of the gating
+Scatterplot matrix of a single flowSet with outlines of the gating
 regions identified by an automated gating operation.}
@@
 
-Static gating for all samples in a high-throughput FCM experiment is
-often not an option, since the measured variables tend to vary between
-different treatments or over time. Automated or data-driven gating
-tries to estimate the gating regions from the underlying data, thus
-providing a fast objective solution to the analysis of potentially
-very large and diverse data sets \cite{lo2008agf}. One of the
-automated gating methods implemented in flowCore is based on
-identifying areas of significant curvature in a kernel density
-estimate of the data \citep{wand2008}. Assuming that the regions of
-interest are of high density, the software is able to reliably detect
-them in either the one or the two-dimensional density
-landscape. Kernel density estimation is a well-known problem in
-statistical computing, and a lot of effort has been invested in the
-development of good software to address it. The modular design of R,
-Bioconductor and flowCore allows to easily integrate these existing
-solutions into our framework. In this example, we directly use R code
-from the feature package developed by \cite{wand2008}. Instead of
-re-writing existing code, we are able to include it via the well
-tested distribution mechanism provided by R's software package
-system. This process is bi-directional, and all functionality
-implemented in flowCore is available to other package authors, as we
-have seen with the afore-mentioned flowClust package.
-
-\section*{Conclusions}
+\section*{Discussion}
 
 Through flowCore, we have provided the FCM community a freely
-available, highly functional, development and analysis platform for high
-throughput data analysis. Our intent is to provide a platform
-for the collaborative development of new analysis methods that
-will facilitate the transition of these new methods to the larger 
-flow community. Our experience has been that the 
-collaborative effort we enable for devising new methodology has been proven
-beneficial for a number of different biological and computational biology challenges. 
-We hope that our
-framework will be the foundation for fruitful shared research by many
-collaborators from multiple scientific fields and will help resolve
-bottlenecks that currently prevent the further development of and deployment of
-FC-HCS to increasingly complex and important scientific and clinical 
-applications.
+available, highly functional, development and analysis platform for
+high throughput data analysis. Our intent is to provide a platform for
+the collaborative development of new analysis methods that will
+facilitate the transition of these new methods to the larger flow
+community. Our experience has been that the collaborative effort we
+enable for devising new methodology has been proven beneficial for a
+number of different biological and computational biology challenges.
+We hope that our framework will be the foundation for fruitful shared
+research by many collaborators from multiple scientific fields and
+will help resolve bottlenecks that currently prevent the further
+development of and deployment of FC-HCS to increasingly complex and
+important scientific and clinical applications.
 

### 2008-07-25 paper/manuscript.tex
Changes Nolwenn suggested
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32910 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 clinical research applications \citep{brinkman2007hcf}. A serious
-bottleneck in the intepretation of existing studies and the
+bottleneck in the interpretation of existing studies and the
 application of FC-HCS to even larger, more complex problems is that
@@
 datasets, assessing measurement quality, developing validated assays,
-controling the accuracy of gating techniques, automating complex gating
+controlling the accuracy of gating techniques, automating complex gating
 strategies, and aggregating statistical results across large study
@@
 computational tools to efficiently analyze FC-HCS. This set of tools
-is based on the open source statistical software R, in conjunction
-with the Bioconductor Project. Our intent is to provide a shared
-research platform that enables bioinformaticians, computer scientists,
-and statisticians to work collaboratively with biologists and
-clinicians to develop novel methods \citep{lizard2007fca} for FCM data
-analysis. A critical component of this approach is the development and
-implementation of standards that will facilitate the adoption of these
-methods by both the larger research community and commercial software
-vendors.  Consequently, we are optimistic that this software platform
-will play a critical role in addressing the bottlenecks described
-above that limit further advances in the application of FC-HCS.
+is based on the open source statistical software R \citep{Rmain}, in
+conjunction with the Bioconductor Project \citep{BIOC}. Our intent is
+to provide a shared research platform that enables bioinformaticians,
+computer scientists, and statisticians to work collaboratively with
+biologists and clinicians to develop novel methods
+\citep{lizard2007fca} for FCM data analysis. A critical component of
+this approach is the development and implementation of standards that
+will facilitate the adoption of these methods by both the larger
+research community and commercial software vendors.  Consequently, we
+are optimistic that this software platform will play a critical role
+in addressing the bottlenecks described above that limit further
+advances in the application of FC-HCS.
 
 The computational tools we have developed are distributed in the R
-software language \citep{Rmain} as the Bioconductor
-\citep{BIOC} package flowCore. flowCore is a freely available,
-highly functional and extensible FCM data analysis platform that
-enables researchers to efficiently handle FC-HCS data and encourages
-open development of tools for their coherent analysis. Among the
-issues addressed by flowCore are computationally efficient data
-structures and a range of specialized methods for compensation,
-transformation, and gating.  flowCore runs on Windows, Mac OS X and
-Linux/Unix operating systems. Conceptually, the underlying data
-structures that build the core of our software can be implemented in
-any programming language, and we chose the R language for its obvious
-advantages in statistical programming. In this paper, we describe the
-functionality and methods underlying flowCore, and provide examples of
-their use.
+software language as the Bioconductor package flowCore. flowCore is a
+freely available, highly functional and extensible FCM data analysis
+platform that enables researchers to efficiently handle FC-HCS data
+and encourages open development of tools for their coherent
+analysis. Among the issues addressed by flowCore are computationally
+efficient data structures and a range of specialized methods for
+compensation, transformation, and gating.  flowCore runs on Windows,
+Mac OS X and Linux/Unix operating systems. Conceptually, the
+underlying data structures that build the core of our software can be
+implemented in any programming language, and we chose the R language
+for its obvious advantages in statistical programming. In this paper,
+we describe the functionality and methods underlying flowCore, and
+provide examples of their use.
 
@@
 fields of application. In particular, the Bioconductor project aims to
-provide R software modules for biolocial and clinical data analysis.
+provide R software modules for biological and clinical data analysis.
 It is widely used to process such data, spanning a diversity of
@@
 a high-throughput technology requires suitable software infrastructure
-to facilitate the rapid handling of the data obtained. We rely on two
-important lessons learned from the field of gene expression data
-analysis: the first being the importance of data structures that
-reflect the underlying data and facilitate the manipulations that are
-of most interest, while the second lesson being the importance of a
-modular architecture that allows for many developers to extend and use
-the underlying infrastructure and to combine its tools in complex
-workflows.
+to facilitate the rapid handling of the data. We rely on two important
+lessons learned from the field of gene expression data analysis: the
+first being the importance of data structures that reflect the
+underlying data and facilitate the manipulations that are of most
+interest, while the second lesson being the importance of a modular
+architecture that allows for many developers to extend and use the
+underlying infrastructure and to combine its tools in complex
+work flows.
 
@@
 the subjective character of manual analyses are a major obstacle to
-reproducibility. For FC-HCS data, unassisted manual inspection is too
-time consuming, and robust statistical methods need to be developed to
-point investigators to interesting aspects of the data, or to
-potential problems. While the expert knowledge of immunologists and
-researchers remains crucial for the understanding of FCM data, we
+reproducibility. For FC-HCS data, unassisted manual inspection is
+extremely time consuming, and robust statistical methods need to be
+developed to point investigators to interesting aspects of the data,
+or to potential problems. While the expert knowledge of immunologists
+and researchers remains crucial for the understanding of FCM data, we
 believe that collaboration with other research fields such as
@@
 Standards Task Force.  For example, the Gating-ML Candidate
-Recommendation (CR) represents a proposal on how to form unambiguous
-XML-based gate definitions, which can facilitate the interchange and
-validation of data between different software packages. Gates may be
-ordered into a hierarchical structure to describe a gating
-strategy. This enables the encoding of a simple analytical workflow,
-in a way that allows us to reconstruct the analysis programmatically.
+Recommendation (CR) \textit{Fixme Ryan or Joseph: How do we cite
+  this?}  represents a proposal on how to form unambiguous XML-based
+gate definitions, which can facilitate the interchange and validation
+of data between different software packages. Gates may be ordered into
+a hierarchical structure to describe a gating strategy. This enables
+the encoding of a simple analytical work flow, in a way that allows us
+to reconstruct the analysis programmatically.
 
@@
 implementation of the Gating-ML CR, which opens the possibility to
-integrate flowCore in existing workflows and to communicate with every
+integrate flowCore in existing work flows and to communicate with every
 other FCM tool that adheres to the proposed standard. Adherence to
@@
 based on flowCore to find there way back into the standard practices
-for FCM data analysis. Note that the data strucures defined in
+for FCM data analysis. Note that the data structures defined in
 flowCore aim to model large sets of FCM data in a natural and coherent
@@
 taking the burden of data management from the programmer, and by
-providing well-defined APIs and a structured class hierarchy, it is
-possible to readily test new ideas and to easily extend the
-framework's functionality.
+providing well-defined programming interfaces (APIs) and a structured
+class hierarchy, it is possible to readily test new ideas and to
+easily extend the framework's functionality.
 
@@
 single FCS file, i.e., the collection of events and the accompanying
-metadata, is stored in one single container. We call the structure
+meta data, is stored in one single container. We call the structure
 that hold this data a flowFrame. Raw data values as well as associated
-metadata of a flowFrame can be accessed programmatically. Most
+meta data of a flowFrame can be accessed programmatically. Most
 commonly, these are descriptors of the stains used in the experiment
@@
 structures and methods has already enabled the development of new
-Bioconductor packages, such as plateCore, that faciliate the analysis
+Bioconductor packages, such as plateCore, that facilitate the analysis
 of plate-based FC-HCS data and make handling the annotation and
@@
 \caption{\label{fig1:FrameWork}{For each experiment, FCS files (as
-    individual flowFrames), phenotypic and metadata are stored in a
+    individual flowFrames), phenotypic and meta data are stored in a
     flowSet. Each flowFrame in a flowSet corresponds to one FCS
@@
     gating) can be applied to either single flowFrames or a flowSet
-    simulataneously.}}
+    simultaneously.}}
 \end{figure}
@@
 own as first-class citizens in the analysis framework, without
-breaking the workflow or the basic infrastructure. This design allows
+breaking the work flow or the basic infrastructure. This design allows
 for the straightforward extension of flowCore's capabilities, and has
 already fostered the development of a number of valuable add-ons
-\citep{lo2008agf,sarkar2008ufv} \textit{(plateCore)}.
+\citep{lo2008agf,sarkar2008ufv} \textit{(Fixme Errol: Will plateCore
+  be in Bioconductor soon. It should be if we want to cite it.)}.
 
@@
 natural logarithm & $ln(x)(r/d)$ \\
-logarithm & log(x)(r/d) \\
+logarithm & $log(x)(r/d)$ \\
 biexponential & need equation \\
@@
 truncate & $x_{x<=a} = a$ \\
-scale & (x-a)/(b-a) \\
-split-scale & need equation \\
+scale & $(x-a)/(b-a)$ \\
+split-scale & combination of linear and logarithm \\
 arcsinh & $asinh(a + bx)+c$ \\
@@
 \hline
-rectangleGate & n-dinensional rectangular regions \\
+rectangleGate & n-dimensional rectangular regions \\
 quadGate & quadrant regions in two dimensions \\
@@
 polytopeGate & generalization of polygon in n dimensions \\
-ellipsoidGate & n-dimesnional ellipsoid region \\
+ellipsoidGate & n-dimensional ellipsoid region \\
 \hline
@@
 \hline
-sampleFilter & random subsampling\\
-expressionFilter & filter by resuts of a boolean expression \\
-kmeansFilter & based on K-means clustering \\
-norm2Filter & based on bivariate normal distribution \\
-curv1Filter & based on local density regions in 1D \\
-curv2Filter & based on local density regions in 2D \\
-timeFilter & based on abnormal data acquisition over time \\
+sampleFilter & random sub-sampling\\
+expressionFilter & results of a boolean expression \\
+kmeansFilter & K-means clustering \\
+norm2Filter & bivariate normal distribution \\
+curv1Filter & local density regions in 1D \\
+curv2Filter & density regions in 2D \\
+timeFilter & abnormal data acquisition over time \\
 \hline
@@
 and quality control. The objects annotation information can be used to
-arange the layout and composition of the plots. Furthermore, the
+arrange the layout and composition of the plots. Furthermore, the
 design and the API of the visualization software is very generic, and
@@
 exemplify some of the software's key features, that is, the coherent
-treatment of all samples in a potentially large expreriment, the
+treatment of all samples in a potentially large experiment, the
 concept of data-driven automated gating, the integration of existing
@@
 
-The report is interactive and provides drilldown to more detailed
+The report is interactive and provides drill-down to more detailed
 aspects of the analysis, starting from a concise overview
@@
 \caption{\label{flowQ}% 
-  HTML quality assesment report generated by
+  HTML quality assessment report generated by
   the flowQ package. Rows correspond to samples, columns to different

### 2008-08-09 paper/manuscript.tex
many comments - some modifications
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33145 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \usepackage{graphicx}
+
+%%FIXME: I would use some markup for the package name(s)
 
@@
 sufficiently far from the methods developed for traditional
-applications of flow cytometry (FCM) to small-scale, tube-based
-studies.
+small-scale, tube-based studies.
+%%FIXME: RG
+%applications of flow cytometry (FCM) to small-scale, tube-based
+%studies.
 
@@
 biologists and clinicians to develop novel methods  
+%%this reference is a bit odd here - you better explain what they say
+%%and why you quote them 
 \citep{lizard2007fca} for FCM data analysis. A critical component of
@@
 implemented in any programming language, and we chose the R language
-for its obvious advantages in statistical programming. In this paper,
-we describe the functionality and methods underlying flowCore, and
-provide examples of their use.
+for its advantages in statistical programming and visualization. In
+this paper, we describe the functionality and methods underlying
+flowCore, and provide examples of their use.
 
@@
 range of statistical and visualization methods developed for various
-fields of application. In particular, the Bioconductor project aims to
-provide R software modules for biological and clinical data analysis.
+fields of application. In particular, the Bioconductor project
+provides R software modules for biological and clinical data analysis.
 It is widely used to process such data, spanning a diversity of
-research fields from genomics to proteomics to cell biology
+research fields including genomics, proteomics and cell biology
 \citep{gentleman2006bos}. The R language and Bioconductor are
@@
 a high-throughput technology requires suitable software infrastructure
-to facilitate the rapid handling of the data. We rely on two important
-lessons learned from the field of gene expression data analysis: the
-first being the importance of data structures that reflect the
-underlying data and facilitate the manipulations that are of most
-interest, while the second lesson being the importance of a modular
-architecture that allows for many developers to extend and use the
-underlying infrastructure and to combine its tools in complex
-work flows.
+to facilitate the rapid handling of the data. In our implementation of
+flowCore we rely on two important lessons learned from the field of
+gene expression data analysis: the first being the importance of data
+structures that reflect the underlying data and facilitate the
+manipulations that are of most interest, while the second is the
+importance of a modular architecture that allows for many developers
+to extend and use the underlying infrastructure and to combine tools
+in complex work flows.
 
 \subsection*{Existing data standards and conventions}
+%%FIXME: I think that this section is mixing two ideas - the need for
+%%data representation - eg the class hierarchy, and the analysis
+%%tools. RG suggests starting by saying what they do -then point out
+%%that for complex experiments there is data on the samples (time
+%%points, drugs etc - we have examples to cite) and the need to
+%%combine these.  The notion of classes from an OOP language provide
+%%one coherent way to describe these richer data structures and
+%%functions/methods can work on them.  We will describe the classes
+%%and their implementation in R, but they could just as easily be
+%%written in any other language (eg Java, C++), and if
+%%programs/analysis tools written in those languages use similar
+%%structures then communication is simplified, and users could more
+%%easily move data between analysis tools.  I doubt that reflection is
+%%needed here - but if so, it would allow a naive function to get
+%%information about the data structure (it is hard for it to use that
+%%information though)
+
 Currently, data from FCM experiments are stored in single files
@@
 large and complex experimental designs with dozens of covariates and
-influencing factors. Self-contained and self-reflective data
+influencing factors. Self-contained reflective data
+%%FIXME: might need to explain reflective - or at least how you are
+%%going to use this - we may need a graphic that shows the flow data,
+%%data about the dyes antibodies but also about the samples etc - and
+%%the need to keep this coordinated through an analysis
 structures are a prerequisite to allow for coordinated analysis of
@@
 amenable to programmatical access. In addition, the closed-source
-nature of these products makes it impractical to integrate them into
+nature of these products often makes it impractical to integrate them into
 analysis pipelines.
@@
 
-The flowCore framework presented here offers import functionality for
-raw data FCS files along with their complete set of file-specific meta
-information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
-implementation of the Gating-ML CR, which opens the possibility to
-integrate flowCore in existing work flows and to communicate with every
-other FCM tool that adheres to the proposed standard. Adherence to
-standards also plays a critical role in the ability of new methods
+The flowCore framework presented here can import raw data FCS files
+along with their complete set of file-specific meta information
+(Figure~\ref{fig1:FrameWork}).  Moreover, it is a software
+implementation of the Gating-ML CR, which makes it possibile to
+integrate flowCore in existing work flows and to communicate with
+any other FCM tool that adheres to the proposed standard. Adherence
+to standards also plays a critical role in the ability of new methods
 based on flowCore to find there way back into the standard practices
@@
 
+%%FIXME: class hierarchy is probably not going to be well understood
+%%by readers - you might need to work that into a more coherent
+%%general discussion
 flowCore is not a GUI-driven software, and all operations are done
-using a command line interface. Conceptually, it is possible to add a
-more elaborate user interface on top of that, however the focus is on
-a programmatic approach to enable the convenient development of novel
-analysis methods and automation of complex analysis approaches. By
-taking the burden of data management from the programmer, and by
-providing well-defined programming interfaces (APIs) and a structured
-class hierarchy, it is possible to readily test new ideas and to
-easily extend the framework's functionality.
+using a command line interface.  It is possible to add a more
+elaborate user interface on top of this infrastructure, however the
+focus in this paper is on a programmatic approach to enable the
+convenient development of novel analysis methods and automation of
+complex analysis approaches.  By taking the burden of data management
+from the programmer, and by providing well-defined programming
+interfaces (APIs) and a structured class hierarchy, it is possible to
+readily test new ideas and to easily extend the framework's
+functionality.
 
@@
 \subsubsection*{flowFrame}
+
 FlowCore's primary task is the representation and basic manipulation
@@
 that hold this data a flowFrame. Raw data values as well as associated
-meta data of a flowFrame can be accessed programmatically. Most
-commonly, these are descriptors of the stains used in the experiment
-and the respective measurement channels, information about
-compensation performed at the instrument side and any additional
+metadata of a flowFrame can be accessed programmatically. Most
+commonly, the metadata consist of descriptors of the stains used in
+the experiment and the respective measurement channels, information
+about compensation performed at the instrument side and any additional
 keywords the user deems to be important to annotate the data. A number
@@
 
+%%again some markup for the class names would help the reader
+%% stick with annotation or metadata not both
 \subsubsection*{flowSet}
@@
 introduce the concept of a collection of flowFrames called a
-flowSet. One can envision a flowSet to be a container for multiple
-flowFrames along with relevant information associated with each
-individual frame such as descriptions of the cell sample or the
-treatment to which the sample was subjected. FlowSets manage the
-consistent application of operations on the individual flowFrames and
-shift the burden of keeping score of the annotation from the user to
-the infrastructure, thus reducing the risk of errors (\textit{e.g.},
-mixups of sample labels). The flowSet structure can be readily
-extended to incorporate the potentially complex annotation associated
-with even larger FC-HCS experiments. Access to flowCore data
-structures and methods has already enabled the development of new
-Bioconductor packages, such as plateCore, that facilitate the analysis
-of plate-based FC-HCS data and make handling the annotation and
-processing the data easier.
+flowSet. A flowSet is a container for multiple flowFrames along with
+relevant information associated with each individual frame such as
+descriptions of the cell sample, the treatment to which the sample was
+subjected, or the location of that sample in a 96 or 384 well plate.
+FlowSets manage the consistent application of operations on the
+individual flowFrames and shift the burden of keeping score of the
+annotation from the user to the infrastructure, thus reducing the risk
+of errors (\textit{e.g.}, mixups of sample labels). The flowSet
+structure can be readily extended to incorporate the potentially
+complex annotation associated with even larger FC-HCS experiments such
+as clinical trials where hundreds patients might provide samples at
+different points in time over the course of the experiment.  Access to
+flowCore data structures and methods has already enabled the
+development of new Bioconductor packages, such as plateCore, that
+facilitate the analysis of plate-based FC-HCS data and make handling
+the annotation and processing the data easier.
 
@@
 
+%%FIXME: if possible I would make the box that has Sample Meta-data in
+%%it, be more of a table view (Age Sex Time with numbers and then it
+%%might be more obvious what is happening
+
 \subsection*{Standard flow operations}
+
 The basic operations in FCM analyses are typically the same: the data
-need to be compensated (if not already done on the instrument side)
+need to be compensated (if that was not already done on the instrument)
 and transformed, and sup-populations of interest need to be selected
 based on a set of (predominantly sequential) gates. All software
-solutions for FCM analysis offer more or less elaborate support for
-these kinds of operation in an interactive, GUI-driven interface. The
+solutions for FCM analysis offer support for
+these operations, typically in an interactive, GUI-driven interface. The
 approach we have taken in flowCore is to abstractly describe these
@@
 \subsubsection*{Transformation and compensation}
+
 Data transformation is essential for both data visualization and
@@
 \subsubsection*{Gating}
+%%FIXME: I would make more of the filters and possibly try to make it
+%%clearer how much stuff thre is for composition of gates/filters etc
+%%
+%% somewhere notions of apply and summary methods should appear and if
+%% we have anything figured out yet on how to do the subsetting etc
+%% that too would go nicely in here (here I am thinking of our
+%% discussion of the sort of workflow stuff)
+
 In flowCore, gating operations are represented by classes that can be
@@
 to subset data sets as well as to create summary statistics, for
-instance the proportions of events falling in single gates or in the
-combinations of gates. Complex combinations and hierarchies of gates
+example the proportions of events falling in a single gate or in a
+combination of gates. Complex combinations and hierarchies of gates
 can be captured in objects of class filterSet, allowing to apply
@@
 \subsection*{Related flow packages}
+
 In addition to the flowCore package that offers basic infrastructure,
@@
 package \citep{sarkar2008ufv} provides sophisticated data
-visualization tools, that make use of multivariate trellis
-plotting. These functions can be used to quickly generate customized
-plots for extended cytometry data sets for both direct data inspection
-and quality control. The objects annotation information can be used to
-arrange the layout and composition of the plots. Furthermore, the
-design and the API of the visualization software is very generic, and
-users can readily extend its capabilities by providing self-defined
-plotting functions.  The flowQ package offers more advanced quality
-assurance (QA) methodology and a framework to create interactive
-web-based reports of QA results. Most of flowUtil's content deals with
-data import and output from and to flowCore and the flow-cytometry
-specific standard markup languages. Finally, flowStats provides more
+visualization tools, that make use of multivariate trellis plotting
+\citep{FIXME:citeDeepayansBook}.  These functions can be used to
+quickly generate customized plots for extended cytometry data sets for
+both direct data inspection and quality control.  The objects
+annotation information can be used to arrange the layout and
+composition of the plots.  Furthermore, the design and the API of the
+visualization software is very generic, and users can readily extend
+its capabilities by providing self-defined plotting functions.  The
+flowQ package offers more advanced quality assurance (QA) methodology
+and a framework to create interactive web-based reports of QA results.
+Most of flowUtil's content deals with data import and export using the
+flow-cytometry specific standard markup languages (FIXME:it implements
+the standards doesn't it?).  Finally, flowStats provides more
 elaborate statistical methods that are relevant in the context of flow
@@
 \section*{Results}
+%%FIXME: we seem to mention other packages a lot - some reduction and
+%%cohesion is going to be needed
+
 The flowCore package has been successfully applied in the analysis of
@@
 
-A complete documentation of the flowCore software is way beyond the
-scope of this publication. Much more comprehensive documentation and
-users guide information with programmatic examples is available online
+A complete documentation of the flowCore software is beyond the scope
+of this publication.  Much more comprehensive documentation and users
+guide information with programmatic examples is available online
 (\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
-and as part of the package distribution. Here, we only want to briefly
+and as part of the package distribution. Here, we want to briefly
 exemplify some of the software's key features, that is, the coherent
@@
 Through flowCore, we have provided the FCM community a freely
-available, highly functional, development and analysis platform for
-high throughput data analysis. Our intent is to provide a platform for
-the collaborative development of new analysis methods that will
-facilitate the transition of these new methods to the larger flow
-community. Our experience has been that the collaborative effort we
-enable for devising new methodology has been proven beneficial for a
-number of different biological and computational biology challenges.
-We hope that our framework will be the foundation for fruitful shared
-research by many collaborators from multiple scientific fields and
-will help resolve bottlenecks that currently prevent the further
-development of and deployment of FC-HCS to increasingly complex and
-important scientific and clinical applications.
+available, highly functional, standards compliant development and
+analysis platform for high throughput data analysis.  Our intention is
+to provide a platform for the collaborative development of new
+analysis methods that will facilitate the transition of these new
+methods to the larger flow community.  Our experience has been that the
+collaborative effort we enable for devising new methodology has been
+proven beneficial for a number of different biological and
+computational biology challenges.  We hope that our framework will be
+the foundation for fruitful shared research by many collaborators from
+multiple scientific fields and will help resolve bottlenecks that
+currently prevent the further development of and deployment of FC-HCS
+to increasingly complex and important scientific and clinical
+applications.
 

### 2008-08-11 paper/manuscript.tex
changes to the manuscript based on suggestions by Robert
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33188 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \usepackage{graphicx} 
-
-%%FIXME: I would use some markup for the package name(s)
+\newcommand{\Rpackage}[1]{{\textsf{#1}}}
+\newcommand{\Rclass}[1]{{\textit{#1}}}
 
@@
 
+\section*{Abstract}
+\subsubsection*{Background}
+Recent advances in automation technologies have enabled the use of
+flow-cytometry high content screening (FH-HCS). However, data
+management and data analysis methods have not advanced sufficiently
+far from the inital small-scale studies to support the large complex
+data sets generated in FH-HCS applications.
+\subsubsection*{Methods}
+In this manuscript, we present \Rpackage{flowCore} a set of flexible
+and well structured computational tools for FC-HCS data analysis based
+on the open source statistical software R in conjunction with the
+Bioconductor Project.
+\subsubsection*{Results}
+Through \Rpackage{flowCore}, we provide the FCM community a shared
+research platform that enables bioinformaticians, computer scientists,
+and statisticians to work collaboratively with biologists and
+clinicians to develop novel methods for FCM data analysis.
+\subsubsection*{Conclusions}
+We hope that our framework will be the foundation for fruitful shared
+research by many collaborators from multiple scientific fields and
+will help resolve bottlenecks that currently prevent the further
+development of and deployment of FC-HCS to increasingly complex and
+important scientific and clinical applications.
+
 \section*{Introduction}
@@
 data management and data analysis methods have not advanced
-sufficiently far from the methods developed for traditional
-small-scale, tube-based studies.
-%%FIXME: RG
-%applications of flow cytometry (FCM) to small-scale, tube-based
-%studies.
+sufficiently far from the methods developed for applications of flow
+cytometry (FCM) to small-scale, tube-based studies.
 
@@
 datasets, assessing measurement quality, developing validated assays,
-controlling the accuracy of gating techniques, automating complex gating
-strategies, and aggregating statistical results across large study
-sets for further analysis.  Among the most serious consequences of the
-current situation is, however, that it is very difficult for new
-analysis approaches to find there way into standard practice. We
+controlling the accuracy of gating techniques, automating complex
+gating strategies, and aggregating statistical results across large
+study sets for further analysis.  Among the most serious consequences
+of the current situation is, however, that it is very difficult for
+new analysis approaches to find there way into standard practice. We
 believe that this barrier to the development and dissemination of new
@@
 computer scientists, and statisticians to work collaboratively with
-biologists and clinicians to develop novel methods  
-%%this reference is a bit odd here - you better explain what they say
-%%and why you quote them 
-\citep{lizard2007fca} for FCM data analysis. A critical component of
-this approach is the development and implementation of standards that
-will facilitate the adoption of these methods by both the larger
-research community and commercial software vendors.  Consequently, we
-are optimistic that this software platform will play a critical role
-in addressing the bottlenecks described above that limit further
-advances in the application of FC-HCS.
+biologists and clinicians to develop novel methods for FCM data
+analysis, a process deemed crucial by many for the further development
+of the technology \citep{lizard2007fca}. A critical component of this
+approach is the development and implementation of standards that will
+facilitate the adoption of these methods by both the larger research
+community and commercial software vendors.  Consequently, we are
+optimistic that this software platform will play a critical role in
+addressing the bottlenecks described above that limit further advances
+in the application of FC-HCS.
 
 The computational tools we have developed are distributed in the R
-software language as the Bioconductor package flowCore. flowCore is a
-freely available, highly functional and extensible FCM data analysis
-platform that enables researchers to efficiently handle FC-HCS data
-and encourages open development of tools for their coherent
-analysis. Among the issues addressed by flowCore are computationally
-efficient data structures and a range of specialized methods for
-compensation, transformation, and gating.  flowCore runs on Windows,
-Mac OS X and Linux/Unix operating systems. Conceptually, the
-underlying data structures that build the core of our software can be
-implemented in any programming language, and we chose the R language
-for its advantages in statistical programming and visualization. In
-this paper, we describe the functionality and methods underlying
-flowCore, and provide examples of their use.
+software language as the Bioconductor package
+\Rpackage{flowCore}. \Rpackage{flowCore} is a freely available, highly
+functional and extensible FCM data analysis platform that enables
+researchers to efficiently handle FC-HCS data and encourages open
+development of tools for their coherent analysis. Among the issues
+addressed by \Rpackage{flowCore} are computationally efficient data
+structures and a range of specialized methods for compensation,
+transformation, and gating.  \Rpackage{flowCore} runs on Windows, Mac
+OS X and Linux/Unix operating systems. Conceptually, the underlying
+data structures that build the core of our software can be implemented
+in any programming language, and we chose the R language for its
+advantages in statistical programming and visualization. In this
+paper, we describe the functionality and methods underlying
+\Rpackage{flowCore}, and provide examples of their use.
 
@@
 research fields including genomics, proteomics and cell biology
-\citep{gentleman2006bos}. The R language and Bioconductor are
+\citep{BIOC}. The R language and Bioconductor are
 particularly geared towards the analysis of large complex datasets
@@
 to facilitate the rapid handling of the data. In our implementation of
-flowCore we rely on two important lessons learned from the field of
+\Rpackage{flowCore} we rely on two important lessons learned from the field of
 gene expression data analysis: the first being the importance of data
@@
 \subsection*{Existing data standards and conventions}
-%%FIXME: I think that this section is mixing two ideas - the need for
-%%data representation - eg the class hierarchy, and the analysis
-%%tools. RG suggests starting by saying what they do -then point out
-%%that for complex experiments there is data on the samples (time
-%%points, drugs etc - we have examples to cite) and the need to
-%%combine these.  The notion of classes from an OOP language provide
-%%one coherent way to describe these richer data structures and
-%%functions/methods can work on them.  We will describe the classes
-%%and their implementation in R, but they could just as easily be
-%%written in any other language (eg Java, C++), and if
-%%programs/analysis tools written in those languages use similar
-%%structures then communication is simplified, and users could more
-%%easily move data between analysis tools.  I doubt that reflection is
-%%needed here - but if so, it would allow a naive function to get
-%%information about the data structure (it is hard for it to use that
-%%information though)
+Traditionally, the majority of FCM experiments have been analyzed by
+manual data inspection in one or two dimensions, or by very basic
+comparisons of summary statistics and most of the currently available
+analysis tools are designed to reflect this workflow.  We believe that
+these approaches, in addition to being expensive and labor intensive,
+do not fully address the highly complex nature of FCM data; in
+particular, they disregard many of the fundamental aspects of the
+data, such as its underlying distribution and its high-dimensional
+nature. Furthermore, the subjective character of manual analyses are a
+major obstacle to reproducibility. For FC-HCS data, unassisted manual
+inspection is extremely time consuming, and robust statistical methods
+need to be developed to point investigators to interesting aspects of
+the data, or to potential problems. While the expert knowledge of
+immunologists and researchers remains crucial for the understanding of
+FCM data, we believe that collaboration with other research fields
+such as statistics and computer science can greatly improve the
+relevance of FCM in today's high-throughput paradigm.
 
@@
 large and complex experimental designs with dozens of covariates and
-influencing factors. Self-contained reflective data
-%%FIXME: might need to explain reflective - or at least how you are
-%%going to use this - we may need a graphic that shows the flow data,
-%%data about the dyes antibodies but also about the samples etc - and
-%%the need to keep this coordinated through an analysis
-structures are a prerequisite to allow for coordinated analysis of
-such experiments.  Many of the currently available software solutions
-offer only limited support for such structures, or make use of
-containers like XML or binary storage formats that are designed
+influencing factors. For example, experiments consist of large numbers
+of samples from different patients, measured at different timepoints
+\citep{brinkman2007hcf} or following different drug treatments
+\citep{gasparetto2004ice}. Modern FCM data analysis tools have to deal
+with an additional layer of sample metadata and they need to provide
+infrastructure to process and to compare groups of samples in a
+concise and coordinated manner.
+
+The notion of classes from an object-oriented programming language
+provides one coherent way to describe these richer data structures,
+and functions or methods that work on them allow for interaction and
+manipulation. Many of the currently available software solutions offer
+only limited support for such self-contained structures, or make use
+of containers like XML or binary storage formats that are designed
 specifically for particular user interfaces and hence are not easily
 amenable to programmatical access. In addition, the closed-source
-nature of these products often makes it impractical to integrate them into
-analysis pipelines.
+nature of these products often makes it impractical to integrate them
+into analysis pipelines. In this manuscript we describe classes for
+FCM data analysis and their implementation in R, however, they could
+just as easily be implemented in any other language (eg Java,
+C++). Software written in those languages could use similar data
+structures, thereby simplifying communication and the interchange of
+data between analysis tools.
 
-Traditionally, the majority of FCM experiments have been analyzed by
-manual data inspection in one or two dimensions, or by very basic
-comparisons of summary statistics. We believe that these approaches,
-in addition to being extremely expensive and labor intensive, do not
-fully address the highly complex nature of FCM data; in particular,
-they discard many of the fundamental aspects of the data, such as its
-underlying distribution and its high-dimensional nature. Furthermore,
-the subjective character of manual analyses are a major obstacle to
-reproducibility. For FC-HCS data, unassisted manual inspection is
-extremely time consuming, and robust statistical methods need to be
-developed to point investigators to interesting aspects of the data,
-or to potential problems. While the expert knowledge of immunologists
-and researchers remains crucial for the understanding of FCM data, we
-believe that collaboration with other research fields such as
-statistics and computer science can greatly improve the relevance of
-FCM in today's high-throughput paradigm.
-
-These issues are currently being addressed by emerging new flow
-cytometry data standards developed in collaboration with the ISAC Data
-Standards Task Force.  For example, the Gating-ML Candidate
-Recommendation (CR) \textit{Fixme Ryan or Joseph: How do we cite
-  this?}  represents a proposal on how to form unambiguous XML-based
-gate definitions, which can facilitate the interchange and validation
-of data between different software packages. Gates may be ordered into
-a hierarchical structure to describe a gating strategy. This enables
-the encoding of a simple analytical work flow, in a way that allows us
-to reconstruct the analysis programmatically.
-
-The flowCore framework presented here can import raw data FCS files
-along with their complete set of file-specific meta information
-(Figure~\ref{fig1:FrameWork}).  Moreover, it is a software
-implementation of the Gating-ML CR, which makes it possibile to
-integrate flowCore in existing work flows and to communicate with
-any other FCM tool that adheres to the proposed standard. Adherence
-to standards also plays a critical role in the ability of new methods
-based on flowCore to find there way back into the standard practices
-for FCM data analysis.  Note that the data structures defined in
-flowCore aim to model large sets of FCM data in a natural and coherent
-way; they are not specific to the R language and can be implemented in
-any other programming language. Ideally, different software programs
-would communicate on the level of these data structures rather than a
-high-level markup language.
-
-%%FIXME: class hierarchy is probably not going to be well understood
-%%by readers - you might need to work that into a more coherent
-%%general discussion
-flowCore is not a GUI-driven software, and all operations are done
-using a command line interface.  It is possible to add a more
-elaborate user interface on top of this infrastructure, however the
-focus in this paper is on a programmatic approach to enable the
-convenient development of novel analysis methods and automation of
-complex analysis approaches.  By taking the burden of data management
-from the programmer, and by providing well-defined programming
-interfaces (APIs) and a structured class hierarchy, it is possible to
-readily test new ideas and to easily extend the framework's
+The \Rpackage{flowCore} framework presented here can import and
+process raw data FCS files along with their complete set of
+file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
+a software implementation of the Gating-ML CR, an emerging new flow
+cytometry data standard developed in collaboration with the ISAC Data
+Standards Task Force, which makes it possibile to integrate
+\Rpackage{flowCore} in existing work flows and to communicate with any
+other FCM tool that adheres to the proposed standard. Adherence to
+standards also plays a critical role in the ability of new methods
+based on \Rpackage{flowCore} to find their way back into the standard
+practices for FCM data analysis. \Rpackage{flowCore} is not a
+GUI-driven software, and all operations are done using a command line
+interface.  It is possible to add a more elaborate user interface on
+top of this infrastructure, however the focus in this paper is on a
+programmatic approach to enable the convenient development of novel
+analysis methods and automation of complex analysis approaches.  By
+taking the burden of data management from the programmer, and by
+providing well-defined programming interfaces (APIs), it is possible
+to readily test new ideas and to easily extend the framework's
 functionality.
@@
 
-FlowCore's primary task is the representation and basic manipulation
-of FCM data. This is accomplished through a data model very close to
-that adopted by other Bioconductor packages. All information from a
-single FCS file, i.e., the collection of events and the accompanying
-meta data, is stored in one single container. We call the structure
-that hold this data a flowFrame. Raw data values as well as associated
-metadata of a flowFrame can be accessed programmatically. Most
-commonly, the metadata consist of descriptors of the stains used in
-the experiment and the respective measurement channels, information
-about compensation performed at the instrument side and any additional
-keywords the user deems to be important to annotate the data. A number
-of quality checks are performed during the creation of a flowFrame to
-ensure data integrity.
+\Rpackage{flowCore}'s primary task is the representation and basic
+manipulation of FCM data. This is accomplished through a data model
+very close to that adopted by other Bioconductor packages. All
+information from a single FCS file, i.e., the collection of events and
+the accompanying metadata, is stored in one single container. We call
+the structure that hold this data a \Rclass{flowFrame}. Raw data
+values as well as associated metadata of a \Rclass{flowFrame} can be
+accessed programmatically. Most commonly, the metadata consist of
+descriptors of the stains used in the experiment and the respective
+measurement channels, information about compensation performed at the
+instrument side and any additional keywords the user deems to be
+important to annotate the data. A number of quality checks are
+performed during the creation of a \Rclass{flowFrame} to ensure data
+integrity.
 
-%%again some markup for the class names would help the reader
-%% stick with annotation or metadata not both
 \subsubsection*{flowSet}
@@
 performed consistently across multiple flow measurements, hence we
-introduce the concept of a collection of flowFrames called a
-flowSet. A flowSet is a container for multiple flowFrames along with
-relevant information associated with each individual frame such as
-descriptions of the cell sample, the treatment to which the sample was
-subjected, or the location of that sample in a 96 or 384 well plate.
-FlowSets manage the consistent application of operations on the
-individual flowFrames and shift the burden of keeping score of the
-annotation from the user to the infrastructure, thus reducing the risk
-of errors (\textit{e.g.}, mixups of sample labels). The flowSet
-structure can be readily extended to incorporate the potentially
-complex annotation associated with even larger FC-HCS experiments such
-as clinical trials where hundreds patients might provide samples at
-different points in time over the course of the experiment.  Access to
-flowCore data structures and methods has already enabled the
-development of new Bioconductor packages, such as plateCore, that
-facilitate the analysis of plate-based FC-HCS data and make handling
-the annotation and processing the data easier.
+introduce the concept of a collection of \Rclass{flowFrames} called a
+\Rclass{flowSet}. A \Rclass{flowSet} is a container for multiple
+\Rclass{flowFrames} along with relevant information associated with
+each individual frame such as descriptions of the cell sample, the
+treatment to which the sample was subjected, or the location of that
+sample in a 96 or 384 well plate.  \Rclass{flowSets} manage the
+consistent application of operations on the individual
+\Rclass{flowFrames} and shift the burden of keeping score of the
+metadata from the user to the infrastructure, thus reducing the risk
+of errors (\textit{e.g.}, mixups of sample labels). The
+\Rclass{flowSet} structure can be readily extended to incorporate the
+potentially complex metadata associated with even larger FC-HCS
+experiments such as clinical trials where hundreds patients might
+provide samples at different points in time over the course of the
+experiment.
 
@@
 \caption{\label{fig1:FrameWork}{For each experiment, FCS files (as
-    individual flowFrames), phenotypic and meta data are stored in a
-    flowSet. Each flowFrame in a flowSet corresponds to one FCS
-    file. All basic operations (e.g., compensation, transformation,
-    gating) can be applied to either single flowFrames or a flowSet
+    individual \Rclass{flowFrames}), phenotypic and metadata are
+    stored in a \Rclass{flowSet}. Each \Rclass{flowFrame} in a
+    \Rclass{flowSet} corresponds to one FCS file. All basic operations
+    (e.g., compensation, transformation, gating) can be applied to
+    either single \Rclass{flowFrames} or a \Rclass{flowSet}
     simultaneously.}}
 \end{figure}
-
-%%FIXME: if possible I would make the box that has Sample Meta-data in
-%%it, be more of a table view (Age Sex Time with numbers and then it
-%%might be more obvious what is happening
 
@@
 The basic operations in FCM analyses are typically the same: the data
-need to be compensated (if that was not already done on the instrument)
-and transformed, and sup-populations of interest need to be selected
-based on a set of (predominantly sequential) gates. All software
-solutions for FCM analysis offer support for
-these operations, typically in an interactive, GUI-driven interface. The
-approach we have taken in flowCore is to abstractly describe these
-operations and build a set of tools to perform them on both flowFrames
-and flowSets. While transformation, and to a certain extent
-compensation, are fairly routine operations with only limited
-potential for improvement, being able to implement new methodologies
-for gating of FCM data, and extending the capabilities of flowCore
-through object oriented programming are features that clearly sets our
-framework apart from other FCM analysis tools. By factoring out as
-much of the bookkeeping as possible, programmers can focus on the
-actual operations rather then having to deal with the tedious details
-of data integration and access. Third-party methods can act on their
-own as first-class citizens in the analysis framework, without
-breaking the work flow or the basic infrastructure. This design allows
-for the straightforward extension of flowCore's capabilities, and has
-already fostered the development of a number of valuable add-ons
+need to be compensated (if that was not already done on the
+instrument) and transformed, and sup-populations of interest need to
+be selected based on a set of (predominantly sequential) gates. All
+software solutions for FCM analysis offer support for these
+operations, typically in an interactive, GUI-driven interface. The
+approach we have taken in \Rpackage{flowCore} is to abstractly
+describe these operations and build a set of tools to perform them on
+both \Rclass{flowFrames} and \Rclass{flowSets}. While transformation,
+and to a certain extent compensation, are fairly routine operations
+with only limited potential for improvement, being able to implement
+new methodologies for gating of FCM data, and extending the
+capabilities of \Rpackage{flowCore} through object oriented
+programming are features that clearly sets our framework apart from
+other FCM analysis tools. By factoring out as much of the bookkeeping
+as possible, programmers can focus on the actual operations rather
+then having to deal with the tedious details of data integration and
+access. Third-party methods can act on their own as first-class
+citizens in the analysis framework, without breaking the work flow or
+the basic infrastructure. This design allows for the straightforward
+extension of \Rpackage{flowCore}'s capabilities, and has already
+fostered the development of a number of valuable add-ons
 \citep{lo2008agf,sarkar2008ufv} \textit{(Fixme Errol: Will plateCore
@@
 modelling \citep{lo2008agf}. The major transformations that are
-routinely used in FCM analysis have been implemented in flowCore
-(\textit{e.g.,} quadratic, log or arcsinh, see Table~\ref{table1} for a
-complete list). Furthermore, the design of the R language makes it
-easy to define arbitrary functions to apply to the data of individual
-flowFrames or entire flowSets, respectively. Compensation is
-available for both flowFrames and flowSets, and the software also
+routinely used in FCM analysis have been implemented in
+\Rpackage{flowCore} (\textit{e.g.,} quadratic, log or arcsinh, see
+Table~\ref{table1} for a complete list). Furthermore, the design of
+the R language makes it easy to define arbitrary functions to apply to
+the data of individual \Rclass{flowFrames} or entire
+\Rclass{flowSets}, respectively. Compensation is available for both
+\Rclass{flowFrames} and \Rclass{flowSets}, and the software also
 offers functionality to compute spillover or compensation matrices
@@
 logarithm & $log(x)(r/d)$ \\
-biexponential & need equation \\
-logicle& need equation \\
+biexponential & $root(a*e^{(b*y)}-c*e^{(-d+y)}+f)$ \\
+logicle& $root(Te^{-(m-w)}*(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
 truncate & $x_{x<=a} = a$ \\
@@
 \end{tabular}
-\caption{\label{table1}Data transformations implemented in flowCore.}
+\caption{\label{table1}Data transformations implemented in
+  \Rpackage{flowCore}.}
 \end{center}
@@
 
-In flowCore, gating operations are represented by classes that can be
-extended in an object-oriented manner (Table~\ref{table2}). Basic gate
-types such as rectangular gates, ellipses and polygon gates are
-implemented as part of the framework. In addition, we introduce the
-notion of data-driven gates, or filters, for which the necessary
-parameters are computed based on the properties of the underlying
-data, for instance by modelling data distribution or by density
-estimation. The ability to programmatically access gates is a
-prerequisite for automated or semi-automated gating. By utilizing a
-unified interface for all different types of gates, the user is able
-to subset data sets as well as to create summary statistics, for
-example the proportions of events falling in a single gate or in a
-combination of gates. Complex combinations and hierarchies of gates
-can be captured in objects of class filterSet, allowing to apply
-multi-step gating strategies. The definition of gates in flowCore
-follows the Gating-ML CR, thus any flowCore gating strategy can be
-reproduced by any other software that also adheres to the standard.
+In \Rpackage{flowCore}, gating operations are represented by classes
+that can be extended in an object-oriented manner
+(Table~\ref{table2}). Basic gate types such as rectangular gates,
+ellipses and polygon gates are implemented as part of the
+framework. In addition, we introduce the notion of data-driven gates,
+or filters, for which the necessary parameters are computed based on
+the properties of the underlying data, for instance by modelling data
+distribution or by density estimation. This approach is fundamentally
+different from the traditional application of static gating regions
+across samples, as it is able to take into accounts unforeseen changes
+in signal intensities. \textit{e.g.} drifts in the instrumentation
+over time or sample variablity.
+
+The ability to programmatically access gates is a prerequisite for
+automated or semi-automated gating. By utilizing a unified interface
+for all different types of gates, the user is able to subset data sets
+as well as to create summary statistics, for example the proportions
+of events falling in a single gate or in a combination of
+gates. Complex combinations and hierarchies of gates can be captured
+in objects of class \Rclass{filterSet}, allowing to apply multi-step
+gating strategies. The definition of gates in \Rpackage{flowCore}
+follows the Gating-ML CR, thus any \Rpackage{flowCore} gating strategy
+can be reproduced by any other software that also adheres to the
+standard.
+
+Gating, as well as all other operations in flowCore, can be applied
+over each individual frame in a \Rclass{flowSet}, and summary methods
+provide information about the outcome of these operations. In
+addition, the result of a gating operation can be used to subset the
+input \Rclass{flowFrame} or \Rclass{flowSet}, either by filtering out
+negative events or by splitting in multiple subpopulations. This
+design allows to easily combine all of \Rpackage{flowCore}'s
+components into complex work flows.
 
@@
 \end{tabular}
-\caption{\label{table2}Filter and gate classes implemented in flowCore.}
+\caption{\label{table2}Filter and gate classes implemented in
+  \Rpackage{flowCore}.}
 \end{center}
@@
 
-In addition to the flowCore package that offers basic infrastructure,
-we have implemented a range of additional Bioconductor packages that
-are dedicated to more specific tasks of FCM data analysis. The flowViz
-package \citep{sarkar2008ufv} provides sophisticated data
-visualization tools, that make use of multivariate trellis plotting
-\citep{FIXME:citeDeepayansBook}.  These functions can be used to
-quickly generate customized plots for extended cytometry data sets for
-both direct data inspection and quality control.  The objects
-annotation information can be used to arrange the layout and
-composition of the plots.  Furthermore, the design and the API of the
+In addition to the \Rpackage{flowCore} package that offers basic
+infrastructure, we have implemented a range of additional Bioconductor
+packages that are dedicated to more specific tasks of FCM data
+analysis. The \Rpackage{flowViz} package \citep{sarkar2008ufv}
+provides sophisticated data visualization tools, that make use of
+multivariate trellis plotting \citep{lattice}.  These functions can be
+used to quickly generate customized plots for extended cytometry data
+sets for both direct data inspection and quality control.  The objects
+metadata information can be used to arrange the layout and composition
+of the plots.  Furthermore, the design and the API of the
 visualization software is very generic, and users can readily extend
 its capabilities by providing self-defined plotting functions.  The
-flowQ package offers more advanced quality assurance (QA) methodology
-and a framework to create interactive web-based reports of QA results.
-Most of flowUtil's content deals with data import and export using the
-flow-cytometry specific standard markup languages (FIXME:it implements
-the standards doesn't it?).  Finally, flowStats provides more
+\Rpackage{flowQ} package offers more advanced quality assurance (QA)
+methodology and a framework to create interactive web-based reports of
+QA results.  Most of \Rpackage{flowUtil}'s content deals with data
+import and export, implementing the flow-cytometry specific standard
+markup language.  Finally, \Rpackage{flowStats} provides more
 elaborate statistical methods that are relevant in the context of flow
-cytometry data analysis.
+cytometry data analysis. \cite{lo2008agf} have recently developed an
+automatic gating approach via robust model-based clustering using
+\Rpackage{flowCore}'s data model and infrastructure which is
+implemented in the Bioconductor package \Rpackage{flowClust} Another
+package, \Rpackage{plateCore}, providing more specialized support for
+experiments conducted on microtitre plates and facilitiating the
+handling of spatial metadata, is under development.
 
 \section*{Results}
-%%FIXME: we seem to mention other packages a lot - some reduction and
-%%cohesion is going to be needed
-
-The flowCore package has been successfully applied in the analysis of
-several datasets \citep{gasparetto2004ice,brinkman2007hcf}, and
-several other Bioconductor packages are making use of its code
-base. As an example, \cite{lo2008agf} have recently developed an
-automatic gating approach via robust model-based clustering using
-flowCore's data model and infrastructure and implemented this in the
-Bioconductor package flowClust. Another package, plateCore, providing
-more specialized support for experiments conducted on microtitre
-plates is under development.
-
-A complete documentation of the flowCore software is beyond the scope
-of this publication.  Much more comprehensive documentation and users
-guide information with programmatic examples is available online
+The \Rpackage{flowCore} package has been successfully applied in the
+analysis of several datasets
+\citep{gasparetto2004ice,brinkman2007hcf}. A complete documentation of
+the \Rpackage{flowCore} software is beyond the scope of this
+publication.  Much more comprehensive documentation and users guide
+information with programmatic examples is available online
 (\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
@@
 Data analysis for most experiments usually begins with a quality
-assurance step, and we can use functionality from the flowQ package to
-create an HTML report that highlights potential quality
+assurance step, and we can use functionality from the \Rpackage{flowQ}
+package to create an HTML report that highlights potential quality
 issues. Assuming that the data has already been imported as the
-flowSet object ``dat'', the following simple line of code produces the
-quite complex output:
+\Rclass{flowSet} object ``dat'', the following simple line of code
+produces the quite complex output:
 
@@
 aspects of the analysis, starting from a concise overview
-(Figure~\ref{flowQ}). The design of flowCore's data model allows for a
-coherent treatment of all the samples, hence we are able to compare
-features between individuals, or between groups of individuals, based
-on the available annotation information.
+(Figure~\ref{flowQ}). The design of \Rpackage{flowCore}'s data model
+allows for a coherent treatment of all the samples, hence we are able
+to compare features between individuals, or between groups of
+individuals, based on the available metadata information.
 
@@
 \caption{\label{flowQ}%
-  HTML quality assessment report generated by
-  the flowQ package. Rows correspond to samples, columns to different
-  quality checks.}
+  HTML quality assessment report generated by the flowQ package. Rows
+  correspond to samples, columns to different quality checks.}
 \end{figure}
@@
 sets \cite{lo2008agf}. One of the automated gating methods implemented
-in flowCore is based on identifying areas of significant curvature in
-a kernel density estimate of the data \citep{wand2008}. Assuming that
-the regions of interest are of high density, the software is able to
-reliably detect them in a one or two-dimensional density landscape.
+in \Rpackage{flowCore} is based on identifying areas of significant
+curvature in a kernel density estimate of the data
+\citep{wand2008}. Assuming that the regions of interest are of high
+density, the software is able to reliably detect them in a one or
+two-dimensional density landscape.
 
@@
 computing, and a lot of effort has been invested in the development of
-good software to address it. The modular design of flowCore allows to
-easily integrate these existing solutions into our framework. In this
-example, we directly use R code from the feature package developed by
-\cite{wand2008}. Instead of re-writing existing code, we are able to
-include it via the well tested distribution mechanism provided by R's
-software package system. This process is bi-directional, and all
-functionality implemented in flowCore is available to other package
-authors, as we have seen with the afore-mentioned flowClust package.
+good software to address it. The modular design of \Rpackage{flowCore}
+allows to easily integrate these existing solutions into our
+framework. In this example, we directly use R code from the
+\Rpackage{feature} package developed by \cite{wand2008}. Instead of
+re-writing existing code, we are able to include it via the well
+tested distribution mechanism provided by R's software package
+system. This process is bi-directional, and all functionality
+implemented in \Rpackage{flowCore} is available to other package
+authors, as we have seen with the afore-mentioned \Rpackage{flowClust}
+package.
 
@@
 \caption{\label{xyplot}%
-Scatterplot matrix of a single flowSet with outlines of the gating
-regions identified by an automated gating operation.}
+  Scatterplot matrix of a single \Rclass{flowSet} with outlines of the
+  gating regions identified by an automated gating operation.}
 \end{figure}
@@
 
-Through flowCore, we have provided the FCM community a freely
-available, highly functional, standards compliant development and
-analysis platform for high throughput data analysis.  Our intention is
-to provide a platform for the collaborative development of new
-analysis methods that will facilitate the transition of these new
-methods to the larger flow community.  Our experience has been that the
-collaborative effort we enable for devising new methodology has been
-proven beneficial for a number of different biological and
+Through \Rpackage{flowCore}, we have provided the FCM community a
+freely available, highly functional, standards compliant development
+and analysis platform for high throughput data analysis.  Our
+intention is to provide a platform for the collaborative development
+of new analysis methods that will facilitate the transition of these
+new methods to the larger flow community.  Our experience has been
+that the collaborative effort we enable for devising new methodology
+has been proven beneficial for a number of different biological and
 computational biology challenges.  We hope that our framework will be

### 2008-08-12 paper/manuscript.tex
couple of quick edits, will read on bus tomorrow in depth. Looking great! See email for suggestions for Figures.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33194 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 FCM data analysis and their implementation in R, however, they could
-just as easily be implemented in any other language (eg Java,
+just as easily be implemented in any other language (e.g., Java,
 C++). Software written in those languages could use similar data
@@
 file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
-a software implementation of the Gating-ML CR, an emerging new flow
+a software implementation of the Gating-ML CR, an new flow
 cytometry data standard developed in collaboration with the ISAC Data
@@
 \Rpackage{flowCore} in existing work flows and to communicate with any
-other FCM tool that adheres to the proposed standard. Adherence to
+other FCM tool that adheres to the proposed standard\citep{SpidlenInPressCytometryA}. Adherence to
 standards also plays a critical role in the ability of new methods
@@
 analysis. The \Rpackage{flowViz} package \citep{sarkar2008ufv}
-provides sophisticated data visualization tools, that make use of
+provides sophisticated data visualization tools, that make use off
 multivariate trellis plotting \citep{lattice}.  These functions can be
@@
 \Rpackage{flowCore}'s data model and infrastructure which is
-implemented in the Bioconductor package \Rpackage{flowClust} Another
+implemented in the Bioconductor package \Rpackage{flowClust}. Another
 package, \Rpackage{plateCore}, providing more specialized support for

### 2008-08-12 paper/manuscript.tex
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33197 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\title{Outline flowCore paper}
+\title{flowCore: A Bioconductor software package for high throughput flow cytometry analysis}
 
@@
   Byron Ellis\\
-  Ryan Brinkman\\
+  Ryan R. Brinkman\\
   Perry Haaland\\
@@
 During the last several years, automation technologies have been
-developed that enable the use of flow cytometry high content screening
+developed that enable the use of FCM high content screening
 (FC-HCS) to generate large, complex datasets in both basic and
@@
 that typically arise from high throughput experiments such as
-microarray gene expression analysis or FC-HCS. Using flow cytometry as
+microarray gene expression analysis or FC-HCS. Using FCM as
 a high-throughput technology requires suitable software infrastructure
@@
 across samples, as it is able to take into accounts unforeseen changes
-in signal intensities. \textit{e.g.} drifts in the instrumentation
-over time or sample variablity.
+in signal intensities (\textit{e.g.,} drifts in the instrumentation
+over time or sample variablity).
 
@@
 \Rclass{flowSet} object ``dat'', the following simple line of code
-produces the quite complex output:
+produces the output shown in Figure~\ref{flowQ}:
 
@@
 The report is interactive and provides drill-down to more detailed
-aspects of the analysis, starting from a concise overview
-(Figure~\ref{flowQ}). The design of \Rpackage{flowCore}'s data model
+aspects of the analysis, starting from a concise overview. The design of \Rpackage{flowCore}'s data model
 allows for a coherent treatment of all the samples, hence we are able
@@
 
-Through \Rpackage{flowCore}, we have provided the FCM community a
+Through \Rpackage{flowCore}, we have provided the FCM community an open source,
 freely available, highly functional, standards compliant development
@@
 \section*{Acknowledgements}
-This work was supported by NIH grant EB005034. RRB is a Michael Smith
-Foundation for Health Research Scholar and a International Society for
-Analytical Cytology Scholar.
+This work was supported by NIH grant EB005034 and by the Michael Smith
+Foundation for Health Research. RRB is an ISAC Scholar.
 

### 2008-08-12 paper/manuscript.tex
tidying up the abstract a bit
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33198 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \subsubsection*{Methods}
-In this manuscript, we present \Rpackage{flowCore} a set of flexible
-and well structured computational tools for FC-HCS data analysis based
-on the open source statistical software R in conjunction with the
-Bioconductor Project.
+In this manuscript, we describe a set of flexible
+computational tools for FC-HCS data analysis written 
+in R and freely available from the Bioconductor Project.
+
 \subsubsection*{Results}
-Through \Rpackage{flowCore}, we provide the FCM community a shared
+The \Rpackage{flowCore} package provides the FCM community a shared
 research platform that enables bioinformaticians, computer scientists,
@@
 \subsubsection*{Conclusions}
-We hope that our framework will be the foundation for fruitful shared
-research by many collaborators from multiple scientific fields and
-will help resolve bottlenecks that currently prevent the further
-development of and deployment of FC-HCS to increasingly complex and
-important scientific and clinical applications.
+We hope that our framework will be the foundation for fruitful 
+collaborative research involving multiple scientific fields and
+that it will help facilitate the
+development and deployment of FC-HCS methods to complex 
+scientific and clinical applications.
 

### 2008-08-15 paper/manuscript.tex
I don't believe I'll be prepared to cite plateCore by the time the paper is submitted.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33211 bc3139a8-67e5-0310-9ffc-ced21a209358
higher resolution for the flowQ image
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33225 bc3139a8-67e5-0310-9ffc-ced21a209358
changes suggested by Ryan and Josef
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33253 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \subsubsection*{Results}
-The \Rpackage{flowCore} package provides the FCM community a shared
-research platform that enables bioinformaticians, computer scientists,
-and statisticians to work collaboratively with biologists and
-clinicians to develop novel methods for FCM data analysis.
+The \Rpackage{flowCore} package provides the flow cytometry community
+with a shared research platform that enables bioinformaticians,
+computer scientists, and statisticians to work collaboratively with
+biologists and clinicians to develop novel methods for FCM data
+analysis.
 \subsubsection*{Conclusions}
@@
 of the current situation is, however, that it is very difficult for
-new analysis approaches to find there way into standard practice. We
+new analysis approaches to find their way into standard practice. We
 believe that this barrier to the development and dissemination of new
@@
 researchers to efficiently handle FC-HCS data and encourages open
-development of tools for their coherent analysis. Among the issues
-addressed by \Rpackage{flowCore} are computationally efficient data
-structures and a range of specialized methods for compensation,
-transformation, and gating.  \Rpackage{flowCore} runs on Windows, Mac
-OS X and Linux/Unix operating systems. Conceptually, the underlying
-data structures that build the core of our software can be implemented
-in any programming language, and we chose the R language for its
-advantages in statistical programming and visualization. In this
-paper, we describe the functionality and methods underlying
-\Rpackage{flowCore}, and provide examples of their use.
+development of tools for their coherent analysis. \Rpackage{flowCore}
+implements computationally efficient data structures and a range of
+specialized methods addressing all components of a typical FCM
+analysis work flow, including compensation, transformation, and
+gating. \Rpackage{flowCore} runs on Windows, Mac OS X and Linux/Unix
+operating systems. Conceptually, the underlying data structures that
+build the core of our software can be implemented in any programming
+language. We chose the R language for its advantages in statistical
+programming and visualization. In this paper, we describe the
+functionality and methods underlying \Rpackage{flowCore}, and provide
+examples of their use.
 
@@
 research fields including genomics, proteomics and cell biology
-\citep{BIOC}. The R language and Bioconductor are
-particularly geared towards the analysis of large complex datasets
-that typically arise from high throughput experiments such as
-microarray gene expression analysis or FC-HCS. Using FCM as
-a high-throughput technology requires suitable software infrastructure
-to facilitate the rapid handling of the data. In our implementation of
-\Rpackage{flowCore} we rely on two important lessons learned from the field of
-gene expression data analysis: the first being the importance of data
-structures that reflect the underlying data and facilitate the
-manipulations that are of most interest, while the second is the
-importance of a modular architecture that allows for many developers
-to extend and use the underlying infrastructure and to combine tools
-in complex work flows.
+\citep{BIOC}. The R language and Bioconductor are particularly geared
+towards the analysis of large complex datasets that typically arise
+from high throughput experiments such as microarray gene expression
+analysis or FC-HCS. Using FCM as a high-throughput technology requires
+suitable software infrastructure to facilitate the rapid handling of
+the data. In our implementation of \Rpackage{flowCore} we rely on two
+important lessons learned from the field of gene expression data
+analysis: the first being the importance of data structures that
+reflect the underlying data and facilitate the manipulations that are
+of most interest, while the second is the importance of a modular
+architecture that allows for many developers to extend and use the
+underlying infrastructure and to combine tools in complex work flows.
 
@@
 manual data inspection in one or two dimensions, or by very basic
-comparisons of summary statistics and most of the currently available
+comparisons of summary statistics. Most of the currently available
 analysis tools are designed to reflect this workflow.  We believe that
@@
 nature. Furthermore, the subjective character of manual analyses are a
-major obstacle to reproducibility. For FC-HCS data, unassisted manual
-inspection is extremely time consuming, and robust statistical methods
-need to be developed to point investigators to interesting aspects of
-the data, or to potential problems. While the expert knowledge of
-immunologists and researchers remains crucial for the understanding of
-FCM data, we believe that collaboration with other research fields
-such as statistics and computer science can greatly improve the
-relevance of FCM in today's high-throughput paradigm.
+major obstacle to reproducibility. In a recent study of flow
+cytometric standardization involving 15 institutions, the mean
+inter-laboratory coefficient of variation ranged from 17 to 44\%, even
+though preparation was standardized and performed using the same
+samples and reagents at each site \citep{Maecker2005}. For FC-HCS
+data, unassisted manual inspection is extremely time consuming, and
+robust statistical methods need to be developed to point investigators
+to interesting aspects of the data, or to potential problems. While
+the expert knowledge of immunologists and researchers remains crucial
+for the understanding of FCM data, we believe that collaboration with
+other research fields such as statistics and computer science can
+greatly improve the relevance of FCM in today's high-throughput
+paradigm.
 
@@
 of containers like XML or binary storage formats that are designed
-specifically for particular user interfaces and hence are not easily
+specifically for particular user interfaces and hence not easily
 amenable to programmatical access. In addition, the closed-source
-nature of these products often makes it impractical to integrate them
+nature of these products often makes them impractical to integrate
 into analysis pipelines. In this manuscript we describe classes for
@@
 file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
-a software implementation of the Gating-ML CR, an new flow
-cytometry data standard developed in collaboration with the ISAC Data
-Standards Task Force, which makes it possibile to integrate
+a software implementation of the Gating-ML Candidate Recommendation
+(CR), an emerging ISAC standard developed in collaboration with the
+ISAC Data Standards Task Force, which makes it possibile to integrate
 \Rpackage{flowCore} in existing work flows and to communicate with any
-other FCM tool that adheres to the proposed standard\citep{SpidlenInPressCytometryA}. Adherence to
-standards also plays a critical role in the ability of new methods
-based on \Rpackage{flowCore} to find their way back into the standard
-practices for FCM data analysis. \Rpackage{flowCore} is not a
-GUI-driven software, and all operations are done using a command line
-interface.  It is possible to add a more elaborate user interface on
-top of this infrastructure, however the focus in this paper is on a
-programmatic approach to enable the convenient development of novel
-analysis methods and automation of complex analysis approaches.  By
-taking the burden of data management from the programmer, and by
-providing well-defined programming interfaces (APIs), it is possible
-to readily test new ideas and to easily extend the framework's
-functionality.
+other FCM tool that adheres to the proposed standard
+\citep{SpidlenInPressCytometryA}. Adherence to standards also plays a
+critical role in the ability of new methods based on
+\Rpackage{flowCore} to find their way back into the standard practices
+for FCM data analysis.
+
+\Rpackage{flowCore} is not a GUI-driven software, and all operations
+are done using a command line interface.  It is possible to add a more
+elaborate user interface on top of this infrastructure, however the
+focus in this paper is on a programmatic approach to enable the
+convenient development of novel analysis methods and automation of
+complex analysis approaches.  By taking the burden of data management
+from the programmer, and by providing well-defined programming
+interfaces (APIs), it is possible to readily test new ideas and to
+easily extend the framework's functionality.
 
@@
 instrument side and any additional keywords the user deems to be
-important to annotate the data. A number of quality checks are
-performed during the creation of a \Rclass{flowFrame} to ensure data
-integrity.
+important to annotate the data. During the creation of a
+\Rclass{flowFrame}, a number of quality checks are performed to ensure
+data integrity.
 
@@
 In high-throughput FCM, many of the analysis tasks need to be
-performed consistently across multiple flow measurements, hence we
-introduce the concept of a collection of \Rclass{flowFrames} called a
+performed consistently across multiple samples, hence we introduce the
+concept of a collection of \Rclass{flowFrames} called a
 \Rclass{flowSet}. A \Rclass{flowSet} is a container for multiple
@@
 potentially complex metadata associated with even larger FC-HCS
-experiments such as clinical trials where hundreds patients might
-provide samples at different points in time over the course of the
+experiments such as clinical trials, where hundreds of patients might
+provide samples at different time points over the course of the
 experiment.
@@
 \includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
-\caption{\label{fig1:FrameWork}{For each experiment, FCS files (as
-    individual \Rclass{flowFrames}), phenotypic and metadata are
-    stored in a \Rclass{flowSet}. Each \Rclass{flowFrame} in a
-    \Rclass{flowSet} corresponds to one FCS file. All basic operations
-    (e.g., compensation, transformation, gating) can be applied to
-    either single \Rclass{flowFrames} or a \Rclass{flowSet}
-    simultaneously.}}
+\caption{\label{fig1:FrameWork}{For each experiment, the content of
+    the FCS files, phenotypic and metadata are stored in a
+    \Rclass{flowSet}. Each \Rclass{flowFrame} in a \Rclass{flowSet}
+    corresponds to one FCS file. All basic operations (e.g.,
+    compensation, transformation, gating) can be applied to either
+    single \Rclass{flowFrames} or a \Rclass{flowSet} simultaneously.}}
 \end{figure}
@@
 
-The basic operations in FCM analyses are typically the same: the data
-need to be compensated (if that was not already done on the
-instrument) and transformed, and sup-populations of interest need to
-be selected based on a set of (predominantly sequential) gates. All
-software solutions for FCM analysis offer support for these
-operations, typically in an interactive, GUI-driven interface. The
-approach we have taken in \Rpackage{flowCore} is to abstractly
-describe these operations and build a set of tools to perform them on
-both \Rclass{flowFrames} and \Rclass{flowSets}. While transformation,
-and to a certain extent compensation, are fairly routine operations
-with only limited potential for improvement, being able to implement
-new methodologies for gating of FCM data, and extending the
-capabilities of \Rpackage{flowCore} through object oriented
-programming are features that clearly sets our framework apart from
-other FCM analysis tools. By factoring out as much of the bookkeeping
-as possible, programmers can focus on the actual operations rather
-then having to deal with the tedious details of data integration and
+Typically, the basic operations in FCM analyses adhere to the
+following common work flow: the data need to be compensated (if that
+was not already done on the instrument) and transformed, and
+sup-populations of interest need to be selected based on a set of
+(predominantly sequential) gates. All software solutions for FCM
+analysis offer support for these operations, typically in an
+interactive, GUI-driven interface. In \Rpackage{flowCore} we have
+taken the approach to abstractly describe these operations and build a
+set of tools to perform them on both \Rclass{flowFrames} and
+\Rclass{flowSets}. While transformation, and to a certain extent
+compensation, are fairly routine operations with only limited
+potential for improvement, being able to implement new methodologies
+for gating of FCM data, and extending the capabilities of
+\Rpackage{flowCore} through object oriented programming are features
+that clearly sets our framework apart from other FCM analysis
+tools. By factoring out as much of the bookkeeping as possible,
+programmers can focus on the actual operations rather then having to
+deal with the tedious details of data integration and
 access. Third-party methods can act on their own as first-class
@@
 fostered the development of a number of valuable add-ons
-\citep{lo2008agf,sarkar2008ufv} \textit{(Fixme Errol: Will plateCore
-  be in Bioconductor soon. It should be if we want to cite it.)}.
+\citep{lo2008agf,sarkar2008ufv}.
 
@@
 routinely used in FCM analysis have been implemented in
-\Rpackage{flowCore} (\textit{e.g.,} quadratic, log or arcsinh, see
-Table~\ref{table1} for a complete list). Furthermore, the design of
-the R language makes it easy to define arbitrary functions to apply to
-the data of individual \Rclass{flowFrames} or entire
+\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponantioal or arcsinh,
+see Table~\ref{table1} for a complete list). Furthermore, the design
+of the R language makes it easy to define arbitrary functions to apply
+to the data of individual \Rclass{flowFrames} or entire
 \Rclass{flowSets}, respectively. Compensation is available for both
@@
 offers functionality to compute spillover or compensation matrices
-from a set of appropriate compensation samples.
+from a set of appropriate single stain controls.
 
@@
 \caption{\label{table1}Data transformations implemented in
-  \Rpackage{flowCore}.}
+  \Rpackage{flowCore}. Within these formulas, $x$ is the variable
+  corresponding to value being transformed, $y$ is the result of the
+  transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
+  constants affecting the transformation function, $e$ is the base of
+  the natural logarithm, and $root$ is a root-finding algorithm.  }
 \end{center}
@@
 \subsubsection*{Gating}
-%%FIXME: I would make more of the filters and possibly try to make it
-%%clearer how much stuff thre is for composition of gates/filters etc
-%%
-%% somewhere notions of apply and summary methods should appear and if
-%% we have anything figured out yet on how to do the subsetting etc
-%% that too would go nicely in here (here I am thinking of our
-%% discussion of the sort of workflow stuff)
-
 In \Rpackage{flowCore}, gating operations are represented by classes
@@
 across samples, as it is able to take into accounts unforeseen changes
-in signal intensities (\textit{e.g.,} drifts in the instrumentation
-over time or sample variablity).
+in signal intensities, such as drifts in the instrumentation
+over time or sample variablity.
 
@@
 
-Gating, as well as all other operations in flowCore, can be applied
-over each individual frame in a \Rclass{flowSet}, and summary methods
-provide information about the outcome of these operations. In
+Gating, as well as all other operations in \Rclass{flowCore}, can be
+applied over each individual frame in a \Rclass{flowSet}, and summary
+methods provide information about the outcome of these operations. In
 addition, the result of a gating operation can be used to subset the
@@
 methodology and a framework to create interactive web-based reports of
-QA results.  Most of \Rpackage{flowUtil}'s content deals with data
-import and export, implementing the flow-cytometry specific standard
-markup language.  Finally, \Rpackage{flowStats} provides more
-elaborate statistical methods that are relevant in the context of flow
+QA results. The \Rpackage{flowUtil} package implements data import and
+export including flow-cytometry specific standard markup
+language. Finally, the \Rpackage{flowStats} package provides elaborate
+statistical methods that are relevant in the context of flow
 cytometry data analysis. \cite{lo2008agf} have recently developed an
@@
 \centering
-\includegraphics[width=0.75\textwidth]{flowQ.jpg}
+\includegraphics[width=0.98\textwidth]{flowQ.jpg}
 \caption{\label{flowQ}%
-  HTML quality assessment report generated by the flowQ package. Rows
-  correspond to samples, columns to different quality checks.}
+  HTML quality assessment report generated by the flowQ package for a
+  subset of data from an experiment focussing on Graft-Versus-Host
+  Disease \citep{brinkman2007hcf}. Rows correspond to the samples in
+  the set, columns to different quality checks.}
 \end{figure}
@@
 Static gating for all samples in a high-throughput FCM experiment is
-often not an option, since the measured variables tend to vary between
+often impossible, since the measured variables tend to vary between
 different treatments, over time or between different experiment
-batches. Automated or data-driven gating tries to estimate the gating
-regions from the underlying data, thus providing a fast objective
-solution to the analysis of potentially very large and diverse data
-sets \cite{lo2008agf}. One of the automated gating methods implemented
-in \Rpackage{flowCore} is based on identifying areas of significant
-curvature in a kernel density estimate of the data
-\citep{wand2008}. Assuming that the regions of interest are of high
-density, the software is able to reliably detect them in a one or
+batches. Automated or data-driven gating has the potential to estimate
+the gating regions from the underlying data, thus providing a fast
+objective solution to the analysis of potentially very large and
+diverse data sets \citep{lo2008agf}. One of the automated gating
+methods implemented in \Rpackage{flowCore} is based on identifying
+areas of significant curvature in a kernel density estimate of the
+data \citep{wand2008}. Assuming that the regions of interest are of
+high density, the software is able to reliably detect them in a one or
 two-dimensional density landscape.
@@
 \caption{\label{xyplot}%
-  Scatterplot matrix of a single \Rclass{flowSet} with outlines of the
-  gating regions identified by an automated gating operation.}
+  Scatterplot matrix of a single \Rclass{flowSet} from an experiement
+  focussing on immune tolerance following kidney
+  transplantation. Outlines of the gating regions identified by an
+  automated gating operation are added on top of the density
+  representation of the data.}
 \end{figure}
@@
 
-Through \Rpackage{flowCore}, we have provided the FCM community an open source,
-freely available, highly functional, standards compliant development
-and analysis platform for high throughput data analysis.  Our
-intention is to provide a platform for the collaborative development
-of new analysis methods that will facilitate the transition of these
-new methods to the larger flow community.  Our experience has been
-that the collaborative effort we enable for devising new methodology
-has been proven beneficial for a number of different biological and
-computational biology challenges.  We hope that our framework will be
-the foundation for fruitful shared research by many collaborators from
-multiple scientific fields and will help resolve bottlenecks that
-currently prevent the further development of and deployment of FC-HCS
-to increasingly complex and important scientific and clinical
-applications.
+Through \Rpackage{flowCore}, we have provided the FCM community an
+open source, freely available, highly functional, standards compliant
+development and analysis platform for high throughput data analysis.
+Our intention is to provide a platform for the collaborative
+development of new analysis methods that will facilitate the
+transition of these new methods to the larger flow community.  Our
+experience has been that the collaborative effort we enable for
+devising new methodology has been proven beneficial for a number of
+different biological and computational biology challenges.  We hope
+that our framework will be the foundation for fruitful shared research
+by many collaborators from multiple scientific fields and will help
+resolve bottlenecks that currently prevent further development and
+deployment of FC-HCS to increasingly complex and important scientific
+and clinical applications.
 

### 2008-08-15 paper/manuscript.tex
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33254 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-Through \Rpackage{flowCore}, we have provided the FCM community an
-open source, freely available, highly functional, standards compliant
-development and analysis platform for high throughput data analysis.
-Our intention is to provide a platform for the collaborative
-development of new analysis methods that will facilitate the
-transition of these new methods to the larger flow community.  Our
+Through \Rpackage{flowCore}, we have provided the FCM community with
+an open source, freely available, highly functional, standards
+compliant development and analysis platform for high throughput data
+analysis.  Our intention is to provide a platform for the
+collaborative development of new analysis methods that will facilitate
+the transition of these new methods to the larger flow community.  Our
 experience has been that the collaborative effort we enable for

### 2008-08-15 paper/manuscript.tex
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33255 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \subsection*{Computational tools for the analysis of FC-HCS data}
-During the last several years, automation technologies have been
-developed that enable the use of FCM high content screening
-(FC-HCS) to generate large, complex datasets in both basic and
-clinical research applications \citep{brinkman2007hcf}. A serious
-bottleneck in the interpretation of existing studies and the
-application of FC-HCS to even larger, more complex problems is that
-data management and data analysis methods have not advanced
-sufficiently far from the methods developed for applications of flow
-cytometry (FCM) to small-scale, tube-based studies.
+Automation technologies developed during the last several years have
+enabled the use of FCM high content screening (FC-HCS) to generate
+large, complex datasets in both basic and clinical research
+applications \citep{brinkman2007hcf}. A serious bottleneck in the
+interpretation of existing studies and the application of FC-HCS to
+even larger, more complex problems is that data management and data
+analysis methods have not advanced sufficiently far from the methods
+developed for applications of flow cytometry (FCM) to small-scale,
+tube-based studies.
 
-Some of the consequences of this lag in development are difficulties
-in maintaining the integrity and documentation of extremely large
-datasets, assessing measurement quality, developing validated assays,
-controlling the accuracy of gating techniques, automating complex
-gating strategies, and aggregating statistical results across large
-study sets for further analysis.  Among the most serious consequences
-of the current situation is, however, that it is very difficult for
-new analysis approaches to find their way into standard practice. We
-believe that this barrier to the development and dissemination of new
-analysis methods is one of the fundamental restraints on the future
-expansion of FC-HCS in both clinical and research applications.
+Some of the consequences of this lag are difficulties in maintaining
+the integrity and documentation of extremely large datasets, assessing
+measurement quality, developing validated assays, controlling the
+accuracy of gating techniques, automating complex gating strategies,
+and aggregating statistical results across large study sets for
+further analysis. The most serious consequence, however, is that new
+analysis approaches face difficulty in finding their way into standard
+practice. We believe that this barrier to the development and
+dissemination of new analysis methods is one of the fundamental
+restraints on the future expansion of FC-HCS in both clinical and
+research applications.
 
 In this paper, we describe a set of flexible and well structured
-computational tools to efficiently analyze FC-HCS. This set of tools
-is based on the open source statistical software R \citep{Rmain}, in
-conjunction with the Bioconductor Project \citep{BIOC}. Our intent is
-to provide a shared research platform that enables bioinformaticians,
-computer scientists, and statisticians to work collaboratively with
-biologists and clinicians to develop novel methods for FCM data
-analysis, a process deemed crucial by many for the further development
-of the technology \citep{lizard2007fca}. A critical component of this
-approach is the development and implementation of standards that will
-facilitate the adoption of these methods by both the larger research
-community and commercial software vendors.  Consequently, we are
-optimistic that this software platform will play a critical role in
-addressing the bottlenecks described above that limit further advances
-in the application of FC-HCS.
+computational tools to efficiently analyze FC-HCS data. This set of
+tools is based on the open source statistical software R
+\citep{Rmain}, in conjunction with the Bioconductor Project
+\citep{BIOC}. Our intent is to provide a shared research platform that
+enables bioinformaticians, computer scientists, and statisticians to
+work collaboratively with biologists and clinicians to develop novel
+methods for FCM data analysis, a process deemed crucial by many for
+the further development of the technology \citep{lizard2007fca}. A
+critical component of this approach is the development and
+implementation of standards that will facilitate the adoption of these
+methods by both the larger research community and commercial software
+vendors.  Consequently, we are optimistic that this software platform
+will play a critical role in addressing the bottlenecks described
+above that limit advances in the application of FC-HCS.
 
@@
 \Rpackage{flowCore}. \Rpackage{flowCore} is a freely available, highly
-functional and extensible FCM data analysis platform that enables
+functional, and extensible FCM data analysis platform that enables
 researchers to efficiently handle FC-HCS data and encourages open
@@
 analysis work flow, including compensation, transformation, and
-gating. \Rpackage{flowCore} runs on Windows, Mac OS X and Linux/Unix
+gating. \Rpackage{flowCore} runs on Windows, Mac OS X, and Linux/Unix
 operating systems. Conceptually, the underlying data structures that

### 2008-08-21 paper/manuscript.tex
title page requirement for cytometry
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33315 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \documentclass[12pt]{article}
+\usepackage{setspace}
 \usepackage{url}
@@
 
-\title{flowCore: A Bioconductor software package for high throughput flow cytometry analysis}
+\title{flowCore: A Bioconductor software package for high throughput
+  flow cytometry data analysis \footnote{This work was supported by
+    NIH grant EB005034 and by the Michael Smith Foundation for Health
+    Research. RRB is an ISAC Scholar.}}
 
-\author{Florian Hahne*\\
-  Nolwenn LeMeur*\\
-  Byron Ellis\\
-  Ryan R. Brinkman\\
-  Perry Haaland\\
-  Errol Strain\\
-  Deepayan Sarkar\\
-  Josef Spidlen\\
-  Robert Gentleman
+\author{Florian Hahne (\textit{Fred Hutchinson Cancer Research
+    Center})\\
+  Nolwenn LeMeur* (\textit{Fred Hutchinson Cancer Research Center})\\
+  Ryan R. Brinkman (\textit{BC Cancer Research Centre})\\
+  Byron Ellis (\textit{AdBrite, Inc})\\
+  Perry Haaland (\textit{BD Biosciences})\\
+  Deepayan Sarkar (\textit{Fred Hutchinson Cancer Research
+    Center})\\
+  Josef Spidlen (\textit{BC Cancer Research Centre})\\
+  Errol Strain (\textit{BD Biosciences})\\
+  Robert Gentleman (\textit{Fred Hutchinson Cancer Research
+    Center})\\
 }
 
+\doublespacing
 \begin{document}
+\begin{singlespace}
+\maketitle
 
-\maketitle
+Corresponding author:\\
+\textit{
+  Florian Hahne, PhD\\
+  Computational Biology Program\\
+  Division of Public Health Sciences\\
+  Fred Hutchinson Cancer Research Center\\
+  1100 Fairview Ave. N, M2-B876\\
+  PO Box 19024\\
+  Seattle, Washington 98109-1024\\
+  phone: 206-667-3148\\
+  fax: 206-667-1319\\
+  fhahne@fhcrc.org
+}
+
+\end{singlespace}
 
@@
 %%Figure1 Diagram of the data structures and the basic operations
+\begin{singlespace}
 \begin{figure}
@@
 \end{figure}
+\end{singlespace}
 
@@
 
+\begin{singlespace}
 \begin{figure}[htbp]

### 2008-09-03 R/filterSummary-accessors.R
updated filterSummary class and available methods
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33537 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1,22 +1,70 @@
-setMethod("[[",signature("filterSummary","numeric"),function(x,i,j,...) {
-	list(name=x@name[i],true=x@true[i],false=x@count[i]-x@true[i],count=x@count[i],p=x@p[i],q=1-x@p[i])
-})
-setMethod("[[",signature("filterSummary","character"),function(x,i,j,...) {
-	i = which(i,names(x))
-	x[[i]]
-})
-setMethod("length",signature("filterSummary"),function(x) length(x@name))
-setMethod("names",signature("filterSummary"),function(x) x@name)
+## The filterSummary class provides a container for the output of summary
+## on a filterResult
 
-setMethod("$",signature("filterSummary","ANY"),function(x,name) {
-	switch(name,"n"=x@count,"true"=x@true,"false"=x@count-x@true,"p"=x@p,"q"=1-x@p)
-})
-setMethod("show",signature("filterSummary"),function(object) {
-    if(length(object@name) == 1) {
-		cat(sprintf("%s: %d of %d (%.2f%%)\n",object@name,object@true,object@count,object@p*100))
-	} else {
-		for(i in seq(along=x@name)) {
-			cat(sprintf("%s: %d of %d (%.2f%%)\n",object@name[i],object@true[i],object@count[i],object@p[i]*100))
-		}
-	}
-})
+## We need this to subset filterSummaries of multipleFilterResults
+setMethod("[[",
+          signature=signature("filterSummary","numeric"),
+          definition=function(x,i,j,...)
+      {
+          list(name=as.vector(x@name[i]), true=as.vector(x@true[i]),
+               false=as.vector(x@count[i]-x@true[i]),
+               count=x@count[i], p=as.vector(x@p[i]),
+               q=as.vector(1-x@p[i]))
+      })
+
+setMethod("[[",
+          signature=signature("filterSummary","character"),
+          definition=function(x,i,j,...)
+      {
+	i <- which(i==names(x))
+	x[[i]]
+    })
+
+
+## Essentially the number of populations (in a multipleFilterResult)
+setMethod("length",
+          signature=signature("filterSummary"),
+          definition=function(x) length(x@name))
+
+
+## The names of the populations (in a multipleFilterResult)
+setMethod("names",
+          signature=signature("filterSummary"),
+          definition=function(x) x@name)
+
+
+## This allows for a list-like accessor to the slots (and more...)
[…]

### 2008-09-03 R/filter-functions.R
updated filterSummary class and available methods
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33537 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1,28 +1,0 @@
-## ==========================================================================
-## function to print filter summary
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-print.filterSummary <- function(x,...) {
-    
-    if(length(x$name) == 1) {
-		with(x,cat(sprintf("%s: %d of %d (%.2f%%)\n",name,true,n,100*p)))
-	} else {
-	for(i in seq(along=x$name))
-			with(x,cat(sprintf("%s: %d of %d (%.2f%%)\n",name[i],true[i],n[i],100*p[i])))
-	}
-}
-
-
-
-# quadGate = function(.split,...,ids=names(.split)) {
-# 	if(missing(.split)) .split = list(...)
-# 	n = names(.split)
-# 	makemat = function(i,pos)
-# 		matrix(if(pos) c(.split[i],Inf) else c(-Inf,.split[i]),2,dimnames=list(c("min","max"),n[i]))
-# 	
-# 	gates = list(rectangleGate(makemat(1,FALSE),filterId=paste(ids[1],"-",sep="")),rectangleGate(makemat(1,TRUE),filterId=paste(ids[1],"+",sep="")))
-# 	for(i in 2:length(.split)) {
-# 		cur = list(rectangleGate(makemat(i,FALSE),filterId=paste(ids[i],"-",sep="")),rectangleGate(makemat(i,TRUE),filterId=paste(ids[i],"+",sep="")))
-# 		gates = outer(gates,cur)
-# 	}
-# 	gates
-# }

### 2008-09-18 R/actionItem-accessors.R
A first draft of the new workflow infrastructure and all the little changes that were necessary to make that work. This might eventually move to some other place. Also updated the documentation a bit. More to come soon...
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33791 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1,0 +1,88 @@
+## ==========================================================================
+## Accessor to parentView slot. This returns the view object after
+## resolving the reference
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("parent",
+          signature=signature("actionItem"),
+          definition=function(object) get(object@parentView))
+
+
+
+## ==========================================================================
+## Accessor to name slot.
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("names",
+          signature=signature("actionItem"),
+          definition=function(x) x@name)
+
+
+
+## ==========================================================================
+## Accessor to gate slot. This returns the filter object after resolving
+## the reference
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("gate",
+          signature=signature("gateActionItem"),
+          definition=function(object) get(object@gate))
+
+
+
+## ==========================================================================
+## Remove an actionItem object from a workFlow. This will traverse down the tree
+## and also remove all dependent objects to free their memory.
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+## For all actionItems we need to remove the associated views.
+setMethod("Rm",
+          signature=signature(symbol="actionItem",
+                              envir="workFlow",
+                              subSymbol="character"),
+          definition=function(symbol, envir, subSymbol)
+      {
+          curAction <- identifier(symbol)
+          aiMatch <- sapply(edgeData(get(envir@tree), attr="actionItem"), identifier)
+          depNodes <- gsub(".*\\|", "", names(aiMatch[aiMatch == curAction]))
+          lapply(mget(depNodes, envir), Rm, envir)
+          return(invisible(NULL))
+      })
+
+## For gateActionItems we need to remove the gate and the filterResult
+setMethod("Rm",
+          signature=signature(symbol="gateActionItem",
+                              envir="workFlow",
+                              subSymbol="character"),
+          definition=function(symbol, envir, subSymbol)
+      {
+          Rm(symbol@gate)
+          selectMethod("Rm", signature("actionItem",
+                                       "workFlow", "character"))(symbol, envir)
+          rm(list=identifier(symbol), envir=envir@env)
+          return(invisible(NULL))         
+      })
+
+## For transformActionItems we need to remove the transformation object
+setMethod("Rm",
+          signature=signature(symbol="transformActionItem",
[…]

### 2008-09-18 R/coerce.R

@@ -1,32 +1,56 @@
 ## ==========================================================================
+## We make heavy use of coercion methods to unify the resolts that
+## produced by the many different filter and transformation classes.
+## ==========================================================================
+
+
+
+
+
+
+## ==========================================================================
 ## We can convert a factor, logical or a numeric into a filterResult by
 ## selecting a specific filterResult type. This is done through the standard
-## R coercion techniques.
+## R coercion techniques. We use these methods to create the appropriate
+## filterResults from the results of the various %in% methods.
 ## --------------------------------------------------------------------------
-setAs("factor", "filterResult",function(from)
-      new("multipleFilterResult",filterId="",subSet=from))
+setAs(from="factor", to="filterResult", def=function(from)
+      new("multipleFilterResult", filterId="", subSet=from))
 
-setAs("logical","filterResult",function(from)
-      new("logicalFilterResult",filterId="",subSet=from))
+setAs(from="logical", to="filterResult", def=function(from)
+      new("logicalFilterResult", filterId="", subSet=from))
 
-setAs("numeric","filterResult",function(from)
-      new("randomFilterResult",filterId="",subSet=from))
+setAs(from="numeric", to="filterResult", def=function(from)
+      new("randomFilterResult", filterId="", subSet=from))
+
+setAs(from="gateView", to="filterResult", def=function(from)
+  {
+      fres <- as(from@indices, "filterResult")
+      ofres <- get(from@filterResult)
+      filter <- filterDetails(get(from@filterResult), 1)[["filter"]]
+      identifier(filter) <- from@frEntry
+      filterDetails(fres, from@frEntry) <- list(filter=filter)
+      identifier(fres) <- from@frEntry
+      return(fres)
+  })
 
 
 
 ## ==========================================================================
 ## We can also convert some filterResult types directly to logical types,
 ## though in general it is not possible. We provide the means for logical 
-## and random filter types. 
+## and random filter types. For the rest we cast useful error messages.
 ## --------------------------------------------------------------------------
-setAs("filterResult","logical",function(from)
[…]

### 2008-09-18 R/complementFilter-accessors.R

@@ -1,17 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'complementFilter'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-
-
-## ==========================================================================
-## show method
-## ---------------------------------------------------------------------------
-setMethod("show",signature("complementFilter"),
-          function(object)
-      {
-          cat("filter '", identifier(object),
-              "', the complement of\n", sep="")
-          print(object@filters[[1]])
-    
-      })

### 2008-09-18 R/curv2Filter-accessors.R

@@ -1,33 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'curv2Filter'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-## FIXME: What do summarizeFilter methods do and why are they not
-## exported in the name space?
-setMethod("summarizeFilter",signature("filterResult","curv2Filter"),
-          function(result,filter) {
-	ret = callNextMethod()
-	ret$polygons = attr(result@subSet, "polygon")
-	ret
-})
-
-
-
-
-## ==========================================================================
-## show method
-## ---------------------------------------------------------------------------
-setMethod("show",signature("curv2Filter"),function(object) {
-	msg = paste("2D curvature filter '",object@filterId,"' in dimensions ",
-        paste(object@parameters, collapse=" and "), "\nwith settings:",
-        "\n  bwFac=", object@bwFac, "\n  gridsize=",
-        paste(object@gridsize, collapse=",", sep=""), sep="")
-	cat(msg)
-	cat("\n")
-	invisible(msg)
-})
-
-
-
-
-

### 2008-09-18 R/ellipsoidGate-accessors.R

@@ -1,31 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'ellipsoidGate'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-
-
-## ==========================================================================
-## show method
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("show",
-          signature(object="ellipsoidGate"),
-          function(object)
-      {
-          cat("Ellipsoid gate '", identifier(object),
-              "' in dimensions ", sep="")
-          cat(paste(object@parameters, sep="", collapse=" and "))
-          cat("\n")
-      })
-
-
-
-plotEll <- function(cov, mean, ...)
-{
-    chol.cov <- t(chol(cov))
-    radius <- 1
-    t <- seq(0, 2 * base::pi, length = 50)
-    ans <- mean +
-        (chol.cov %*% rbind(x = radius * cos(t),
-                            y = radius * sin(t)))
-    lines(t(ans), ...)
-}

### 2008-09-18 R/expressionFilter-accessors.R

@@ -1,23 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'expressionFilter'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-
-
-          
-## ==========================================================================
-## show method
-## ---------------------------------------------------------------------------
-setMethod("show",
-          signature("expressionFilter"),
-          function(object)
-      {
-	msg <- paste("expression filter '", identifier(object),
-                     "' evaluating the expression:\n",
-                     paste(object@deparse, collapse="\n"), sep="")
-	cat(msg)
-	cat("\n")
-	invisible(msg)
-})
-
-

### 2008-09-18 R/filter-accessors.R

@@ -1,182 +1,0 @@
-## ==========================================================================
-## This implementation of filter does not require a per-filter implementation
-## anything filter specific has been moved to either filterDetails or the %in%
-## implementation. A filter operation is now about assembling those two calls 
-## AND associating a filterResult with a particular flowFrame if possible.
-## ==========================================================================
-
-
-
-## ==========================================================================
-## accessor method for slot parameters
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("parameters",
-          signature=signature("filter"),
-          definition=function(object) character(0))
-
-setMethod("parameters",
-          signature=signature("parameterFilter"),
-          definition=function(object) object@parameters)
-
-setMethod("parameters",
-          signature=signature("setOperationFilter"),
-          definition=function(object)
-          unique(unlist(lapply(object@filters, parameters))))
-
-
-
-## ==========================================================================
-## summary method for filters
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("summary",
-          signature=signature("filter"),
-          definition=function(object, result, ...)
-      {
-          if(missing(result))
-              stop("Only resolved filters may be summarized.\nTry something ",
-                   "like 'filter(myFlowFrame, myFilter)'\nbefore calling ",
-                   "'summary'.", call.=FALSE)
-          else
-              l <- as(result, "logical")
-          true <- sum(l)
-          count <- length(l)
-          new("filterSummary", name=identifier(object), true=true,
-              count=count, p=true/count)
-      })
-
-
-
-## ==========================================================================
-## Printing out a filter should give us something at least mildly sensible.
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("show",
-          signature=signature("filter"),
-          definition=function(object) 
-          cat(paste("A filter named '", object@filterId, "'\n", sep="")))
-
-
-## Most filters define only a single population
-## setMethod("length","filter",function(x) 1)
-
-
-
-## ==========================================================================
-## We need to be able to get an ID for a filter
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("identifier",
-          signature="filter",
-          definition=function(object) object@filterId)
-
-setReplaceMethod("identifier",
-                 signature=signature("filter","character"),
-                 definition=function(object,value)
[…]

### 2008-09-18 R/filterSummary-accessors.R

@@ -1,43 +1,66 @@
+## ==========================================================================
 ## The filterSummary class provides a container for the output of summary
 ## on a filterResult
+## ==========================================================================
 
+
+
+
+
+
+## ==========================================================================
+## Subsetting methods
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 ## We need this to subset filterSummaries of multipleFilterResults
 setMethod("[[",
-          signature=signature("filterSummary","numeric"),
-          definition=function(x,i,j,...)
+          signature=signature(x="filterSummary",
+                              i="numeric"),
+          definition=function(x, i, j, ...)
       {
           list(name=as.vector(x@name[i]), true=as.vector(x@true[i]),
                false=as.vector(x@count[i]-x@true[i]),
-               count=x@count[i], p=as.vector(x@p[i]),
+               count=x@count, p=as.vector(x@p[i]),
                q=as.vector(1-x@p[i]))
       })
 
+## By name
 setMethod("[[",
-          signature=signature("filterSummary","character"),
+          signature=signature(x="filterSummary",
+                              i="character"),
           definition=function(x,i,j,...)
       {
-	i <- which(i==names(x))
-	x[[i]]
-    })
+          i <- which(i==names(x))
+          x[[i]]
+      })
 
 
+
+## ==========================================================================
 ## Essentially the number of populations (in a multipleFilterResult)
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("length",
-          signature=signature("filterSummary"),
+          signature=signature(x="filterSummary"),
           definition=function(x) length(x@name))
 
 
+
+## ==========================================================================
 ## The names of the populations (in a multipleFilterResult)
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("names",
           signature=signature("filterSummary"),
           definition=function(x) x@name)
 
 
+
+## ==========================================================================
 ## This allows for a list-like accessor to the slots (and more...)
 ## Valid values are 'n', 'count', 'true', 'false', 'name', 'p' and 'q'
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("$",
-          signature=signature("filterSummary","ANY"),
-          definition=function(x,name)
+          signature=signature(x="filterSummary",
+                              name="ANY"),
[…]

### 2008-09-18 R/intersectFilter-accessors.R

@@ -1,23 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'intersectFilter'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-
-
-## ==========================================================================
-## show method
-## ---------------------------------------------------------------------------
-setMethod("show",signature("intersectFilter"),
-          function(object)
-      {
-          cat("filter '", identifier(object),
-              "'\nthe intersection between the ", length(object@filters),
-              " filters\n\n", sep="")
-          for(i in 1:length(object@filters)){
-              print(object@filters[[i]])
-              cat("\n")
-          }
-      })
-
-
-

### 2008-09-18 R/kmeansFilter-accessors.R

@@ -2,16 +2,7 @@
 ## Methods for objects of type 'kmeansFilter'
 ## Note: All filtering methods are stored in file 'in-methods.R'
 ## ==========================================================================
-## Method to gather information about the filtering operation which is
-## supposed to be stored in the filterDetails slot. 
-setMethod("summarizeFilter",
-          signature("filterResult","kmeansFilter"),
-          function(result,filter)
-      {
-          ret <- callNextMethod()
-          ret$populations <- filter@populations
-          ret
-      })
+
 
 
 ## ==========================================================================
@@ -21,20 +12,3 @@ setMethod("length",
           signature("kmeansFilter"),
           function(x) length(x@populations))
 
-
-## ==========================================================================
-## show method
-## ---------------------------------------------------------------------------
-setMethod("show",
-          signature("kmeansFilter"),
-          function(object)
-      {
-          msg <- paste("k-means filter '", object@filterId,
-                       "' in dimension ", object@parameters[1],
-                       "\nwith ", length(object), " populations (",
-                       paste(object@populations, collapse=","),
-                       ")", sep="")
-          cat(msg)
-          cat("\n")
-          invisible(msg)
-      })

### 2008-09-18 R/norm2Filter-accessors.R

@@ -1,38 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'norm2Filter'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-
-
-## ==========================================================================
-##  summarize the results of a norm2Filter operation
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-## FIXME: What do summarizeFilter methods do and why are they not
-## exported in the name space?
-setMethod("summarizeFilter",signature("filterResult","norm2Filter"),
-          function(result,filter) {
-              ret = callNextMethod()
-	ret$cov    = attr(result@subSet,'cov')
-	ret$center = attr(result@subSet,'center')
-	ret$radius = attr(result@subSet,'radius')
-	ret
-})
-
-
-## ==========================================================================
-## show method
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("show",
-          signature(object="norm2Filter"),
-          function(object)
-      {
-          cat(ifelse(length(object@transformation), "transformed", ""),
-              "norm2Filter '", identifier(object),
-              "' in dimensions ", sep="")
-          cat(paste(object@parameters, sep="", collapse=" and "),
-              "with parameters:\n")
-          cat("  method:", object@method, "\n")
-          cat("  scale.factor:", object@scale.factor, "\n")
-          cat("  n:", object@n, "\n")
-          cat("\n")
-      })

### 2008-09-18 R/rectangleGate-accessors.R

@@ -4,25 +4,8 @@
 ## ==========================================================================
 
 
-## ==========================================================================
-## show method
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("show",signature(object="rectangleGate"),function(object)
-      {
-          parms <- parameters(object)
-          cat("Rectangular gate '", identifier(object),
-              "' with dimensions:\n", sep="")
-          for(i seq_along(parms)) {
-              cat("  ")
-              if(is.character(parms[i]))
-                  cat(parms[i])
-              else
-                  cat("anonymous parameter")
-              cat(": (")
-              cat(paste(object@min[],object@max[i],sep=","))
-		cat(")\n")
-	}
-})
+
+
 
 
 ## ==========================================================================

### 2008-09-18 R/AllClasses.R
A first draft of the new workflow infrastructure and all the little changes that were necessary to make that work. This might eventually move to some other place. Also updated the documentation a bit. More to come soon...
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33791 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1,9 +1,29 @@
 ## =========================================================================##
 ## =========================================================================##
-##              Class definitions and contructors if available              ##
+##                    Class definitions and contructors                     ##
 ## =========================================================================##
 ## =========================================================================##
 
+
+
+
+
+
+## ===========================================================================
+##  Some helpers
+## ---------------------------------------------------------------------------
+## Check for the class of object x and its length and cast error if wrong
+checkClass <- function(x, class, length=NULL)
+{
+    msg <- paste("'", substitute(x), "' must be object of class '",
+                 class, "'", sep="")
+    fail <- !is(x, class)
+    if(!is.null(length) && length(x) != length){
+        fail <- TRUE
+        msg <- paste(msg, "of length", length)
+    }
+    if(fail) stop(msg, call.=FALSE) else invisible(NULL)     
+}
 
 
 
@@ -13,24 +33,26 @@
 ## A container for flow cytometry measurements with slots exprs, parameters
 ## and description. exprs contains measurement values, description contains 
 ## information from file headers of FCS file and parameters contains
-## information about the FCS measurement parameters (i.e. channels) available
+## information about the FCS measurement parameters (i.e. channels) available.
 ## ---------------------------------------------------------------------------
 setClass("flowFrame",                
-         representation(exprs="matrix",
-                        parameters="AnnotatedDataFrame",
-                        description="list"),
-         prototype=list(exprs=matrix(numeric(0), nrow=0, ncol=0),
-         parameters=new("AnnotatedDataFrame"),
-         description=list(note="empty")))
-
+         representation=representation(exprs="matrix",
+                                       parameters="AnnotatedDataFrame",
+                                       description="list"),
+         prototype=list(exprs=matrix(numeric(0),
+                                     nrow=0,
+                                     ncol=0),
+                        parameters=new("AnnotatedDataFrame"),
+                        description=list(note="empty")))
 
 ## helper function to create empty AnnotatedDataFrame for the parameters slot
-parDefault <- function(exp){
+parDefault <- function(exp)
+{
     vm <- data.frame(labelDescription=c(name="Name of Parameter",
-                     desc="Description of Parameter",
-                     range="Range of Parameter",
[…]

### 2008-09-18 R/fcReference-accessors.R

@@ -1,0 +1,117 @@
+## ==========================================================================
+## fcReferences provide a means to reference into workFlow environments.
+## We provide methods to get and assign object from references, as well
+## as methods to remove them along with the referenced objects
+## ==========================================================================
+
+
+
+
+
+
+## ==========================================================================
+## Remove an object referenced to by an fcReference object as well as the
+## reference object itself (if possible)
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+## Just the object and the reference
+setMethod("Rm",
+          signature=signature(symbol="fcReference",
+                              envir="missing",
+                              subSymbol="character"),
+          definition=function(symbol, subSymbol, rmRef=TRUE)
+      {
+          rm(list=identifier(symbol), envir=symbol@env)
+          ## This doesn't work because the name space is messing up the
+          ## evaluation environment tree.
+          ## rm(list=subSymbol, inherits=TRUE)
+          ## Instead we are rather bold and remove the reference from the
+          ## global env if it exists.
+          if(subSymbol %in% ls(globalenv(), all.names=TRUE) &&
+             is(get(subSymbol), globalenv()))
+              rm(list=subSymbol, envir=globalenv())
+      })
+
+## Object and reference and all dependent objects (further down in the
+## workFlow tree). This is only recursive for actionItems and views. All
+## other objects can be removed without further side effects, but we
+## strongly discourage from doing so since the consequences might be
+## catastrophic...
+setMethod("Rm",
+          signature=signature(symbol="fcReference",
+                              envir="workFlow",
+                              subSymbol="character"),
+          definition=function(symbol, envir, subSymbol, rmRef=TRUE)
+      {
+          ##callGeneric(symbol)
+          selectMethod("Rm", signature("fcReference", "missing", "character"))(symbol,
+                                                                               identifier(symbol))
+      })
+
+
+## For nullReferences we don't need to do anything
+setMethod("Rm",
+          signature=signature(symbol="fcNullReference",
+                              envir="missing",
+                              subSymbol="character"),
+          definition=function(symbol, envir, subSymbol, ...){})
+
+
+
+
+
+## ==========================================================================
+## Test for a NULL reference
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("isNull",
[…]

### 2008-09-18 R/filterResult-accessors.R

@@ -1,12 +1,26 @@
 ## ==========================================================================
-## Add or Replacement method to complete or replace the results
-## coming out of a filtering operation
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+## filterResults are the output of a filtering operation. 
+## ==========================================================================
 
-## simple accessor for the whole slot
+
+
+
+
+
+## ==========================================================================
+## Add or replacement method to complete or replace the results coming out
+## of a filtering operation. The filter replacement method (along with it
+## the summarizeFilter method) gets called during each filtering operation,
+## making sure that the filter item in the filterDetails slot contains all
+## necessary information
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+## Simple accessor for the whole slot. All filterResult classes except for
+## manyFilterResults only contain a single filterDetails item, and those are
+## returned directly.
 setMethod("filterDetails",
-          signature("filterResult", "missing"),
-          function(result, filterId)
+          signature=signature(result="filterResult",
+                              filterId="missing"),
+          definition=function(result, filterId)
       {
           res <- result@filterDetails
           if(length(res)==1)
@@ -14,94 +28,58 @@ setMethod("filterDetails",
           return(res)
       })
 
-## access only a single filter of the filterDetails list
+## Access only a single filter of the filterDetails list either by name
+## or by index. This is only useful for manyFilterResults.
 setMethod("filterDetails",
-          signature("filterResult", "ANY"),
-          function(result,filterId)
+          signature=signature(result="filterResult",
+                              filterId="ANY"),
+          definition=function(result, filterId)
       {
           result@filterDetails[[filterId]]
       })
 
-## replace a single filter in the list
+## Replace a single filterDetails item in the list. 
 setReplaceMethod("filterDetails",
-                 signature("filterResult", "character", value="ANY"),
-                 function(result,filterId,...,value)
+                 signature=signature(result="filterResult",
+                                     filterId="character",
+                 value="ANY"),
+                 definition=function(result, filterId, ... ,value)
              {
-                 result@filterDetails[[filterId]] = value
+                 result@filterDetails[[filterId]] <- value
                  result
              })
 
-## replace the filter, this actually calls summarizeFilter
+## Replace the filter item in a single filterDetails list. This is actually
+## only a wrapper around summarizeFilter which has to be defined for each
[…]

### 2008-09-18 R/flowSet-accessors.R

@@ -1,4 +1,13 @@
 ## ==========================================================================
+## flowSets are basically lists flowFrames
+## ==========================================================================
+
+
+
+
+
+
+## ==========================================================================
 ## subsetting methods
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
 ## to flowSet
@@ -205,19 +214,6 @@ setMethod("length","flowSet",function(x) nrow(pData(phenoData(x))))
 
 
 
-## ==========================================================================
-## show method for flowSet
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("show","flowSet",function(object) {
-	cat("A flowSet with",length(object),"experiments.\n\n")
-        if(any(varMetadata(phenoData(object))$labelDescription != "Name")){
-            show(phenoData(object))
-            cat("\n")
-        }
-	cat("  column names:\n  ")
-	cat(paste(object@colnames,sep=","))
-	cat("\n")
-})
 
 
 
@@ -252,7 +248,7 @@ setMethod("fsApply",signature("flowSet","ANY"),function(x,FUN,...,simplify=TRUE,
 ## ===========================================================================
 ## compensate method
 ## ---------------------------------------------------------------------------
-setMethod("compensate",signature("flowSet","matrix"),
+setMethod("compensate",signature("flowSet","ANY"),
           function(x,spillover,inv=TRUE, ...)
           fsApply(x,compensate,spillover,inv=inv, ...))
 
@@ -282,17 +278,11 @@ setMethod("transform",signature(`_data`="missing"),function(...) {
 
 
 ## ==========================================================================
-## General %on% implementation for a flowSet. Basically a wrapper around
-## fsApply
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("%on%",
-          signature(e2="flowSet"),
-          function(e1, e2) fsApply(e2, "%on%", e1=e1))
-
-
-
-## ==========================================================================
 ## filter methods
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+## These methods apply single filters, filterSets or lists of filters to
+## flowSet object. In all cases, the output of the filtering operation is
+## a filterResultList
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 ## for filters
 setMethod("filter",signature=signature(x="flowSet", filter="filter"),
[…]

### 2008-09-18 R/in-methods.R

@@ -1,4 +1,35 @@
 ## ==========================================================================
+## %in% methods are the workhorses to evaluate a filter. Each filter class
+## needs to define its own %in% method. Two types of return values are
+## allowed: logical vectors for logicalFilterResults and factors for
+## multipleFilterResults. In both cases, the length of the return vector
+## has to be equal to the number of events (i.e., rows) in the flowFrame.
+## In order to provide useful names, factor levels should be chosen
+## approriately.
+## ==========================================================================
+
+
+
+
+
+
+## ==========================================================================
+## Figure out useful population names for logical filters from the parameter
+## names. First choice are channel descriptors, then channel names.
+## ---------------------------------------------------------------------------
+popNames <- function(x, table)
+{
+    mt <- match(table@parameters, parameters(x, names=TRUE))
+    desc <- pData(parameters(x))[mt, "desc"]
+    names(desc) <- table@parameters
+    noName <- which(is.na(desc) | desc=="")
+    desc[noName] <- parameters(x)$name[mt][noName]
+    return(desc)
+}
+
+
+
+## ==========================================================================
 ## quadGate -- this is not a logical filter so we return a vector
 ## of factors indicating a population. Factor levels are later used
 ## as population names, e.g. when splitting. The evaluation of the
@@ -9,13 +40,7 @@ setMethod("%in%",
           definition=function(x,table)
       {
           e <-  exprs(x)[,table@parameters, drop=FALSE]
-          ## create useful names for the populations:
-          ## first choice are channel descriptors, then
-          ## channel names.
-          mt <- match(table@parameters, parameters(x)$name)
-          desc <- pData(parameters(x))[mt, "desc"]
-          noName <- which(is.na(desc) | desc=="")
-          desc[noName] <- parameters(x)$name[mt][noName]
+          desc <- popNames(x,table)
           lev <- c(sprintf("%s+%s+", desc[1], desc[2]),
                    sprintf("%s-%s+", desc[1], desc[2]),
                    sprintf("%s+%s-", desc[1], desc[2]),
@@ -35,7 +60,7 @@ setMethod("%in%",
 ## ---------------------------------------------------------------------------
 setMethod("%in%",
           signature=signature(x="flowFrame", table="rectangleGate"),
-          definition=function(x,table)
[…]

### 2008-09-18 R/multipleFilterResult-accessors.R

@@ -1,15 +1,5 @@
 
 
-## ==========================================================================
-## summary method
-## ---------------------------------------------------------------------------
-setMethod("summary","multipleFilterResult",function(object, ...) {
-	true <- summary(object@subSet[!is.na(object@subSet)])
-	count <-  rep(sum(!is.na(object@subSet)),nlevels(object@subSet))
-        name=levels(object@subSet)
-        new("filterSummary", name=levels(object@subSet),true=true,
-            count=count,p=true/count)
-})
 
 
 ## ==========================================================================
@@ -70,11 +60,3 @@ setMethod("[",
       })
 
 
-
-setMethod("show",
-          signature("multipleFilterResult"),
-          function(object) 
-          cat(paste("A filterResult produced by the filter named '",
-                    object@filterId, "'\n resulting in multiple ",
-                    "populations:\n", paste("\t", names(object), collapse="\n"),
-                    "\n", sep="")))

### 2008-09-18 R/polygonGate-accessors.R

@@ -1,20 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'polygoneGate'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-
-
-## ==========================================================================
-## show method
-## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-setMethod("show",
-          signature(object="polygonGate"),
-          function(object)
-      {
-          nb <-  nrow(object@boundaries)
-          cat("Polygonal gate '", identifier(object) ,"' with ",
-              ifelse(all(is.na(object@boundaries)), 0, nb),
-              " vertices in dimensions ", sep="")
-          cat(paste(object@parameters, sep="", collapse=" and "))
-          cat("\n")
-      })

### 2008-09-18 R/split-methods.R

@@ -39,8 +39,8 @@
 ## We actually split on filterResults and multipleFilterResults, so filters
 ## have to be evaluated first. Note that the 'drop' argument is silently
 ## ignored when splitting by filter since it doesn't have a clear meaning in
-## this application. It does have the expected meaning when splitting by
-## factors.
+## this application. It does have the expected behaviour when splitting by
+## factors, though.
 
 ## Evaluate the filter first and split on the filterResult
 setMethod("split",
@@ -173,14 +173,10 @@ setMethod("split", signature("flowFrame","manyFilterResult"),
                            names=if(is.null(prefix)) nn else
                            paste(prefix,nn,sep=""))
           if(length(flowSet) > 0 && flowSet) {
-              print(data.frame(name=nn, as.data.frame(f)[nn, ], row.names=nn))
-              out <- flowSet(out, phenoData=new("AnnotatedDataFrame",
-                                  data=data.frame(name=I(nn),
-                                  as.data.frame(f)[nn, ],
-                                  row.names=nn),
-                                  varMetadata=data.frame(labelDescription=I(
-                                                         c("Name", "Filter")),
-                                  row.names=c("name", "filter"))))
+              df <- data.frame(name=I(nn), row.names=nn)
+              vm <- data.frame(labelDescription=I("Name"), row.names="name")
+              out <- flowSet(out, phenoData=new("AnnotatedDataFrame", data=df,
+                                                varMetadata=vm))
           }
           return(if(is.list(out) && length(out)==1) out[[1]] else out)
       })
@@ -238,7 +234,10 @@ setMethod("split",
 ## filterResults, but we also need to make sure that this list is valid, i.e.,
 ## only contains filters or filterResults of the same class, using the same
 ## parameters, etc.
-## Splitting a lowSet always returns a list of flowSets.
+## Splitting a flowSet always returns a list of flowSets.
+## FIXME: How do we treat the cases in which multipleFilterResult produce
+## different numbers of populations? We can't collapse to a flowSet any more,
+## do we want lists of lists? Or should this be disallowed completely?
 
 
 ## Try and split a flowSet by whatever comes your way...
@@ -291,6 +290,8 @@ compatibleFilters <- function(f1, f2)
 
 ## Split a flowSet according to a list of filters or filterResults
 ## of equal length, We make sure that this list makes sense.
+## FIXME: Eventually, this function should be deprecated since we represent
+## filterResult for a flowSet in filterResultLists now.
[…]

### 2008-09-18 R/subsetFilter-accessors.R

@@ -1,46 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'intersectFilter'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-
-
-## ==========================================================================
-## show method
-## ---------------------------------------------------------------------------
-setMethod("show",signature("subsetFilter"),
-          function(object)
-      {
-          cat("filter '", identifier(object),
-              "'\nthe filtering operation defined by\n", sep="")
-          print(object@filters[[1]])
-          cat("after subsetting by\n")
-          print(object@filters[[2]])
-      })
-
-
-
-# FIXME: What do summarizeFilter methods do and why are they not
-## exported in the name space?
-setMethod("summarizeFilter",signature("filterResult","subsetFilter"),function(result,filter) {
-	ret = callNextMethod()
-	ret$subsetCount = attr(result@subSet,'subsetCount')
-	ret
-})
-
-
-## ==========================================================================
-## summary method for subsetFilter
-## --------------------------------------------------------------------------
-setMethod("summary","subsetFilter",function(object,result,...) {
-	if(missing(result)) {
-		e1 = as(object@filters[[1]],"logical")
-		e2 = as(object@filters[[2]],"logical")
-		true = sum(e1&e2)
-		count= sum(e2)		
-		new("filterSummary",name=identifier(object),true=true,count=count,p=true/count)			
-	} else {
-		true = sum(as(result,"logical"))
-		count = filterDetails(result,identifier(object))$subsetCount
-		new("filterSummary",name=identifier(object),true=true,count=count,p=true/count)			
-	}
-})

### 2008-09-18 R/timeFilter-accessors.R

@@ -1,24 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'timeFilter'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-
-
-## ==========================================================================
-## show method
-## ---------------------------------------------------------------------------
-setMethod("show",signature("timeFilter"),function(object)
-          {
-            msg <- paste("time filter '",object@filterId,
-                         "' with settings:\n  bandwidth=",
-                         object@bandwidth, sep="")
-            cat(msg)
-            if(length(object@binSize))
-              cat("\n  binSize=", object@binSize, sep="")
-            if(length(object@timeParameter))
-              cat("\n  timeParameter=", object@timeParameter, sep="")
-            cat("\n")
-            invisible(msg)
-          })
-
-

### 2008-09-18 R/transformFilter-accessors.R

@@ -1,31 +1,0 @@
-## ==========================================================================
-## Methods for objects of type 'transformFilter'
-## Note: All filtering methods are stored in file 'in-methods.R'
-## ==========================================================================
-
-
-## ==========================================================================
-## show method
-## ---------------------------------------------------------------------------
-setMethod("show",signature("transformFilter"),
-          function(object)
-      {
-          cat("transformed filter '", identifier(object), "'\n", sep="")
-      })
-
-
-
-## ===========================================================================
-## Constructors
-## ---------------------------------------------------------------------------
-setMethod("%on%",
-          signature("filter", "transformList"),
-          function(e1,e2)
-      {
-          new("transformFilter",
-              filterId=paste(e1@filterId,"on transformed values of",
-              paste(colnames(e2),collapse=",")),
-              transforms=e2,
-              filter=e1)
-      })
-

### 2008-09-18 R/workFlow-accessors.R

@@ -1,0 +1,209 @@
+## ==========================================================================
+## workFlow objects are basically evaluation environments and an associated
+## tree that binds the objects in the environment in a work flow structure.
+## All the usual operations for environments (get, assign, ls, rm) are also
+## available for workFlow objects.
+## ==========================================================================
+
+
+
+
+
+
+## ==========================================================================
+## Resolve a reference, i.e., get the symbol 'ID' from the evaluation
+## environment in 'workspace'.
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("get",
+          signature=signature(x="character", pos="workFlow",
+          envir="missing", mode="missing", inherits="missing"),
+          definition=function(x, pos)
+      {
+          if(!x %in% ls(pos@env)){
+              mess <- paste("Unable to resolve reference to object '",
+                            x, "'", sep="")
+              if(!is.na(w <- pmatch(tolower(x), 
+                                    tolower(ls(pos@env))))) 
+                  mess <- paste(mess, sprintf("Perhaps you meant %s?",
+                            sQuote(ls(pos@env)[w])), sep="\n")
+              stop(mess, call.=FALSE)
+          }
+          get(x, envir=pos@env)
+      })
+
+## The same behaviour as above, but allow workflow to be the 'envir' argument
+setMethod("get",
+          signature=signature(x="character", pos="missing",
+          envir="workFlow", mode="missing", inherits="missing"),
+          definition=function(x, envir) get(x, envir=envir@env))
+
+## get multiple objects
+setMethod("mget",
+          signature=signature(x="character", envir="workFlow",
+          mode="missing", ifnotfound="missing", inherits="missing"),
+          definition=function(x, envir) sapply(x, get, envir))
+
+
+
+## ==========================================================================
+## List the content of 'env' in a workFlow object
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("ls",
+          signature=signature(name="workFlow", pos="missing",
+          envir="missing", all.names="missing", pattern="missing"),
+          definition=function(name) ls(name@env))
+
+## Allow for pattern search
+setMethod("ls",
+          signature=signature(name="workFlow", pos = "missing",
+          envir="missing", all.names="missing", pattern="character"),
+          definition=function(name, pattern) ls(name@env, pattern=pattern))
+
+
+
[…]

### 2008-10-29 inst/doc/HowTo-flowCore.Rnw
updated filterSummary class and available methods
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33537 bc3139a8-67e5-0310-9ffc-ced21a209358
A first draft of the new workflow infrastructure and all the little changes that were necessary to make that work. This might eventually move to some other place. Also updated the documentation a bit. More to come soon...
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@33791 bc3139a8-67e5-0310-9ffc-ced21a209358
added workFlow to vignette
git-svn-id: file:///home/git/hedgehog.fhcrc.

@@
 will the base for the development of many tools for the analysis of
-high throughput flowcytometry.\\
+high throughput flow cytometry.\\
 
-\noindent \textbf{keywords} Flow cytometry, high throughtput,
+\noindent \textbf{keywords} Flow cytometry, high throughput,
 software, standard
@@
 the results have been equivalent or superior to traditional manual
-multiparameter staining and analysis techniques.
+multi-parameter staining and analysis techniques.
 
@@
 The aim is to merge both prada and rflowcyt \citep{LeMeur2006} into
-one core package wich is compliant with the data exchange standards
+one core package which is compliant with the data exchange standards
 that are currently developed in the community \citep{Spidlen2006}.
@@
 Finally, only a sample of lines can be read in case you need a quick
-overview of a large serie of files.
+overview of a large series of files.
 <<RedFiles3, echo=true, results=verbatim>>= 
@@
 rectGate <- rectangleGate(filterId="Fluorescence Region", 
-                          "FL1-H"=c(50, 100), "FL2-H"=c(50, 100))
+                          "FL1-H"=c(0, 12), "FL2-H"=c(0, 12))
 @
@@
 Though they are not explicitly filters, meaning that they cannot play
-a role in the construction of subfilters via operators,
+a role in the construction of sub filters via operators,
 \Rclass{filterSet} objects do define \Rfunction{filter} and
@@
 ensures that each parent filter is only evaluated once per call to
-\Rfunction{filter} so that if many subfilters rely on a particular,
+\Rfunction{filter} so that if many sub filters rely on a particular,
 likely expensive, filtering operation they use the original result
@@
 
-split(fs[[1]],fset1,flowSet=TRUE)
+split(fs[[1]], fset1, flowSet=TRUE)
 
@@
 
+\section{Work flows}
+\Rclass{filterSets} are very limited in their use for complex
+analysis work flows. They are  result-centric and it is hard to access
+intermediate results. \Rpackage{flowCore} offers much more versatile
+tools for such tasks though the \Rclass{workFlow} class. The general
+idea is to let the software handle the organization of intermediate
+results, naming schemes and operations and to provide a unified API
+to access and summarize these operations.
+
+\subsection{Abstraction of work flows}
+There are three classes in \Rpackage{flowCore} that are used to
+abstract work flows: \Rclass{workFlow} objects are the basic container
+holding all the necessary bits and pieces and they are the main
+structure for user interaction. \Rclass{actionItem} objects are
+abstractions of data analysis operations like gating, transformation
+or normalization. There are sub-classes for different types of
+operations. Finally, \Rclass{view} objects are created when applying
+\Rclass{actionItems} to a \Rclass{workFlow}. One can think of
+\Rclass{views} as separate data sets, for instance a subset that is
+created by a gating operation or the modified data following a
+transformation. Again, different types of \Rclass{views} are available
+through various sub-classes. This structure allows for a unified API
+in which the user interacts with either \Rclass{views} (the most
+common application) or \Rclass{actionItems}.
+
+It is important to know that \Rclass{workFlows} use a reference
+semantic instead of the pass-by-value semantic that is usually found
+in the R language. This design was chosen to minimize memory
+consumption and to keep the framework as flexible as possible. The
+only main consequence on the user-level is the fact that direct
+assignments to a \Rclass{workFlow} object are usually not necessary;
+i.e., functions that operate on the \Rclass{workFlow} have the
+potential side-effect of modifying the object.
+
+\subsection{Naming schemes}
+One major obstacle of using scripts as a work flow representation is
+the need to keep track of object symbols or names of list
+items. Choosing arbitrary names makes it hard to access objects and
+there is no way to easily query the structure of the work flow other
+than reading through the complete script. It is also very hard to
+appreciate the hierarchical structure of the work flow from a linear
+listing of R commands. Most objects in \Rpackage{flowCore} provide
+useful names or identifiers, and the work flow framework tries to
+utilize these names as much as possible. For example, a
+\Rclass{rectangleGate} called ``Lymphocytes'' will create two
+subpopulations, namely the cells within the gate and their
+complement. Intuitive names for these subpopulation would therefore be
+``Lymphocyte+'' and ``Lymphocyte-''. These names don't necessarily
+need to be unique, and in order to be able to unambiguously address
+each item in a work flow we need additional unique identifiers. Our
+software keeps an alias table that maps these internal unique
+identifiers to more human-readable and user friendly names, and unless
+there are naming collisions, those names will also be the primary
+identifiers that are exposed in the API. Only if and alias is not
+unique, objects need to be addressed by their internal unique
+identifier.
+
+\subsection{Creating \Rclass{workFlow} objects}
+Objects of class \Rclass{workFlow} have to be created using the
+constructor \Rfunction{workFlow}. The single mandatory argument is a
+flow data object, either a \Rclass{flowFrame} or a \Rclass{flowSet}.
+<<createWorkFlow>>=
+data(GvHD)
+wf <- workFlow(GvHD[1:5], name="myWorkflow")
+wf
+@
+%
+Printing the objects shows its structure: the work flow contains a
+single \Rclass{view} ``base view'' which essentially is the
+\Rclass{flowSet} \Robject{fs}. Since the data was added by the
+constructor and is not the result of a particular data analysis
+operation, no \Rclass{actionItem} is associated to the
+\Rclass{view}. We can query the available views in the
+\Rclass{workFlow} using the \Rfunction{views} method:
+<<views>>=
+views(wf)
+@
+
+\subsection{Adding  \Rclass{actionItems}}
+Currently there are four different types of \Rclass{actionItems} that
+can be added to a \Rclass{workFlow}:
+
+\begin{itemize}
+\item{\Rclass{compensateActionItem}: A \Rclass{compensation} object}
+\item{\Rclass{transformActionItem}: A \Rclass{transformList} object}
+\item{\Rclass{normalizeActionItem}: A \Rclass{normalization} object}
+\item{\Rclass{gateActionItem}: A \Rclass{filter} object}
+\end{itemize}
+
+The user doesn't have to bother with the details of these classes,
+they will be created internally when adding one of the above source
+objects to the \Rclass{workFlow} using the \Rfunction{add} method.
+<<addItems>>=
+tf <- transformList(colnames(GvHD[[1]])[3:6], asinh, transformationId="asinh")
+add(wf, tf)
+wf
+views(wf)
+@ 
+%
+There are several things to note here. First, we didn't assign the
+return value of \Rfunction{add} back to \Robject{wf}. This is the
+aforementioned feature of the reference semantic. Internally,
+\Rclass{workFlows} are to the most part \Rclass{environments}, and all
+modifications change the content of these \Rclass{environments} rather
+than the object itself. We also used the \Rfunction{add} methods with
+two arguments only: the \Rclass{workFlow} and the
+\Rclass{transformList}. In a hierarchical work flow we usually want to
+control to which subset (or rather to which \Rclass{view}, to stick to
+the work flow lingo) we want to add the \Rclass{actionItem} to. The
+default is to use the base view, and alternative parent \Rclass{views}
+have to be specified using the \Robject{parent} argument;
+\Robject{add(wf,tf,parent="base view")} would have been
+equivalent. Adding the transformation created a new \Rclass{view}
+called ``asinh'' and we can see that this name was automatically taken
+from the \Rclass{transformationList} object. If we want to control the
+name of the new \Rclass{view}, we can do so using the \Robject{name} argument:
+<<itemName>>=
+add(wf, tf, name="another asinh transform")
+wf
+@
+%
+This operation created a warning since we didn't use a leaf node to
+add the transformation. The software can't be sure if any of the
+downstream leafs in the work flow hierarchy are affected by the
+transformation (as it would certainly be true for most gates). One
+solution would be to update all downstream leafs in the work flow,
+but this feature is not supported yet. Also note that we used the same
+\Rclass{transformList} object twice. While we controlled the name that
+was used for the second \Rclass{view}, the alias for the
+\Rclass{actionItem} is the same in both cases, and we would have to
+use the internal unique identifier to unambiguously access the
+respective object.
+
+Transformation operations (and also compensations or normalizations)
+only create single \Rclass{views}. This is not true for gating
+operations. Even the most simple gate creates two populations: those
+cells in the gate and the complement. Accordingly, two separate
+\Rclass{views} have to be created as well. 
+<<addGate>>=
+rg <- rectangleGate("FSC-H"=c(200,400), "SSC-H"=c(250, 400),
+                    filterId="rectangle")
+add(wf, rg, parent="asinh")
+wf
+@
+%
+As we see, there are now two new \Rclass{views} under the ``asinh''
+node called ``rectangle+'' and ``rectangle-''. Other filter types may
+create more than two populations, for instance a \Rclass{quadrantGate}
+always results in four sub-populations. There is no restriction on the
+number of populations that are allowed to be created by a single
+gating operation, however, when gating a \Rclass{flowSet}, the number
+of populations have to be the same for each frame in the set.
+<<addQuadGate>>=
+qg <- quadGate("FL1-H"=2, "FL2-H"=4)
+add(wf,qg,parent="rectangle+")
+wf
+@ 
+%
+For complex work flows it becomes increasingly hard to convey all
+information by printing on the screen. There is a plotting function
+for \Rclass{workFlow} objects which plots the underlying tree. The
+function depends on the \Rpackage{Rgraphviz} package.
+<<plotwf, fig=true>>=
+plot(wf)
+@
+
+\subsection{Accessing items in the \Rclass{workFlow} object}
+The main reason for having the \Rclass{workFlow} class is to be able
+to easily access all elements of a potentially complex analysis work
+flow. The predominant use case here is to access individual
+\Rclass{views}, either for plotting, to create summary statistics or
+to use the underlying data for subsequent analysis steps. You can do
+that using the familiar list subsetting syntax:
+<<getView>>=
+wf[["rectangle+"]]
+wf$asinh
+@
+%
+This also works for \Rclass{actionItems}:
+<<getAction>>=
+wf[["action_rectangle"]]
+@ 
+% 
+In order to retreive the underlying data for a \Rclass{view} you
+can use the \Rfunction{Data} method.
+<<getData>>=
+Data(wf[["rectangle+"]])
+@ 
+%
+There are \Rfunction{summary} methods defined for the different
+\Rclass{view} subclasses, which basically create the respective
+summaries of the associated \Rpackage{flowCore} objects, e.g. a
+\Rclass{filterResult} for a filtering operation.
+<<summaries>>=
+summary(wf[["action_rectangle"]])
+summary(wf[["CD15 FITC+CD45 PE+"]])
+@ 
+The \Rpackage{flowViz} package also defines \Rfunction{xyplot} and
+\Rfunction{densityplot} methods for \Rclass{view} objects. Sensible defaults
+are usually chosen automatically, for instance to add the gate
+boundaries to the plot of a \Rclass{gateView}.
+<<plotViews>>=
+densityplot(wf[["base view"]])
+xyplot(`FL1-H` ~ `FL2-H`, wf[["CD15 FITC+CD45 PE+"]])
+@
+
+\subsection{Removing items from \Rclass{workFlow} object}
+Still need to document Rm, undo, etc.
+
 \clearpage

### 2008-11-01 inst/doc/HowTo-flowCore.Rnw
applyParentFilter is gone for good now
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35030 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \item[biexponentialTransform]{ $ f^{-1}(x) = ae^{bx}-ce^{dx}+f$}
-\item[logicleTransform]{ A special form of the biexponential transform with parameters selected by the data.}
+\item[logicleTransform]{ A special form of the biexponential transform
+    with parameters selected by the data.}
 \item[arcsinhTransform]{ $ f(x) = asinh\left(a+bx\right)+c$}
@@
 
-To use a standard transform, first we create a transform function via the constructors supplied by \Rpackage{flowCore}:
+To use a standard transform, first we create a transform function via
+the constructors supplied by \Rpackage{flowCore}:
 
@@
 <<getData>>=
-Data(wf[["rectangle+"]])
+Data(wf[["rectangle-"]])
 @ 
@@
 \subsection{Removing items from \Rclass{workFlow} object}
-Still need to document Rm, undo, etc.
+The hierarchical structur of the \Rclass{workFlow} object introduces
+dependencies between \Rclass{views} or between \Rclass{views} and
+\Rclass{actionItems}. Thus, removing a particular view means also
+removing all of its associated child \Rclass{views} and
+\Rclass{actionItems}. The easiest way to remove objects from a
+\Rclass{workFlow} is through the \Robject{undo} mechanism. The
+function takes a \Rclass{workFlow} object as first argument and an
+optional second integer argument giving the number of operartions that
+are supposed to be roled back.
+<<undo>>=
+undo(wf)
+wf
+@ 
+%
+Sometimes it is more convenient to specify a particular \Rclass{view}
+or \Rclass{action} item that is to be removed. This can be archived
+using the \Rfunction{Rm} methods for \Rclass{views} or
+\Rclass{actionItems}. As a side-effect these methods will also remove
+all dependent items and the user should be well aware of this
+potentially dangerous behaviour.
+<<RmView>>=
+Rm('rectangle-', wf)
+@ 
+<<RmAction>>=
+Rm('asinh', wf)
+@
 

### 2008-11-01 R/view-accessors.R

@@ -24,7 +24,46 @@ setMethod("action",
 ## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
 setMethod("Data",
           signature=signature(object="view"),
-          definition=function(object) get(object@data))
+          definition=function(object){
+              dat <- get(object@data)
+              if(is.null(dat)){
+                  parent <- parent(object)
+                  ## The parent gate is not be evaluated yet, we do that now
+                  if(is(object, "gateView")){
+                      parentData <- Data(parent)
+                      if(is(parentData, "flowFrame")){
+                          newData <- parentData[object@indices[[1]],]
+                      }else{
+                          newData <- parentData[1:length(parentData)]
+                          for(i in seq_along(object@indices))
+                              newData[[i]] <- newData[[i]][object@indices[[i]],]
+                      }
+                      newDataID <- refName(newData)
+                      env <- object@env
+                      ## We also need to update the appropriate journal entry
+                      jid <- grep("journalRef", ls(env), value=TRUE)
+                      if(length(jid) != 1)
+                          stop("Unable to find journal in this environment.",
+                               call.=FALSE)
+                      journal <- journal(env)
+                      assign(x=newDataID, value=newData, envir=env)
+                      actionId <- identifier(action(object))
+                      if(actionId %in% names(journal)){
+                          journal[[actionId]] <- c(journal[[actionId]],
+                                                          newDataID)
+                          assign(jid, journal, env)
+                      }
+                      object@data <- new("fcDataReference", ID=newDataID,
+                                         env=env)
+                      assign(identifier(object), value=object, envir=env)
+                      return(newData)
+                  }else{
+                      stop("Unable to find data for this view.\n This workFlow object ",
+                           "is corrupted beyond repair.", call.=FALSE)
+                  }
+              }
+              return(dat)
+          })
 
 
 

### 2008-11-01 R/AllClasses.R
applyParentFilter is gone for good now
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35030 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -19,9 +19,9 @@ checkClass <- function(x, class, length=NULL, verbose=FALSE,
     if(mandatory && missing(x))
         stop("Argument '", substitute(x), "' missing with no default",
              call.=verbose)
-    msg <- paste("'", substitute(x), "' must be object of class '",
-                 class, "'", sep="")
-    fail <- !is(x, class)
+    msg <- paste("'", substitute(x), "' must be object of class ",
+                 paste("'", class, "'", sep="", collapse=" or "), sep="")
+    fail <- !any(sapply(class, function(c, y) is(y, c), x))
     if(!is.null(length) && length(x) != length){
         fail <- TRUE
         msg <- paste(msg, "of length", length)
@@ -1553,9 +1553,9 @@ setAlias <- function(alias, value, workflow)
 {
     checkClass(alias, "character", 1)
     checkClass(value, "character", 1)
-    checkClass(workflow, "workFlow")
-    workflow <- alias(workflow)
-    workflow[[alias]] <- unique(c(workflow[[alias]], value))
+    checkClass(workflow, c("workFlow", "environment"))
+    aliasEnv <- alias(workflow)
+    aliasEnv[[alias]] <- unique(c(aliasEnv[[alias]], value))
     return(invisible(NULL))
 }
 
@@ -1563,7 +1563,7 @@ setAlias <- function(alias, value, workflow)
 getAlias <- function(alias, workflow)
 {
     checkClass(alias, "character")
-    checkClass(workflow, "workFlow")
+    checkClass(workflow, c("workFlow", "environment"))
     fun <- function(x)
         if(x %in% ls(workflow)) x else alias(workflow)[[x]]
     return(as.vector(sapply(alias, fun)))
@@ -1611,8 +1611,9 @@ fcReference <- function(ID=paste("genericRef", guid(), sep="_"),
                         env=new.env(parent=emptyenv()))
 {
     checkClass(ID, "character", 1)
-    checkClass(env, "environment")
-    ref <- new("fcReference", ID=ID, env=env)
+    checkClass(env, c("workFlow", "environment"))
+    envir <- if(is(env, "workflow")) env@env else env
+    ref <- new("fcReference", ID=ID, env=envir)
     setAlias(substitute(get(ref)), identifier(ref), env)
     return(ref)
 }
@@ -1646,9 +1647,10 @@ fcTreeReference <- function(ID=paste("treeRef", guid(), sep="_"),
[…]

### 2008-11-21 paper/bmc_article.tex
flowCore to BMC bioinfo
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35431 bc3139a8-67e5-0310-9ffc-ced21a209358

--- 2008-08-15 paper/manuscript.tex
@@
-\documentclass[12pt]{article}
-\usepackage{url}
-\usepackage[authoryear,round]{natbib}
-\usepackage{graphicx} 
-\newcommand{\Rpackage}[1]{{\textsf{#1}}}
-\newcommand{\Rclass}[1]{{\textit{#1}}}
+%% BioMed_Central_Tex_Template_v1.05
+%%                                      %
+%  bmc_article.tex            ver: 1.05 %
+%                                       %
 
-\title{flowCore: A Bioconductor software package for high throughput flow cytometry analysis}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                     %%
+%%  LaTeX template for BioMed Central  %%
+%%     journal article submissions     %%
+%%                                     %%
+%%         <27 January 2006>           %%
+%%                                     %%
+%%                                     %%
+%% Uses:                               %%
+%% cite.sty, url.sty, bmc_article.cls  %%
+%% ifthen.sty. multicol.sty		       %%
+%%									   %%
+%%                                     %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\author{Florian Hahne*\\
-  Nolwenn LeMeur*\\
-  Byron Ellis\\
-  Ryan R. Brinkman\\
-  Perry Haaland\\
-  Errol Strain\\
-  Deepayan Sarkar\\
-  Josef Spidlen\\
-  Robert Gentleman
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                                                 %%	
+%% For instructions on how to fill out this Tex template           %%
+%% document please refer to Readme.pdf and the instructions for    %%
+%% authors page on the biomed central website                      %%
+%% http://www.biomedcentral.com/info/authors/                      %%
+%%                                                                 %%
+%% Please do not use \input{...} to include other tex files.       %%
+%% Submit your LaTeX manuscript as one .tex document.              %%
+%%                                                                 %%
+%% All additional figures and files should be attached             %%
+%% separately and not embedded in the \TeX\ document itself.       %%
+%%                                                                 %%
+%% BioMed Central currently use the MikTex distribution of         %%
+%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
+%% http://www.miktex.org                                           %%
+%%                                                                 %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\NeedsTeXFormat{LaTeX2e}[1995/12/01]
+\documentclass[10pt]{bmc_article}
+
+% Load packages
+\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
+\usepackage{url}  % Formatting web addresses  
+\usepackage{ifthen}  % Conditional 
+\usepackage{multicol}   %Columns
+\usepackage[utf8]{inputenc} %unicode support
+%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
+%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+\urlstyle{rm}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
+%%                                             %%
+%%  If you wish to display your graphics for   %%
+%%  your own use using includegraphic or       %%
+%%  includegraphics, then comment out the      %%
+%%  following two lines of code.               %%   
+%%  NB: These line *must* be included when     %%
+%%  submitting to BMC.                         %% 
+%%  All figure files must be submitted as      %%
+%%  separate graphics through the BMC          %%
+%%  submission process, not included in the    %% 
+%%  submitted article.                         %% 
+%%                                             %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\def\includegraphic{}
+\def\includegraphics{}
+
+\setlength{\topmargin}{0.0cm}
+\setlength{\textheight}{21.5cm}
+\setlength{\oddsidemargin}{0cm} 
+\setlength{\textwidth}{16.5cm}
+\setlength{\columnsep}{0.6cm}
+
+\newboolean{publ}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                              %%
+%% You may change the following style settings  %%
+%% Should you wish to format your article       %%
+%% in a publication style for printing out and  %%
+%% sharing with colleagues, but ensure that     %%
+%% before submitting to BMC that the style is   %%
+%% returned to the Review style setting.        %%
+%%                                              %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%Review style settings
+\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+
+%Publication style settings
+%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+
+% Begin ...
+\begin{document}
+\begin{bmcformat}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the title of your article here     %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\title{flowCore: A Bioconductor software package for high throughput
+  flow cytometry data analysis}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors here                   %%
+%%                                          %%
+%% Ensure \and is entered between all but   %%
+%% the last two authors. This will be       %%
+%% replaced by a comma in the final article %%
+%%                                          %%
+%% Ensure there are no trailing spaces at   %% 
+%% the ends of the lines                    %%     	
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\author{Florian Hahne\correspondingauthor$^{1}$%
+       \email{Florian Hahne\correspondingauthor - fhahne@fhcrc.org}%
+      \and
+         Nolwenn LeMeur\correspondingauthor$^{1,2}$%
+         \email{Nolwenn LeMeur\correspondingauthor - nlemeur@irisa.fr}
+       and 
+         Ryan R. Brinkman$^3$%
+         \email{Ryan R. Brinkman - rbrinkman@bccrc.ca}%
+       and
+          Byron Ellis$^4$%
+          \email{Byron Ellis - byron.ellis@gmail.com}
+       and
+          Perry Haaland$^5$%
+          \email{Perry Haaland - Perry_D_Haaland@bd.com}
+       and
+          Deepayan Sarkar$^{1}$%
+          \email{Deepayan Sarkar - dsarkar@fhcrc.org}
+       and
+         Josef Spidlen$^3$%
+         \email{Josef Spidlen - jspidlen@bccrc.ca}
+       and
+         Errol Strain$^5$%
+         \email{Errol Strain - Errol_Strain@bd.com  }  
+       and
+          Robert Gentleman$^{1}$%   
+          \email{Robert Gentleman - rgentlem@fhcrc.org}
       }
 
-\begin{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% Enter the authors' addresses here        %%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\address{%
+    \iid(1)Life Sciences Department, Computational Biology Program,%
+    Division of Public Health Sciences, Fred Hutchinson Cancer Research Center,% 
+    1100 Fairview Ave. N, M2-B876, PO Box 19024,  Seattle, Washington 98109-1024\\
+    \iid(2)INSERM, IRISA - Symbiose, Campus Beaulieu, Université de Rennes I, %
+    35042 Rennes Cedex,France \\
+    \iid(3) Terry Fox Laboratory, British Columbia Cancer Agency Research Centre, %
+    675 West 10th Avenue, Vancouver, BC V5Z 1L3  Canada, Canada\\
+    \iid(4) AdBrite Inc\\
+    \iid(5) BD Biosciences\\
+
+}%
 
@@
 
-\section*{Abstract}
-\subsubsection*{Background}
-Recent advances in automation technologies have enabled the use of
-flow-cytometry high content screening (FH-HCS). However, data
-management and data analysis methods have not advanced sufficiently
-far from the inital small-scale studies to support the large complex
-data sets generated in FH-HCS applications.
-\subsubsection*{Methods}
-In this manuscript, we describe a set of flexible
-computational tools for FC-HCS data analysis written 
-in R and freely available from the Bioconductor Project.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Abstract begins here                 %%
+%%                                          %%
+%% The Section headings here are those for  %%
+%% a Research article submitted to a        %%
+%% BMC-Series journal.                      %%  
+%%                                          %%
+%% If your article is not of this type,     %%
+%% then refer to the Instructions for       %%
+%% authors on http://www.biomedcentral.com  %%
+%% and change the section headings          %%
+%% accordingly.                             %%   
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsubsection*{Results}
-The \Rpackage{flowCore} package provides the flow cytometry community
-with a shared research platform that enables bioinformaticians,
-computer scientists, and statisticians to work collaboratively with
-biologists and clinicians to develop novel methods for FCM data
-analysis.
-\subsubsection*{Conclusions}
-We hope that our framework will be the foundation for fruitful 
-collaborative research involving multiple scientific fields and
-that it will help facilitate the
-development and deployment of FC-HCS methods to complex 
-scientific and clinical applications.
+\begin{abstract}
+        % Do not use inserted blank lines (ie \\) until main body of text.
+  \paragraph*{Background:} Recent advances in automation technologies
+  have enabled the use of flow-cytometry high content screening
+  (FH-HCS). However, data management and data analysis methods have
+  not advanced sufficiently far from the inital small-scale studies to
+  support the large complex data sets generated in FH-HCS
+  applications.
 
-\section*{Introduction}
+  \paragraph*{Results:} The \Rpackage{flowCore} package provides the
+  flow cytometry community with a shared research platform that
+  enables bioinformaticians, computer scientists, and statisticians to
+  work collaboratively with biologists and clinicians to develop novel
+  methods for FCM data analysis.
+
+  \paragraph*{Conclusions:} We hope that our framework will be the
+  foundation for fruitful collaborative research involving multiple
+  scientific fields and that it will help facilitate the development
+  and deployment of FC-HCS methods to complex scientific and clinical
+  applications.
+\end{abstract}
+
+\ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                          %%
+%% The Main Body begins here                %%
+%%                                          %%
+%% The Section headings here are those for  %%
+%% a Research article submitted to a        %%
+%% BMC-Series journal.                      %%  
+%%                                          %%
+%% If your article is not of this type,     %%
+%% then refer to the instructions for       %%
+%% authors on:                              %%
+%% http://www.biomedcentral.com/info/authors%%
+%% and change the section headings          %%
+%% accordingly.                             %% 
+%%                                          %%
+%% See the Results and Discussion section   %%
+%% for details on how to create sub-sections%%
+%%                                          %%
+%% use \cite{...} to cite references        %%
+%%  \cite{koon} and                         %%
+%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
+%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
+%%                                          %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%
+%% Background %%
+%%
+\section*{Background}
  \subsection*{Computational tools for the analysis of FC-HCS data}
@@
 
-\section*{Materials and Methods}
+%%%%%%%%%%%%%%%%%%
+\section*{Implementation}
  \subsection*{Basic data structures}
@@
 experiment.
-
-%%Figure1 Diagram of the data structures and the basic operations
-\begin{figure}
-\centering
-\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
-\caption{\label{fig1:FrameWork}{For each experiment, the content of
-    the FCS files, phenotypic and metadata are stored in a
-    \Rclass{flowSet}. Each \Rclass{flowFrame} in a \Rclass{flowSet}
-    corresponds to one FCS file. All basic operations (e.g.,
-    compensation, transformation, gating) can be applied to either
-    single \Rclass{flowFrames} or a \Rclass{flowSet} simultaneously.}}
-\end{figure}
 
@@
 
-\begin{table}[ht]
-\begin{center}
-\begin{tabular}{|l|l|}
-\hline
-\multicolumn{2}{|c|}{Data Transformations} \\
-\hline
-linear & $ax + b$ \\
-quadratic & $ax^2 + bx + c$ \\
-natural logarithm & $ln(x)(r/d)$ \\
-logarithm & $log(x)(r/d)$ \\
-biexponential & $root(a*e^{(b*y)}-c*e^{(-d+y)}+f)$ \\
-logicle& $root(Te^{-(m-w)}*(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
-truncate & $x_{x<=a} = a$ \\
-scale & $(x-a)/(b-a)$ \\
-split-scale & combination of linear and logarithm \\
-arcsinh & $asinh(a + bx)+c$ \\
-\hline
-\end{tabular}
-\caption{\label{table1}Data transformations implemented in
-  \Rpackage{flowCore}. Within these formulas, $x$ is the variable
-  corresponding to value being transformed, $y$ is the result of the
-  transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
-  constants affecting the transformation function, $e$ is the base of
-  the natural logarithm, and $root$ is a root-finding algorithm.  }
-\end{center}
-\end{table}
-
 \subsubsection*{Gating}
@@
 
-\begin{table}[ht]
-\begin{center}
-\begin{tabular}{|l|l|}
-\hline
-\multicolumn{2}{|c|}{Gates} \\
-\hline
-rectangleGate & n-dimensional rectangular regions \\
-quadGate & quadrant regions in two dimensions \\
-polygonGate & polygonal regions in two dimensions \\
-polytopeGate & generalization of polygon in n dimensions \\
-ellipsoidGate & n-dimensional ellipsoid region \\
-\hline
-\multicolumn{2}{|c|}{Filters} \\
-\hline
-sampleFilter & random sub-sampling\\
-expressionFilter & results of a boolean expression \\
-kmeansFilter & K-means clustering \\
-norm2Filter & bivariate normal distribution \\
-curv1Filter & local density regions in 1D \\
-curv2Filter & density regions in 2D \\
-timeFilter & abnormal data acquisition over time \\
-\hline
-filterSet & gating strategies \\
-\hline
-\end{tabular}
-\caption{\label{table2}Filter and gate classes implemented in
-  \Rpackage{flowCore}.}
-\end{center}
-\end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Results and Discussion %%
+%%
+\section*{Results and Discussion}
+  \subsection*{From quality assessment to batch gating}
+  The \Rpackage{flowCore} package has been successfully applied in the
+  analysis of several datasets
+  \citep{gasparetto2004ice,brinkman2007hcf}. A complete documentation
+  of the \Rpackage{flowCore} software is beyond the scope of this
+  publication.  Much more comprehensive documentation and users guide
+  information with programmatic examples is available online
+  (\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
+  and as part of the package distribution. Here, we want to briefly
+  exemplify some of the software's key features, that is, the coherent
+  treatment of all samples in a potentially large experiment, the
+  concept of data-driven automated gating, the integration of existing
+  software into the framework, and the generation of
+  publication-quality graphics for data visualization.
+
+  Data analysis for most experiments usually begins with a quality
+  assurance step, and we can use functionality from the
+  \Rpackage{flowQ} package to create an HTML report that highlights
+  potential quality issues. Assuming that the data has already been
+  imported as the \Rclass{flowSet} object ``dat'', the following
+  simple line of code produces the output shown in Figure~\ref{flowQ}:
+
+\begin{verbatim}
+> library(flowCore)
+> library(flowQ)
+> qaReport(dat, c("qaProcess.timeline", "qaProcess.timeflow", 
+                  "qaProcess.cellnumber"))
+\end{verbatim}
+
+  The report is interactive and provides drill-down to more detailed
+  aspects of the analysis, starting from a concise overview. The
+  design of \Rpackage{flowCore}'s data model allows for a coherent
+  treatment of all the samples, hence we are able to compare features
+  between individuals, or between groups of individuals, based on the
+  available metadata information.
+
+  Static gating for all samples in a high-throughput FCM experiment is
+  often impossible, since the measured variables tend to vary between
+  different treatments, over time or between different experiment
+  batches. Automated or data-driven gating has the potential to
+  estimate the gating regions from the underlying data, thus providing
+  a fast objective solution to the analysis of potentially very large
+  and diverse data sets \citep{lo2008agf}. One of the automated gating
+  methods implemented in \Rpackage{flowCore} is based on identifying
+  areas of significant curvature in a kernel density estimate of the
+  data \citep{wand2008}. Assuming that the regions of interest are of
+  high density, the software is able to reliably detect them in a one
+  or two-dimensional density landscape.
+
+\begin{verbatim}
+> cf <- curv2Filter("FL1-H", "FL3-H")
+> fres <- filter(dat, cf)
+\end{verbatim}
+
+Kernel density estimation is a well-known problem in statistical
+computing, and a lot of effort has been invested in the development of
+good software to address it. The modular design of \Rpackage{flowCore}
+allows to easily integrate these existing solutions into our
+framework. In this example, we directly use R code from the
+\Rpackage{feature} package developed by \cite{wand2008}. Instead of
+re-writing existing code, we are able to include it via the well
+tested distribution mechanism provided by R's software package
+system. This process is bi-directional, and all functionality
+implemented in \Rpackage{flowCore} is available to other package
+authors, as we have seen with the afore-mentioned \Rpackage{flowClust}
+package.
+
+We can chose one of the many visualization option from the flowViz
+package to plot the results of the recent filtering operation. A very
+basic matrix of density plots is shown in Figure \ref{xyplot}, where
+each panel in the matrix represents the data of one individual
+patient.
+
+\begin{verbatim}
+> xyplot(`FL1-H` ~ `FL3-H` | SampleID, data=dat, filter=fres)
+\end{verbatim}
 
@@
 
-\section*{Results}
-The \Rpackage{flowCore} package has been successfully applied in the
-analysis of several datasets
-\citep{gasparetto2004ice,brinkman2007hcf}. A complete documentation of
-the \Rpackage{flowCore} software is beyond the scope of this
-publication.  Much more comprehensive documentation and users guide
-information with programmatic examples is available online
-(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
-and as part of the package distribution. Here, we want to briefly
-exemplify some of the software's key features, that is, the coherent
-treatment of all samples in a potentially large experiment, the
-concept of data-driven automated gating, the integration of existing
-software into the framework, and the generation of publication-quality
-graphics for data visualization.
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Conclusions}
+  Text for this section \ldots
 
-Data analysis for most experiments usually begins with a quality
-assurance step, and we can use functionality from the \Rpackage{flowQ}
-package to create an HTML report that highlights potential quality
-issues. Assuming that the data has already been imported as the
-\Rclass{flowSet} object ``dat'', the following simple line of code
-produces the output shown in Figure~\ref{flowQ}:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Authors contributions}
+    Text for this section \ldots
 
-\begin{verbatim}
-> library(flowCore)
-> library(flowQ)
-> qaReport(dat, c("qaProcess.timeline", "qaProcess.timeflow", 
-                  "qaProcess.cellnumber"))
-\end{verbatim}
-
-The report is interactive and provides drill-down to more detailed
-aspects of the analysis, starting from a concise overview. The design of \Rpackage{flowCore}'s data model
-allows for a coherent treatment of all the samples, hence we are able
-to compare features between individuals, or between groups of
-individuals, based on the available metadata information.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=0.98\textwidth]{flowQ.jpg}
-\caption{\label{flowQ}%
-  HTML quality assessment report generated by the flowQ package for a
-  subset of data from an experiment focussing on Graft-Versus-Host
-  Disease \citep{brinkman2007hcf}. Rows correspond to the samples in
-  the set, columns to different quality checks.}
-\end{figure}
-
-Static gating for all samples in a high-throughput FCM experiment is
-often impossible, since the measured variables tend to vary between
-different treatments, over time or between different experiment
-batches. Automated or data-driven gating has the potential to estimate
-the gating regions from the underlying data, thus providing a fast
-objective solution to the analysis of potentially very large and
-diverse data sets \citep{lo2008agf}. One of the automated gating
-methods implemented in \Rpackage{flowCore} is based on identifying
-areas of significant curvature in a kernel density estimate of the
-data \citep{wand2008}. Assuming that the regions of interest are of
-high density, the software is able to reliably detect them in a one or
-two-dimensional density landscape.
-
-\begin{verbatim}
-> cf <- curv2Filter("FL1-H", "FL3-H")
-> fres <- filter(dat, cf)
-\end{verbatim}
-
-Kernel density estimation is a well-known problem in statistical
-computing, and a lot of effort has been invested in the development of
-good software to address it. The modular design of \Rpackage{flowCore}
-allows to easily integrate these existing solutions into our
-framework. In this example, we directly use R code from the
-\Rpackage{feature} package developed by \cite{wand2008}. Instead of
-re-writing existing code, we are able to include it via the well
-tested distribution mechanism provided by R's software package
-system. This process is bi-directional, and all functionality
-implemented in \Rpackage{flowCore} is available to other package
-authors, as we have seen with the afore-mentioned \Rpackage{flowClust}
-package.
-
-We can chose one of the many visualization option from the flowViz
-package to plot the results of the recent filtering operation. A very
-basic matrix of density plots is shown in Figure \ref{xyplot}, where
-each panel in the matrix represents the data of one individual
-patient.
-
-\begin{verbatim}
-> xyplot(`FL1-H` ~ `FL3-H` | SampleID, data=dat, filter=fres)
-\end{verbatim}
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=0.75\textwidth]{xyplot}
-\caption{\label{xyplot}%
-  Scatterplot matrix of a single \Rclass{flowSet} from an experiement
-  focussing on immune tolerance following kidney
-  transplantation. Outlines of the gating regions identified by an
-  automated gating operation are added on top of the density
-  representation of the data.}
-\end{figure}
-
-\section*{Discussion}
-
-Through \Rpackage{flowCore}, we have provided the FCM community with
-an open source, freely available, highly functional, standards
-compliant development and analysis platform for high throughput data
-analysis.  Our intention is to provide a platform for the
-collaborative development of new analysis methods that will facilitate
-the transition of these new methods to the larger flow community.  Our
-experience has been that the collaborative effort we enable for
-devising new methodology has been proven beneficial for a number of
-different biological and computational biology challenges.  We hope
-that our framework will be the foundation for fruitful shared research
-by many collaborators from multiple scientific fields and will help
-resolve bottlenecks that currently prevent further development and
-deployment of FC-HCS to increasingly complex and important scientific
-and clinical applications.
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Acknowledgements}
@@
 
-\bibliographystyle{plainnat}  
-\bibliography{flowCoreRef} 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                  The Bibliography                       %%
+%%                                                         %%              
+%%  Bmc_article.bst  will be used to                       %%
+%%  create a .BBL file for submission, which includes      %%
+%%  XML structured for BMC.                                %%
+%%                                                         %%
+%%                                                         %%
+%%  Note that the displayed Bibliography will not          %% 
+%%  necessarily be rendered by Latex exactly as specified  %%
+%%  in the online Instructions for Authors.                %% 
+%%                                                         %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
+ \bibliographystyle{bmc_article}  % Style BST file
+  \bibliography{flowCoreRef} }     % Bibliography file (usually '*.bib' )
+
+%%%%%%%%%%%
+
+\ifthenelse{\boolean{publ}}{\end{multicols}}{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Figures                       %%
+%%                               %%
+%% NB: this is for captions and  %%
+%% Titles. All graphics must be  %%
+%% submitted separately and NOT  %%
+%% included in the Tex document  %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%
+%% Do not use \listoffigures as most will included as separate files
+
+\section*{Figures}
+  \subsection*{Figure 1 - Sample figure title}
+      A short description of the figure content
+      should go here.
+
+  \subsection*{Figure 2 - Sample figure title}
+      Figure legend text.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Tables                        %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Use of \listoftables is discouraged.
+%%
+\section*{Tables}
+  \subsection*{Table 1 - Sample table title}
+    Here is an example of a \emph{small} table in \LaTeX\ using  
+    \verb|\tabular{...}|. This is where the description of the table 
+    should go. \par \mbox{}
+    \par
+    \mbox{
+      \begin{tabular}{|c|c|c|}
+        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
+        A1 & B2  & C3 \\ \hline
+        A2 & ... & .. \\ \hline
+        A3 & ..  & .  \\ \hline
+      \end{tabular}
+      }
+  \subsection*{Table 2 - Sample table title}
+    Large tables are attached as separate files but should
+    still be described here.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Additional Files              %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Additional Files}
+  \subsection*{Additional file 1 --- Sample additional file title}
+    Additional file descriptions text (including details of how to
+    view the file, if it is in a non-standard format or the file extension).  This might
+    refer to a multi-page table or a figure.
+
+  \subsection*{Additional file 2 --- Sample additional file title}
+    Additional file descriptions text.
+
+\end{bmcformat}
 \end{document}

### 2008-11-25 paper/bmc_article.tex
work on the new layout
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35475 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 \urlstyle{rm}
+
+\newcommand{\Rpackage}[1]{{\textsf{#1}}}
+\newcommand{\Rclass}[1]{{\textit{#1}}}
 
@@
 
-\title{flowCore: A Bioconductor software package for high throughput
-  flow cytometry data analysis}
+\title{flowCore a Bioconductor package for high throughput flow cytometry}
 
@@
 
-\author{Florian Hahne\correspondingauthor$^{1}$%
+\author{Florian Hahne\correspondingauthor$^1$%
        \email{Florian Hahne\correspondingauthor - fhahne@fhcrc.org}%
@@
          \email{Nolwenn LeMeur\correspondingauthor - nlemeur@irisa.fr}
-       and 
+       \and 
          Ryan R. Brinkman$^3$%
          \email{Ryan R. Brinkman - rbrinkman@bccrc.ca}%
-       and
+        \and
           Byron Ellis$^4$%
           \email{Byron Ellis - byron.ellis@gmail.com}
-       and
+        \and
           Perry Haaland$^5$%
-          \email{Perry Haaland - Perry_D_Haaland@bd.com}
-       and
-          Deepayan Sarkar$^{1}$%
+          \email{Perry Haaland - Perry\_D\_Haaland@bd.com}
+        \and
+          Deepayan Sarkar$^1$%
           \email{Deepayan Sarkar - dsarkar@fhcrc.org}
-       and
+        \and
          Josef Spidlen$^3$%
          \email{Josef Spidlen - jspidlen@bccrc.ca}
+        \and
+         Errol Strain$^5$%
+         \email{Errol Strain - estrain@gmail.com}  
        and
-         Errol Strain$^5$%
-         \email{Errol Strain - Errol_Strain@bd.com  }  
-       and
-          Robert Gentleman$^{1}$%   
+          Robert Gentleman$^1$%   
           \email{Robert Gentleman - rgentlem@fhcrc.org}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \address{%
@@
   1100 Fairview Ave. N, M2-B876, PO Box 19024,  Seattle, Washington 98109-1024\\
-    \iid(2)INSERM, IRISA - Symbiose, Campus Beaulieu, Université de Rennes I, %
-    35042 Rennes Cedex,France \\
-    \iid(3) Terry Fox Laboratory, British Columbia Cancer Agency Research Centre, %
+  \iid(2)EA SeRAIC INSERM, IRISA - Symbiose, Campus Beaulieu, Universit\'{e} de Rennes I, % 
+  35042 Rennes Cedex, France \\
+  \iid(3)Terry Fox Laboratory, British Columbia Cancer Agency Research Centre,% 
   675 West 10th Avenue, Vancouver, BC V5Z 1L3  Canada, Canada\\
-    \iid(4) AdBrite Inc\\
-    \iid(5) BD Biosciences\\
-
+  \iid(4)AdBrite Inc\\
+  \iid(5)BD Biosciences
 }%
-
 \maketitle
@@
 
-  \paragraph*{Results:} The \Rpackage{flowCore} package provides the
-  flow cytometry community with a shared research platform that
-  enables bioinformaticians, computer scientists, and statisticians to
-  work collaboratively with biologists and clinicians to develop novel
-  methods for FCM data analysis.
+  \paragraph*{Results:} We developped a set of flexible computational
+  tools for FC-HCS data analysis written in R and freely available
+  from the Bioconductor Project. The \Rpackage{flowCore} package
+  provides the flow cytometry community with a shared research
+  platform that enables bioinformaticians, computer scientists, and
+  statisticians to work collaboratively with biologists and clinicians
+  to develop novel methods for flow cytometry data analysis.
 
@@
 \section*{Background}
- \subsection*{Computational tools for the analysis of FC-HCS data}
 Automation technologies developed during the last several years have
-enabled the use of FCM high content screening (FC-HCS) to generate
-large, complex datasets in both basic and clinical research
-applications \citep{brinkman2007hcf}. A serious bottleneck in the
+enabled the use of flow cytometry high content screening (FC-HCS) to
+generate large, complex datasets in both basic and clinical research
+applications \cite{brinkman2007hcf}. A serious bottleneck in the
 interpretation of existing studies and the application of FC-HCS to
@@
 developed for applications of flow cytometry (FCM) to small-scale,
-tube-based studies.
+tube-based studies \cite{mahnke2007omi}.
+%% add one of the last roederer's paper that "request" algo and
+%% automation for FC-HCS data analysis.
 
@@
 
-In this paper, we describe a set of flexible and well structured
-computational tools to efficiently analyze FC-HCS data. This set of
-tools is based on the open source statistical software R
-\citep{Rmain}, in conjunction with the Bioconductor Project
-\citep{BIOC}. Our intent is to provide a shared research platform that
-enables bioinformaticians, computer scientists, and statisticians to
-work collaboratively with biologists and clinicians to develop novel
-methods for FCM data analysis, a process deemed crucial by many for
-the further development of the technology \citep{lizard2007fca}. A
-critical component of this approach is the development and
-implementation of standards that will facilitate the adoption of these
-methods by both the larger research community and commercial software
-vendors.  Consequently, we are optimistic that this software platform
-will play a critical role in addressing the bottlenecks described
-above that limit advances in the application of FC-HCS.
-
-The computational tools we have developed are distributed in the R
-software language as the Bioconductor package
-\Rpackage{flowCore}. \Rpackage{flowCore} is a freely available, highly
-functional, and extensible FCM data analysis platform that enables
-researchers to efficiently handle FC-HCS data and encourages open
-development of tools for their coherent analysis. \Rpackage{flowCore}
-implements computationally efficient data structures and a range of
-specialized methods addressing all components of a typical FCM
-analysis work flow, including compensation, transformation, and
-gating. \Rpackage{flowCore} runs on Windows, Mac OS X, and Linux/Unix
-operating systems. Conceptually, the underlying data structures that
-build the core of our software can be implemented in any programming
-language. We chose the R language for its advantages in statistical
-programming and visualization. In this paper, we describe the
-functionality and methods underlying \Rpackage{flowCore}, and provide
-examples of their use.
-
-\subsection*{R and Bioconductor}
-R is a robust statistical programing environment which offers a wide
-range of statistical and visualization methods developed for various
-fields of application. In particular, the Bioconductor project
-provides R software modules for biological and clinical data analysis.
-It is widely used to process such data, spanning a diversity of
-research fields including genomics, proteomics and cell biology
-\citep{BIOC}. The R language and Bioconductor are particularly geared
-towards the analysis of large complex datasets that typically arise
-from high throughput experiments such as microarray gene expression
-analysis or FC-HCS. Using FCM as a high-throughput technology requires
-suitable software infrastructure to facilitate the rapid handling of
-the data. In our implementation of \Rpackage{flowCore} we rely on two
-important lessons learned from the field of gene expression data
-analysis: the first being the importance of data structures that
-reflect the underlying data and facilitate the manipulations that are
-of most interest, while the second is the importance of a modular
-architecture that allows for many developers to extend and use the
-underlying infrastructure and to combine tools in complex work flows.
-
-\subsection*{Existing data standards and conventions}
+Another critical component is the development and implementation of
+standards that will facilitate the adoption of these methods by both
+the larger research community and commercial software vendors.
 Traditionally, the majority of FCM experiments have been analyzed by
@@
 though preparation was standardized and performed using the same
-samples and reagents at each site \citep{Maecker2005}. For FC-HCS
-data, unassisted manual inspection is extremely time consuming, and
-robust statistical methods need to be developed to point investigators
-to interesting aspects of the data, or to potential problems. While
-the expert knowledge of immunologists and researchers remains crucial
-for the understanding of FCM data, we believe that collaboration with
+samples and reagents at each site \cite{Maecker2005}. For FC-HCS data,
+unassisted manual inspection is extremely time consuming, and robust
+statistical methods need to be developed to point investigators to
+interesting aspects of the data, or to potential problems. While the
+expert knowledge of immunologists and researchers remains crucial for
+the understanding of FCM data, we believe that collaboration with
 other research fields such as statistics and computer science can
@@
 
+In this paper, we describe a set of flexible and well structured
+computational tools to efficiently analyze FC-HCS data. This set of
+tools is based on the open source statistical software R \cite{Rmain}
+and distributed as the Bioconductor package \Rpackage{flowCore}
+\cite{BIOC}. Our intent is to provide a shared research platform that
+enables bioinformaticians, computer scientists, and statisticians to
+work collaboratively with biologists and clinicians to develop novel
+methods for FCM data analysis, a process deemed crucial by many for
+the further development of the technology \cite{lizard2007fca}.
+
+%% Perhaps in Conclusion
+%% Consequently, we are optimistic that this
+%% software platform will play a critical role in addressing the
+%% bottlenecks described above that limit advances in the application of
+%% FC-HCS.
+%%
+
+%%%%%%%%%%%%%%%%%%
+\section*{Implementation}
+%% The computational tools we have developed are distributed in the R
+%% software language as the Bioconductor package
+%% \Rpackage{flowCore}. 
+The Bioconductor package \Rpackage{flowCore} is a freely available,
+highly functional, and extensible FCM data analysis platform that
+enables researchers to efficiently handle FC-HCS data and encourages
+open development of tools for their coherent
+analysis. \Rpackage{flowCore} implements computationally efficient
+data structures and a range of specialized methods addressing all
+components of a typical FCM analysis work flow, including
+compensation, transformation, and gating. \Rpackage{flowCore} runs on
+Windows, Mac OS X, and Linux/Unix operating systems. Conceptually, the
+underlying data structures that build the core of our software can be
+implemented in any programming language. We chose the R language for
+its advantages in statistical programming and visualization. 
+%% In this paper, we describe the functionality and methods underlying
+%% \Rpackage{flowCore}, and provide examples of their use.
+
+\subsection*{Existing data standards and conventions}
 Currently, data from FCM experiments are stored in single files
-according to the Flow Cytometry Standard (FCS) \citep{seamer1997pnd}.
+according to the Flow Cytometry Standard (FCS) \cite{seamer1997pnd}.
 However, recent developments in high-throughput FCM are shifting the
@@
 of samples from different patients, measured at different timepoints
-\citep{brinkman2007hcf} or following different drug treatments
-\citep{gasparetto2004ice}. Modern FCM data analysis tools have to deal
+\cite{brinkman2007hcf} or following different drug treatments
+\cite{gasparetto2004ice}. Modern FCM data analysis tools have to deal
 with an additional layer of sample metadata and they need to provide
@@
 other FCM tool that adheres to the proposed standard
-\citep{SpidlenInPressCytometryA}. Adherence to standards also plays a
+\cite{SpidlenInPressCytometryA}. Adherence to standards also plays a
 critical role in the ability of new methods based on
@@
 
-%%%%%%%%%%%%%%%%%%
-\section*{Implementation}
- \subsection*{Basic data structures}
+\subsection*{flowCore basic data structures}
 \subsubsection*{flowFrame}
-
 \Rpackage{flowCore}'s primary task is the representation and basic
@@
 fostered the development of a number of valuable add-ons
-\citep{lo2008agf,sarkar2008ufv}.
+\cite{lo2008agf,sarkar2008ufv}.
 
@@
 Data transformation is essential for both data visualization and
-modelling \citep{lo2008agf}. The major transformations that are
+modelling \cite{lo2008agf}. The major transformations that are
 routinely used in FCM analysis have been implemented in
@@
   analysis of several datasets
-  \citep{gasparetto2004ice,brinkman2007hcf}. A complete documentation
+  \cite{gasparetto2004ice,brinkman2007hcf}. A complete documentation
   of the \Rpackage{flowCore} software is beyond the scope of this
@@
   a fast objective solution to the analysis of potentially very large
-  and diverse data sets \citep{lo2008agf}. One of the automated gating
+  and diverse data sets \cite{lo2008agf}. One of the automated gating
   methods implemented in \Rpackage{flowCore} is based on identifying
   areas of significant curvature in a kernel density estimate of the
-  data \citep{wand2008}. Assuming that the regions of interest are of
+  data \cite{wand2008}. Assuming that the regions of interest are of
   high density, the software is able to reliably detect them in a one
@@
 
+\subsection*{Workflow proposal(??)}
+
+Florian?
+
 \subsection*{Related flow packages}
@@
 packages that are dedicated to more specific tasks of FCM data
-analysis. The \Rpackage{flowViz} package \citep{sarkar2008ufv}
+analysis. The \Rpackage{flowViz} package \cite{sarkar2008ufv}
 provides sophisticated data visualization tools, that make use off
-multivariate trellis plotting \citep{lattice}.  These functions can be
+multivariate trellis plotting \cite{lattice}.  These functions can be
 used to quickly generate customized plots for extended cytometry data
@@
 \section*{Conclusions}
-  Text for this section \ldots
+Through \Rpackage{flowCore}, we have provided the FCM community with
+an open source, freely available, highly functional, standards
+compliant development and analysis platform for high throughput data
+analysis.  Our intention is to provide a platform for the
+collaborative development of new analysis methods that will facilitate
+the transition of these new methods to the larger flow community.  Our
+experience has been that the collaborative effort we enable for
+devising new methodology has been proven beneficial for a number of
+different biological and computational biology challenges.  We hope
+that our framework will be the foundation for fruitful shared research
+by many collaborators from multiple scientific fields and will help
+resolve bottlenecks that currently prevent further development and
+deployment of FC-HCS to increasingly complex and important scientific
+and clinical applications.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Availability and requirements}
+Project name: flowCore
+Project home page: http://bioconductor.org 
+Operating system(s): The flowCore package can be installed on all the 
+platforms supporting R. These include a wide variety of UNIX platforms, Windows 
+and MacOS.  
+Programming language: R 
+Licence: The Artistic License, Version 2.0
 
@@
 \section*{Authors contributions}
-    Text for this section \ldots
+The project was conceived by RB, RG, PH. It was designed, and
+developed by all the authors.  NLM, FH, DS, and BE wrote most of the
+software’s code. The manuscript was prepared by FH and NLM, then
+revised and approved by all authors.
 
@@
 \section*{Figures}
-  \subsection*{Figure 1 - Sample figure title}
-      A short description of the figure content
-      should go here.
+  \subsection*{Figure 1 - flowCore Framework}
+  \begin{figure}[hp]
+    \centering
+    \includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
+    \caption{\label{fig1:FrameWork}{For each experiment, the content of
+        the FCS files, phenotypic and metadata are stored in a
+        \Rclass{flowSet}. Each \Rclass{flowFrame} in a \Rclass{flowSet}
+        corresponds to one FCS file. All basic operations (e.g.,
+        compensation, transformation, gating) can be applied to either
+        single \Rclass{flowFrames} or a \Rclass{flowSet} simultaneously.}}
+  \end{figure}
 
-  \subsection*{Figure 2 - Sample figure title}
-      Figure legend text.
+  \subsection*{Figure 2 - Quality assessment}
+    \begin{figure}[hp]
+      \centering
+      \includegraphics[width=0.98\textwidth]{flowQ.jpg}
+      \caption{\label{flowQ}%
+        HTML quality assessment report generated by the flowQ package for a
+        subset of data from an experiment focussing on Graft-Versus-Host
+        Disease \citep{brinkman2007hcf}. Rows correspond to the samples in
+        the set, columns to different quality checks.}
+    \end{figure}
+
+    \subsection*{Figure 3 - Quality assessment}
+    \begin{figure}[hp]
+      \centering
+      \includegraphics[width=0.75\textwidth]{xyplot}
+      \caption{\label{xyplot}%
+        Scatterplot matrix of a single \Rclass{flowSet} from an experiement
+        focussing on immune tolerance following kidney
+        transplantation. Outlines of the gating regions identified by an
+        automated gating operation are added on top of the density
+        representation of the data.}
+    \end{figure}
 
@@
 \section*{Tables}
-  \subsection*{Table 1 - Sample table title}
-    Here is an example of a \emph{small} table in \LaTeX\ using  
-    \verb|\tabular{...}|. This is where the description of the table 
-    should go. \par \mbox{}
-    \par
-    \mbox{
-      \begin{tabular}{|c|c|c|}
-        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
-        A1 & B2  & C3 \\ \hline
-        A2 & ... & .. \\ \hline
-        A3 & ..  & .  \\ \hline
+  \subsection*{Table 1 - Data transformations}
+    \begin{table}[ht]
+      \caption{\label{table1} Within these formulas, $x$ is the variable
+          corresponding to value being transformed, $y$ is the result of the
+          transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
+          constants affecting the transformation function, $e$ is the base of
+          the natural logarithm, and $root$ is a root-finding algorithm.  }
+      \begin{center}
+        \begin{tabular}{|l|l|}
+          \hline
+          \multicolumn{2}{|c|}{Data Transformations} \\
+          \hline
+          linear & $ax + b$ \\
+          quadratic & $ax^2 + bx + c$ \\
+          natural logarithm & $ln(x)(r/d)$ \\
+          logarithm & $log(x)(r/d)$ \\
+          biexponential & $root(a*e^{(b*y)}-c*e^{(-d+y)}+f)$ \\
+          logicle& $root(Te^{-(m-w)}*(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
+          truncate & $x_{x<=a} = a$ \\
+          scale & $(x-a)/(b-a)$ \\
+          split-scale & combination of linear and logarithm \\
+          arcsinh & $asinh(a + bx)+c$ \\
+          \hline
         \end{tabular}
-      }
-  \subsection*{Table 2 - Sample table title}
-    Large tables are attached as separate files but should
-    still be described here.
+      \end{center}
+    \end{table}
+
+  \subsection*{Table 2 - Filter and gate classes}
+  \begin{table}[ht]
+    \begin{center}
+      \begin{tabular}{|l|l|}
+        \hline
+        \multicolumn{2}{|c|}{Gates} \\
+        \hline
+        rectangleGate & n-dimensional rectangular regions \\
+        quadGate & quadrant regions in two dimensions \\
+        polygonGate & polygonal regions in two dimensions \\
+        polytopeGate & generalization of polygon in n dimensions \\
+        ellipsoidGate & n-dimensional ellipsoid region \\
+        \hline
+        \multicolumn{2}{|c|}{Filters} \\
+        \hline
+        sampleFilter & random sub-sampling\\
+        expressionFilter & results of a boolean expression \\
+        kmeansFilter & K-means clustering \\
+        norm2Filter & bivariate normal distribution \\
+        curv1Filter & local density regions in 1D \\
+        curv2Filter & density regions in 2D \\
+        timeFilter & abnormal data acquisition over time \\
+        \hline
+        filterSet & gating strategies \\
+        \hline
+      \end{tabular}
+    \end{center}
+  \end{table}
 

### 2008-12-07 R/AllClasses.R
edits to the vignette
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35702 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1488,9 +1488,11 @@ compensation <- function(..., spillover, inv=FALSE,
        !all(sapply(parms$parameters, parameters) %in% colnames(spillover)))
         stop("Parameters and column names of the spillover matrix ",
              "don't match.", call.=FALSE)
-    if(inv)
+    if(inv){
       ## spillover <- solve(spillover/max(spillover))
       spillover <- solve(spillover)
+        colnames(spillover) <- rownames(spillover)
+    }
     new("compensation", spillover=spillover, 
         compensationId=compensationId,
         parameters=new("parameters", parms$parameters))

### 2008-12-12 paper/bmc_article.tex
some changes of the paper and addition of work flows
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35870 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 % Load packages
+%\usepackage[authoryear,round]{natbib}
 \usepackage{cite} % Make references as [1-4], not [1,2,3,4]
@@
 \usepackage[utf8]{inputenc} %unicode support
+\usepackage{graphicx}
+%\usepackage{setspace}
 %\usepackage[applemac]{inputenc} %applemac support if unicode package fails
@@
 %Review style settings
-\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
 
 %Publication style settings
-%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
 
@@
 enables researchers to efficiently handle FC-HCS data and encourages
-open development of tools for their coherent
-analysis. \Rpackage{flowCore} implements computationally efficient
-data structures and a range of specialized methods addressing all
-components of a typical FCM analysis work flow, including
-compensation, transformation, and gating. \Rpackage{flowCore} runs on
-Windows, Mac OS X, and Linux/Unix operating systems. Conceptually, the
-underlying data structures that build the core of our software can be
-implemented in any programming language. We chose the R language for
-its advantages in statistical programming and visualization. 
+open development of tools for their coherent analysis. In our
+implementation of \Rpackage{flowCore} we rely on two important lessons
+learned from the field of gene expression data analysis: the first
+being the importance of data structures that reflect the underlying
+data and facilitate the manipulations that are of most interest, while
+the second is the importance of a modular architecture that allows for
+many developers to extend and use the underlying infrastructure and to
+combine tools in complex work flows.  \Rpackage{flowCore} implements
+such computationally efficient data structures and a range of
+specialized methods addressing all components of a typical FCM
+analysis work flow, including compensation, transformation, and
+gating. \Rpackage{flowCore} runs on Windows, Mac OS X, and Linux/Unix
+operating systems. Conceptually, the underlying data structures that
+build the core of our software can be implemented in any programming
+language. We chose the R language for its advantages in statistical
+programming and visualization.
 %% In this paper, we describe the functionality and methods underlying
@@
 (CR), an emerging ISAC standard developed in collaboration with the
-ISAC Data Standards Task Force, which makes it possibile to integrate
+ISAC Data Standards Task Force, which makes it possible to integrate
 \Rpackage{flowCore} in existing work flows and to communicate with any
@@
 manipulation of FCM data. This is accomplished through a data model
-very close to that adopted by other Bioconductor packages. All
-information from a single FCS file, i.e., the collection of events and
-the accompanying metadata, is stored in one single container. We call
-the structure that hold this data a \Rclass{flowFrame}. Raw data
-values as well as associated metadata of a \Rclass{flowFrame} can be
-accessed programmatically. Most commonly, the metadata consist of
-descriptors of the stains used in the experiment and the respective
-measurement channels, information about compensation performed at the
-instrument side and any additional keywords the user deems to be
-important to annotate the data. During the creation of a
-\Rclass{flowFrame}, a number of quality checks are performed to ensure
-data integrity.
+very close to that adopted by other successful Bioconductor
+packages. All information from a single FCS file, i.e., the collection
+of events and the accompanying metadata, is stored in one single
+container. We call the structure that hold this data a
+\Rclass{flowFrame}. Raw data values as well as associated metadata of
+a \Rclass{flowFrame} can be accessed programmatically. Most commonly,
+the metadata consist of descriptors of the stains used in the
+experiment and the respective measurement channels, information about
+compensation performed at the instrument side and any additional
+keywords the user deems to be important to annotate the data. During
+the creation of a \Rclass{flowFrame}, a number of quality checks are
+performed to ensure data integrity.
 
@@
 was not already done on the instrument) and transformed, and
-sup-populations of interest need to be selected based on a set of
+sub-populations of interest need to be selected based on a set of
 (predominantly sequential) gates. All software solutions for FCM
@@
 gates. Complex combinations and hierarchies of gates can be captured
-in objects of class \Rclass{filterSet}, allowing to apply multi-step
-gating strategies. The definition of gates in \Rpackage{flowCore}
-follows the Gating-ML CR, thus any \Rpackage{flowCore} gating strategy
-can be reproduced by any other software that also adheres to the
-standard.
+in multi-step gating strategies. The definition of gates in
+\Rpackage{flowCore} follows the Gating-ML CR, thus any
+\Rpackage{flowCore} gating strategy can be reproduced by any other
+software that also adheres to the standard and vice versa.
 
@@
 \begin{verbatim}
-> library(flowCore)
 > library(flowQ)
-> qaReport(dat, c("qaProcess.timeline", "qaProcess.timeflow", 
+> qaReport(dat, c("qaProcess.timeline", 
+                  "qaProcess.timeflow", 
                   "qaProcess.cellnumber"))
@@
 framework. In this example, we directly use R code from the
-\Rpackage{feature} package developed by \cite{wand2008}. Instead of
+\Rpackage{feature} package \cite{wand2008}. Instead of
 re-writing existing code, we are able to include it via the well
@@
 implemented in \Rpackage{flowCore} is available to other package
-authors, as we have seen with the afore-mentioned \Rpackage{flowClust}
-package.
+authors.
 
-We can chose one of the many visualization option from the flowViz
-package to plot the results of the recent filtering operation. A very
-basic matrix of density plots is shown in Figure \ref{xyplot}, where
-each panel in the matrix represents the data of one individual
-patient.
+We can chose one of the many visualization option from the
+\Rpackage{flowViz} package to plot the results of the recent filtering
+operation. A very basic matrix of density plots is shown in Figure
+\ref{xyplot}, where each panel in the matrix represents the data of
+one individual patient.
 
 \begin{verbatim}
-> xyplot(`FL1-H` ~ `FL3-H` | SampleID, data=dat, filter=fres)
+> xyplot(`FL1-H` ~ `FL3-H` | SampleID, 
+         data=dat, filter=fres)
 \end{verbatim}
 
-\subsection*{Workflow proposal(??)}
-
-Florian?
+Scripting languages like R provide a natural representation of work
+flows through a sequence of code instructions in regular text
+files. This allows for the rapid development and testing of new ideas,
+however it is not very well suited for routine data analysis
+tasks. Furthermore, the overhead of data management and variable
+tracking can be considerable. To that end, \Rpackage{flowCore} also
+provides data structures that help organize sequences of typical FCM
+data analysis operations and complex gating strategies into concise
+work flows. These structures are self-reflective, they contain all
+intermediate results and offer a unified user interface to assess the
+progress and the outcome of an anlysis.
 
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
+    \includegraphics{Figure1-flowCoreFrameWork.jpg}
     \caption{\label{fig1:FrameWork}{For each experiment, the content of
@@
       \centering
-      \includegraphics[width=0.98\textwidth]{flowQ.jpg}
+      \includegraphics{flowQ.jpg}
       \caption{\label{flowQ}%
@@
         subset of data from an experiment focussing on Graft-Versus-Host
-        Disease \citep{brinkman2007hcf}. Rows correspond to the samples in
+        Disease \cite{brinkman2007hcf}. Rows correspond to the samples in
         the set, columns to different quality checks.}
@@
       \centering
-      \includegraphics[width=0.75\textwidth]{xyplot}
+      \includegraphics{xyplot.pdf}
       \caption{\label{xyplot}%
@@
   \begin{table}[ht]
+    \label{table2}
     \begin{center}

### 2008-12-12 paper/bmc_article.tex
last check for the bmc version of the paper
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35899 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 %Review style settings
-%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
 
 %Publication style settings
-\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
 
@@
   Division of Public Health Sciences, Fred Hutchinson Cancer Research Center,%
-  1100 Fairview Ave. N, M2-B876, PO Box 19024,  Seattle, Washington 98109-1024\\
+  1100 Fairview Ave. N, M2-B876, PO Box 19024,  Seattle, Washington 98109-1024, USA\\
   \iid(2)EA SeRAIC INSERM, IRISA - Symbiose, Campus Beaulieu, Universit\'{e} de Rennes I, % 
@@
   675 West 10th Avenue, Vancouver, BC V5Z 1L3  Canada, Canada\\
-  \iid(4)AdBrite Inc\\
-  \iid(5)BD Biosciences
+  \iid(4)AdBrite Inc., 731 Market St., 5th Floor, San Francisco, California 94103, USA\\
+  \iid(5)BD Biosciences, Research Triangle Park, North Carolina 27709, USA
 }%
@@
   (FH-HCS). However, data management and data analysis methods have
-  not advanced sufficiently far from the inital small-scale studies to
+  not advanced sufficiently far from the initial small-scale studies to
   support the large complex data sets generated in FH-HCS
@@
 
-  \paragraph*{Results:} We developped a set of flexible computational
+  \paragraph*{Results:} We developed a set of flexible computational
   tools for FC-HCS data analysis written in R and freely available
@@
 \section*{Implementation}
-%% The computational tools we have developed are distributed in the R
-%% software language as the Bioconductor package
-%% \Rpackage{flowCore}. 
+The computational tools we have developed are distributed in the R
+software language as the Bioconductor package \Rpackage{flowCore}.
 The Bioconductor package \Rpackage{flowCore} is a freely available,
@@
 influencing factors. For example, experiments consist of large numbers
-of samples from different patients, measured at different timepoints
+of samples from different patients, measured at different time points
 \cite{brinkman2007hcf} or following different drug treatments
@@
 process raw data FCS files along with their complete set of
-file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
+file-specific metadata (Figure~1\ref{fig1:FrameWork}).  Moreover, it is
 a software implementation of the Gating-ML Candidate Recommendation
@@
 routinely used in FCM analysis have been implemented in
-\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponantioal or arcsinh,
+\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponantial or arcsinh,
 see Table~\ref{table1} for a complete list). Furthermore, the design
@@
 in signal intensities, such as drifts in the instrumentation
-over time or sample variablity.
+over time or sample variability.
 
@@
 \Rpackage{flowCore} gating strategy can be reproduced by any other
-software that also adheres to the standard and vice versa.
+software that also adheres to the standard and \textit{vice versa}.
 
@@
 input \Rclass{flowFrame} or \Rclass{flowSet}, either by filtering out
-negative events or by splitting in multiple subpopulations. This
+negative events or by splitting in multiple sub-populations. This
 design allows to easily combine all of \Rpackage{flowCore}'s
@@
   imported as the \Rclass{flowSet} object ``dat'', the following
-  simple line of code produces the output shown in Figure~\ref{flowQ}:
+  simple line of code produces the output shown in Figure~2\ref{flowQ}:
 
@@
 \Rpackage{flowViz} package to plot the results of the recent filtering
-operation. A very basic matrix of density plots is shown in Figure
+operation. A very basic matrix of density plots is shown in Figure~3
 \ref{xyplot}, where each panel in the matrix represents the data of
@@
 intermediate results and offer a unified user interface to assess the
-progress and the outcome of an anlysis.
+progress and the outcome of an analysis.
 
@@
 package, \Rpackage{plateCore}, providing more specialized support for
-experiments conducted on microtitre plates and facilitiating the
+experiments conducted on microtitre plates and facilitating the
 handling of spatial metadata, is under development.
@@
   \subsection*{Figure 1 - flowCore Framework}
-  \begin{figure}[hp]
-    \centering
-    \includegraphics{Figure1-flowCoreFrameWork.jpg}
-    \caption{\label{fig1:FrameWork}{For each experiment, the content of
+ \label{fig1:FrameWork}For each experiment, the content of
         the FCS files, phenotypic and metadata are stored in a
@@
         compensation, transformation, gating) can be applied to either
-        single \Rclass{flowFrames} or a \Rclass{flowSet} simultaneously.}}
-  \end{figure}
+        single \Rclass{flowFrames} or a \Rclass{flowSet} simultaneously.
 
   \subsection*{Figure 2 - Quality assessment}
-    \begin{figure}[hp]
-      \centering
-      \includegraphics{flowQ.jpg}
-      \caption{\label{flowQ}%
+    \label{flowQ}%
         HTML quality assessment report generated by the flowQ package for a
@@
         Disease \cite{brinkman2007hcf}. Rows correspond to the samples in
-        the set, columns to different quality checks.}
-    \end{figure}
+        the set, columns to different quality checks.
 
     \subsection*{Figure 3 - Quality assessment}
-    \begin{figure}[hp]
-      \centering
-      \includegraphics{xyplot.pdf}
-      \caption{\label{xyplot}%
+    \label{xyplot}{
         Scatterplot matrix of a single \Rclass{flowSet} from an experiement
@@
         representation of the data.}
-    \end{figure}
 
@@
     \begin{table}[ht]
-      \caption{\label{table1} Within these formulas, $x$ is the variable
+      \caption{\label{table1} Data transformations implemented in \Rpackage{flowCore}. Within these formulas, $x$ is the variable
         corresponding to value being transformed, $y$ is the result of the
@@
         constants affecting the transformation function, $e$ is the base of
-          the natural logarithm, and $root$ is a root-finding algorithm.  }
+        the natural logarithm, and $root$ is a root-finding algorithm. Other transformations can easily be implemented in R.}
       \begin{center}
@@
   \begin{table}[ht]
-    \label{table2}
+    \caption{\label{table2} Filter and gate classes implemented in
+      \Rpackage{flowCore}.} 
     \begin{center}
@@
 
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
-
+%% \section*{Additional Files}
+%%   \subsection*{Additional file 1 --- Sample additional file title}
+%%     Additional file descriptions text (including details of how to
+%%     view the file, if it is in a non-standard format or the file extension).  This might
+%%     refer to a multi-page table or a figure.
+%% 
+%%   \subsection*{Additional file 2 --- Sample additional file title}
+%%     Additional file descriptions text.
+%% 
+%% 
 \end{bmcformat}

### 2008-12-16 paper/bmc_article.tex
more changes
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35947 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
   have enabled the use of flow-cytometry high content screening
-  (FH-HCS). However, data management and data analysis methods have
-  not advanced sufficiently far from the initial small-scale studies to
-  support the large complex data sets generated in FH-HCS
-  applications.
+  (FH-HCS), generating large complex data sets often in clinical
+  trials or drug discovery settings. However, data management and data
+  analysis methods have not advanced sufficiently far from the initial
+  small-scale studies to support modeling in the presence of multiple
+  covariates.
 
-  \paragraph*{Results:} We developed a set of flexible computational
-  tools for FC-HCS data analysis written in R and freely available
-  from the Bioconductor Project. The \Rpackage{flowCore} package
-  provides the flow cytometry community with a shared research
-  platform that enables bioinformaticians, computer scientists, and
-  statisticians to work collaboratively with biologists and clinicians
-  to develop novel methods for flow cytometry data analysis.
+  \paragraph*{Results:} We developed a set of flexible open source
+  computational tools (in the R package \Rpackage{flowCore}) to
+  facilitate the analysis of these complex data. A key component of
+  which is having suitable data structures that support the
+  application of similar operations to a collection of samples or a
+  clinical cohort. In addition, our software constitutes a shared and
+  extensible research platform that enables colaboration between
+  bioinformaticians, computer scientists, statisticians, biologists
+  and clinicians. This platform will foster the developement of novel
+  analytic methods for flow cytometry.
 
-  \paragraph*{Conclusions:} We hope that our framework will be the
-  foundation for fruitful collaborative research involving multiple
-  scientific fields and that it will help facilitate the development
-  and deployment of FC-HCS methods to complex scientific and clinical
-  applications.
+  \paragraph*{Conclusions:} 
+  The software has been applied in the analysis of various data sets
+  and its data structures have been proven to be highly efficient in
+  capturing and organizing the analytic work flow. A number of
+  additional Bioconductor packages successfully build on the
+  infrastructure provided by \Rpackage{flowCore}, opening new avenues
+  for flow data analysis.
+
 \end{abstract}
@@
 enabled the use of flow cytometry high content screening (FC-HCS) to
-generate large, complex datasets in both basic and clinical research
+generate large, complex data sets in both basic and clinical research
 applications \cite{brinkman2007hcf}. A serious bottleneck in the
@@
 developed for applications of flow cytometry (FCM) to small-scale,
-tube-based studies \cite{mahnke2007omi}.
+tube-based studies \cite{mahnke2007omi}. In particular, the data often
+need to be organized into groups of samples based on combinations of
+additional covariates and similar operations need to be applied to
+these groups in a transparent and reproducible manner. Furthermore,
+the growing depth of knowledge in the field of immunology, for
+instance the charactarization of distinct human T-cell subpopulation
+\cite{Appay}, clearly argues for more systematic approaches.
 %% add one of the last roederer's paper that "request" algo and
@@
 
-Some of the consequences of this lag are difficulties in maintaining
-the integrity and documentation of extremely large datasets, assessing
-measurement quality, developing validated assays, controlling the
-accuracy of gating techniques, automating complex gating strategies,
-and aggregating statistical results across large study sets for
-further analysis. The most serious consequence, however, is that new
-analysis approaches face difficulty in finding their way into standard
-practice. We believe that this barrier to the development and
+Some of the consequences of the lag of efficient software solutions
+are difficulties in maintaining the integrity and documentation of
+large data sets, assessing measurement quality, developing validated
+assays, controlling the accuracy of gating techniques, automating
+complex gating strategies, and aggregating statistical results across
+large study sets for further analysis. In addition, new analysis
+approaches face difficulty in finding their way into standard
+practice. We believe that these barriers to the development and
 dissemination of new analysis methods is one of the fundamental
@@
 
-Another critical component is the development and implementation of
-standards that will facilitate the adoption of these methods by both
-the larger research community and commercial software vendors.
 Traditionally, the majority of FCM experiments have been analyzed by
@@
 comparisons of summary statistics. Most of the currently available
-analysis tools are designed to reflect this workflow.  We believe that
-these approaches, in addition to being expensive and labor intensive,
-do not fully address the highly complex nature of FCM data; in
-particular, they disregard many of the fundamental aspects of the
-data, such as its underlying distribution and its high-dimensional
-nature. Furthermore, the subjective character of manual analyses are a
-major obstacle to reproducibility. In a recent study of flow
-cytometric standardization involving 15 institutions, the mean
-inter-laboratory coefficient of variation ranged from 17 to 44\%, even
-though preparation was standardized and performed using the same
-samples and reagents at each site \cite{Maecker2005}. For FC-HCS data,
-unassisted manual inspection is extremely time consuming, and robust
-statistical methods need to be developed to point investigators to
-interesting aspects of the data, or to potential problems. While the
-expert knowledge of immunologists and researchers remains crucial for
-the understanding of FCM data, we believe that collaboration with
+analysis tools are designed to reflect this work flow.  We believe
+that these approaches, in addition to being expensive and labor
+intensive, do not fully address the highly complex nature of FCM data;
+in particular, they disregard many of the fundamental aspects of the
+data, such as sample groups or cohorts, the underlying distribution or
+its high-dimensional nature. Furthermore, the subjective character of
+manual analyses are a major obstacle to reproducibility. In a recent
+study of flow cytometric standardization involving 15 institutions,
+the mean inter-laboratory coefficient of variation ranged from 17 to
+44\%, even though preparation was standardized and performed using the
+same samples and reagents at each site \cite{Maecker2005}. For FC-HCS
+data, unassisted manual inspection is extremely time consuming, and
+robust statistical methods need to be developed to point investigators
+to interesting aspects of the data, or to potential problems. While
+the expert knowledge of immunologists and researchers remains crucial
+for the understanding of FCM data, we believe that collaboration with
 other research fields such as statistics and computer science can
@@
 In this paper, we describe a set of flexible and well structured
-computational tools to efficiently analyze FC-HCS data. This set of
-tools is based on the open source statistical software R \cite{Rmain}
-and distributed as the Bioconductor package \Rpackage{flowCore}
-\cite{BIOC}. Our intent is to provide a shared research platform that
-enables bioinformaticians, computer scientists, and statisticians to
-work collaboratively with biologists and clinicians to develop novel
-methods for FCM data analysis, a process deemed crucial by many for
-the further development of the technology \cite{lizard2007fca}.
+computational tools to efficiently analyze FC-HCS data. Our intent is
+to provide a shared research platform that enables bioinformaticians,
+computer scientists, and statisticians to work collaboratively with
+biologists and clinicians to develop novel methods for FCM data
+analysis, a process deemed crucial by many for the further development
+of the technology \cite{lizard2007fca}.
 
@@
 The computational tools we have developed are distributed in the R
-software language as the Bioconductor package \Rpackage{flowCore}.
-The Bioconductor package \Rpackage{flowCore} is a freely available,
-highly functional, and extensible FCM data analysis platform that
-enables researchers to efficiently handle FC-HCS data and encourages
-open development of tools for their coherent analysis. In our
-implementation of \Rpackage{flowCore} we rely on two important lessons
-learned from the field of gene expression data analysis: the first
-being the importance of data structures that reflect the underlying
-data and facilitate the manipulations that are of most interest, while
-the second is the importance of a modular architecture that allows for
-many developers to extend and use the underlying infrastructure and to
-combine tools in complex work flows.  \Rpackage{flowCore} implements
-such computationally efficient data structures and a range of
-specialized methods addressing all components of a typical FCM
-analysis work flow, including compensation, transformation, and
-gating. \Rpackage{flowCore} runs on Windows, Mac OS X, and Linux/Unix
-operating systems. Conceptually, the underlying data structures that
-build the core of our software can be implemented in any programming
-language. We chose the R language for its advantages in statistical
-programming and visualization.
+software language \cite{Rmain} as the Bioconductor \cite{BIOC} package
+\Rpackage{flowCore}.  The package \Rpackage{flowCore} is a freely
+available, highly functional, and extensible FCM data analysis
+platform that enables researchers to efficiently handle FC-HCS data
+and encourages open development of tools for their coherent
+analysis. In our implementation of \Rpackage{flowCore} we rely on two
+important lessons learned from the field of gene expression data
+analysis: the first being the importance of data structures that
+reflect the underlying data and facilitate the manipulations that are
+of most interest, while the second is the importance of a modular
+architecture that allows for many developers to extend and use the
+underlying infrastructure and to combine tools in complex work flows.
+\Rpackage{flowCore} implements such computationally efficient data
+structures and a range of specialized methods addressing all
+components of a typical FCM analysis work flow, including
+compensation, transformation, and gating. \Rpackage{flowCore} runs on
+Windows, Mac OS X, and Linux/Unix operating systems. 
 %% In this paper, we describe the functionality and methods underlying
@@
 
-\subsection*{flowCore basic data structures}
+\subsection*{Basic Data Structures}
 \subsubsection*{flowFrame}
@@
 treatment to which the sample was subjected, or the location of that
-sample in a 96 or 384 well plate.  \Rclass{flowSets} manage the
-consistent application of operations on the individual
-\Rclass{flowFrames} and shift the burden of keeping score of the
-metadata from the user to the infrastructure, thus reducing the risk
-of errors (\textit{e.g.}, mixups of sample labels). The
-\Rclass{flowSet} structure can be readily extended to incorporate the
-potentially complex metadata associated with even larger FC-HCS
+sample in a microtitre plate.  \Rclass{flowSets} manage the consistent
+application of operations on the individual \Rclass{flowFrames} and
+shift the burden of keeping score of the metadata from the user to the
+infrastructure, thus reducing the risk of errors (\textit{e.g.},
+mixups of sample labels). Crucial operations like taking subsets, data
+transformations and gating or computation of summary statistics are
+greatly facilitated and all relevant annotation information is
+constantly passed on along the analysis pipeline.
+
+The \Rclass{flowSet} structure can be readily extended to incorporate
+the potentially complex metadata associated with even larger FC-HCS
 experiments such as clinical trials, where hundreds of patients might
 provide samples at different time points over the course of the
-experiment.
+experiment. The \Rclass{flowSet} data structure is one of the key
+features in the \Rpackage{flowCore} package and it is fundamental to
+the implementation of many of the high level functionalities such as
+quality assessment and control, visualization and automateds gating.
 
 \subsection*{Standard flow operations}
-
 Typically, the basic operations in FCM analyses adhere to the
@@
 \subsubsection*{Transformation and compensation}
-
 Data transformation is essential for both data visualization and
@@
 The \Rpackage{flowCore} package has been successfully applied in the
-  analysis of several datasets
-  \cite{gasparetto2004ice,brinkman2007hcf}. A complete documentation
-  of the \Rpackage{flowCore} software is beyond the scope of this
-  publication.  Much more comprehensive documentation and users guide
-  information with programmatic examples is available online
+analysis of several data sets, both originating from clinical trials
+\cite{brinkman2007hcf} and drug discovery experiments
+\cite{gasparetto2004ice}. A complete description of the
+\Rpackage{flowCore} software is beyond the scope of this publication.
+Much more comprehensive documentation and users guide information with
+programmatic examples is available online
 (\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
@@
 concept of data-driven automated gating, the integration of existing
-  software into the framework, and the generation of
-  publication-quality graphics for data visualization.
+software into the framework, and the generation of publication-quality
+graphics for data visualization.
 
 Data analysis for most experiments usually begins with a quality
-  assurance step, and we can use functionality from the
-  \Rpackage{flowQ} package to create an HTML report that highlights
-  potential quality issues. Assuming that the data has already been
-  imported as the \Rclass{flowSet} object ``dat'', the following
-  simple line of code produces the output shown in Figure~2\ref{flowQ}:
+assurance step, and we can use functionality from the \Rpackage{flowQ}
+package to create an HTML report that highlights potential quality
+issues. Assuming that the data has already been imported as the
+\Rclass{flowSet} object ``dat'', the following simple line of code
+produces the output shown in Figure~2\ref{flowQ}:
 
@@
 The report is interactive and provides drill-down to more detailed
-  aspects of the analysis, starting from a concise overview. The
-  design of \Rpackage{flowCore}'s data model allows for a coherent
-  treatment of all the samples, hence we are able to compare features
-  between individuals, or between groups of individuals, based on the
-  available metadata information.
+aspects of the analysis, starting from a concise overview. The design
+of \Rpackage{flowCore}'s data model allows for a coherent treatment of
+all the samples, hence we are able to compare features between
+individuals, or between groups of individuals, based on the available
+metadata information.
 
@@
 different treatments, over time or between different experiment
-  batches. Automated or data-driven gating has the potential to
-  estimate the gating regions from the underlying data, thus providing
-  a fast objective solution to the analysis of potentially very large
-  and diverse data sets \cite{lo2008agf}. One of the automated gating
+batches. Automated or data-driven gating has the potential to estimate
+the gating regions from the underlying data, thus providing a fast
+objective solution to the analysis of potentially very large and
+diverse data sets \cite{lo2008agf}. One of the automated gating
 methods implemented in \Rpackage{flowCore} is based on identifying
@@
 data \cite{wand2008}. Assuming that the regions of interest are of
-  high density, the software is able to reliably detect them in a one
-  or two-dimensional density landscape.
+high density, the software is able to reliably detect them in a one or
+two-dimensional density landscape.
 
@@
 \subsection*{Related flow packages}
-
 In addition to the \Rpackage{flowCore} package that offers basic
@@
 compliant development and analysis platform for high throughput data
-analysis.  Our intention is to provide a platform for the
-collaborative development of new analysis methods that will facilitate
-the transition of these new methods to the larger flow community.  Our
-experience has been that the collaborative effort we enable for
-devising new methodology has been proven beneficial for a number of
-different biological and computational biology challenges.  We hope
-that our framework will be the foundation for fruitful shared research
-by many collaborators from multiple scientific fields and will help
-resolve bottlenecks that currently prevent further development and
-deployment of FC-HCS to increasingly complex and important scientific
-and clinical applications.
+analysis.  We hope to foster collaborative development of new analysis
+methods and to facilitate the transition of these new methods to the
+larger flow community.  Our experience has been that such
+collaborative effort has proven beneficial for a number of different
+biological and computational biology challenges, greatly elevating
+their applicability.  We hope that our framework will be the
+foundation for fruitful shared research by many collaborators from
+multiple scientific fields and will help resolve bottlenecks that
+currently prevent further development and deployment of FC-HCS to
+increasingly complex and important scientific and clinical
+applications.
 

### 2008-12-16 paper/bmc_article.tex
more changes
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35948 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 set of tools to perform them on both \Rclass{flowFrames} and
-\Rclass{flowSets}. While transformation, and to a certain extent
-compensation, are fairly routine operations with only limited
-potential for improvement, being able to implement new methodologies
-for gating of FCM data, and extending the capabilities of
-\Rpackage{flowCore} through object oriented programming are features
-that clearly sets our framework apart from other FCM analysis
-tools. By factoring out as much of the bookkeeping as possible,
-programmers can focus on the actual operations rather then having to
-deal with the tedious details of data integration and
-access. Third-party methods can act on their own as first-class
-citizens in the analysis framework, without breaking the work flow or
-the basic infrastructure. This design allows for the straightforward
-extension of \Rpackage{flowCore}'s capabilities, and has already
-fostered the development of a number of valuable add-ons
+\Rclass{flowSets}. Typically, the results of these operations are
+again \Rclass{flowFrame} or \Rclass{flwoSet} objects. While
+transformation, and to a certain extent compensation, are fairly
+routine operations with only limited potential for improvement, being
+able to implement new methodologies for gating of FCM data, and
+extending the capabilities of \Rpackage{flowCore} through object
+oriented programming are features that clearly sets our framework
+apart from other FCM analysis tools. By factoring out as much of the
+bookkeeping as possible, programmers can focus on the actual
+operations rather then having to deal with the tedious details of data
+integration and access. Third-party methods can act on their own as
+first-class citizens in the analysis framework, without breaking the
+work flow or the basic infrastructure. This design allows for the
+straightforward extension of \Rpackage{flowCore}'s capabilities, and
+has already fostered the development of a number of valuable add-ons
 \cite{lo2008agf,sarkar2008ufv}.

### 2008-12-17 paper/bmc_article.tex
more changes
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35949 bc3139a8-67e5-0310-9ffc-ced21a209358
add elements in the papaer, especially in the Results and Discussion section
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@36080 bc3139a8-67e5-0310-9ffc-ced21a209358

--- 2008-12-16 paper/bmc_article.tex
@@
   clinical cohort. In addition, our software constitutes a shared and
-  extensible research platform that enables colaboration between
+  extensible research platform that enables collaboration between
   bioinformaticians, computer scientists, statisticians, biologists
-  and clinicians. This platform will foster the developement of novel
+  and clinicians. This platform will foster the development of novel
   analytic methods for flow cytometry.
@@
   The software has been applied in the analysis of various data sets
-  and its data structures have been proven to be highly efficient in
-  capturing and organizing the analytic work flow. A number of
-  additional Bioconductor packages successfully build on the
-  infrastructure provided by \Rpackage{flowCore}, opening new avenues
-  for flow data analysis.
+  and its data structures have proven to be highly efficient in
+  capturing and organizing the analytic work flow. Finally, a number
+  of additional Bioconductor packages successfully build on the
+  infrastructure provided by \Rpackage{flowCore}, open new avenues for
+  flow data analysis.
 
@@
 Automation technologies developed during the last several years have
-enabled the use of flow cytometry high content screening (FC-HCS) to
-generate large, complex data sets in both basic and clinical research
-applications \cite{brinkman2007hcf}. A serious bottleneck in the
-interpretation of existing studies and the application of FC-HCS to
-even larger, more complex problems is that data management and data
-analysis methods have not advanced sufficiently far from the methods
-developed for applications of flow cytometry (FCM) to small-scale,
+enabled the use of flow cytometry (FCM) to generate large, complex
+data sets in both basic and clinical research applications
+\cite{brinkman2007hcf}. A serious bottleneck in the interpretation of
+existing studies and the application of flow cytometry high content
+screening (FC-HCS) to even larger, more complex problems is that data
+management and data analysis methods have not advanced sufficiently
+far from the methods developed for applications of FCM
+ to small-scale,
 tube-based studies \cite{mahnke2007omi}. In particular, the data often
@@
 the growing depth of knowledge in the field of immunology, for
-instance the charactarization of distinct human T-cell subpopulation
+instance the characterization of distinct human T-cell sub-population
 \cite{Appay}, clearly argues for more systematic approaches.
@@
 
-Traditionally, the majority of FCM experiments have been analyzed by
-manual data inspection in one or two dimensions, or by very basic
+Traditionally, for the majority, FCM experiments were being analyzed
+by manual data inspection in one or two dimensions, or by very basic
 comparisons of summary statistics. Most of the currently available
@@
 
-%% Perhaps in Conclusion
-%% Consequently, we are optimistic that this
-%% software platform will play a critical role in addressing the
-%% bottlenecks described above that limit advances in the application of
-%% FC-HCS.
-%%
-
 %%%%%%%%%%%%%%%%%%
@@
 Windows, Mac OS X, and Linux/Unix operating systems.
-%% In this paper, we describe the functionality and methods underlying
-%% \Rpackage{flowCore}, and provide examples of their use.
 
@@
 The notion of classes from an object-oriented programming language
-provides one coherent way to describe these richer data structures,
-and functions or methods that work on them allow for interaction and
-manipulation. Many of the currently available software solutions offer
-only limited support for such self-contained structures, or make use
-of containers like XML or binary storage formats that are designed
-specifically for particular user interfaces and hence not easily
-amenable to programmatical access. In addition, the closed-source
-nature of these products often makes them impractical to integrate
-into analysis pipelines. In this manuscript we describe classes for
-FCM data analysis and their implementation in R, however, they could
-just as easily be implemented in any other language (e.g., Java,
-C++). Software written in those languages could use similar data
-structures, thereby simplifying communication and the interchange of
-data between analysis tools.
+provides one coherent way to describe these richer data structures.
+In addition, functions or methods that work on those classes allow for
+interaction and manipulation. Many of the currently available software
+solutions offer only limited support for such self-contained
+structures, or make use of containers like XML or binary storage
+formats that are designed specifically for particular user interfaces
+and hence not easily amenable to programmatical access. In addition,
+the closed-source nature of these products often makes them
+impractical to integrate into analysis pipelines. In this manuscript
+we describe classes for FCM data analysis and their implementation in
+R, however, they could just as easily be implemented in any other
+language (e.g., Java, C++). Software written in those languages could
+use similar data structures, thereby simplifying communication and the
+interchange of data between analysis tools.
+
+%% I think switching the last 2 paragraphs make more sense as the 2nd
+%% one of this section talk about other closed-source soft etc.
+%% But up to you
+\Rpackage{flowCore} does not provide a graphical user interface
+(\textit{i.e.}, is not a GUI-driven software) and all operations are
+done using a command line interface.  It is possible to add a more
+elaborate user interface on top of this infrastructure, however the
+focus in this paper is on a programmatic approach to enable the
+convenient development of novel analysis methods and automation of
+complex analysis approaches.  By taking the burden of data management
+from the programmer, and by providing well-defined application
+programming interfaces (APIs), it is possible to readily test new
+ideas and to easily extend the framework's functionality.
 
@@
 
-\Rpackage{flowCore} is not a GUI-driven software, and all operations
-are done using a command line interface.  It is possible to add a more
-elaborate user interface on top of this infrastructure, however the
-focus in this paper is on a programmatic approach to enable the
-convenient development of novel analysis methods and automation of
-complex analysis approaches.  By taking the burden of data management
-from the programmer, and by providing well-defined programming
-interfaces (APIs), it is possible to readily test new ideas and to
-easily extend the framework's functionality.
-
 \subsection*{Basic Data Structures}
-\subsubsection*{flowFrame}
+\subsubsection*{flowFrame: sample unit}
 \Rpackage{flowCore}'s primary task is the representation and basic
@@
 very close to that adopted by other successful Bioconductor
-packages. All information from a single FCS file, i.e., the collection
-of events and the accompanying metadata, is stored in one single
-container. We call the structure that hold this data a
-\Rclass{flowFrame}. Raw data values as well as associated metadata of
-a \Rclass{flowFrame} can be accessed programmatically. Most commonly,
-the metadata consist of descriptors of the stains used in the
-experiment and the respective measurement channels, information about
-compensation performed at the instrument side and any additional
-keywords the user deems to be important to annotate the data. During
-the creation of a \Rclass{flowFrame}, a number of quality checks are
-performed to ensure data integrity.
+packages. All information from a single FCS file, \textit{i.e.}, the
+collection of events and the accompanying metadata, is stored in one
+single container. We call the structure that hold this data a
+\Rclass{flowFrame} (Figure~1\ref{fig1:FrameWork}). Raw data values as
+well as associated metadata of a \Rclass{flowFrame} can be accessed
+programmatically. Most commonly, the metadata consist of descriptors
+of the stains used in the experiment and the respective measurement
+channels, information about compensation performed at the instrument
+side and any additional keywords the user deems to be important to
+annotate the data. During the creation of a \Rclass{flowFrame}, a
+number of quality checks are performed to ensure data integrity.
 
-\subsubsection*{flowSet}
+\subsubsection*{flowSet: a collection of  \Rclass{flowFrames}}
 In high-throughput FCM, many of the analysis tasks need to be
@@
 concept of a collection of \Rclass{flowFrames} called a
-\Rclass{flowSet}. A \Rclass{flowSet} is a container for multiple
-\Rclass{flowFrames} along with relevant information associated with
-each individual frame such as descriptions of the cell sample, the
-treatment to which the sample was subjected, or the location of that
-sample in a microtitre plate.  \Rclass{flowSets} manage the consistent
-application of operations on the individual \Rclass{flowFrames} and
-shift the burden of keeping score of the metadata from the user to the
-infrastructure, thus reducing the risk of errors (\textit{e.g.},
-mixups of sample labels). Crucial operations like taking subsets, data
-transformations and gating or computation of summary statistics are
-greatly facilitated and all relevant annotation information is
-constantly passed on along the analysis pipeline.
+\Rclass{flowSet} (Figure~1\ref{fig1:FrameWork}). A \Rclass{flowSet} is
+a container for multiple \Rclass{flowFrames} along with relevant
+information associated with each individual frame such as descriptions
+of the cell sample, the treatment to which the sample was subjected,
+or the location of that sample in a microtitre plate.  The objects are
+self-contained and can be shipped to other computers, platform
+independently. \Rclass{flowSets} manage the consistent application of
+operations on the individual \Rclass{flowFrames} and shift the burden
+of keeping score of the metadata from the user to the infrastructure,
+thus reducing the risk of errors (\textit{e.g.}, mixups of sample
+labels). Crucial operations like taking subsets, data transformations
+and gating, or computation of summary statistics are greatly
+facilitated and all relevant annotation information is constantly
+passed on along the analysis pipeline.
 
@@
 the implementation of many of the high level functionalities such as
-quality assessment and control, visualization and automateds gating.
+quality assessment and control, visualization and automated gating.
 
@@
 (predominantly sequential) gates. All software solutions for FCM
-analysis offer support for these operations, typically in an
+analysis offer support for these operations, most often in an
 interactive, GUI-driven interface. In \Rpackage{flowCore} we have
@@
 set of tools to perform them on both \Rclass{flowFrames} and
-\Rclass{flowSets}. While transformation, and to a certain extent
-compensation, are fairly routine operations with only limited
-potential for improvement, being able to implement new methodologies
-for gating of FCM data, and extending the capabilities of
-\Rpackage{flowCore} through object oriented programming are features
-that clearly sets our framework apart from other FCM analysis
-tools. By factoring out as much of the bookkeeping as possible,
-programmers can focus on the actual operations rather then having to
-deal with the tedious details of data integration and
-access. Third-party methods can act on their own as first-class
-citizens in the analysis framework, without breaking the work flow or
-the basic infrastructure. This design allows for the straightforward
-extension of \Rpackage{flowCore}'s capabilities, and has already
-fostered the development of a number of valuable add-ons
+\Rclass{flowSets}. Typically, the results of these operations are
+again \Rclass{flowFrame} or \Rclass{flowSet} objects. While
+transformation, and to a certain extent compensation, are fairly
+routine operations with only limited potential for improvement, being
+able to implement new methodologies for gating of FCM data, and
+extend the capabilities of \Rpackage{flowCore} through object
+oriented programming are features that clearly sets our framework
+apart from other FCM analysis tools. By factoring out as much of the
+bookkeeping as possible, programmers can focus on the actual
+operations rather then having to deal with the tedious details of data
+integration and access. Third-party methods can act on their own as
+first-class citizens in the analysis framework, without breaking the
+work flow or the basic infrastructure. This design allows for the
+straightforward extension of \Rpackage{flowCore}'s capabilities, and
+has already fostered the development of a number of valuable add-ons
 \cite{lo2008agf,sarkar2008ufv}.
@@
 routinely used in FCM analysis have been implemented in
-\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponantial or arcsinh,
+\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponential or arcsinh,
 see Table~\ref{table1} for a complete list). Furthermore, the design
@@
 to the data of individual \Rclass{flowFrames} or entire
-\Rclass{flowSets}, respectively. Compensation is available for both
-\Rclass{flowFrames} and \Rclass{flowSets}, and the software also
-offers functionality to compute spillover or compensation matrices
-from a set of appropriate single stain controls.
+\Rclass{flowSets}, respectively. Compensation, that corrects for
+fluorescence spillover originating from the inherent overlap of
+emission spectra from antibody fluorescent labels, is available for
+both \Rclass{flowFrames} and \Rclass{flowSets}. In addition, the
+software offers functionality to compute spillover or compensation
+matrices from a set of appropriate single stain controls.
 
@@
 The ability to programmatically access gates is a prerequisite for
-automated or semi-automated gating. By utilizing a unified interface
+semi-automated or automated gating. By utilizing an unified interface
 for all different types of gates, the user is able to subset data sets
-as well as to create summary statistics, for example the proportions
-of events falling in a single gate or in a combination of
+as well as to create summary statistics, as for instance the
+proportions of events falling in a single gate or in a combination of
 gates. Complex combinations and hierarchies of gates can be captured
@@
 \cite{brinkman2007hcf} and drug discovery experiments
-\cite{gasparetto2004ice}. A complete description of the
-\Rpackage{flowCore} software is beyond the scope of this publication.
-Much more comprehensive documentation and users guide information with
-programmatic examples is available online
-(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
-and as part of the package distribution. Here, we want to briefly
+\cite{gasparetto2004ice}. A complete description of the methods
+implemented in \Rpackage{flowCore} is beyond the scope of this
+publication.  Much more comprehensive documentation and users guide
+information with programmatic examples are available online
+(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html}),
+as part of the package distribution. Here, we want to briefly
 exemplify some of the software's key features, that is, the coherent
@@
 Data analysis for most experiments usually begins with a quality
-assurance step, and we can use functionality from the \Rpackage{flowQ}
-package to create an HTML report that highlights potential quality
+assurance step. In a FCS analysis work flow, we can use functionality
+from the \Rpackage{flowQ} package, build upon \Rpackage{flowCore}, to
+create an HTML report that highlights potential quality
 issues. Assuming that the data has already been imported as the
-\Rclass{flowSet} object ``dat'', the following simple line of code
-produces the output shown in Figure~2\ref{flowQ}:
+\Rclass{flowSet} object ``dat'', using for instance the
+\Rclass{read.flowSet} method, the following simple lines of code
+produce the output shown in Figure~2\ref{flowQ}:
 
@@
 
-Static gating for all samples in a high-throughput FCM experiment is
-often impossible, since the measured variables tend to vary between
-different treatments, over time or between different experiment
-batches. Automated or data-driven gating has the potential to estimate
-the gating regions from the underlying data, thus providing a fast
-objective solution to the analysis of potentially very large and
-diverse data sets \cite{lo2008agf}. One of the automated gating
-methods implemented in \Rpackage{flowCore} is based on identifying
-areas of significant curvature in a kernel density estimate of the
-data \cite{wand2008}. Assuming that the regions of interest are of
-high density, the software is able to reliably detect them in a one or
+According quality of the measurements, the next steps of a FCS
+analysis work flow are potentially the compensation and transformation
+of the data. Once again \Rpackage{flowCore}'s data structure and its
+methods allow a flexible processing. One can use the basic
+transformation and compensation functions implemented in
+\Rpackage{flowCore} (Table~1) or develop its own approaches that could
+then be apply to \Rclass{flowFrame} or \Rclass{flowSet}.
+
+Ultimately, a flow cytometry experiment aims at identifying and
+characterizing cell population of biological interest, using static
+gating or data-driven procedures (Table~2). Static gating for all
+samples in a high-throughput FCM experiment is often impossible, since
+the measured variables tend to vary between different treatments, over
+time or between different experiment batches. Automated or data-driven
+gating has the potential to estimate the gating regions from the
+underlying data, thus providing a fast objective solution to the
+analysis of potentially very large and diverse data sets
+\cite{lo2008agf}. One of the automated gating methods implemented in
+\Rpackage{flowCore} is based on identifying areas of significant
+curvature in a kernel density estimate of the data
+\cite{wand2008}. Assuming that the regions of interest are of high
+density, the software is able to reliably detect them in a one- or
 two-dimensional density landscape.
@@
 tested distribution mechanism provided by R's software package
-system. This process is bi-directional, and all functionality
-implemented in \Rpackage{flowCore} is available to other package
+system. This process is bi-directional, and all functionalities
+implemented in \Rpackage{flowCore} are available to other package
 authors.
 
-We can chose one of the many visualization option from the
+Finally, we can chose one of the many visualization options from the
 \Rpackage{flowViz} package to plot the results of the recent filtering
 operation. A very basic matrix of density plots is shown in Figure~3
-\ref{xyplot}, where each panel in the matrix represents the data of
-one individual patient.
+\ref{xyplot}, where each panel in the matrix represents the
+fluorescent measurements of two channels for one individual patient.
 
@@
 packages that are dedicated to more specific tasks of FCM data
-analysis. The \Rpackage{flowViz} package \cite{sarkar2008ufv}
-provides sophisticated data visualization tools, that make use off
-multivariate trellis plotting \cite{lattice}.  These functions can be
-used to quickly generate customized plots for extended cytometry data
-sets for both direct data inspection and quality control.  The objects
-metadata information can be used to arrange the layout and composition
-of the plots.  Furthermore, the design and the API of the
-visualization software is very generic, and users can readily extend
-its capabilities by providing self-defined plotting functions.  The
+analysis. As exemplified in the previous section, the
+\Rpackage{flowViz} package \cite{sarkar2008ufv} provides sophisticated
+data visualization tools, that make use off multivariate trellis
+plotting \cite{lattice}.  These functions can be used to quickly
+generate customized plots for extended cytometry data sets for both
+direct data inspection and quality control.  The objects metadata
+information can be used to arrange the layout and composition of the
+plots.  Furthermore, the design and the API of the visualization
+software is very generic, and users can readily extend its
+capabilities by providing self-defined plotting functions.  The
 \Rpackage{flowQ} package offers more advanced quality assurance (QA)
@@
 language. Finally, the \Rpackage{flowStats} package provides elaborate
-statistical methods that are relevant in the context of flow
-cytometry data analysis. \cite{lo2008agf} have recently developed an
-automatic gating approach via robust model-based clustering using
-\Rpackage{flowCore}'s data model and infrastructure which is
-implemented in the Bioconductor package \Rpackage{flowClust}. Another
-package, \Rpackage{plateCore}, providing more specialized support for
-experiments conducted on microtitre plates and facilitating the
-handling of spatial metadata, is under development.
+statistical methods that are relevant in the context of flow cytometry
+data analysis.
+
+More recently, \cite{lo2008agf} have developed an automatic gating
+approach via robust model-based clustering using \Rpackage{flowCore}'s
+data model and infrastructure which is implemented in the Bioconductor
+package \Rpackage{flowClust}. Another package, \Rpackage{plateCore},
+providing more specialized support for experiments conducted on
+microtitre plates and facilitating the handling of spatial metadata,
+is under development.
 
@@
 Through \Rpackage{flowCore}, we have provided the FCM community with
-an open source, freely available, highly functional, standards
-compliant development and analysis platform for high throughput data
+an open source, freely available, highly functional, and standards
+compliant, development and analysis platform for high throughput data
 analysis.  We hope to foster collaborative development of new analysis
-methods and to facilitate the transition of these new methods to the
+methods and to facilitate the transition of these new methods to a
 larger flow community.  Our experience has been that such
@@
 \section*{Availability and requirements}
-Project name: flowCore
-Project home page: http://bioconductor.org 
+Project name: flowCore ;
+Project home page: http://bioconductor.org ;
 Operating system(s): The flowCore package can be installed on all the 
-platforms supporting R. These include a wide variety of UNIX platforms, Windows 
-and MacOS.  
-Programming language: R 
-Licence: The Artistic License, Version 2.0
+platforms supporting R. These include a wide variety of UNIX platforms, Windows and MacOS.;  
+Programming language: R ;
+Licence: The Artistic License, Version 2.0 .
 
@@
         HTML quality assessment report generated by the flowQ package for a
-        subset of data from an experiment focussing on Graft-Versus-Host
+        subset of data from an experiment focusing on Graft-Versus-Host
         Disease \cite{brinkman2007hcf}. Rows correspond to the samples in
@@
     \label{xyplot}{
-        Scatterplot matrix of a single \Rclass{flowSet} from an experiement
-        focussing on immune tolerance following kidney
+        Scatterplot matrix of a single \Rclass{flowSet} from an experiment
+        focusing on immune tolerance following kidney
         transplantation. Outlines of the gating regions identified by an
@@
     \caption{\label{table2} Filter and gate classes implemented in
-      \Rpackage{flowCore}.} 
+      \Rpackage{flowCore}. Filters are automated, data driven procedures. Gates are static, user-defined methods.} 
     \begin{center}

### 2009-01-20 paper/bmc_article.tex
fix the call to featureSignif function from the feature package as some parameters have been redifined: addSignifCurv=TRUE now replaces addSignifCurvRegion=TRUE, plotFS=FALSE
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@36811 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
           Perry Haaland$^5$%
-          \email{Perry Haaland - Perry\_D\_Haaland@bd.com}
+          \email{Perry D. Haaland - Perry\_D\_Haaland@bd.com}
         \and
@@
   \paragraph*{Background:} Recent advances in automation technologies
-  have enabled the use of flow-cytometry high content screening
+  have enabled the use of flow cytometry high content screening
   (FH-HCS), generating large complex data sets often in clinical
@@
   \paragraph*{Results:} We developed a set of flexible open source
-  computational tools (in the R package \Rpackage{flowCore}) to
+  computational tools in the R package \Rpackage{flowCore} to
   facilitate the analysis of these complex data. A key component of

### 2009-03-18 paper/bmc_article.tex
response2reviewers and manuscript amendements
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@38020 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 
-\title{flowCore a Bioconductor package for high throughput flow cytometry}
+\title{flowCore: a Bioconductor package for high throughput flow cytometry}
 
@@
 %% But up to you
-\Rpackage{flowCore} does not provide a graphical user interface
-(\textit{i.e.}, is not a GUI-driven software) and all operations are
-done using a command line interface.  It is possible to add a more
-elaborate user interface on top of this infrastructure, however the
-focus in this paper is on a programmatic approach to enable the
-convenient development of novel analysis methods and automation of
-complex analysis approaches.  By taking the burden of data management
-from the programmer, and by providing well-defined application
-programming interfaces (APIs), it is possible to readily test new
-ideas and to easily extend the framework's functionality.
+\Rpackage{flowCore} does not provide a graphical user interface and
+all operations are done using a command line interface.  It is
+possible to add a more elaborate user interface on top of this
+infrastructure, however the focus in this paper is on a programmatic
+approach to enable the convenient development of novel analysis
+methods and automation of complex analysis approaches.  By taking the
+burden of data management from the programmer, and by providing
+well-defined application programming interfaces (APIs), it is possible
+to readily test new ideas and to easily extend the framework's
+functionality.
 
@@
 file-specific metadata (Figure~1\ref{fig1:FrameWork}). Moreover, it is
-a software implementation of the Gating-ML Candidate Recommendation
-(CR), an emerging ISAC standard developed in collaboration with the
-ISAC Data Standards Task Force, which makes it possible to integrate
+a software implementation of the Gating-Markup Language Candidate
+Recommendation (Gating-ML CR), an emerging standard developed in collaboration
+with the International Society for Analytical Cytology (ISAC) Data
+Standards Task Force, which makes it possible to integrate
 \Rpackage{flowCore} in existing work flows and to communicate with any
@@
 analysis offer support for these operations, most often in an
-interactive, GUI-driven interface. In \Rpackage{flowCore} we have
+interactive, graphical user interface. In \Rpackage{flowCore} we have
 taken the approach to abstractly describe these operations and build a
@@
 routine operations with only limited potential for improvement, being
-able to implement new methodologies for gating of FCM data, and
-extend the capabilities of \Rpackage{flowCore} through object
-oriented programming are features that clearly sets our framework
-apart from other FCM analysis tools. By factoring out as much of the
-bookkeeping as possible, programmers can focus on the actual
-operations rather then having to deal with the tedious details of data
-integration and access. Third-party methods can act on their own as
-first-class citizens in the analysis framework, without breaking the
-work flow or the basic infrastructure. This design allows for the
-straightforward extension of \Rpackage{flowCore}'s capabilities, and
-has already fostered the development of a number of valuable add-ons
+able to implement new methodologies for gating of FCM data, and extend
+the capabilities of \Rpackage{flowCore} through object oriented
+programming are features that clearly sets our framework apart from
+other FCM analysis tools. By factoring out as much of the bookkeeping
+as possible, programmers can focus on the actual operations rather
+then having to deal with the tedious details of data integration and
+access. Third-party methods can act on their own as first-class
+citizens in the analysis framework, without breaking the work flow or
+the basic infrastructure. This design allows for the straightforward
+extension of \Rpackage{flowCore}'s capabilities, and has already
+fostered the development of a number of valuable add-ons
 \cite{lo2008agf,sarkar2008ufv}.
@@
 capabilities by providing self-defined plotting functions.  The
-\Rpackage{flowQ} package offers more advanced quality assurance (QA)
+\Rpackage{flowQ} package offers more advanced quality assurance
 methodology and a framework to create interactive web-based reports of
-QA results. The \Rpackage{flowUtil} package implements data import and
-export including flow-cytometry specific standard markup
-language. Finally, the \Rpackage{flowStats} package provides elaborate
-statistical methods that are relevant in the context of flow cytometry
-data analysis.
+quality assurance results. The \Rpackage{flowUtil} package implements
+data import and export including flow-cytometry specific standard
+markup language. Finally, the \Rpackage{flowStats} package provides
+elaborate statistical methods that are relevant in the context of flow
+cytometry data analysis.
 
@@
 Project home page: http://bioconductor.org ;
-Operating system(s): The flowCore package can be installed on all the 
-platforms supporting R. These include a wide variety of UNIX platforms, Windows and MacOS.;  
+Operating system(s): A wide variety of UNIX platforms, Windows and MacOS.;  
 Programming language: R ;
 Licence: The Artistic License, Version 2.0 .
+
+The flowCore package and its associated packages are part of the
+R/Bioconductor project, an environment for statistical computing and
+bioinformatics. The R software environment is freely available at
+http://www.r-project.org. flowCore and its dependencies (flowQ,
+flowViz) are available on the Bioconductor project website
+(http://bioconductor.org) as freely distributed and open source
+software packages with an Artistic license. They are fully integrated
+into the R/Bioconductor environment for statistical computing and
+bioinformatics and run on operating systems Windows, Mac OS X, and
+Unix.
 
@@
 
-    \subsection*{Figure 3 - Quality assessment}
+    \subsection*{Figure 3 - Batch gating}
     \label{xyplot}{ 
-        Scatterplot matrix of a single \Rclass{flowSet} from an experiment
-        focusing on immune tolerance following kidney
-        transplantation. Outlines of the gating regions identified by an
-        automated gating operation are added on top of the density
-        representation of the data.}
+      Scatterplot matrix of a single \Rclass{flowSet}
+      from an experiment focusing on immune tolerance following kidney
+      transplantation. Outlines of the gating regions identified by a
+      \Rclass{curve2Filter} automated gating operation are added on
+      top of the density representation of the data.}
 
@@
           quadratic & $ax^2 + bx + c$ \\
-          natural logarithm & $ln(x)(r/d)$ \\
-          logarithm & $log(x)(r/d)$ \\
-          biexponential & $root(a*e^{(b*y)}-c*e^{(-d+y)}+f)$ \\
-          logicle& $root(Te^{-(m-w)}*(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
-          truncate & $x_{x<=a} = a$ \\
+          natural logarithm & $log_e(x)(r/d)$ \\
+          logarithm & $log_b(x)(r/d)$ \\
+          biexponential & $root(ae^{(b*y)}-ce^{(-d+y)}+f)$ \\
+          logicle& $root(Te^{-(m-w)}(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
+          truncate & $x_{x{\leq}a} = a$ \\
           scale & $(x-a)/(b-a)$ \\
           split-scale & combination of linear and logarithm \\
-          arcsinh & $asinh(a + bx)+c$ \\
+          arcsinh & $arcsinh(a + bx)+c$ \\
           \hline

### 2009-03-23 paper/bmc_article.tex
reviewer recommended updates
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@38132 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
   \paragraph*{Background:} Recent advances in automation technologies
-  have enabled the use of flow cytometry high content screening
-  (FH-HCS), generating large complex data sets often in clinical
+  have enabled the use of flow cytometry for high throughput
+  screening, generating large complex data sets often in clinical
   trials or drug discovery settings. However, data management and data
@@
 \cite{brinkman2007hcf}. A serious bottleneck in the interpretation of
-existing studies and the application of flow cytometry high content
-screening (FC-HCS) to even larger, more complex problems is that data
-management and data analysis methods have not advanced sufficiently
-far from the methods developed for applications of FCM
- to small-scale,
-tube-based studies \cite{mahnke2007omi}. In particular, the data often
-need to be organized into groups of samples based on combinations of
-additional covariates and similar operations need to be applied to
-these groups in a transparent and reproducible manner. Furthermore,
-the growing depth of knowledge in the field of immunology, for
-instance the characterization of distinct human T-cell sub-population
-\cite{Appay}, clearly argues for more systematic approaches.
+existing studies and the application of high throughput FCM to even
+larger, more complex problems is that data management and data
+analysis methods have not advanced sufficiently far from the methods
+developed for applications of FCM to small-scale, tube-based studies
+\cite{mahnke2007omi}. In particular, the data often need to be
+organized into groups of samples based on combinations of additional
+covariates and similar operations need to be applied to these groups
+in a transparent and reproducible manner. Furthermore, the growing
+depth of knowledge in the field of immunology, for instance the
+characterization of distinct human T-cell sub-population \cite{Appay},
+clearly argues for more systematic approaches.
 %% add one of the last roederer's paper that "request" algo and
@@
 solutions offer only limited support for such self-contained
-structures, or make use of containers like XML or binary storage
-formats that are designed specifically for particular user interfaces
-and hence not easily amenable to programmatical access. In addition,
-the closed-source nature of these products often makes them
-impractical to integrate into analysis pipelines. In this manuscript
-we describe classes for FCM data analysis and their implementation in
-R, however, they could just as easily be implemented in any other
-language (e.g., Java, C++). Software written in those languages could
-use similar data structures, thereby simplifying communication and the
-interchange of data between analysis tools.
+structures, or make use of binary storage containers that are designed
+specifically for the needs of particular user interfaces and hence are
+not easily amenable to programmatical access. In addition, the
+closed-source nature of these products often makes them impractical to
+integrate into analysis pipelines. In this manuscript we describe
+classes for FCM data analysis and their implementation in R, however,
+they could just as easily be implemented in any other language (e.g.,
+Java, C++). Software written in those languages could use similar data
+structures, thereby simplifying communication and the interchange of
+data between analysis tools.
 
@@
 file-specific metadata (Figure~1\ref{fig1:FrameWork}). Moreover, it is
-a software implementation of the Gating-Markup Language Candidate
-Recommendation (Gating-ML CR), an emerging standard developed in collaboration
+a software implementation of the Gating Markup Language Candidate
+Recommendation, an emerging standard developed in collaboration
 with the International Society for Analytical Cytology (ISAC) Data
@@
           logarithm & $log_b(x)(r/d)$ \\
-          biexponential & $root(ae^{(b*y)}-ce^{(-d+y)}+f)$ \\
-          logicle& $root(Te^{-(m-w)}(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
+          biexponential & $ae^{(b*x)}-ce^{(-d+x)}+f$ \\
+          logicle& $Te^{-(m-w)}(e^{(x-w)}-p^2e^{-(x-w)/p}+p^2-1)$ \\
           truncate & $x_{x{\leq}a} = a$ \\
           scale & $(x-a)/(b-a)$ \\
-          split-scale & combination of linear and logarithm \\
           arcsinh & $arcsinh(a + bx)+c$ \\

### 2009-03-24 inst/doc/HowTo-flowCore.Rnw
edits to the vignette
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35702 bc3139a8-67e5-0310-9ffc-ced21a209358
reduce the package dependencies and optimize the search path strategy
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@37386 bc3139a8-67e5-0310-9ffc-ced21a209358
update for revised version of the manuscript: logicle reference, flowCore vignette...
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@38141 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 %\VignetteIndexEntry{Basic Functions for Flow Cytometry Data}
-%\VignetteDepends{flowCore}
+%\VignetteDepends{flowCore, flowViz}
 %\VignetteKeywords{}
@@
 transforms commonly found in flow cytometry and defined in the
-Transformation Markup Language (Transformation-ML). Briefly, the
+Transformation Markup Language (Transformation-ML, see \url{http://www.ficcs.org/} and \cite{Spidlen2006} for more details). Briefly, the
 predefined transforms are:
@@
 
-\section{Filtering}
-
-The most common task in the analysis of flow cytometry data is usual
-some form of filtering operation, either to obtain summary statistics
-about the number of events that meet a certain criteria or to perform
-further analysis on a subset of the data.
-
-\subsection{Standard Filters}
-
-Most filtering operations are a composition of one or more common
-filtering operations. Like transformations, \Rpackage{flowCore}
-includes a number of built-in common flow cytometry filters.
-
-The simplest of these filters are the geometric filters, which
-correspond to those typically found in interactive flow cytometry
-software:
+\section{Gating}
+
+The most common task in the analysis of flow cytometry data is some
+form of filtering operation, also known as gating, either to obtain
+summary statistics about the number of events that meet a certain
+criteria or to perform further analysis on a subset of the data.  Most
+filtering operations are a composition of one or more common filtering
+operations. The definition of gates in \Rpackage{flowCore} follows the
+Gating Markup Language Candidate Recommendation \cite{Spidlen2008},
+thus any \Rpackage{flowCore} gating strategy can be reproduced by any
+other software that also adheres to the standard and \textit{vice
+  versa}.
+
+\subsection{Standard gates and filters}
+Like transformations, \Rpackage{flowCore} includes a number of
+built-in common flow cytometry gates. The simplest of these gates are
+the geometric gates, which correspond to those typically found in
+interactive flow cytometry software:
 \begin{description}
@@
 
-We also include some data-driven filters not usually found in flow
-cytometry software:
+In addition, we introduce the notion of data-driven gates, or filters,
+not usually found in flow cytometry software. In these approaches, the
+necessary parameters are computed based on the properties of the
+underlying data, for instance by modelling data distribution or by
+density estimation :
 
@@
 function depends on the \Rpackage{Rgraphviz} package.
-<<plotwf, fig=true>>=
+<<plotwf, eval=FALSE>>=
 plot(wf)
+@ 
+<<plotwfdo, fig=true, echo=FALSE, results=hide>>=
+if(suppressWarnings(require(Rgraphviz))){
+    plot(wf)
+}else{
+    plot(1,1, type="n", axes=FALSE, ann=FALSE)
+    text(1,1,"Need to install Rgraphviz")
+}
 @

### 2013-02-06 inst/doc/HowTo-flowCore.Rnw
Fixed typos in the vignette.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@43485 bc3139a8-67e5-0310-9ffc-ced21a209358
initial checkin

@@
 \noindent \textbf{Results} We hope that those proposed data structures
-will the base for the development of many tools for the analysis of
+will be the base for the development of many tools for the analysis of
 high throughput flow cytometry.\\
@@
 the usual way using the \Rfunction{[[} or \Rfunction{\$} extraction
-  operators. On the other hand using he \Rfunction{[} extraction
+  operators. On the other hand using the \Rfunction{[} extraction
   operator returns a new \Rclass{flowSet} by \textbf{copying} the
@@
 @
-
+%
 which is equivalent to the less readable
@@
 @
-
+%
 In this case, the \Robject{use.exprs} argument is not required in the

### 2013-02-06 paper/bmc_article.tex
initial checkin

@@
 routinely used in FCM analysis have been implemented in
-\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponential or arcsinh,
-see Table~\ref{table1} for a complete list). Furthermore, the design
-of the R language makes it easy to define arbitrary functions to apply
-to the data of individual \Rclass{flowFrames} or entire
-\Rclass{flowSets}, respectively. Compensation, that corrects for
-fluorescence spillover originating from the inherent overlap of
-emission spectra from antibody fluorescent labels, is available for
-both \Rclass{flowFrames} and \Rclass{flowSets}. In addition, the
-software offers functionality to compute spillover or compensation
-matrices from a set of appropriate single stain controls.
+\Rpackage{flowCore} (\textit{e.g.,} log, bi-exponential, arcsinh or
+logicle \cite{Parks:2006}, see Table~\ref{table1} for a complete
+list). Furthermore, the design of the R language makes it easy to
+define arbitrary functions to apply to the data of individual
+\Rclass{flowFrames} or entire \Rclass{flowSets},
+respectively. Compensation, that corrects for fluorescence spillover
+originating from the inherent overlap of emission spectra from
+antibody fluorescent labels, is available for both \Rclass{flowFrames}
+and \Rclass{flowSets}. In addition, the software offers functionality
+to compute spillover or compensation matrices from a set of
+appropriate single stain controls.
 
@@
 in multi-step gating strategies. The definition of gates in
-\Rpackage{flowCore} follows the Gating-ML CR, thus any
+\Rpackage{flowCore} follows the Gating Markup Language Candidate
+Recommendation \cite{SpidlenInPressCytometryA}, thus any
 \Rpackage{flowCore} gating strategy can be reproduced by any other
@@
       \caption{\label{table1} Data transformations implemented in \Rpackage{flowCore}. Within these formulas, $x$ is the variable
-        corresponding to value being transformed, $y$ is the result of the
-        transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
-        constants affecting the transformation function, $e$ is the base of
-        the natural logarithm, and $root$ is a root-finding algorithm. Other transformations can easily be implemented in R.}
+        corresponding to value being transformed, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
+        constants affecting the transformation function, $e$ is the base of the natural logarithm (see \cite{Parks:2006} for details on the logicle transformation).
+Other transformations can easily be implemented in R.}
       \begin{center}

### 2013-02-06 R/AllGenerics.R

@@ -1,0 +1,279 @@
+## =========================================================================##
+## =========================================================================##
+##                       Generic methods definitions                        ##
+## =========================================================================##
+## =========================================================================##
+
+
+
+
+
+
+## ===========================================================================
+## Generics for our operators
+## %in% asks a question whereas %subset% forms a set.
+## (default for %in% in base) 
+## ---------------------------------------------------------------------------
+setGeneric("%in%",function(x,table) standardGeneric("%in%"))
+setGeneric("%subset%",function(e1,e2) standardGeneric("%subset%"))
+setGeneric("%&%",function(e1,e2) standardGeneric("%&%"))
+setGeneric("%on%",function(e1,e2) standardGeneric("%on%"))
+
+
+
+## ===========================================================================
+## Generic for sorting (default in base)
+## ---------------------------------------------------------------------------
+setGeneric("sort",function(x,decreasing=FALSE,...) standardGeneric("sort"))
+
+
+
+## ===========================================================================
+## Generic for printing details of an object
+## ---------------------------------------------------------------------------
+setGeneric("print",function(x,...) standardGeneric("print"))
+
+
+
+## ===========================================================================
+## Generic for range should be present via the 'Summary' group generic
+## ---------------------------------------------------------------------------
+
+
+
+## ===========================================================================
+## Generic for head and tail (already S3 in utils)
+## ---------------------------------------------------------------------------
+setGeneric("head", function(x, ...) standardGeneric("head"))
+setGeneric("tail", function(x, ...) standardGeneric("tail"))
+
+
+
+## ===========================================================================
+## Generic for keyword
+## ---------------------------------------------------------------------------
+setGeneric("keyword",function(object,keyword) standardGeneric("keyword"))
+setGeneric("keyword<-",function(object,value) standardGeneric("keyword<-"))
+
+
+
+## ===========================================================================
+## Generic to access parameters in a flowframe
+## ---------------------------------------------------------------------------
+setGeneric("parameters", function(object, ...) standardGeneric("parameters"))
+setGeneric("parameters<-",
+           function(object, value) standardGeneric("parameters<-"))
+
+
+
+## ===========================================================================
+## Generic to access colnames (defaults in base)
+## ---------------------------------------------------------------------------
+setGeneric("colnames<-",
+           function(x, value) standardGeneric("colnames<-"))
+
+
+
+## ===========================================================================
+## Generic for identifier
[…]

### 2013-02-06 R/on-methods.R

@@ -1,0 +1,135 @@
+## ==========================================================================
+## %on% methods are basically constructors for transformFilter or
+## transformList objects. They allow for an inline-syntax for transformation
+## operations.
+## ==========================================================================
+
+
+
+
+
+
+## ===========================================================================
+## Constructor for a transformFilter for a transform
+## ---------------------------------------------------------------------------
+setMethod("%on%",
+          signature=signature(e1="filter",
+                              e2="transform"),
+          definition=function(e1, e2)
+      {
+          new("transformFilter",
+              filterId=paste(identifier(e1), "on transformed values"),
+              transforms=parameterTransform(e2, parameters(e1)),
+              filter=e1, parameters=parameters(e1))
+      })
+
+
+
+## ===========================================================================
+## Wrapper for a flowFrame. We transform on the actual raw data
+## ---------------------------------------------------------------------------
+setMethod("%on%",
+          signature=signature(e1="transform",
+                              e2="flowFrame"),
+          definition=function(e1, e2)
+      {
+          exprs(e2) <- e1(exprs(e2))
+          e2
+      })
+
+
+
+## ===========================================================================
+## Constructor for a transformFilter from a parameterTransform
+## ---------------------------------------------------------------------------
+setMethod("%on%",
+          signature=signature(e1="filter",
+                              e2="parameterTransform"),
+          definition=function(e1, e2)
+      {
+          new("transformFilter",
+              filterId=paste(e1@filterId," on transformed values of ",
+              paste(parameters(e2), sep=","),collapse=" "),
+              transforms=e2, filter=e1,
+              parameters=unique(c(parameters(e1), parameters(e1))))
+})
+
+
+
+## ===========================================================================
+## Wrapper for a flowFrame. We transform on the actual raw data
+## ---------------------------------------------------------------------------
+setMethod("%on%",
+          signature=signature(e1="parameterTransform",
+                              e2="flowFrame"),
+          definition=function(e1, e2)
+          {
+            params <- if(length(parameters(e1))) colnames(e2) else
+            parameters(e1)
+            exprs(e2)[,params] <- e1(exprs(e2))[,params]
+            e2
+          })
+
+
+
+## ===========================================================================
[…]

### 2013-02-06 inst/unitTests/runit.00classes.R

@@ -1,0 +1,10 @@
+#Make sure our special generics are actually defined as generic.
+test.generics = function() {
+	checkTrue(isGeneric("transform"))
+	checkTrue(isGeneric("%in%"))
+	#Actually resolve a method
+	checkTrue(hasMethod("%in%",c("flowFrame","rectangleGate")))
+	print(hasMethod("%in%",c("flowFrame","rectangleGate")))
+	print(getMethods("%in%"))
+	print(get("%in%"))
+}

### 2013-02-06 paper/figure.R

@@ -1,0 +1,47 @@
+## Loading library
+library("flowCore")
+library("flowQ")
+library("flowViz")
+
+## Read a batch of FCS files (find another data set with time channel - Rituximab to change from the GvHD?)
+fs = read.flowSet(path = system.file("extdata", package="flowCore"), pattern="0877408774")
+
+fs
+
+## Tranform
+
+## Assess Quality (using fs give me en error) 
+qaReport(GvHD[1:3], c("qaProcess.timeline",
+                      "qaProcess.timeflow",
+                      "qaProcess.cellnumber"))
+
+## Gating 
+?polygonGate
+
+## Defining the gate
+sqrcut <- matrix(c(300,300,600,600,50,300,300,50),ncol=2,nrow=4)
+colnames(sqrcut) <- c("FSC-H","SSC-H")
+pg <- polygonGate(filterId="nonDebris", .gate= sqrcut)
+pg
+
+fres <- filter(GvHD[1:3], pg)
+fres
+summary(fres)
+
+## Filtering
+## Define filter approach
+n2f <- norm2Filter("FSC-H", "SSC-H", filterId="myNorm2Filter")
+
+## Filtering using curv2Filter
+fres <- filter(GvHD[1:7], n2f)
+fres
+summary(fres)
+
+## Visualizing
+pdf("n2f-output.pdf")
+print(xyplot(`SSC-H` ~ `FSC-H` | Patient:Visit, data = GvHD,
+            filter=n2f, subset=Patient==5))
+dev.off()
+
+## Select subpobulation for futher analysis
+nonDebris= Subset(GvHD[1:3], fres)

### 2013-02-06 src/flowCore.h

@@ -1,0 +1,9 @@
+#include <R.h>
+#include <Rinternals.h>
+
+SEXP biexponential_transform(SEXP input,SEXP A,SEXP B,SEXP C,SEXP D,SEXP F,SEXP G,SEXP tol,SEXP maxit);
+SEXP logicle_transform(SEXP input,SEXP M,SEXP W,SEXP P,SEXP T,SEXP A,SEXP tol,SEXP maxit);
+SEXP invLogicle_transform(SEXP input,SEXP M,SEXP W,SEXP P,SEXP T,SEXP A);
+SEXP inPolygon(SEXP _data, SEXP _vertices);
+SEXP inPolytope(SEXP _data, SEXP _A, SEXP _b);
+

### 2013-02-06 unitTests/InOutEventsTests/RUnitScript_Files/performGateTest.R

@@ -1,0 +1,99 @@
+####################################################################################
+# This R script file is part RUnit tests of Bioconductor/flowCore package intended #
+# to test Gating-ML compatibility.                                                 #
+#                                                                                  #
+# File includes the performGateTest function                                       #
+# Parametes:                                                                       #
+# - gateID - the ID of the gate; corresponds to filterId slot the particular gate  #
+#            object and to the column name of the table with expected results      #
+# - testGates - list of gates                                                      #
+# - fcs - the FCS data as an instance of the flowFrame                             #
+# - expectedResults - table of expected results                                    #
+#   - columns correspond to events in FCS file                                     #
+#   - rows correspond to events in FCS file                                        #
+#   - individual cells: 1 = the event is expected to be found in the gate          #
+#                       0 = the event is NOT expected to be found in the gate      #
+# - numberOfGates - the number of gates in testgates object, default is 0, which   #
+#                   means that the functions computes the real numberOfGates.      #
+#                                                                                  #
+# @author Stanley Wong, Josef Spidlen, Ryan Brinkman                               #
+#                                                                                  #
+# File distributed under terms of the GNU Lesser General Public License (LGPL)     #
+####################################################################################
+
+# Ensure loading of the required libraries
+library(flowCore)
+library(flowUtils)
+library(RUnit)
+library(XML)
+
+# Function performing the actual test of a single gate.
+#
+# The gate to be tested is identified by gateID. The gate is taken from the testgates list
+# (list of unknown gates) and is applied on the FCS data. The results are compared against
+# the appropriate column of the expectedResults table. Any discrepancy between the computed
+# and expected results (gate membership of each event) is reported within the RUnit framework.
+#
+# Parametes:
+# - gateID - the ID of the gate; corresponds to filterId slot the particular gate object
+#            and to the column name of the table with expected results
+# - testGates - list of gates
+# - fcs - the FCS data as an instance of the flowFrame
+# - expectedResults - table of expected results
+#   - columns correspond to events in FCS file
+#   - rows correspond to events in FCS file
+#   - individual cells: 1 = the event is expected to be found in the gate
[…]

### 2013-02-06 unitTests/InvalidGateTests/RUnitScript_Files/runit.InvalidMinMax.R

@@ -1,0 +1,19 @@
+library(flowCore)
+library(flowUtils)
+library(RUnit)
+library(XML)
+
+fcsFile <- system.file("unitTests/standardsCompliance/Gating-ML_ComplianceTests/InvalidGateTests/FCS_Files","testfile_17parameters.fcs",package="flowCore")
+
+gateFile <- system.file("unitTests/standardsCompliance/Gating-ML_ComplianceTests/InvalidGateTests/Gating-ML_Files","InvalidMinMax.xml",package="flowCore")
+
+source(system.file("unitTests/standardsCompliance/Gating-ML_ComplianceTests/InvalidGateTests/RUnitScript_Files","performSimpleGateTest.R",package="flowCore"))
+
+test.1 <- function() {
+  performSimpleGateTest("CD3",fcsFile,gateFile)
+}
+
+test.2 <- function() {
+  performSimpleGateTest("RectGate1",fcsFile,gateFile)
+}
+

### 2013-11-19 R/IO.R
merge from git
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@83412 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -69,6 +69,7 @@ read.FCS <- function(filename,
     on.exit(close(con))
 
     ## transform or scale data?
+    fcsPnGtransform <- FALSE
     if(is.logical(transformation) && transformation ||
        !is.null(transformation) && transformation == "linearize") {
         transformation <- TRUE
@@ -76,6 +77,10 @@ read.FCS <- function(filename,
     } else if ( !is.null(transformation) && transformation == "scale") {
         transformation <- TRUE
         scale <- TRUE
+    } else if ( !is.null(transformation) && transformation == "linearize-with-PnG-scaling") {
+        transformation <- TRUE
+        scale <- FALSE
+        fcsPnGtransform <- TRUE
     } else if (is.null(transformation) || is.logical(transformation) &&
                !transformation) {
         transformation <- FALSE 
@@ -107,6 +112,7 @@ read.FCS <- function(filename,
     txt <- readFCStext(con, offsets,emptyValue=emptyValue)
     ## We only transform if the data in the FCS file hasn't already been
     ## transformed before
+    if (fcsPnGtransform) txt[["flowCore_fcsPnGtransform"]] <- "linearize-with-PnG-scaling"
     if("transformation" %in% names(txt) &&
        txt[["transformation"]] %in% c("applied", "custom"))
        transformation <- FALSE
@@ -656,7 +662,18 @@ readFCSdata <- function(con, offsets, x, transformation, which.lines,
             dat[dat<min.limit] <- min.limit
     }
 
-    ## transform or scale if necessary
+    ## Transform or scale if necessary
+    # J.Spidlen, Nov 13, 2013: added the flowCore_fcsPnGtransform keyword, which is 
+    # set to "linearize-with-PnG-scaling" when transformation="linearize-with-PnG-scaling"
+    # in read.FCS(). This does linearization for log-stored parameters and also division by
+    # gain ($PnG value) for linearly stored parameters. This is how the channel-to-scale
+    # transformation should be done according to the FCS specification (and according to 
+    # Gating-ML 2.0), but lots of software tools are ignoring the $PnG division. I added it
+    # so that it is only done when specifically asked for so that read.FCS remains backwards
+    # compatible with previous versions.
[…]

### 2013-11-19 inst/doc/HowTo-flowCore.Rnw
merge from git
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@83412 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
 the \Rfunction{transformation} parameter, which can either
-``linearize'' (the default) or ``scale'' our data. To
-see how this works, first we will examine an FCS file
+``linearize'' (the default) or ``linearize-with-PnG-scaling'' 
+or ``scale'' our data. To see how this works, 
+first we will examine an FCS file
 without any transformation at all:
@@
 @
+
+The ``linearize-with-PnG-scaling'' option will perform the previous
+transformation and it will also apply a ``division by gain''
+to pamameters stored on linear scale with specified gain.
+The gain is specified in the \$PnG keywords. This option has been introduced
+as part of Gating-ML 2.0 compliance.
 

### 2014-04-11 R/eval-methods.R
Creating 2.14 release branch.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/branches/RELEASE_2_14/madman/Rpacks/flowCore@88839 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1,0 +1,626 @@
+##===========================================================================
+## Eval methods for evaluating transformation objects 
+## ---------------------------------------------------------------------------
+
+
+## Helper function to fully resolve all transformationReferences
+resolve <- function(x, df)
+{
+    if(!is(x, "transformReference"))
+        eval(x)(df) else resolveTransformReference(x, df)  
+}
+
+
+
+## ===========================================================================
+## Unity transformation
+## ---------------------------------------------------------------------------
+setMethod("eval",
+          signature=signature(expr="unitytransform", envir="missing"),
+          definition=function (expr, envir = parent.frame(),
+              enclos=if (is.list(envir) ||
+                is.pairlist(envir)) parent.frame() else baseenv())
+          {        
+              function(df) 
+              {
+                  df <- flowFrameToMatrix(df)
+                  return(df[, expr@parameters, drop=FALSE]) 
+              }
+          })
+
+
+## ===========================================================================
+## Polynomial transformation of degree 1
+## ---------------------------------------------------------------------------
+
+setMethod("eval",
+	  signature=signature(expr="dg1polynomial", envir="missing"),
+          definition=function (expr, envir = parent.frame(),
+              enclos=if (is.list(envir) ||
+                is.pairlist(envir)) parent.frame() else baseenv())
+          {    
+            function(df)
+              {   
+                par <- expr@parameters
+                temp <- sapply(par,function(i){
+                  if(is(i, "function")) return(i)
+                  resolve(i, df)
+                })
+                temp <- flowFrameToMatrix(temp)
+                coeff <- expr@a
+                res <- 0
+                for(i in seq_along(expr@parameters))
+                  res <- res + temp[,i,drop=FALSE]*coeff[i]
+                res <- res+expr@b
+              }
+          })
+
+
+## ===========================================================================
+## Ratio transformation of two arguments
+## ---------------------------------------------------------------------------
+setMethod("eval", 
+          signature=signature(expr="ratio", envir="missing"),
+          definition=function (expr, envir = parent.frame(),
+              enclos=if (is.list(envir) ||
+                is.pairlist(envir)) parent.frame() else baseenv())
+      { 
+          function(df)
+          {  
+              num <- resolve(expr@numerator, df)
+              den <- resolve(expr@denominator, df)
[…]

### 2014-04-11 R/gatingML.R

@@ -1,0 +1,106 @@
+## Recursively resolve transformation references until we get to a
+## realized transform object
+##
+# Josef Spidlen, Oct 18, 2013:
+# Changed df to x since now there may be a flowFrame object instead of just 
+# the data matrix. Gating-ML 2.0 includes the option of specifying "FCS" as 
+# the "spillover matrix", which means that parameters are supposed to be 
+# compensated as per compensation description in FCS. This is why the
+# whole flowFrame is passed here and to eval methods, which then have
+# the option of extracting the spillover matrix from the FCS keywords.
+resolveTransformReference<-function(trans, x)
+{  
+    if (is(trans, "transformReference")) 
+    {
+        trans <- eval(trans)
+        if(!is(trans, "compensatedParameter"))
+          trans <- resolveTransformReference(trans, x)
+        else
+          trans <- eval(trans)(x)
+    }
+    else
+    { 
+        if(!is(trans, "function")) 
+            trans <- eval(trans)(x)
+    }
+    trans
+}
+
+## FIXME NG: Please document
+EH <- function(y, a, b, argument)
+{
+    if(y>=0)
+        return(10^(y/a)+b*y/a-1-argument)
+    else
+        return(-10^(-y/a)+b*y/a+1-argument)
+}
+
+solveEH<-function(args,a,b)
+{  
+    #temp=eval(args)(df)
+    len=length(args)
+    result=0;
+    
+    while(len!=0)
+    {
+        step<-2
+        checkme<-EH(-step,a,b,args[len])*EH(step,a,b,args[len])
+        while(checkme>0)
+        {
+          step<-step*2
+          checkme<-EH(-step,a,b,args[len])*EH(step,a,b,args[len])
+        } 
+        result[len]=uniroot(EH,c(-step,step),tol=0.001,a=a,b=b,args[len])$root
+        len=len-1
+            
+    }
+
+    return(matrix(result))
+}
+
+## ===========================================================================
+## Evaluates all the transformations in the parameter slot of the filter 
+## and returns a list containing the modified flowFrame and a modified filter 
+## with slot parameter referencing the columns in the modified flowframe
+## ---------------------------------------------------------------------------
+resolveTransforms <- function(x, filter)
+{    
+    ## Filters need to be fully realized for this to work and we have to
+    ## resolve all filterReferences
+    recCoerce <- function(y){
+        if(is(y, "filterReference")){
+            y <- as(y, "concreteFilter")
[…]

### 2014-04-11 R/rectangleGate-accessors.R

@@ -1,0 +1,53 @@
+## ==========================================================================
+## Methods for objects of type 'rectangleGate'
+## Note: All filtering methods are stored in file 'in-methods.R'
+## ==========================================================================
+
+
+
+
+
+
+## ==========================================================================
+# Compose two rectangle gates together into a higher dimensional cube.
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("*",
+          signature(e1="rectangleGate",
+                    e2="rectangleGate"),
+          function(e1, e2)
+      {
+          if(any(parameters(e1) %in% parameters(e2)))
+              stop("Rectangle gate parameters overlap.", call.=FALSE)
+          new("rectangleGate", parameters=c(parameters(e1), parameters(e2)),
+              min=c(e1@min, e2@min), max=c(e1@max, e2@max))
+      })
+
+
+## ==========================================================================
+## subsetting by parameter name 
+## - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+setMethod("[",
+          signature=signature(x="rectangleGate",
+                              i="character"),
+          definition=function(x, i, j, ..., drop=FALSE)
+      {
+          if(drop)
+              warning("Argument 'drop' ignored for subsetting of ",
+                      "rectangelGates", call.=FALSE)
+          mt <- i %in% parameters(x)
+          if(!all(mt))
+              stop("The following parameter(s) are not defined in this gate:",
+                   paste("  ", i[!mt], collapse="\n"), call.=FALSE)
+          x@min <- x@min[i]
+          x@max <- x@max[i]
+          parameters(x) <- i
+          x
+      })
+
+
+setMethod("[",
+          signature=signature(x="rectangleGate"),
+          definition=function(x, i, j, ..., drop=FALSE)
+          stop("rectangleGates may only be subset by parameter name",
+               call.=FALSE)
+          )

### 2014-04-11 inst/scripts/smoothScatter/blendFuns.R

@@ -1,0 +1,218 @@
+## Here are a number of basic image processing and blending
+## algorithms. Most of them are based on the documentation to
+## the GIMP software but they should be rather general (although
+## image processing is a hell of a complex field!!!)
+
+require(RColorBrewer)
+
+## A wrapper around blending functions. Converts color matrices to rgb,
+## applies the blending and resets the dimensions of the resulting,
+## color vector. This allows to feed bitmaps to the blending functions.
+blendWrapper <- function(background, foreground, blendFun, ...){
+  d1 <- dim(background)
+  d2 <- dim(foreground)
+  stopifnot(all(d1==d2))
+  bgrgb <- col2rgb(background)
+  fgrgb <- col2rgb(foreground)
+  blendcol <- blendFun(bg=bgrgb, fg=fgrgb, ...)
+  if(!is.null(d1) && !is.null(d2))
+    res <- matrix(rgb(blendcol[1,], blendcol[2,], blendcol[3,], max=255),
+                  nrow=d1[1], ncol=d1[2])
+  else
+    res <- rgb(blendcol[1,], blendcol[2,], blendcol[3,], max=255) 
+  return(res)
+}
+
+##------------------------------------------------------------------------
+## Some usefull image processing functions:
+
+## The brightness of a color: a percentage
+brightness <- function(col){
+  signif((colSums(col)/3)/2.55, 3)
+}
+
+## The lightness of a pixel: a percentage
+lightness <- function(col){
+  signif(((apply(col, 2, max)+apply(col, 2, min))/2)/2.55, 3)
+}
+
+## Best reflects the human perception of
+## the brighntess of a color: a percentage
+luminance <- function(col){
+  signif(colSums(col*c(0.3, 0.59, 0.11))/2.55, 3)
+}
+
+## The actual color component in hsv colorspace:
+## a value from 0 to 360
+hue <- function(col){
+  col <- col/255
+  cmax <- apply(col, 2, max)
+  cmin <- apply(col, 2, min)
+  delta <- cmax - cmin
+  ##h <- rep(0, ncol(col))
+  undef <- cmax == 0
+  ym <- col[1,]==cmax
+  cm <- col[2,]==cmax
+  h <- 4+(col[1,]-col[2,])/delta
[…]

### 2014-04-11 unitTests/InOutEventsTests/RUnitScript_Files/runit.LargeDecisionTrees.R

@@ -1,0 +1,57 @@
+####################################################################################
+# This R script file is part RUnit tsets of Bioconductor/flowCore package intended #
+# to test Gating-ML compatibility.                                                 #
+#                                                                                  #
+# Tests of expected vs. detected events in gates of                                #
+#  - "LargeDecisionTrees.xml" Gating-ML file                                       #
+# applied to                                                                       #
+#  - "int-10000_events_random.fcs" FCS data file                                   #
+#                                                                                  #
+# @author Stanley Wong, Josef Spidlen, Ryan Brinkman                               #
+#                                                                                  #
+# File distributed under terms of the GNU Lesser General Public License (LGPL)     #
+####################################################################################
+
+# Ensure loading of the required libraries
+library(flowCore)
+library(flowUtils)
+library(RUnit)
+library(XML)
+
+# Variables shared for all the functions within this script file
+#
+# Table of expected results:
+# - columns correspond to gate IDs
+# - rows correspond to events in FCS file
+# - individual cells: 1 = the event is expected to be found in the gate
+#                     0 = the event is NOT expected to be found in the gate
+csvFile <- system.file("unitTests/standardsCompliance/Gating-ML_ComplianceTests/InOutEventsTests/ExpectedResult_Files","int-10000_events_random.csv",package="flowCore")
+
+# Full file name of the FCS data file used in tests within this script file
+fcsFile <- system.file("unitTests/standardsCompliance/Gating-ML_ComplianceTests/InOutEventsTests/FCS_Files","int-10000_events_random.fcs",package="flowCore")
+
+# Full file name of the Gating-ML file used in tests within this script file
+gateFile <- system.file("unitTests/standardsCompliance/Gating-ML_ComplianceTests/InOutEventsTests/Gating-ML_Files","LargeDecisionTrees.xml",package="flowCore")
+
+# The expected results read in from the CSV file
+expectedResults <- read.table(csvFile,sep=",",header=TRUE) 
+
+# The FCS data (an instance of the flowFrame class) read in from the FCS data file 
+fcs <- read.FCS(fcsFile)
+
+# A list of gates read in from the Gating-ML file
+testGates <- read.gatingML(gateFile)
+
+# Change the type of the "fcs@exprs" slot (the actual parameter values) to numeric
+fcs@exprs[,1:ncol(fcs@exprs)] <- as.numeric(fcs@exprs)
[…]

### 2014-09-22 R/AllClasses.R
defunct filterSet class

@@ -902,6 +902,7 @@ setClass("filterSet",
 
 ## constructor
 filterSet <- function(..., name="default") {
+    .Defunct("flowWorkspace::GatingSet")
     filters <- list(...)
     ## Allow the list(x, y, z) format as well.
     if(length(filters)==1 && is.list(filters[[1]]))
@@ -2267,6 +2268,7 @@ copyFlowFrame <- function(x) x[1:nrow(x)]
 ## graph and the alias table in the environment
 workFlow <- function(data, name="default", env=new.env(parent=emptyenv()))
 {
+  .Defunct("flowWorkspace::GatingSet")
     ## some sanity checks up front
     checkClass(name, "character", 1)
     checkClass(env, "environment")

### 2014-09-22 inst/tests/test-filterSet.R
demo sequential quadrant gating using  OpenCyto in vignette

@@ -1,47 +1,0 @@
-test_that("filterSet", {
-    b08 = read.FCS(system.file("extdata","0877408774.B08",package="flowCore"),transformation="scale")
-    fs <- new("filterSet")
-    fs[["filter1"]] <- rectangleGate("FSC-H"=c(.2,.8),"SSC-H"=c(0,.8))
-    
-    expect_is(fs[["filter1"]], "rectangleGate")
-    expect_is(fs["filter1"], "filterReference")
-    
-    expect_equal(sum(b08 %in% fs["filter1"]), 8291)
-    
-    b08.result1 <- filter(b08,fs["filter1"])
-    expect_is(b08.result1, "filterResult")
-    expect_equal(summary(b08.result1)@p, 0.8291)
-    
-    fs[[""]] <- norm2Filter("FSC-H","SSC-H",scale.factor=2,filterId="Live Cells")
-    
-    filter1 <- rectangleGate("FSC-H"=c(.2,.8),"SSC-H"=c(0,.8))
-    filter2 <- norm2Filter("FSC-H","SSC-H",scale.factor=2,filterId="Live Cells")
-    
-    #these can't be evaluated within local session
-#    expect_is(as("filter1","filter"), "filterReference")
-#    expect_is(as(as.name("filter1"),"filter"), "filterReference")
-    
-#    expect_is(as(~ filter1 %subset% filter2,"filter"), "subsetFilter")
-    
-    fs[["Combined"]] <- ~ `Live Cells` %subset% filter1
-    expect_is(as(fs,"list"), "list")
-    
-    expect_is(sort(fs,dependencies=TRUE), "character")
-    f <- filter(b08,fs)
-    expect_is(f, "manyFilterResult")
-    expect_is(as.data.frame(f), "data.frame")
-    expect_is(f@subSet[1:10,], "matrix")
-    expect_equal(colSums(f@subSet), c(filter1 = 8291, `Live Cells` =  6717, "Combined" = 6496))
-
-    expect_is(f[[1]], "filterResult")
-    expect_is(f[[2]], "filterResult")
-    expect_is(f[[3]], "filterResult")
-    
-    expect_equal(rownames(f@dependency)[rowSums(f@dependency)==0], "Combined")
-    
-    expect_is(split(b08,f,flowSet=TRUE), "flowSet")
-    expect_is(split(b08,f,drop=TRUE,flowSet=TRUE), "flowSet")
-    
-
-})
-

### 2016-08-12 R/AllClasses.R
create S3 generic and method of estimateLogicle for flowFrame. RGLab/flowWorkspace#206
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@120102 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1473,19 +1473,27 @@ inverseLogicleTransform <- function(transformationId, trans) {
     logicleTransform( transformationId = transId, w=w, t = t, m = m, a = a)
 }
 
-estimateLogicle <- function(x, channels,...){
+estimateLogicle <- function(x, channels, ...)UseMethod("estimateLogicle")
+estimateLogicle.flowFrame <- function(x, channels, ...){
+  trans <- .estimateLogicle(x, channels, ...)
+  channels <- names(trans)
+  transformList(channels, trans)
+}
+.estimateLogicle <- function(x, channels,...){
             if(!is(x,"flowFrame"))
                 stop("x has to be an object of class \"flowFrame\"")
             if(missing(channels))
                 stop("Please specify the channels to be logicle transformed");
-            indx <- channels %in% colnames(x)
-            if(!all(indx))
-                stop(paste("Channels", channels[!indx] , "were not found in x ",
-                            sep = " "))
-            trans <- lapply(channels, function(p) {
+#            indx <- channels %in% colnames(x)
+#            if(!all(indx))
+#                stop(paste("Channels", channels[!indx] , "were not found in x ",
+#                            sep = " "))
+            channels <- sapply(channels, function(channel)getChannelMarker(x, channel)[["name"]], USE.NAMES = FALSE)
+            
+            sapply(channels, function(p) {
                         .lgclTrans(x, p, ...)               
                     })
-            transformList( channels, trans)   
+              
         }
 
 ## Truncation transformation constructor
